### 2015-11-09 Paper Sigatoka - 2015.tex
primera

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\modulolinenumbers[1]

\journal{Journal of \LaTeX\ Templates}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
%\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}

\title{Forecasting the black Sigatoka development rate: A machine learning methods comparison 
%\tnoteref{mytitlenote}
}
%\tnotetext[mytitlenote]{Fully documented templates are %available in the elsarticle package on \href{http://%www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}%{CTAN}.}

%% Group authors per affiliation:
\author[afiLuisAlex]{Luis-Alexander Calvo-Valverde\fnref{myfootnote}}
\ead{lualcava.sa@gmail.com}
\fntext[myfootnote]{Corresponding author. (506)70104420}

\author[afiCorbana] {Mauricio Guzm\'an-Quesada}

\author[afiCorbana]{Jos\'e-Antonio Guzm\'an-Alvarez}

\author[afiPablo]{Pablo Alvarado-Moya}

\address[afiLuisAlex]{DOCINADE, Instituto Tecnol\'ogico de Costa Rica, 
Computer Research Center, Multidisciplinar program eScience, 
CNCA/CeNAT, Cartago, Costa Rica}

\address[afiCorbana]{Direcci\'on de Investigaciones, Corporaci\'on Bananera Nacional S.A., Gu\'apiles, Costa Rica}

\address[afiPablo]{DOCINADE, Instituto Tecnol\'ogico de Costa Rica, Cartago, Costa Rica}

\begin{abstract}
Pending.
\end{abstract}

\begin{keyword}
\texttt{Machine learning \sep Black Sigatoka \sep Support vector regression \sep
Banana disease prediction \sep Biological warning system }
\end{keyword}

\end{frontmatter}

\linenumbers

\section{Introduction}
The Black Sigatoka disease caused by the fungus {\it Mycosphaerella fijiensis Morelet} is the major pathological problem of banana and plantain crops in Central America, Panama, Colombia and Ecuador, as in many parts of Africa and Asia \citep{MarinVargas1995}.\\
This disease attacks the plant leaves producing a rapid deterioration of the leaf area, affects the growth and productivity of plants as the ability of photosynthesis decreases, causes a reduction in the quality of the fruit, and promotes premature maturation of bunches, which is the major cause of product losses due to this disease. Fig 1 shows three stages of this disease. \\
Phytopathological studies point out that precipitation, temperature, relative humidity and wind are the main climatic variables that affect the development of this disease \citep{MarinVargas1995}.\\

Fig 1. Examples of three disease stages of the black Sigatoka. (a) Initial stage. (b) Intermediate stage, and (c) Advanced stage.\\
According to studies by the National Banana Corporation of Costa Rica (Corbana) made in 2013, considering on average between 53 thru 57 cycles of fungicide applications per farm, the cost per hectare per year ranged between 1800 USD and thru 1900 USD. This represents about 0.76 cents of the price of a box of 18.14 kilograms. Overall, this represents 10 to 12 of the total production cost (Bresciani).\\
The past and present disease development rate can in principle be used to predict its future behavior, tendencies and to determine whether particular fungicide spray schedules will be able to effectively and economically control the disease (Chuang and Jeger, 1987).\\
There are efforts to apply machine learning methods for decision-making in agriculture, including the control of crop diseases. For example, Camargo et al. (2012) present an intelligent systems for the assessment of crop disorders, Huang et al. (2010) introduce a plant virus identification method based on neural networks with an evolutionary preprocessing stage, Kim et al. (2014) summarize in their survey crop pests prediction methods using regression and machine learning approaches, while Zhao et al. (2013) present an intelligent agricultural forecasting system based on wireless sensor networks.\\
In this work, we compare four machine learning methods: support vector regression (SVR), echo state networks (ESN), ridge regression, and ordinary least squares linear regression, to predict the black Sigatoka disease development rate.\\
The main contribution of this work is a comparison between machine learning methods to forecast black Sigatoka development rate.\\

\paragraph{Installation} If the document class \emph{elsarticle} is not available on your computer, you can download and install the system package \emph{texlive-publishers} (Linux) or install the \LaTeX\ package \emph{elsarticle} using the package manager of your \TeX\ installation, which is typically \TeX\ Live or Mik\TeX.

\paragraph{Usage} Once the package is properly installed, you can use the document class \emph{elsarticle} to create a manuscript. Please make sure that your manuscript follows the guidelines in the Guide for Authors of the relevant journal. It is not necessary to typeset your manuscript in exactly the same way as an article, unless you are submitting to a camera-ready copy (CRC) journal.

\paragraph{Functionality} The Elsevier article class is based on the standard article class and supports almost all of the functionality of that class. In addition, it features commands and options to format the
\begin{itemize}
\item document style
\item baselineskip
\item front matter
\item keywords and MSC codes
\item theorems, definitions and proofs
\item lables of enumerations
\item citation style and labeling.
\end{itemize}

\section{Front matter}

The author names and affiliations could be formatted in two ways:
\begin{enumerate}[(1)]
\item Group the authors per affiliation.
\item Use footnotes to indicate the affiliations.
\end{enumerate}
See the front matter of this document for examples. You are recommended to conform your choice to the journal you are submitting to.

\section{Bibliography styles}

There are various bibliography styles available. You can select the style of your choice in the preamble of this document. These styles are Elsevier styles based on standard styles like Harvard and Vancouver. Please use Bib\TeX\ to generate your bibliography and include DOIs whenever available.

Here are two sample references: \citep{Zhao2013}.

\begin{thebibliography}{1}

\bibitem[Marin and Romero, 1995]{MarinVargas1995} Marin Vargas, D., Romero Calderón, R. 1995. El combate de la Sigatoka Negra. Boletín Departamento de Investigaciones, Corbana Costa Rica.

\bibitem[Zhao et al., 2013]{Zhao2013}Zhao, L., He, L., Harry, W., Jin, X. 2013. Intelligent Agricultural Forecasting System Based on Wireless Sensor. Journal of Networks(8), 1817–1824. doi:10.4304/jnw.8.8.1817-1824.

\end{thebibliography}

\end{document}

### 2015-11-09 Paper Sigatoka - 2015.tex
martes

@@
 
-%% APA style
-\bibliographystyle{model5-names}\biboptions{authoryear}
-
 %% AMA style
@@
 %\bibliographystyle{elsarticle-num}
+
+%% APA style
+% \bibliographystyle{model5-names}\biboptions{authoryear}
+
 %%%%%%%%%%%%%%%%%%%%%%%
+
+\usepackage{subcaption}
 
@@
 \author[afiCorbana] {Mauricio Guzm\'an-Quesada}
-
 \author[afiCorbana]{Jos\'e-Antonio Guzm\'an-Alvarez}
-
 \author[afiPablo]{Pablo Alvarado-Moya}
@@
 This disease attacks the plant leaves producing a rapid deterioration of the leaf area, affects the growth and productivity of plants as the ability of photosynthesis decreases, causes a reduction in the quality of the fruit, and promotes premature maturation of bunches, which is the major cause of product losses due to this disease.
-Fig 1 shows three stages of this disease.
-\\
+\figurename $.$\ref{figura1} shows three stages of this disease.\\
 Phytopathological studies point out that precipitation, temperature, relative humidity and wind are the main climatic variables that affect the development of this disease \citep{MarinVargas1995}.\\
-
-Fig 1. Examples of three disease stages of the black Sigatoka. (a) Initial stage. (b) Intermediate stage, and (c) Advanced stage.\\
-According to studies by the National Banana Corporation of Costa Rica (Corbana) made in 2013, considering on average between 53 thru 57 cycles of fungicide applications per farm, the cost per hectare per year ranged between 1800 USD and thru 1900 USD.
+\begin{figure}[h] 
+\begin{subfigure}{.3\textwidth}
+  \centering
+  \includegraphics[width=.8\linewidth]{Roya_a}
+  \caption{}
+  \label{fig:sfig1}
+\end{subfigure}
+\begin{subfigure}{.3\textwidth}
+  \centering
+  \includegraphics[width=.8\linewidth]{Roya_b}
+  \caption{}
+  \label{fig:sfig2}
+\end{subfigure}
+\begin{subfigure}{.3\textwidth}
+  \centering
+  \includegraphics[width=.8\linewidth]{Roya_c}
+  \caption{}
+  \label{fig:sfig3}
+\end{subfigure}
+\caption{Examples of three disease stages of the black Sigatoka. (a) Initial stage. (b) Intermediate stage, and (c) Advanced stage.} 
+\label{figura1} 
+\end{figure}\\
+According to studies by the National Banana Corporation of Costa Rica (Corbana) made in 2013, considering on average between 53 thru 57 cycles of fungicide applications per farm, the cost per hectare per year ranged between \$1800 USD and thru \$1900 USD.
 This represents about 0.76 cents of the price of a box of 18.14 kilograms.
-Overall, this represents 10 to 12 of the total production cost (Bresciani).\\
-The past and present disease development rate can in principle be used to predict its future behavior, tendencies and to determine whether particular fungicide spray schedules will be able to effectively and economically control the disease (Chuang and Jeger, 1987).\\
+Overall, this represents 10\% to 12\% of the total production cost \citet{Bresciani2015}.\\
+The past and present disease development rate can in principle be used to predict its future behavior, tendencies and to determine whether particular fungicide spray schedules will be able to effectively and economically control the disease \citet{ChuangJeger1987}.\\
 There are efforts to apply machine learning methods for decision-making in agriculture, including the control of crop diseases.
-For example, Camargo et al.
-(2012) present an intelligent systems for the assessment of crop disorders, Huang et al.
-(2010) introduce a plant virus identification method based on neural networks with an evolutionary preprocessing stage, Kim et al.
-(2014) summarize in their survey crop pests prediction methods using regression and machine learning approaches, while Zhao et al.
-(2013) present an intelligent agricultural forecasting system based on wireless sensor networks.\\
+For example, \cite{Camargo2012} present an intelligent systems for the assessment of crop disorders, \cite{Huang2010} introduce a plant virus identification method based on neural networks with an evolutionary preprocessing stage, \cite{Kim2014} summarize in their survey crop pests prediction methods using regression and machine learning approaches, while \cite{Zhao2013} present an intelligent agricultural forecasting system based on wireless sensor networks.\\
 In this work, we compare four machine learning methods: support vector regression (SVR), echo state networks (ESN), ridge regression, and ordinary least squares linear regression, to predict the black Sigatoka disease development rate.\\
@@
 
-\paragraph{Installation} If the document class \emph{elsarticle} is not available on your computer, you can download and install the system package \emph{texlive-publishers} (Linux) or install the \LaTeX\ package \emph{elsarticle} using the package manager of your \TeX\ installation, which is typically \TeX\ Live or Mik\TeX.
+\section{Materials and methods}
 
-\paragraph{Usage} Once the package is properly installed, you can use the document class \emph{elsarticle} to create a manuscript.
-Please make sure that your manuscript follows the guidelines in the Guide for Authors of the relevant journal.
-It is not necessary to typeset your manuscript in exactly the same way as an article, unless you are submitting to a camera-ready copy (CRC) journal.
+\subsection{Concepts}
 
-\paragraph{Functionality} The Elsevier article class is based on the standard article class and supports almost all of the functionality of that class.
-In addition, it features commands and options to format the
-\begin{itemize}
-\item document style
-\item baselineskip
-\item front matter
-\item keywords and MSC codes
-\item theorems, definitions and proofs
-\item lables of enumerations
-\item citation style and labeling.
-\end{itemize}
+\subsubsection{Biological warning system}
 
-\section{Front matter}
+The system measures the disease development state to determine when to apply fungicides \citep{MarinVargas1995}. 
+This system is based on two components: a climate component, which is given by the Piche evaporation and a biological component, given by the stage of progress or the rate of disease development.
+Originally, this system was designed to work with young plants.
+One selected plant must exhibit a normal growth and be in a place that enforces a healthy development.
+The plant must start with 5 to 6 true leaves.
+The assessments are made at fixed intervals of seven days as long as possible, on the same plant.
+The first observations should consider the leaf emission, also the level of infection on the leaves should be evaluated considering the stages of development \citep{MarinVargas1995}.
 
-The author names and affiliations could be formatted in two ways:
-\begin{enumerate}[(1)]
-\item Group the authors per affiliation.
-\item Use footnotes to indicate the affiliations.
-\end{enumerate}
-See the front matter of this document for examples. You are recommended to conform your choice to the journal you are submitting to.
+\subsubsection{Support Vector Regression (SVR)}
 
-\section{Bibliography styles}
+Other section
+I need to write
 
-There are various bibliography styles available.
-You can select the style of your choice in the preamble of this document.
-These styles are Elsevier styles based on standard styles like Harvard and Vancouver.
-Please use Bib\TeX\ to generate your bibliography and include DOIs whenever available.
+\section{Results}
 
-Here are two sample references: \citep{Zhao2013}.
+\section{Discussion and conclusions}
+
+\section{References}
 
@@
 
-\bibitem[Marin and Romero, 1995]{MarinVargas1995} Marin Vargas, D., Romero Calderón, R. 1995. El combate de la Sigatoka Negra. Boletín Departamento de Investigaciones, Corbana Costa Rica.
+\bibitem[Brescani, 2015]{Bresciani2015} Brescani, XXXXX.
+
+\bibitem[Camargo et al.,2012]{Camargo2012} Camargo, A., Molina, J., Cadena-Torres, J., Jim\'enez, N., Kim, J.
+2012.
+Intelligent systems for the assessment of crop disorders.
+Computers and Electronics in Agriculture(85), 1-7.
+doi:10.1016/j.compag.2012.02.017.
+
+\bibitem[Chuang and Jeger, 1987]{ChuangJeger1987} Chuang, T., Jeger, M.
+1987.
+Predicting the Rate of Development of Black Sigatoka ( Mycosphaerella fijiensis var.
+difformis ) Disease in Southern Taiwan.
+Phytopathology, 77, 1542-1547.
+
+\bibitem[Huang et al., 2010]{Huang2010} Huang, Y., Lan, Y., Thomson, S., Fang, A., Hoffmann, W., Lacey, R.
+2010.
+Development of soft computing and applications in agricultural and biological engineering.
+Computers and Electronics in Agriculture,(71(2)), 107–127.
+doi:10.1016/j.compag.
+
+\bibitem[Kim et al., 2014]{Kim2014} Kim, Y., Yoo, S., Gu, Y., Lim, J., Han, D.,  Baik, S.
+2014.
+Crop Pests Prediction Method Using Regression and Machine Learning Technology: Survey.
+IERI Procedia(6), 52–56.
+doi:10.1016/j.ieri.2014.03.009.
+
+\bibitem[Marin and Romero, 1995]{MarinVargas1995} Marin Vargas, D., Romero Calderón, R. 1995. El combate de la Sigatoka Negra. Bolet\'in Departamento de Investigaciones, Corbana Costa Rica.
 

### 2015-11-14 Paper Sigatoka - 2015.tex
cuatro
Error en puntos y aparte corregido.  Párrafos separados en líneas más cortas para simplificar control de versiones

@@
 \section{Introduction}
-The Black Sigatoka disease caused by the fungus {\it Mycosphaerella fijiensis Morelet} is the major pathological problem of banana and plantain crops in Central America, Panama, Colombia and Ecuador, as in many parts of Africa and Asia \citep{MarinVargas1995}.\\
-This disease attacks the plant leaves producing a rapid deterioration of the leaf area, affects the growth and productivity of plants as the ability of photosynthesis decreases, causes a reduction in the quality of the fruit, and promotes premature maturation of bunches, which is the major cause of product losses due to this disease.
-\figurename $.$\ref{figura1} shows three stages of this disease.\\
-Phytopathological studies point out that precipitation, temperature, relative humidity and wind are the main climatic variables that affect the development of this disease \citep{MarinVargas1995}.\\
+
+The Black Sigatoka disease caused by the fungus {\it Mycosphaerella
+  fijiensis Morelet} is the major pathological problem of banana and
+plantain crops in Central America, Panama, Colombia and Ecuador, as in
+many parts of Africa and Asia \citep{MarinVargas1995}.
+
+This disease attacks the plant leaves producing a rapid deterioration
+of the leaf area, affects the growth and productivity of plants as the
+ability of photosynthesis decreases, causes a reduction in the quality
+of the fruit, and promotes premature maturation of bunches, which is
+the major cause of product losses due to this disease. \figurename
+$.$\ref{figura1} shows three stages of this disease.
+
+Phytopathological studies point out that precipitation, temperature,
+relative humidity and wind are the main climatic variables that affect
+the development of this disease \citep{MarinVargas1995}.
+
 \begin{figure}[h] 
@@
 \label{figura1} 
-\end{figure}\\
-According to studies by the National Banana Corporation of Costa Rica (Corbana) made in 2013, considering on average between 53 thru 57 cycles of fungicide applications per farm, the cost per hectare per year ranged between \$1800 USD and thru \$1900 USD.
-This represents about 0.76 cents of the price of a box of 18.14 kilograms.
-Overall, this represents 10\% to 12\% of the total production cost \citet{Bresciani2015}.\\
-The past and present disease development rate can in principle be used to predict its future behavior, tendencies and to determine whether particular fungicide spray schedules will be able to effectively and economically control the disease \citet{ChuangJeger1987}.\\
-There are efforts to apply machine learning methods for decision-making in agriculture, including the control of crop diseases.
-For example, \cite{Camargo2012} present an intelligent systems for the assessment of crop disorders, \cite{Huang2010} introduce a plant virus identification method based on neural networks with an evolutionary preprocessing stage, \cite{Kim2014} summarize in their survey crop pests prediction methods using regression and machine learning approaches, while \cite{Zhao2013} present an intelligent agricultural forecasting system based on wireless sensor networks.\\
-In this work, we compare four machine learning methods: support vector regression (SVR), echo state networks (ESN), ridge regression, and ordinary least squares linear regression, to predict the black Sigatoka disease development rate.\\
-The main contribution of this work is a comparison between machine learning methods to forecast black Sigatoka development rate.\\
+\end{figure}
+
+According to studies by the National Banana Corporation of Costa Rica
+(Corbana) made in 2013, considering on average between 53 thru 57
+cycles of fungicide applications per farm, the cost per hectare per
+year ranged between \$1800 USD and thru \$1900 USD. This represents
+about 0.76 cents of the price of a box of 18.14 kilograms. Overall,
+this represents 10\% to 12\% of the total production cost
+\citet{Bresciani2015}.
+
+The past and present disease development rate can in principle be used
+to predict its future behavior, tendencies and to determine whether
+particular fungicide spray schedules will be able to effectively and
+economically control the disease \citet{ChuangJeger1987}.
+
+There are efforts to apply machine learning methods for
+decision-making in agriculture, including the control of crop
+diseases. For example, \cite{Camargo2012} present an intelligent
+systems for the assessment of crop disorders, \cite{Huang2010}
+introduce a plant virus identification method based on neural networks
+with an evolutionary preprocessing stage, \cite{Kim2014} summarize in
+their survey crop pests prediction methods using regression and
+machine learning approaches, while \cite{Zhao2013} present an
+intelligent agricultural forecasting system based on wireless sensor
+networks.
+
+In this work, we compare four machine learning methods: support vector
+regression (SVR), echo state networks (ESN), ridge regression, and
+ordinary least squares linear regression, to predict the black
+Sigatoka disease development rate.
+
+The main contribution of this work is a comparison between machine
+learning methods to forecast black Sigatoka development rate.
 
@@
 
-The system measures the disease development state to determine when to apply fungicides \citep{MarinVargas1995}. 
-This system is based on two components: a climate component, which is given by the Piche evaporation and a biological component, given by the stage of progress or the rate of disease development.
-Originally, this system was designed to work with young plants.
-One selected plant must exhibit a normal growth and be in a place that enforces a healthy development.
-The plant must start with 5 to 6 true leaves.
-The assessments are made at fixed intervals of seven days as long as possible, on the same plant.
-The first observations should consider the leaf emission, also the level of infection on the leaves should be evaluated considering the stages of development \citep{MarinVargas1995}.
+The system measures the disease development state to determine when to
+apply fungicides \citep{MarinVargas1995}.  This system is based on two
+components: a climate component, which is given by the Piche
+evaporation and a biological component, given by the stage of progress
+or the rate of disease development. Originally, this system was
+designed to work with young plants. One selected plant must exhibit a
+normal growth and be in a place that enforces a healthy
+development. The plant must start with 5 to 6 true leaves. The
+assessments are made at fixed intervals of seven days as long as
+possible, on the same plant. The first observations should consider
+the leaf emission, also the level of infection on the leaves should be
+evaluated considering the stages of development
+\citep{MarinVargas1995}.
 
@@
 
-Other section
-I need to write
+Revisar
+Ver el tema de la Bibliografìa
 

### 2015-11-17 Paper Sigatoka - 2015.tex
lunes noche

@@
 \subsubsection{Support Vector Regression (SVR)}
+From the perspective of Support Vector Regression (SVR) the regression function $y = f(s)$ for a given dataset $D ={(s_i,y_i ) }_(i=1)^n$ , is represented as a linear function of the form (Wei, Tao, ZhuoShu, and Zio, 2013):
+$f(s)=w^T s+b$
+where 	$w$ and $b$ are respectively the weight vector and the intercept of the model, and they are selected to find an optimal fit to the data available in $D$.
+For nonlinear cases, one proceeds by mapping the input p-dimensional vectors via a nonlinear function $ɸ∶R^p→F$, onto the feature space F.
+After nonlinear mapping, the regression function evolves to a pervasive form:
+$f(s)=w^T ɸ(s)+b$
+SVR uses the є-insensitive loss function:
+$l=|y-f(s)|_є= {█(0,   |y-f(s)|_  ≤ є@|y-f(s)|_ - є,     else)┤$
+which ignores the error if the difference between the prediction value and the actual value is smaller than $є$.
+$є-insensitive loss function$ allows to find the coefficients $w$ and $b$ by solving a convex optimization problem, which balances the empirical error and the generalization ability.
+In SVR, the empirical error is measured by the loss function є-insensitive and the generalization ability is measured by the Euclidean norm of $w$.
+Then, the optimization problem to identify the regression model can be formulated by (Wei, Tao, ZhuoShu, and Zio, 2013):
+$minimize    J(w,ξ_i,ξ_i^* )=  1/2  ||w| |^2+ C ∑_(i=1)^n▒〖(ξ_i,ξ_i^*)〗
+subject to 	{█(y_i- w^T  ɸ(s)-b  ≤   є+ξ_i@w^T  ɸ(s)+b- y_i   ≤   є+ξ_i^*     i=1,…,n@ξ_i,ξ_i^(* )≥0)┤$
+where C denotes the penalty parameter between empirical and generalization errors, and  $ξ_i,ξ_i^(* )$are slack variables, as shown in Fig 2.
 
-Revisar
-Ver el tema de la Bibliografìa
+Fig 2: є-insensitive loss function (Wei, Tao, ZhuoShu, and Zio, 2013)
+The solution of this optimization problem by the Lagrange method is:
+$f(s)=w^T ɸ(s)+b= ∑_(i=1)^n▒〖(α_i-α_i^* )K(s,s_i )+b〗$
+where $α_i,α_i^*$  are the Lagrange multipliers of the optimization problem’s dual form and $K(s_i,s_j )$ is the kernel function satisfying Mercer condition, and can be described by:
+$K(s_i,s_j )=〈ɸ(s_i )  ɸ(s_j)〉  $
+Common kernel functions are: linear, polynomial and sigmoid.
+Operations in the kernel function K(s,s_i ) are performed in the input space rather than in the potentially high dimensional feature space of ɸ.
+An inner product in the feature space has an equivalent kernel in the input space (Alonso, Rodríguez Castañón, and Bahamonde, 2013).
+
+\subsubsection{Ordinary least square}
+This method fits a linear model with coefficients w = (w1,...,wp) to minimize the residual sum of squares between the observed responses in the dataset, and the responses predicted by the linear approximation.
+Mathematically it solves a problem of the form (scikit-learn developer, 2014):
+〖■(min@w) 〗⁡〖‖Xw-y‖_2^2 〗
+
+\subsubsection{Ridge regression}
+
+This addresses some of the problems of Ordinary Least Squares by imposing a penalty on the size of coefficients.
+The ridge coefficients minimize a penalized residual sum of squares (scikit-learn developer, 2014):
+〖■(min@w) 〗⁡〖‖Xw-y‖_2^2 〗+ α‖w‖_2^2
+Here, α ˃ 0 is a complexity parameter that controls the amount of shrinkage: the larger the value of α, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
+\subsubsection{Echo State Networks (ESN)}
+Recurrent Neural Networks (RNN) are useful for temporal patterns, but when they are trained with backpropagation method, they are very slow.
+Echo State Network (ESN) is an alternative training method to solve that problem.
+ESN is based on the observation that if a random RNN possesses certain algebraic properties, training only a linear readout from it is often sufficient to achieve excellent performance in practical applications (Lukoševičius and Jaeger, 2009).
+For a given training input signal u(n)  ϵ R^(N_u ) a desired target output signal y^(target(n))  ϵ R^(N_y )
+is known.
+Here n = 1, .
+.
+.
+,T is the discrete time and T is the number of data points in the training dataset.
+The task is to learn a model with output y(n)  ϵ R^(N_y ), where y(n) matches y^target(n) (n) as well as possible, minimizing an error measure E(y,y^target), and, more importantly, generalizes well to unseen data.
+The untrained RNN part of an ESN is called a dynamical reservoir, and the resulting states x(n) are termed echoes of its input history
+ (Lukoševičius M. , 2012). Finally, these signals are sent to an output layer as shown in the following Fig.
+
+Figure 3: An echo state network (Lukosevicius, 2012)
+The connections between the different elements of an Echo State Network have weights randomly generated.
+The weights of the internal connections of the reservoir (W) as well as the weights of the input layer (Win), after being generated are set statically during all stages of implementation of the algorithm.
+The weights between the reservoir and the output layer (Wout) are subject to changes of a supervised learning algorithm to correct the degree of error generated by the entire system (Lukoševičius M.
+, 2012).
+Related works
+A related work, no machine learning approach, was performed by Romero (1995) who in the third chapter of his doctoral thesis in the field of plant pathology, proposed regression models using stepwise procedure to predict incubation and latency times of black Sigatoka.
+The author performed experiments on two farms located in Costa Rica (Rita and Waldeck, the same as those used in this study but with different names).
+The time intervals used for that study were: December 1993 thru August 1995.
+Romero concluded that the model to predict the incubation period accounted a R2 of 69\% in his observed data but it was not a good predictor when it was validated against an independent dataset (cross validation).
+For latency, he developed two models that accounted a R2 of 78\% in the observed data, however, when validated against an independent dataset (cross validation), the model was incorrect for Weldeck, and for Rita obtained an adjusted R2 of 82\%.
+A machine learning method was proposed by Glezakos, Moschopoulou, Tsiligiridis, Kintzios, and Yialouris (2010), who presented a genetic algorithm as to smooth out the initial information while, the so produced meta-data sets were used in the training and testing of the applied neural network, producing fitter training data.
+Given the features of the acquired virus time-series signals of the problem under study, an evolutionary method was proposed in order to produce meta-data from the original time-series initial information, reduce the dimensionality of the input data space, and eliminating the noise inherent in the initial raw information The method was tested against some of the most commonly used classifiers in machine learning (Bayes, Trees and k-NN) via cross-validation and proved its potential towards assisting virus identification.
+They made their test with CGMM and TR viruses.
+In agricultural area, Alves, de Carvalho, Pozza, Sanches, and Mai (2011) selected the zones that are potentially favorable to coffee, soybean and banana diseases in Brazil according to the spatial-temporal variability of climatic variables and the geographical distribution of hosts.
+Their study applied methodology enabled the visualization of the variation of areas favorable to epidemics under future scenarios of climate change.
+The geoscientific and statistical modeling techniques developed in that study enabled the development of predictive models and the characterization of risk areas for soybean rust, coffee rust and black Sigatoka disease of banana.
+There have been attempts to generate software tools, Camargo, Molina, Cadena-Torres, Jiménez, and Kim (2012) presented an information system for the assessment of plant disorders (Isacrodi).
+They proposed that experts will attain a much better accuracy than the Isacrodi classifier, particularly when provided with samples from the affected crop.
+However, where such expertise is not available, they suggest that Isacrodi can provide valuable support to farmers.
+Isacordi includes 15 crop disorders, but the black Sigatoka no is one.
+The prediction process is based on multi-class Support Vector Machines.
+Regarding black Sigatoka with machine learning methods, Bendini, Moraes, da Silva, Tezuka, and Cruvinel (2013) presented a study about the risk analysis of black Sigatoka occurrence based on polynomial models.
+A case study was developed in a commercial banana plantation located in Jacupiranga, Brazil, it was monitored weekly during the period from February to December 2005.
+Data were the weekly monitoring of the disease’s evolution stage, time series of meteorological data and remote sensing data.
+They obtained a model to estimate the evolution of the disease from satellite imagery.
+This model relates gray levels (NC) of the corresponding image, band 2 of the Landsat-5 satellite, with the progress status or disease severity (EE): Authors express have reach an R2 of 90%.
+Also there are research related to banana fruit, Soares, Pasqual, Lacerda, Silva, and Donato (2014) show in their study that  to the analyses, the neural network proved to be more accurate in forecasting the weight of the bunch in comparison to the multiple linear regressions in terms of the mean prediction-error (MPE = 1.40), mean square deviation (MSD = 2.29) and coefficient of determination (R2 = 91%).
+In general, machine learning methods applied to prediction plant diseases can be classified in two main approaches: 1) Those that their main inputs are images, and 2) Those that their main inputs are environmental and biological variables.
+Our study is focus in the second one.
+
+\subsubsection{Data}
+In this work we used data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
+The banana type is Musa AAA, subgroup Cavendish, cv.
+Grande Naine.
+Table 1 shows the variables considered initially.
+Table 1 Variables used in the study
+Variable	Meaning
+T_(a_max )	Max air temperature
+T_(a_min )	Min air temperature
+T ̅_a	Mean air temperature
+H	Humidity
+H_min	Min humidity
+H_max	Max humidity
+R	Solar radiation
+P ̅	Mean precipitation
+W_max	Max speed wind
+W ̅	Mean speed wind
+L_2	Biological warning system – Leaf 2
+L_3	Biological warning system – Leaf 3
+L_4	Biological warning system – Leaf 4
+ES	Biological warning system – Evolution Stage
+The value to be predicted in all cases was ES, that is the total measure of the biological warning system.
+The data on the biological warning system are collected once a week.
+Although Corbana has meteorological stations that take data every five minutes, for these experiments, weekly averages generated by nearby stations to each of the farms were used.
+The time intervals used for this study were: La Rita, week 48 of 2002 to week 17 of the 2015 (647 weeks) and for 28 Miles, week 37 of 2003 to week 18 of 2015 (605 weeks).
+Data preprocessing
+In 28 Miles farm we detect that 1\% of the data were missing, while in La Rita 2.25\% of the data were missing.
+To complete the missing value we use spline interpolation.
+The data collected did not exhibit outliers.
+Due the fact that the variables measure meteorological or biological process, they are discretized in order to reflect trends in the data rather than the specific continuous values.
+The coefficient of variation C_v (x) of each variable x was used to determine the numbers of discretizations with:
+n= ⌊100 C_v (x)⌋
+Each discretization range was uniformly partitioned.
+Besides enabling the capture of tendencies, the discretization removes the effect of small variations in the data collection, either by inaccuracies of the instruments (meteorological variables) or by subjective bias introduced by the human who collects the data (biological warning system).
+Each feature was scaled in a range 0 between 1. The variable to be predicted was not scaled.
+
+\subsubsection{Evaluation criteria}
+
+Although there are many types of indicators to assess the quality of the prediction, we selected the root mean square error (RMSE) and the determination coefficient (R2).
+This decision is supported by the widespread use in machine learning and agriculture areas (Soares, Pasqual and Lacerda (2013); Soares, Pasqual and Lacerda (2014); Ibrahim and Wibowo (2014) and Demir and Bruzzone (2014)).
+
+\subsubsection{Methods}
+This research had two phases.
+{\bf Phase one }
+
+In the phase one, we did ten-fold-cross-validation and did a lot of proofs with different machine learning methods and different configuration. 
+We proved with several combinations:
+	Patterns: From one week of observed data to predict the next week until nine weeks before to predict two weeks later.
+	Algorithms: Support vector regression with different kernel functions: linear, RBF (Gaussian) and sigmoid; echo state networks; ordinary least squares linear regression and ridge regression.
+	Variables included in the model. We proved the following combinations:
+	All variables.
+	Only variables that according to expert judgment have more impact on the black Sigatoka development: humidity, precipitation, temperature and wind speed (Marin Vargas and Romero Calderón, 1995).
+	From the four variables listed in the previous paragraph, runs were conducted using each of the variables separately, and combining other runs all the possible pairs of those four variables.
+Phase two
+In the second phase, we used the best configurations obtained en la phase one and did validation with the last 52 and 102 weeks. 
+This second phase pretended to show how these methods behaved on a time of important climate change how are 2014 and 2015 years.
+Programming environment
+We use python programming language with the Integrated Development Environment (IDE) Spyder, particularly with libraries: pandas (Comunity, 2014); numpy (numpy.org, 2013); for SVR, ridge and ordinary least squares, we used sklearn (Pedregosa, et al., 2011); and for ESN the python-based code used belongs to Dr.
+Mantas Lukoševičius (2012) from which we made the necessary adjustments for the experiments of this research.
+The computer was a Lenovo ThinkPad, processor Intel(R) Core™ i7-4800MQ CPU @ 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
 

### 2015-11-17 Paper Sigatoka - 2015.tex
lunes sigo actualizando

@@
 
-\title{Forecasting the black Sigatoka development rate: A machine learning methods comparison 
+\title{Forecasting the black Sigatoka development rate: A comparison of machine learning techniques 
 %\tnoteref{mytitlenote}
@@
 
-The Black Sigatoka disease caused by the fungus {\it Mycosphaerella
+The black Sigatoka disease caused by the fungus {\it Mycosphaerella
   fijiensis Morelet} is the major pathological problem of banana and
@@
 This disease attacks the plant leaves producing a rapid deterioration
-of the leaf area, affects the growth and productivity of plants as the
-ability of photosynthesis decreases, causes a reduction in the quality
+of the leaf area, affects the growth and productivity of the plants due to the impairment of their photosinthetic ability  causes a reduction in the quality
 of the fruit, and promotes premature maturation of bunches, which is
-the major cause of product losses due to this disease. \figurename
+the major cause of product losses associated with the black Sigatoka. \figurename
 $.$\ref{figura1} shows three stages of this disease.
@@
 relative humidity and wind are the main climatic variables that affect
-the development of this disease \citep{MarinVargas1995}.
+its development \citep{MarinVargas1995}.
 
@@
 cycles of fungicide applications per farm, the cost per hectare per
-year ranged between \$1800 USD and thru \$1900 USD. This represents
+year ranged between \$1800 USD and \$1900 USD. This represents
 about 0.76 cents of the price of a box of 18.14 kilograms. Overall,
@@
 
-The past and present disease development rate can in principle be used
-to predict its future behavior, tendencies and to determine whether
+The past and present rates of disease development  can in principle be used
+to predict its future behavior and to determine whether
 particular fungicide spray schedules will be able to effectively and
@@
 diseases. For example, \cite{Camargo2012} present an intelligent
-systems for the assessment of crop disorders, \cite{Huang2010}
+system for the assessment of crop disorders, \cite{Huang2010}
 introduce a plant virus identification method based on neural networks
@@
 
-In this work, we compare four machine learning methods: support vector
-regression (SVR), echo state networks (ESN), ridge regression, and
-ordinary least squares linear regression, to predict the black
-Sigatoka disease development rate.
+In this work, we compare four machine learning techniques (support vector
+regression (SVR), echo state networks (ESN), ridge regression and
+ordinary least squares linear regression) to predict the development rate of the black Sigatoka disease.
 
 The main contribution of this work is a comparison between machine
-learning methods to forecast black Sigatoka development rate.
+learning methods to forecast black Sigatoka development rate. (FALTA COMPLETAR)
 
@@
 
-The system measures the disease development state to determine when to
+This system measures the disease development state to determine when to
 apply fungicides \citep{MarinVargas1995}.  This system is based on two
@@
 After nonlinear mapping, the regression function evolves to a pervasive form:
-$f(s)=w^T ɸ(s)+b$
+$$f(s)=w^T ɸ(s)+b$$
 SVR uses the є-insensitive loss function:
-$l=|y-f(s)|_є= {█(0,   |y-f(s)|_  ≤ є@|y-f(s)|_ - є,     else)┤$
+% $$l=|y-f(s)|_є= {(0,   |y-f(s)|_  ≤ є |y-f(s)|_ - є,     else)$$
+
 which ignores the error if the difference between the prediction value and the actual value is smaller than $є$.
@@
 Then, the optimization problem to identify the regression model can be formulated by (Wei, Tao, ZhuoShu, and Zio, 2013):
-$minimize    J(w,ξ_i,ξ_i^* )=  1/2  ||w| |^2+ C ∑_(i=1)^n▒〖(ξ_i,ξ_i^*)〗
-subject to 	{█(y_i- w^T  ɸ(s)-b  ≤   є+ξ_i@w^T  ɸ(s)+b- y_i   ≤   є+ξ_i^*     i=1,…,n@ξ_i,ξ_i^(* )≥0)┤$
+% $minimize    J(w,ξ_i,ξ_i^* )=  1/2  ||w| |^2+ C ∑_(i=1)^n▒〖(ξ_i,ξ_i^*)〗
+% subject to 	{█(y_i- w^T  ɸ(s)-b  ≤   є+ξ_i@w^T  ɸ(s)+b- y_i   ≤   є% 
+% +ξ_i^*     i=1,…,n@ξ_i,ξ_i^(* )≥0)┤$
 where C denotes the penalty parameter between empirical and generalization errors, and  $ξ_i,ξ_i^(* )$are slack variables, as shown in Fig 2.
@@
 The solution of this optimization problem by the Lagrange method is:
-$f(s)=w^T ɸ(s)+b= ∑_(i=1)^n▒〖(α_i-α_i^* )K(s,s_i )+b〗$
+% $f(s)=w^T ɸ(s)+b= ∑_(i=1)^n▒〖(α_i-α_i^* )K(s,s_i )+b〗$
 where $α_i,α_i^*$  are the Lagrange multipliers of the optimization problem’s dual form and $K(s_i,s_j )$ is the kernel function satisfying Mercer condition, and can be described by:
-$K(s_i,s_j )=〈ɸ(s_i )  ɸ(s_j)〉  $
+% $K(s_i,s_j )=〈ɸ(s_i )  ɸ(s_j)〉  $
 Common kernel functions are: linear, polynomial and sigmoid.
-Operations in the kernel function K(s,s_i ) are performed in the input space rather than in the potentially high dimensional feature space of ɸ.
+Operations in the kernel function $K(s,s_i $) are performed in the input space rather than in the potentially high dimensional feature space of $ɸ$.
 An inner product in the feature space has an equivalent kernel in the input space (Alonso, Rodríguez Castañón, and Bahamonde, 2013).
@@
 Mathematically it solves a problem of the form (scikit-learn developer, 2014):
-〖■(min@w) 〗⁡〖‖Xw-y‖_2^2 〗
+% 〖■(min@w) 〗⁡〖‖Xw-y‖_2^2 〗
 
@@
 The ridge coefficients minimize a penalized residual sum of squares (scikit-learn developer, 2014):
-〖■(min@w) 〗⁡〖‖Xw-y‖_2^2 〗+ α‖w‖_2^2
+% 〖■(min@w) 〗⁡〖‖Xw-y‖_2^2 〗+ α‖w‖_2^2
 Here, α ˃ 0 is a complexity parameter that controls the amount of shrinkage: the larger the value of α, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
@@
 ESN is based on the observation that if a random RNN possesses certain algebraic properties, training only a linear readout from it is often sufficient to achieve excellent performance in practical applications (Lukoševičius and Jaeger, 2009).
-For a given training input signal u(n)  ϵ R^(N_u ) a desired target output signal y^(target(n))  ϵ R^(N_y )
+For a given training input signal $u(n)  ϵ R^(N_u )$ a desired target output signal 
+% $y^(target(n))  ϵ R^(N_y )$
 is known.
-Here n = 1, .
+Here $n = 1, .
 .
 .
-,T is the discrete time and T is the number of data points in the training dataset.
-The task is to learn a model with output y(n)  ϵ R^(N_y ), where y(n) matches y^target(n) (n) as well as possible, minimizing an error measure E(y,y^target), and, more importantly, generalizes well to unseen data.
+,T$ is the discrete time and $T$ is the number of data points in the training dataset.
+The task is to learn a model with output $y(n)  ϵ R^(N_y )$, where $y(n)$ matches $y^target(n)$ $(n)$ as well as possible, minimizing an error measure $E(y,y^target)$, and, more importantly, generalizes well to unseen data.
 The untrained RNN part of an ESN is called a dynamical reservoir, and the resulting states x(n) are termed echoes of its input history
@@
 
-\subsubsection{Data}
+\subsection{Data}
 In this work we used data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
@@
 Variable	Meaning
-T_(a_max )	Max air temperature
-T_(a_min )	Min air temperature
-T ̅_a	Mean air temperature
-H	Humidity
-H_min	Min humidity
-H_max	Max humidity
-R	Solar radiation
-P ̅	Mean precipitation
-W_max	Max speed wind
-W ̅	Mean speed wind
-L_2	Biological warning system – Leaf 2
-L_3	Biological warning system – Leaf 3
-L_4	Biological warning system – Leaf 4
-ES	Biological warning system – Evolution Stage
+%
+%T_(a_max )	Max air temperature
+%T_(a_min )	Min air temperature
+%T ̅_a	Mean air temperature
+%H	Humidity
+%H_min	Min humidity
+%H_max	Max humidity
+%R	Solar radiation
+%P ̅	Mean precipitation
+%W_max	Max speed wind
+%W ̅	Mean speed wind
+%L_2	Biological warning system – Leaf 2
+%L_3	Biological warning system – Leaf 3
+%L_4	Biological warning system – Leaf 4
+%ES	Biological warning system – Evolution Stage
+
 The value to be predicted in all cases was ES, that is the total measure of the biological warning system.
@@
 Due the fact that the variables measure meteorological or biological process, they are discretized in order to reflect trends in the data rather than the specific continuous values.
-The coefficient of variation C_v (x) of each variable x was used to determine the numbers of discretizations with:
-n= ⌊100 C_v (x)⌋
+The coefficient of variation $C_v (x)$ of each variable x was used to determine the numbers of discretizations with:
+$$n= ⌊100 C_v (x)⌋$$
 Each discretization range was uniformly partitioned.
@@
 
-\subsubsection{Evaluation criteria}
+\subsection{Evaluation criteria}
 
@@
 
-\subsubsection{Methods}
-This research had two phases.
-{\bf Phase one }
+\subsection{Methods}
+This research had two phases.\\
+{\bf Phase one } \\
 
@@
 	Only variables that according to expert judgment have more impact on the black Sigatoka development: humidity, precipitation, temperature and wind speed (Marin Vargas and Romero Calderón, 1995).
-	From the four variables listed in the previous paragraph, runs were conducted using each of the variables separately, and combining other runs all the possible pairs of those four variables.
-Phase two
+	From the four variables listed in the previous paragraph, runs were conducted using each of the variables separately, and combining other runs all the possible pairs of those four variables.\\
+{\bf Phase two } \\
 In the second phase, we used the best configurations obtained en la phase one and did validation with the last 52 and 102 weeks. 
 This second phase pretended to show how these methods behaved on a time of important climate change how are 2014 and 2015 years.
+
 Programming environment

### 2015-11-18 Paper Sigatoka - 2015.tex
miercoles medio dia

@@
 \usepackage{lineno,hyperref}
+
+\usepackage{amsmath}
+
 \modulolinenumbers[1]
@@
 system for the assessment of crop disorders, \cite{Huang2010}
-introduce a plant virus identification method based on neural networks
-with an evolutionary preprocessing stage, \cite{Kim2014} summarize in
+introduce a plant virus identification method based on neural networks with an evolutionary preprocessing stage, \cite{Kim2014} summarize in
 their survey crop pests prediction methods using regression and
@@
 \subsubsection{Support Vector Regression (SVR)}
-From the perspective of Support Vector Regression (SVR) the regression function $y = f(s)$ for a given dataset $D ={(s_i,y_i ) }_(i=1)^n$ , is represented as a linear function of the form (Wei, Tao, ZhuoShu, and Zio, 2013):
-$f(s)=w^T s+b$
+From the perspective of Support Vector Regression (SVR) the regression function $y = f(s)$ for a given dataset $D=\{(s_i,y_i)\}_{i=1}^n$ , is represented as a linear function of the form \citep{Wei2013}:
+
+$$f(s)=w^Ts+b$$
+
 where $w$ and $b$ are respectively the weight vector and the intercept of the model, and they are selected to find an optimal fit to the data available in $D$.
-For nonlinear cases, one proceeds by mapping the input p-dimensional vectors via a nonlinear function $ɸ∶R^p→F$, onto the feature space F.
+
+For nonlinear cases, one proceeds by mapping the input p-dimensional vectors via a nonlinear function $\phi : R^p\rightarrow F$, onto the feature space $F$.
 After nonlinear mapping, the regression function evolves to a pervasive form:
-$$f(s)=w^T ɸ(s)+b$$
-SVR uses the є-insensitive loss function:
-% $$l=|y-f(s)|_є= {(0,   |y-f(s)|_  ≤ є |y-f(s)|_ - є,     else)$$
 
-which ignores the error if the difference between the prediction value and the actual value is smaller than $є$.
-$є-insensitive loss function$ allows to find the coefficients $w$ and $b$ by solving a convex optimization problem, which balances the empirical error and the generalization ability.
-In SVR, the empirical error is measured by the loss function є-insensitive and the generalization ability is measured by the Euclidean norm of $w$.
-Then, the optimization problem to identify the regression model can be formulated by (Wei, Tao, ZhuoShu, and Zio, 2013):
-% $minimize    J(w,ξ_i,ξ_i^* )=  1/2  ||w| |^2+ C ∑_(i=1)^n▒〖(ξ_i,ξ_i^*)〗
-% subject to 	{█(y_i- w^T  ɸ(s)-b  ≤   є+ξ_i@w^T  ɸ(s)+b- y_i   ≤   є% 
-% +ξ_i^*     i=1,…,n@ξ_i,ξ_i^(* )≥0)┤$
-where C denotes the penalty parameter between empirical and generalization errors, and  $ξ_i,ξ_i^(* )$are slack variables, as shown in Fig 2.
+$$f(s)=w^T \phi (s)+b$$
+
+SVR uses the $ \epsilon -insensitive \ loss \ function$:
+
+\[l= {\Bigr| y - f(s) \Bigr| }_{\epsilon } = 
+\left\{\begin{array}{lcl}
+	0 & \mbox{} & {\Bigr| y - f(s) \Bigr| } \leq \epsilon \\
+	 &  & \\
+	{\Bigr| y - f(s) \Bigr| - \epsilon} & \mbox{} & else\\ \end{array} \right. \]
+
+which ignores the error if the difference between the prediction value and the actual value is smaller than $ \epsilon$.
+The $ \epsilon -insensitive \ loss \ function$ allows to find the coefficients $w$ and $b$ by solving a convex optimization problem, which balances the empirical error and the generalization ability.
+In SVR, the empirical error is measured by the loss function є-insensitive and the generalization ability is measured by the Euclidean norm of $w$ \citep{XXXX201X}.
+Then, the optimization problem to identify the regression model can be formulated by \citep{Wei2013}:
+
+\begin{equation} 
+\begin{aligned}
+& \underset{}{\text{miimize}}
+& & J(w,\xi_i , \xi_i^* ) = \frac{1}{2}   \Bigr| \Bigr| x \Bigr| \Bigr|^2 + C \sum_{i=1}^{n} (\xi_i , \xi_i^* )    \\
+& \text{subject to}
+& & \begin{array}{lcl} 
+y_i - w^T \phi (s) - b  \leq \epsilon + \xi_i   \\
+w^T \phi (s) + b - y_i \leq \epsilon + \xi_i^* & i= 1,2,...,n \\
+\xi_i , \xi_i^* \geq 0  \\
+\end{array}
+\end{aligned}
+\end{equation}  
+where $C$ denotes the penalty parameter between empirical and generalization errors, and  $\xi_i , \xi_i^*$ are slack variables. \figurename
+$.$\ref{figura2} shows this situation.
+
+\begin{figure}[h] 
+ \centering
+ \includegraphics[scale=.9]{SVR}
+ \caption{$\epsilon -insensitive \ loss \ function \citep{Wei2013}.$} 
+ \label{figura2} 
+\end{figure}
 
@@
 
+\bibitem[Wei et al., 2013]{Wei2013} Wei, Z., Tao, T., ZhuoShu, D.,  Zio, E.
+(2013).
+A dynamic particle filter-support vector regression method for reliability prediction.
+Reliability Engineering \& System Safety, 109–116.
+doi:10.1016/j.ress.2013.05.021.
+
+\bibitem[Libro SVM XXXX, 201X]{XXXX201X} Libro SVM XXXX.
+
 \end{thebibliography}

### 2015-11-19 Paper Sigatoka - 2015.tex
miercoles 18 6pm

@@
 \modulolinenumbers[1]
+
+\usepackage{soul}
 
@@
 From the perspective of Support Vector Regression (SVR) the regression function $y = f(s)$ for a given dataset $D=\{(s_i,y_i)\}_{i=1}^n$ , is represented as a linear function of the form \citep{Wei2013}:
-
 $$f(s)=w^Ts+b$$
-
 where $w$ and $b$ are respectively the weight vector and the intercept of the model, and they are selected to find an optimal fit to the data available in $D$.
@@
 & \underset{}{\text{miimize}}
-& & J(w,\xi_i , \xi_i^* ) = \frac{1}{2}   \Bigr| \Bigr| x \Bigr| \Bigr|^2 + C \sum_{i=1}^{n} (\xi_i , \xi_i^* )    \\
+& & J(w,\xi_i , \xi_i^* ) = \frac{1}{2}   \Bigr| \Bigr| w \Bigr| \Bigr|^2 + C \sum_{i=1}^{n} (\xi_i , \xi_i^* )    \\
 & \text{subject to}
@@
 
-Fig 2: є-insensitive loss function (Wei, Tao, ZhuoShu, and Zio, 2013)
-The solution of this optimization problem by the Lagrange method is:
-% $f(s)=w^T ɸ(s)+b= ∑_(i=1)^n▒〖(α_i-α_i^* )K(s,s_i )+b〗$
-where $α_i,α_i^*$  are the Lagrange multipliers of the optimization problem’s dual form and $K(s_i,s_j )$ is the kernel function satisfying Mercer condition, and can be described by:
-% $K(s_i,s_j )=〈ɸ(s_i )  ɸ(s_j)〉  $
-Common kernel functions are: linear, polynomial and sigmoid.
-Operations in the kernel function $K(s,s_i $) are performed in the input space rather than in the potentially high dimensional feature space of $ɸ$.
-An inner product in the feature space has an equivalent kernel in the input space (Alonso, Rodríguez Castañón, and Bahamonde, 2013).
+The solution of this optimization problem by the Lagrange method is given by:\\
+$$ f(s) = w^T \phi (s) + b = \sum_{i=1}^{n} (\alpha_i - \alpha_i^*) K (s,s_i) + b $$ 
+where $\alpha_i - \alpha_i^*$  are the Lagrange multipliers of the optimization problem’s dual form and $K(s_i,s_j )$ is the kernel function satisfying the Mercer condition, and holds:
+$$K(s_i,s_j ) = \big \langle  \phi(s_i) , \phi(s_j)  \big \rangle $$
+Operations in the kernel function $K(s,s_i )$ are performed in the input space rather than in the potentially high dimensional feature space of $\phi$ \citep{Alonso2013}.
 
-\subsubsection{Ordinary least square}
-This method fits a linear model with coefficients w = (w1,...,wp) to minimize the residual sum of squares between the observed responses in the dataset, and the responses predicted by the linear approximation.
-Mathematically it solves a problem of the form (scikit-learn developer, 2014):
-% 〖■(min@w) 〗⁡〖‖Xw-y‖_2^2 〗
+\subsubsection{Ordinary least squares regression}
+This method fits a linear model with coefficients $w = (w1,..,wp)$ to minimize the residual sum of squares between the observed responses in the dataset, and the responses predicted by the linear approximation.
+Mathematically it solves a problem of the form \citep{ scikit-learn2011}:
+
+$$\min_{w} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  $$
+where $X$ denotes the features matriz. \\
+According Pedregosa et al.
+\citep{ scikitlearn2011}  the coefficient estimates for Ordinary Least Squares rely on the independence of the model terms.
+When terms are correlated and the columns of the design matrix $X$ have an approximate linear dependence, the design matrix becomes close to singular and as a result, the least-squares estimate becomes highly sensitive to random errors in the observed response, producing a large variance.
+This situation of multicollinearity can arise, for example, when data are collected without an experimental design
 
 \subsubsection{Ridge regression}
-
-This addresses some of the problems of Ordinary Least Squares by imposing a penalty on the size of coefficients.
-The ridge coefficients minimize a penalized residual sum of squares (scikit-learn developer, 2014):
-% 〖■(min@w) 〗⁡〖‖Xw-y‖_2^2 〗+ α‖w‖_2^2
-Here, α ˃ 0 is a complexity parameter that controls the amount of shrinkage: the larger the value of α, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
+The ridge regression addresses some of the problems of ordinary least squares regression by imposing a penalty on the size of the coefficients.
+The ridge coefficients minimize a penalized residual sum of squares  \citep{scikitlearn2011}:
+$$\min_{w} { \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
+Here, $\alpha \ \textgreater \  0$ is a complexity parameter that controls the amount of shrinkage: the larger the value of $\alpha$, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
 \subsubsection{Echo State Networks (ESN)}
-Recurrent Neural Networks (RNN) are useful for temporal patterns, but when they are trained with backpropagation method, they are very slow.
+Recurrent Neural Networks (RNN) are useful for temporal patterns, but when they are trained with backpropagation methods, they are very slow.
 Echo State Network (ESN) is an alternative training method to solve that problem.
-ESN is based on the observation that if a random RNN possesses certain algebraic properties, training only a linear readout from it is often sufficient to achieve excellent performance in practical applications (Lukoševičius and Jaeger, 2009).
-For a given training input signal $u(n)  ϵ R^(N_u )$ a desired target output signal 
-% $y^(target(n))  ϵ R^(N_y )$
+ESN is based on the observation that if a random RNN possesses certain algebraic properties, training only a linear readout from it is often sufficient to achieve excellent performance in practical applications \citep{Lukose2009}.
+For a given training input signal $u(n)  \in R^{N_u}$ a desired target output signal $y^{target}(n) \in R^{N_y}$
 is known.
@@
 ,T$ is the discrete time and $T$ is the number of data points in the training dataset.
-The task is to learn a model with output $y(n)  ϵ R^(N_y )$, where $y(n)$ matches $y^target(n)$ $(n)$ as well as possible, minimizing an error measure $E(y,y^target)$, and, more importantly, generalizes well to unseen data.
-The untrained RNN part of an ESN is called a dynamical reservoir, and the resulting states x(n) are termed echoes of its input history
- (Lukoševičius M. , 2012). Finally, these signals are sent to an output layer as shown in the following Fig.
+The task is to learn a model with output $y(n) \in R^{N_y}$, where $y(n)$ matches $y^target(n)$ as well as possible, minimizing an error measure $E(y,y^target)$, and, more importantly, generalizes well to unseen data.
+The untrained RNN part of an ESN is called a dynamical reservoir, and the resulting states x(n) are termed echoes of its input history \citep{Lukose2012}.
+Finally, these signals are sent to an output layer as shown in the \figurename
+$.$\ref{figura3}.
+\begin{figure}[h] 
+ \centering
+ \includegraphics[scale=.9]{Reservorio}
+ \caption{An echo state network \citep{Lukose2012}.} 
+ \label{figura3} 
+\end{figure}
 
-Figure 3: An echo state network (Lukosevicius, 2012)
 The connections between the different elements of an Echo State Network have weights randomly generated.
-The weights of the internal connections of the reservoir (W) as well as the weights of the input layer (Win), after being generated are set statically during all stages of implementation of the algorithm.
-The weights between the reservoir and the output layer (Wout) are subject to changes of a supervised learning algorithm to correct the degree of error generated by the entire system (Lukoševičius M.
-, 2012).
-Related works
-A related work, no machine learning approach, was performed by Romero (1995) who in the third chapter of his doctoral thesis in the field of plant pathology, proposed regression models using stepwise procedure to predict incubation and latency times of black Sigatoka.
+The weights of the internal connections of the reservoir $(W)$ as well as the weights of the input layer $(W_in)$, after being generated are set statically during all stages of implementation of the algorithm.
+The weights between the reservoir and the output layer $(W_out)$ are subject to changes of a supervised learning algorithm to correct the degree of error generated by the entire system \citep{Lukose2012}.
+
+\subsubsection{Related works}
+Huang et al.
+\citep{Huang2010}  surveyed the development of soft computing techniques in agricultural and biological engineering, including fuzzy logic, artificial neural networks, genetic algorithms, bayesian inference and decision trees.
+
+A related work, proposed by Romero \citep{Romero1995} relies on regression models using a stepwise procedure to predict incubation and latency times of black Sigatoka.
 The author performed experiments on two farms located in Costa Rica (Rita and Waldeck, the same as those used in this study but with different names).
-The time intervals used for that study were: December 1993 thru August 1995.
-Romero concluded that the model to predict the incubation period accounted a R2 of 69\% in his observed data but it was not a good predictor when it was validated against an independent dataset (cross validation).
-For latency, he developed two models that accounted a R2 of 78\% in the observed data, however, when validated against an independent dataset (cross validation), the model was incorrect for Weldeck, and for Rita obtained an adjusted R2 of 82\%.
-A machine learning method was proposed by Glezakos, Moschopoulou, Tsiligiridis, Kintzios, and Yialouris (2010), who presented a genetic algorithm as to smooth out the initial information while, the so produced meta-data sets were used in the training and testing of the applied neural network, producing fitter training data.
-Given the features of the acquired virus time-series signals of the problem under study, an evolutionary method was proposed in order to produce meta-data from the original time-series initial information, reduce the dimensionality of the input data space, and eliminating the noise inherent in the initial raw information The method was tested against some of the most commonly used classifiers in machine learning (Bayes, Trees and k-NN) via cross-validation and proved its potential towards assisting virus identification.
-They made their test with CGMM and TR viruses.
-In agricultural area, Alves, de Carvalho, Pozza, Sanches, and Mai (2011) selected the zones that are potentially favorable to coffee, soybean and banana diseases in Brazil according to the spatial-temporal variability of climatic variables and the geographical distribution of hosts.
-Their study applied methodology enabled the visualization of the variation of areas favorable to epidemics under future scenarios of climate change.
-The geoscientific and statistical modeling techniques developed in that study enabled the development of predictive models and the characterization of risk areas for soybean rust, coffee rust and black Sigatoka disease of banana.
-There have been attempts to generate software tools, Camargo, Molina, Cadena-Torres, Jiménez, and Kim (2012) presented an information system for the assessment of plant disorders (Isacrodi).
+The study used data from: December 1993 to August 1995.
+Romero concluded that the model to predict the incubation period accounted a $R^2$ of 69\% in his observed data but it was not a good predictor when it was validated against an independent dataset (cross validation).
+For latency, he developed two models that accounted a $R^2$ of 78\% \hl{PONER EL VALOR OBTENIDO} in the observed data, however, when validated against an independent dataset (cross validation), the model was incorrect \hl{PONER EL VALOR OBTENIDO} for Weldeck, and for Rita obtained an adjusted $R^2$ of 82\%.
+
+Glezakos et al.
+\citep{Glezakos2010} proposed to use Genetic Algorithms (GA) and Neural Networks (NN) to identify plant virus (Tobacco Rattle Virus (TRV) and the Cucumber Green Mottle Mosaic Virus (CGMMV)).
+This is achieved by the development of ana- lytical tools of evolutionary adaptive width, propelled by Genetic Algorithms (GAs) and Neural Networks (NNs).
+The method was tested against some of the most commonly used classifiers in machine learning (Bayes, Trees and k-NN) via cross-validation and proved its potential towards the identification.
+
+In the agricultural context, Alves et al.
+\citep{Alves2011} used  geoinformation techniques to develop predictive models to study the areas of risk to soybean rust in soybean, coffee leaf rust in coffee, and black Sigatoka in banana, considering Brazil’s climatic characterization and the distribution of soybean, coffee and banana crops.
+Temperature and rainfall data were obtained for the period from 1950 to 2000, and of simulations for 2020, 2050 and 2080 using the SRES A2 climate change scenarios.
+Using principal components analysis, a single variable was generated based on 57 variables, in order to determine an index explaining 87\%, 88\% and 90\% of the variability of soybean, coffee and banana crops, respectively, in municipal districts across Brazil.
+The climatic model was used to generate the zoning of the three plant diseases, using temperature and leaf wetness as input.
+Areas of favorability for the diseases were plotted against the main coffee, soybean and banana growing areas in Brazil.
+This methodology enabled the visualization of the changes in areas favorable for epidemics under possible future scenarios of climate change.
+
+Other applications of machine learning methods in precision agriculture include the use of support vector regression to predict carcass weight in beef cattle in advance to the slaughter  \citep{Alonso2013}, machine learning assessments of soil drying for agricultural planning \citep{Coopersmith2014}, and early detection and classification of plant diseases with support vector machines based on hyperspectral reflectance \citep{Rumpf2010}.
+
+Furthermore, there have been attempts to generate software tools.
+Camargo et al.
+\citep{Camargo2012} presented an information system for the assessment of plant disorders (Isacrodi).
 They proposed that experts will attain a much better accuracy than the Isacrodi classifier, particularly when provided with samples from the affected crop.
-However, where such expertise is not available, they suggest that Isacrodi can provide valuable support to farmers.
-Isacordi includes 15 crop disorders, but the black Sigatoka no is one.
-The prediction process is based on multi-class Support Vector Machines.
-Regarding black Sigatoka with machine learning methods, Bendini, Moraes, da Silva, Tezuka, and Cruvinel (2013) presented a study about the risk analysis of black Sigatoka occurrence based on polynomial models.
-A case study was developed in a commercial banana plantation located in Jacupiranga, Brazil, it was monitored weekly during the period from February to December 2005.
-Data were the weekly monitoring of the disease’s evolution stage, time series of meteorological data and remote sensing data.
+However, those cases where such expertise is not available, they suggest that Isacrodi can provide valuable support to farmers.
+Isacordi includes 15 crop disorders, but the black Sigatoka no is one of them.
+The prediction process is based on multi-class support vector machines.
+
+Regarding the prediction of the develpment of the black Sigatoka with machine learning methods, Bendini et al.
+\citep{Bendini2013}  presented a study about the risk analysis of black Sigatoka occurrence based on polynomial models.
+A case study was developed in a commercial banana plantation located in Jacupiranga, Brazil.
+It was monitored weekly during the period from February to December 2005.
+Data included the weekly monitoring of the disease’s evolution stage, time series of meteorological data and remote sensing data.
 They obtained a model to estimate the evolution of the disease from satellite imagery.
-This model relates gray levels (NC) of the corresponding image, band 2 of the Landsat-5 satellite, with the progress status or disease severity (EE): Authors express have reach an R2 of 90%.
-Also there are research related to banana fruit, Soares, Pasqual, Lacerda, Silva, and Donato (2014) show in their study that  to the analyses, the neural network proved to be more accurate in forecasting the weight of the bunch in comparison to the multiple linear regressions in terms of the mean prediction-error (MPE = 1.40), mean square deviation (MSD = 2.29) and coefficient of determination (R2 = 91%).
-In general, machine learning methods applied to prediction plant diseases can be classified in two main approaches: 1) Those that their main inputs are images, and 2) Those that their main inputs are environmental and biological variables.
-Our study is focus in the second one.
+This model relates gray levels (NC) of the band 2 images of the Landsat-5 satellite, with the progress status or disease severity (EE).
+The authors claim to reach an $R^2$ of 90\%.
 
-\subsection{Data}
-In this work we used data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
+Also there are works related to banana fruit.
+Soares et al.
+\citep{Soares2014} apply two techniques: artificial neural networks (ANNs) and multiple linear regression (MLR) in banana plant to predict the yield, their results show that the neural network proved to be more accurate in forecasting the weight of the bunch in comparison to the multiple linear regressions in terms of the mean prediction-error $(MPE = 1.40)$, mean square deviation $(MSD = 2.29)$ and coefficient of determination $(R^2 = 91\%)$.
+
+In general, the machine learning methods applied to predict the evolution of plant diseases, can be classified in two main approaches: 1) Those whose main inputs are images, and 2) Those whose main inputs are environmental and biological variables.
+Our study focuses in the second case.
+
+\subsubsection{Data}
+In this work we use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
 The banana type is Musa AAA, subgroup Cavendish, cv.
@@
 
+\bibitem[Alonso et al., 2013]{Alonso2013} Alonso, J., Rodríguez Castañón, Á., Bahamonde, A.
+(2013).
+Support Vector Regression to predict carcass weight in beef cattle in advance of the slaughter.
+Computers and Electronics in Agriculture, 116-120.
+
+\bibitem[Pedregosa et al., 2013] {scikitlearn2011} Pedregosa, F.
+and Varoquaux, G.
+and Gramfort, A.
+and Michel, V.
+and Thirion, B.
+and Grisel, O.
+and Blondel, M.
+and Prettenhofer, P.
+and Weiss, R.
+and Dubourg, V.
+and Vanderplas, J.
+and Passos, A.
+and Cournapeau, D.
+and Brucher, M.
+and Perrot, M.
+and Duchesnay, E.
+(2011) Scikit-learn: Machine Learning in {P}ython.
+Journal of Machine Learning Research, vol.
+12, 2825--2830.
+
+\bibitem[Lukosevicius and Jaeger, 2009] {Lukose2009} Lukosevicius, M.
+and Jaeger, H.
+(2009).
+Reservoir computing approaches to recurrent neural network training.
+Computer Science Review(3), 127–149.
+doi:10.1016/j.cosrev.2009.03.005.
+
+\bibitem[Lukosevicius, 2012] {Lukose2012} Lukosevicius, M. (2012). A Practical Guide to Applying Echo State Networks. Neural Networks: Tricks of the Trade. 1-20
+
+\bibitem[Romero, 1995] {Romero1995}  Romero, R.
+(1995).
+Dynamics of fungicide resistant populations of Mycosphaerella fijiensis and Epidemiology of black Sigatoka of banana.
+Costa Rica: North Carolina State University.
+
+\bibitem[Glezakos et al., 2010] {Glezakos2010} Glezakos, T., Moschopoulou, G., Tsiligiridis, T., Kintzios, S., Yialouris, C.
+(2010).
+Plant virus identification based on neural networks with evolutionary preprocessing.
+Computers and Electronics in Agriculture, 70, 263–275.
+
+\bibitem[Alves et al., 2011] {Alves2011} Alves, M., de Carvalho, L., Pozza, E., Sanches, L., Maia, J.
+(2011).
+Ecological zoning of soybean rust, coffee rust and banana black sigatoka based on Brazilian climate changes.
+Procedia Environmental Sciences, 6, 35-49.
+
+\bibitem[Huang et al., 2010] {Huang2010}  Huang, Y., Lan, Y., Thomson, S., Fang, A., Hoffmann, W., Lacey, R.
+(2010).
+Development of soft computing and applications in agricultural and biological engineering.
+Computers and Electronics in Agriculture,(71(2)), 107–127.
+doi:10.1016/j.compag.
+
+\bibitem[Coopersmith et al., 2010] {Coopersmith2014} Coopersmith, E.
+J., Minsker, B.
+S., Wenzel, C.
+E., Gilmore, B.
+J.
+(2014).
+Machine learning assessments of soil drying for agricultural planning.
+Computers and Electronics in Agriculture, 104, 93–104.
+http://doi.org/10.1016/j.compag.2014.04.004
+
+\bibitem[Rumpf et al., 2010] {Rumpf2010} Rumpf, T., Mahlein, a.-K., Steiner, U., Oerke, E.-C., Dehne, H.-W., Plümer, L.
+(2010).
+Early detection and classification of plant diseases with Support Vector Machines based on hyperspectral reflectance.
+Computers and Electronics in Agriculture, 74(1), 91–99.
+http://doi.org/10.1016/j.compag.2010.06.009
+
+\bibitem[Bendini et al., 2010] {Bendini2013} Bendini, H., Moraes, W., da Silva, S., Tezuka, E., Cruvinel, P.
+(2013).
+Análise de risco da ocorrência de Sigatoka-negra baseada em modelos polinomiais: um estudo de caso.
+Tropical Plant Pathology, 38, 035-043.
+
+\bibitem[Soares et al., 2014] {Soares2014} Soares, J., Pasqual, M., Lacerda, W., Silva, S., Donato, S.
+(2014).
+Comparison of techniques used in the prediction of yield in banana plants.
+Scientia Horticulturae journal, 167, 84-90.
+
 \end{thebibliography}

### 2015-11-19 Paper Sigatoka - 2015.tex
jueves 19 12m - todo el texto - latex por data section

@@
 \subsection{Concepts}
+\subsubsection{Black Sigatoka disease}
+
+Black Sigatoka, disease caused by the fungus Mycosphaerella fijiensis Morelet, is the main problem phytopathologic of banana and plantain crops in Central America \citep{MarinVargas1995}.
+
+This disease attacks the leaves of plants producing a rapid deterioration of the leaf area.
+It affects the growth and productivity of plants by decreasing photosynthetic capacity.
+Also causes a reduction in quality of the fruit \citep{MarinVargas1995}.
+
+The climate has a major effect on the behavior of the black Sigatoka.
+Precipitation, temperature, relative humidity and wind are the main climatic variables affecting the development of this disease \citep{MarinVargas1995}.
 
@@
 Here, $\alpha \ \textgreater \  0$ is a complexity parameter that controls the amount of shrinkage: the larger the value of $\alpha$, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
+
 \subsubsection{Echo State Networks (ESN)}
@@
 A related work, proposed by Romero \citep{Romero1995} relies on regression models using a stepwise procedure to predict incubation and latency times of black Sigatoka.
-The author performed experiments on two farms located in Costa Rica (Rita and Waldeck, the same as those used in this study but with different names).
+The author performed experiments on two farms located in Costa Rica (La Rita and Waldeck, the same as those used in this study but with different names).
 The study used data from: December 1993 to August 1995.
@@
 \subsubsection{Data}
-In this work we use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
+In this work we use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (previously called Waldeck and located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
 The banana type is Musa AAA, subgroup Cavendish, cv.
 Grande Naine.
-Table 1 shows the variables considered initially.
-Table 1 Variables used in the study
-Variable	Meaning
-%
-%T_(a_max )	Max air temperature
-%T_(a_min )	Min air temperature
-%T ̅_a	Mean air temperature
-%H	Humidity
-%H_min	Min humidity
-%H_max	Max humidity
-%R	Solar radiation
-%P ̅	Mean precipitation
-%W_max	Max speed wind
-%W ̅	Mean speed wind
-%L_2	Biological warning system – Leaf 2
-%L_3	Biological warning system – Leaf 3
-%L_4	Biological warning system – Leaf 4
-%ES	Biological warning system – Evolution Stage
+The \tablename $.$\ref{tabla1} shows the variables available.
 
-The value to be predicted in all cases was ES, that is the total measure of the biological warning system.
+\begin{table}[h] 
+\caption{Variables used in the study} 
+\label{tabla1} 
+\centering
+\begin{tabular}{c|c} 
+\hline
+\bfseries Variable & \bfseries Meaning0'0' \\ 
+\hline\hline 
+$T_{a_max}$ &	Max air temperature \\
+$T_{a_min}$ &	Min air temperature \\
+$\overline{T}_{a}$	 & Mean air temperature \\
+$H$	&  Humidity \\
+$H_min$ &	Min humidity \\
+$H_max$	& Max humidity \\
+$R$	& Solar radiation \\
+$\overline{P}$	& Mean precipitation  \\
+$W_max$	& Max speed wind \\
+$\overline{W}$	& Mean speed wind  \\
+$L_2$	& Biological warning system – Leaf 2 \\
+$L_3$	& Biological warning system – Leaf 3 \\
+$L_4$	& Biological warning system – Leaf 4 \\
+$E_s$	& Biological warning system – Evolution Stage \\
+\hline
+\end{tabular} 
+\end{table}
+
+The value to be predicted in all cases was $E_s$, that is the total measure of the biological warning system.
 The data on the biological warning system are collected once a week.

### 2015-11-20 Paper Sigatoka - 2015.tex
Jueves 19 6pm - Actualizado

@@
 \usepackage{soul}
+\usepackage{marginnote}
 
@@
 
+\marginnote{La bibliografia debe ser autor, anno}
+
 The black Sigatoka disease caused by the fungus {\it Mycosphaerella
@@
 \citet{Bresciani2015}.
+\marginnote{Hay que completar la cita de Brescani, o cambiarla}
 
@@
 The main contribution of this work is a comparison between machine
-learning methods to forecast black Sigatoka development rate. (FALTA COMPLETAR)
+learning methods to forecast black Sigatoka development rate. 
+\marginnote{FALTA COMPLETAR esta parte}
 
@@
 \citep{MarinVargas1995}.
+\marginnote{Tendra sentido ampiar esta parte?}
 
@@
 Then, the optimization problem to identify the regression model can be formulated by \citep{Wei2013}:
+\marginnote{Falta completar la cita XXXXX con el libro de SVR}
 
@@
 For latency, he developed two models that accounted a $R^2$ of 78\% \hl{PONER EL VALOR OBTENIDO} in the observed data, however, when validated against an independent dataset (cross validation), the model was incorrect \hl{PONER EL VALOR OBTENIDO} for Weldeck, and for Rita obtained an adjusted $R^2$ of 82\%.
+\marginnote{Revisar si se puede implementar la propuesta de este autor para comparar}
 
@@
 The value to be predicted in all cases was $E_s$, that is the total measure of the biological warning system.  
+\marginnote{Mas detalle aqui?  o ampliar en los conceptos?}
+
 The data on the biological warning system are collected once a week.
 Although Corbana has meteorological stations that take data every five minutes, for these experiments, weekly averages generated by nearby stations to each of the farms were used.
+
 The time intervals used for this study were: La Rita, week 48 of 2002 to week 17 of the 2015 (647 weeks) and for 28 Miles, week 37 of 2003 to week 18 of 2015 (605 weeks).
-Data preprocessing
-In 28 Miles farm we detect that 1\% of the data were missing, while in La Rita 2.25\% of the data were missing.
-To complete the missing value we use spline interpolation.
-The data collected did not exhibit outliers.
-Due the fact that the variables measure meteorological or biological process, they are discretized in order to reflect trends in the data rather than the specific continuous values.
-The coefficient of variation $C_v (x)$ of each variable x was used to determine the numbers of discretizations with:
-$$n= ⌊100 C_v (x)⌋$$
+
+\subsubsection{Data preprocessing}
+
+In 28 Miles farm, 1\% of the data were missing, while in La Rita was 2.25\%. To fill-in the missing values we use spline interpolation. The data collected did not exhibit outliers.
+
+Due the fact that the variables measure meteorological or biological process, they are discretized in order to reflect trends in the data, i.e.
+the continuous values are not directly used.
+The coefficient of variation $C_v(x)$ of each variable x was used to determine the number $n$ of discretization levels.
+$$n= \lfloor 100 \ C_v(x) \rfloor$$
+where $\lfloor \ \rfloor$  is the round operator.
+
 Each discretization range was uniformly partitioned.
 Besides enabling the capture of tendencies, the discretization removes the effect of small variations in the data collection, either by inaccuracies of the instruments (meteorological variables) or by subjective bias introduced by the human who collects the data (biological warning system).
-Each feature was scaled in a range 0 between 1. The variable to be predicted was not scaled.
+\hl{ESTO DEBE ESTAR DESCRITO EN ALGUNA PARTE}
+
+Each feature was scaled to fit in a range between 0 and 1. The variable to be predicted was not scaled.
 
@@
 
-Although there are many types of indicators to assess the quality of the prediction, we selected the root mean square error (RMSE) and the determination coefficient (R2).
-This decision is supported by the widespread use in machine learning and agriculture areas (Soares, Pasqual and Lacerda (2013); Soares, Pasqual and Lacerda (2014); Ibrahim and Wibowo (2014) and Demir and Bruzzone (2014)).
+Although there are many types of indicators to assess the quality of the prediction, we selected the root mean square error $(RMSE)$ and the determination coefficient $(R^2)$.
 
-\subsection{Methods}
-This research had two phases.\\
-{\bf Phase one } \\
+Given $n$ records, let be $y$ the actual value of the series, $\hat{y}$ the predicted value and $\acute{y}$ the mean of the observed data.
+$$RMSE = \sqrt{ \sum_{i=1}^{n} \frac{{(y-\hat{y})}^2}{n} }$$
 
-In the phase one, we did ten-fold-cross-validation and did a lot of proofs with different machine learning methods and different configuration. 
-We proved with several combinations:
-	Patterns: From one week of observed data to predict the next week until nine weeks before to predict two weeks later.
-	Algorithms: Support vector regression with different kernel functions: linear, RBF (Gaussian) and sigmoid; echo state networks; ordinary least squares linear regression and ridge regression.
-	Variables included in the model. We proved the following combinations:
-	All variables.
-	Only variables that according to expert judgment have more impact on the black Sigatoka development: humidity, precipitation, temperature and wind speed (Marin Vargas and Romero Calderón, 1995).
-	From the four variables listed in the previous paragraph, runs were conducted using each of the variables separately, and combining other runs all the possible pairs of those four variables.\\
-{\bf Phase two } \\
-In the second phase, we used the best configurations obtained en la phase one and did validation with the last 52 and 102 weeks. 
-This second phase pretended to show how these methods behaved on a time of important climate change how are 2014 and 2015 years.
+$$ \bar{y} = \frac{1}{n} \sum_{i=1}^{n} y_i $$
 
-Programming environment
-We use python programming language with the Integrated Development Environment (IDE) Spyder, particularly with libraries: pandas (Comunity, 2014); numpy (numpy.org, 2013); for SVR, ridge and ordinary least squares, we used sklearn (Pedregosa, et al., 2011); and for ESN the python-based code used belongs to Dr.
-Mantas Lukoševičius (2012) from which we made the necessary adjustments for the experiments of this research.
-The computer was a Lenovo ThinkPad, processor Intel(R) Core™ i7-4800MQ CPU @ 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
+$$ S_e^2 = \frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}$$
+
+$$ S_R^2 = \frac{\sum_{i=1}^{n} {(\hat{y}_i-\bar{y}_i)}^2 }{n}$$
+
+$$ S_y^2 = S_R^2 + S_e^2$$
+
+$$ R^2 = \frac{S_R^2}{S_y^2}$$
+
+This decision is supported by the widespread use in machine learning and agriculture areas \citep{Soares2014}, \citep{Soares2013}, \citep{Ibrahim2014} and \citep{Demir2014}.
+
+\subsection{Methodology}
+The selection of methods and their parametrisation was performed in two stages.
+
+{\bf Phase one }
+
+In the phase one, we did ten-fold-cross-validation on a set of  machine learning methods and different configurations:
+
+\begin{itemize}
+\item 	Patterns: n by m, where n from 1 to 8 and m from 1 to 3.
+
+\item Methods: support vector regression with the kernels functions: linear, RBF (Gaussian) and sigmoid; echo state networks; ordinary least squares linear regression and ridge regression.
+
+\item Variables included in the model:
+\begin{itemize}
+\item All variables.
+\item from the set $\{ \overline{T}_{a} , H, \overline{P} , \overline{W}  \}$ use the subsets with one, two or four elements.
+These variables are according to experts the ones having most impact on the disease development \citep{MarinVargas1995}.
+\end{itemize}
+
+\end{itemize}
+
+{\bf Phase two }
+
+In the second phase, the best configurations obtained in phase one are used to validate with the last 52 and 102 weeks.
+
+This second phase intents to expose how these methods behave on a considerable climate in the years 2014 and 2015.
+\marginnote{Poner una cita que fundamente este ultimo parrafo}
+
+\subsection{Programming environment}
+
+We use the python programming language with the Integrated Development Environment (IDE) Spyder \citep{Continuum2015}, particularly with the libraries pandas \citep{mckinneypandas2010} and numpy \citep{vanderWalt2011}.
+For SVR, ridge and ordinary least squares regressions, we used sklearn \citep{scikitlearn2011} and for ESN the python-based code of Dr.
+$Luko\breve{s} evi \breve{c} ius$ \citep{Lukose2012} on which the necessary were done for adjustments for the experiments of this work.
+The computer used a processor Intel(R) Core™ i7-4800MQ CPU @ 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
 
@@
 
+\bibitem[Soares et al., 2013] {Soares2013} Soares , J., Pasqual, M., Lacerda, W.
+(2013).
+Utilization of artificial neural networks in the prediction of the bunches’weight in banana plants.
+Scientia Horticulturae(155), 24-29.
+
+\bibitem[Ibrahim and Wibowo, 2014] {Ibrahim2014}  Ibrahim, N.
+and Wibowo, A.
+(2014).
+Time Series Support Vector Regression with Missing Data Treatment Based Variables Selection for Water Level Prediction of Galas River in Kelantan Malaysia.
+International Journal of Applied Research in Engineering and Science, 3, 25-36.
+
+\bibitem[Demir and Bruzzone, 2014] {Demir2014} Demir, B.
+and Bruzzone, L.
+(2014).
+A multiple criteria active learning method for support vector regression.
+Pattern Recognition, 2558–2567.
+doi:10.1016/j.patcog.2014.02.001
+
+\bibitem[Continuum Analitycs, 2015] {Continuum2015} Continuum Analitycs. (2015). Anaconda. Retrieved from https://www.continuum.io/
+
+\bibitem[McKinney, 2010] {mckinneypandas2010} McKinney W. (2010). Data Structures for Statistical Computing in Python, Proceedings of the 9th Python in Science Conference, 51-56
+
+\bibitem[van der Walt et al., 201] {vanderWalt2011} van der Walt S., Colbert C.
+and Varoquaux G.
+(2011).
+The NumPy Array: A Structure for Efficient Numerical Computation, Computing in Science \& Engineering, 13, 22-30 (2011), DOI:10.1109/MCSE.2011.37
+
 \end{thebibliography}

### 2015-11-24 Paper Sigatoka - 2015.tex
Lunes 23 noche, cambios en tema de sigatoka

@@
 
-According to studies by the National Banana Corporation of Costa Rica
-(Corbana) made in 2013, considering on average between 53 thru 57
-cycles of fungicide applications per farm, the cost per hectare per
-year ranged between \$1800 USD and \$1900 USD. This represents
-about 0.76 cents of the price of a box of 18.14 kilograms. Overall,
-this represents 10\% to 12\% of the total production cost
-\citet{Bresciani2015}.
-\marginnote{Hay que completar la cita de Brescani, o cambiarla}
+In Costa Rica for the control of black Sigatoka is necessary the application of chemical fungicides.
+Depending of the zone of production and the weather conditions 45 – 55 cycles/year of fungicides is needed to keep this disease under control and to produce fruit with quality for exportation.
+This represent a cost per hectare per year ranged between \$1600 USD and \$2000 USD, about 0,64 - 0.80 cents of the cost of production of a box of 18.14 kilograms.
+Overall, this represents 10\% to 12\% of the total production cost.
 
@@
 
-This system measures the disease development state to determine when to
-apply fungicides \citep{MarinVargas1995}.  This system is based on two
-components: a climate component, which is given by the Piche
-evaporation and a biological component, given by the stage of progress
-or the rate of disease development. Originally, this system was
-designed to work with young plants. One selected plant must exhibit a
-normal growth and be in a place that enforces a healthy
-development. The plant must start with 5 to 6 true leaves. The
-assessments are made at fixed intervals of seven days as long as
-possible, on the same plant. The first observations should consider
-the leaf emission, also the level of infection on the leaves should be
-evaluated considering the stages of development
-\citep{MarinVargas1995}.
-\marginnote{Tendra sentido ampiar esta parte?}
+The early warning system for black Sigatoka is an adaptation of the yellow Sigatoka warning system developed by Ganry and Meyer and modified by Ganry and Laville to use for controlling yellow Sigatoka in Cameroon.
+Ternesien and Fouré  later improved Ganry and Laville's system.
+The latter system is based on weekly observations of disease symtoms on young leaves of the plant, according to Fouré's symptom (stages) descriptions.
+Arbitrary coefficientes, based on incidence and severity of disease development, are used to calculate two variables: gross sum and state of evolution.
+Gross sum is based on the stage present and an arbitrary coefficient, which increases with the advance of the symptoms and the juvenility of the leaf.
+The state of evolution is calculated using the gross sum and the foliar emission period.
+Although threshold levels were initially suggested as a guide to spray timing, the fluctuation of these two variables was found to better define appropiate times to spray \citep{Marinetal2003}.
 
@@
 
-\bibitem[Brescani, 2015]{Bresciani2015} Brescani, XXXXX.
-
 \bibitem[Camargo et al.,2012]{Camargo2012} Camargo, A., Molina, J., Cadena-Torres, J., Jim\'enez, N., Kim, J.
@@
 \bibitem[Marin and Romero, 1995]{MarinVargas1995} Marin Vargas, D., Romero Calderón, R. 1995. El combate de la Sigatoka Negra. Bolet\'in Departamento de Investigaciones, Corbana Costa Rica.
+
+\bibitem[Marin et al., 2003]{Marinetal2003} Marin, D., Romero, R., Guzman, M, Sutton, T. 2003. Black Sigatoka: An increasing threat to banana cultivation. Plant Disease, 87(3), 208-222.
 
@@
 and Duchesnay, E.
-(2011) Scikit-learn: Machine Learning in {P}ython.
+(2011) Scikit-learn: Machine Learning in Python.
 Journal of Machine Learning Research, vol.
@@
 Procedia Environmental Sciences, 6, 35-49.
-
-\bibitem[Huang et al., 2010] {Huang2010}  Huang, Y., Lan, Y., Thomson, S., Fang, A., Hoffmann, W., Lacey, R.
-(2010).
-Development of soft computing and applications in agricultural and biological engineering.
-Computers and Electronics in Agriculture,(71(2)), 107–127.
-doi:10.1016/j.compag.
 

### 2015-11-24 Paper Sigatoka - 2015.tex
Corregidas unas tildes

@@
 The early warning system for black Sigatoka is an adaptation of the yellow Sigatoka warning system developed by Ganry and Meyer and modified by Ganry and Laville to use for controlling yellow Sigatoka in Cameroon.
-Ternesien and Fouré  later improved Ganry and Laville's system.
-The latter system is based on weekly observations of disease symtoms on young leaves of the plant, according to Fouré's symptom (stages) descriptions.
+Ternesien and Four\'e  later improved Ganry and Laville's system.
+The latter system is based on weekly observations of disease symtoms on young leaves of the plant, according to Four\'e's symptom (stages) descriptions.
 Arbitrary coefficientes, based on incidence and severity of disease development, are used to calculate two variables: gross sum and state of evolution.
@@
 
-which ignores the error if the difference between the prediction value and the actual value is smaller than $ \epsilon$.
-The $ \epsilon -insensitive \ loss \ function$ allows to find the coefficients $w$ and $b$ by solving a convex optimization problem, which balances the empirical error and the generalization ability.
+which ignores the error if the difference between the prediction value and the actual value is smaller than $\epsilon$.
+The $\epsilon -insensitive \ loss \ function$ allows to find the coefficients $w$ and $b$ by solving a convex optimization problem, which balances the empirical error and the generalization ability.
 In SVR, the empirical error is measured by the loss function є-insensitive and the generalization ability is measured by the Euclidean norm of $w$ \citep{XXXX201X}.

### 2015-11-29 Paper Sigatoka - 2015.tex
2015-11-28 Algunos cambios mas la inclusion de elastic-net

@@
 In Costa Rica for the control of black Sigatoka is necessary the application of chemical fungicides.
-Depending of the zone of production and the weather conditions 45 – 55 cycles/year of fungicides is needed to keep this disease under control and to produce fruit with quality for exportation.
+Depending of the zone of production and the weather conditions 45 - 55 cycles/year of fungicides is needed to keep this disease under control and to produce fruit with quality for exportation.
 This represent a cost per hectare per year ranged between \$1600 USD and \$2000 USD, about 0,64 - 0.80 cents of the cost of production of a box of 18.14 kilograms.
@@
 
-In this work, we compare four machine learning techniques (support vector
-regression (SVR), echo state networks (ESN), ridge regression and
+In this work, we compare five machine learning techniques (support vector
+regression (SVR), echo state networks (ESN), ridge regression, elastic-net regression and
 ordinary least squares linear regression) to predict the development rate of the black Sigatoka disease.
@@
 \begin{aligned}
-& \underset{}{\text{miimize}}
+& \underset{}{\text{minimize}}
 & & J(w,\xi_i , \xi_i^* ) = \frac{1}{2}   \Bigr| \Bigr| w \Bigr| \Bigr|^2 + C \sum_{i=1}^{n} (\xi_i , \xi_i^* )    \\
@@
 The ridge coefficients minimize a penalized residual sum of squares  \citep{scikitlearn2011}:
+
 $$\min_{w} { \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
+
 Here, $\alpha \ \textgreater \  0$ is a complexity parameter that controls the amount of shrinkage: the larger the value of $\alpha$, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
+
+\subsubsection{Elastic-Net regression}
+Elastic-Net is a linear regression model trained with $L1$ and $L2$ prior as regularizer.
+This combination allows for learning a sparse model where few of the weights are non-zero like Lasso, while still maintaining the regularization properties of Ridge  \citep{scikitlearn2011}.
+The convex combination of $L1$ and $L2$ is controled by using the $l1_ratio$ parameter.
+
+Elastic-Net is useful when there are multiple features which are correlated with one another. Lasso is likely to pick one of these at random, while elastic-net is likely to pick both.
+A practical advantage of trading-off between Lasso and Ridge is it allows Elastic-Net to inherit some of Ridge’s stability under rotation.
+The objective function to minimize is \citep{scikitlearn2011}:
+
+$$\min_{w} { \frac{1}{2n_{samples}} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \rho \Bigr| \Bigr| w \Bigr| \Bigr|_1 + \frac{\alpha (1- \rho)}{2} \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
 
@@
 
-Although there are many types of indicators to assess the quality of the prediction, we selected the root mean square error $(RMSE)$ and the determination coefficient $(R^2)$.
+Although there are many types of indicators to assess the quality of the prediction, we selected the determination coefficient $(R^2)$.
 
 Given $n$ records, let be $y$ the actual value of the series, $\hat{y}$ the predicted value and $\acute{y}$ the mean of the observed data.
-$$RMSE = \sqrt{ \sum_{i=1}^{n} \frac{{(y-\hat{y})}^2}{n} }$$
 
@@
 \begin{itemize}
-\item 	Patterns: n by m, where n from 1 to 8 and m from 1 to 3.
+\item 	Patterns: n by m, where n from 1 to 8 and m from 1 to 2.
 
-\item Methods: support vector regression with the kernels functions: linear, RBF (Gaussian) and sigmoid; echo state networks; ordinary least squares linear regression and ridge regression.
+\item Methods: support vector regression with the kernels functions: linear, RBF (Gaussian) and sigmoid; echo state networks; ordinary least squares linear regression, ridge regression and elastic-net regression.
 

### 2016-01-17 Paper Sigatoka - 2015.tex
Cambios en formato del los archivos para facilitar el rastreo de cambios.  Una que otra fórmula se editó para facilitar el formato.  Se agregó el codec de entrada UTF-8 y se aseguró que el archivo esté en UTF-8 para poder usar los acentos normales.  Algunos errores reportados por LaTeX fueron corregidos.

@@
 \usepackage{lineno,hyperref}
-
+\usepackage[utf8]{inputenc}  %% Permite utilizar tildes si archivo usa utf-8
 \usepackage{amsmath}
@@
 
-\author[afiCorbana] {Mauricio Guzm\'an-Quesada}
-\author[afiCorbana]{Jos\'e-Antonio Guzm\'an-Alvarez}
+\author[afiCorbana] {Mauricio Guzmán-Quesada}
+\author[afiCorbana]{José-Antonio Guzmán-Alvarez}
 \author[afiPablo]{Pablo Alvarado-Moya}
 
-\address[afiLuisAlex]{DOCINADE, Instituto Tecnol\'ogico de Costa Rica, 
+\address[afiLuisAlex]{DOCINADE, Instituto Tecnológico de Costa Rica, 
 Computer Research Center, Multidisciplinar program eScience, 
@@
 
-\address[afiCorbana]{Direcci\'on de Investigaciones, Corporaci\'on Bananera Nacional S.A., Gu\'apiles, Costa Rica}
+\address[afiCorbana]{Dirección de Investigaciones, Corporación Bananera Nacional S.A., Guápiles, Costa Rica}
 
-\address[afiPablo]{DOCINADE, Instituto Tecnol\'ogico de Costa Rica, Cartago, Costa Rica}
+\address[afiPablo]{DOCINADE, Instituto Tecnológico de Costa Rica, Cartago, Costa Rica}
 
@@
 \begin{keyword}
-\texttt{Machine learning \sep Black Sigatoka \sep Support vector regression \sep
-Banana disease prediction \sep Biological warning system }
+Machine learning \sep Black Sigatoka \sep Support vector regression \sep
+Banana disease prediction \sep Biological warning system 
 \end{keyword}
@@
 
-\marginnote{La bibliografia debe ser autor, anno}
+\marginnote{La bibliografia debe ser autor, año}
 
-The black Sigatoka disease caused by the fungus {\it Mycosphaerella
+The black Sigatoka disease caused by the fungus \emph{Mycosphaerella
   fijiensis Morelet} is the major pathological problem of banana and
@@
 This disease attacks the plant leaves producing a rapid deterioration
-of the leaf area, affects the growth and productivity of the plants due to the impairment of their photosinthetic ability  causes a reduction in the quality
-of the fruit, and promotes premature maturation of bunches, which is
-the major cause of product losses associated with the black Sigatoka. \figurename
-$.$\ref{figura1} shows three stages of this disease.
+of the leaf area, affects the growth and productivity of the plants
+due to the impairment of their photosinthetic ability, causes a
+reduction in the quality of the fruit, and promotes premature
+maturation \marginnote{maturation or ripening?} of bunches, which is
+the major cause of product losses associated with the black
+Sigatoka. \figurename~\ref{fig:diseasestages} shows three stages of
+this disease.
 
@@
 \begin{figure}[h] 
-\begin{subfigure}{.3\textwidth}
 \centering
-  \includegraphics[width=.8\linewidth]{Roya_a}
-  \caption{}
-  \label{fig:sfig1}
-\end{subfigure}
-\begin{subfigure}{.3\textwidth}
-  \centering
-  \includegraphics[width=.8\linewidth]{Roya_b}
-  \caption{}
-  \label{fig:sfig2}
-\end{subfigure}
-\begin{subfigure}{.3\textwidth}
-  \centering
-  \includegraphics[width=.8\linewidth]{Roya_c}
-  \caption{}
-  \label{fig:sfig3}
-\end{subfigure}
-\caption{Examples of three disease stages of the black Sigatoka. (a) Initial stage. (b) Intermediate stage, and (c) Advanced stage.} 
-\label{figura1} 
+\begin{tabular}{c@{\;}c@{\;}c}
+  \includegraphics[width=.32\linewidth]{Roya_a} &
+  \includegraphics[width=.32\linewidth]{Roya_b} &
+  \includegraphics[width=.32\linewidth]{Roya_c} \\
+  (a) & (b) & (c) 
+\end{tabular}
+\caption{Examples of three disease stages of the black Sigatoka. (a)~Initial stage. (b)~Intermediate stage, and (c)~Advanced stage.} 
+\label{fig:diseasestages} 
 \end{figure}
 
-In Costa Rica for the control of black Sigatoka is necessary the application of chemical fungicides.
-Depending of the zone of production and the weather conditions 45 - 55 cycles/year of fungicides is needed to keep this disease under control and to produce fruit with quality for exportation.
-This represent a cost per hectare per year ranged between \$1600 USD and \$2000 USD, about 0,64 - 0.80 cents of the cost of production of a box of 18.14 kilograms.
-Overall, this represents 10\% to 12\% of the total production cost.
+In Costa Rica the control of black Sigatoka makes use of chemical fungicides.
+%
+Depending on the zone of production and the weather conditions, 45--55
+cycles/year of fungicide applications are required to keep this
+disease under control and to produce the fruit quality expected in the
+international markets.
+%
+This represent a cost per hectare per year in the range between US\$1600 and
+US\$2000; about 0.64--0.80 cents of the production costs for a
+18.14\,kg box.
+%
+Overall, this represents 10\%--12\% of the total production cost.
 
@@
 to predict its future behavior and to determine whether
-particular fungicide spray schedules will be able to effectively and
+a particular fungicide spray program will be able to effectively and
 economically control the disease \citet{ChuangJeger1987}.
 
+\marginnote{Esto es parte del ``related works''}
+%
 There are efforts to apply machine learning methods for
@@
 system for the assessment of crop disorders, \cite{Huang2010}
-introduce a plant virus identification method based on neural networks with an evolutionary preprocessing stage, \cite{Kim2014} summarize in
+introduce a plant virus identification method based on neural networks
+with an evolutionary preprocessing stage, \cite{Kim2014} summarize in
 their survey crop pests prediction methods using regression and
@@
 
-In this work, we compare five machine learning techniques (support vector
-regression (SVR), echo state networks (ESN), ridge regression, elastic-net regression and
-ordinary least squares linear regression) to predict the development rate of the black Sigatoka disease.
+In this work, we compare five machine learning techniques (support
+vector regression (SVR), echo state networks (ESN), ridge regression,
+elastic-net regression and ordinary least squares linear regression)
+to predict the development rate of the black Sigatoka disease.
 
@@
 \subsection{Concepts}
+
 \subsubsection{Black Sigatoka disease}
 
-Black Sigatoka, disease caused by the fungus Mycosphaerella fijiensis Morelet, is the main problem phytopathologic of banana and plantain crops in Central America \citep{MarinVargas1995}.
+Black Sigatoka, disease caused by the fungus Mycosphaerella fijiensis
+Morelet, is the main problem phytopathologic of banana and plantain
+crops in Central America \citep{MarinVargas1995}.
 
@@
 The early warning system for black Sigatoka is an adaptation of the yellow Sigatoka warning system developed by Ganry and Meyer and modified by Ganry and Laville to use for controlling yellow Sigatoka in Cameroon.
-Ternesien and Four\'e  later improved Ganry and Laville's system.
-The latter system is based on weekly observations of disease symtoms on young leaves of the plant, according to Four\'e's symptom (stages) descriptions.
+Ternesien and Fouré  later improved Ganry and Laville's system.
+%
+The latter system is based on weekly observations of disease symtoms on young leaves of the plant, according to Fourés symptom (stages) descriptions.
 Arbitrary coefficientes, based on incidence and severity of disease development, are used to calculate two variables: gross sum and state of evolution.
+%
 Gross sum is based on the stage present and an arbitrary coefficient, which increases with the advance of the symptoms and the juvenility of the leaf.
 The state of evolution is calculated using the gross sum and the foliar emission period.
+%
 Although threshold levels were initially suggested as a guide to spray timing, the fluctuation of these two variables was found to better define appropiate times to spray \citep{Marinetal2003}.
@@
 
-SVR uses the $ \epsilon -insensitive \ loss \ function$:
-
-\[l= {\Bigr| y - f(s) \Bigr| }_{\epsilon } = 
-\left\{\begin{array}{lcl}
-	0 & \mbox{} & {\Bigr| y - f(s) \Bigr| } \leq \epsilon \\
-	 &  & \\
-	{\Bigr| y - f(s) \Bigr| - \epsilon} & \mbox{} & else\\ \end{array} \right. \]
-
+SVR uses the $\epsilon$-insensitive loss function:
+\begin{equation*}
+l= {\left| y - f(s) \right| }_{\epsilon } = 
+\begin{cases}
+   0 & {\left| y - f(s) \right| } \leq \epsilon \\
+   \left| y - f(s) \right| - \epsilon & \text{otherwise}
+\end{cases}
+\end{equation*}
 which ignores the error if the difference between the prediction value and the actual value is smaller than $\epsilon$.
-The $\epsilon -insensitive \ loss \ function$ allows to find the coefficients $w$ and $b$ by solving a convex optimization problem, which balances the empirical error and the generalization ability.
-In SVR, the empirical error is measured by the loss function є-insensitive and the generalization ability is measured by the Euclidean norm of $w$ \citep{XXXX201X}.
-Then, the optimization problem to identify the regression model can be formulated by \citep{Wei2013}:
-\marginnote{Falta completar la cita XXXXX con el libro de SVR}
+%
+The $\epsilon$-insensitive loss function allows to find the coefficients $w$ and $b$ by solving a convex optimization problem, which balances the empirical error and the generalization ability.
+%
+In SVR, the empirical error is measured by the loss function $\epsilon$-insensitive and the generalization ability is measured by the Euclidean norm of $w$ \citep{XXXX201X}.
+%
+Then, the optimization problem to identify the regression model can be
+formulated by \citep{Wei2013}: \marginnote{Falta completar la cita
+  XXXXX con el libro de SVR}
 
@@
  \includegraphics[scale=.9]{SVR}
- \caption{$\epsilon -insensitive \ loss \ function \citep{Wei2013}.$} 
+ \caption{$\epsilon$-insensitive loss function \citep{Wei2013}.} 
  \label{figura2} 
@@
 Mathematically it solves a problem of the form \citep{ scikit-learn2011}:
-
 $$\min_{w} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  $$
@@
 
-\bibitem[Camargo et al.,2012]{Camargo2012} Camargo, A., Molina, J., Cadena-Torres, J., Jim\'enez, N., Kim, J.
+\bibitem[Camargo et al.,2012]{Camargo2012} Camargo, A., Molina, J., Cadena-Torres, J., Jiménez, N., Kim, J.
 2012.
@@
 
-\bibitem[Marin and Romero, 1995]{MarinVargas1995} Marin Vargas, D., Romero Calderón, R. 1995. El combate de la Sigatoka Negra. Bolet\'in Departamento de Investigaciones, Corbana Costa Rica.
+\bibitem[Marin and Romero, 1995]{MarinVargas1995} Marin Vargas, D., Romero Calderón, R. 1995. El combate de la Sigatoka Negra. Boletín Departamento de Investigaciones, Corbana Costa Rica.
 

### 2016-02-17 Paper Sigatoka - 2015.tex
2016-02-16 10pm

@@
 \usepackage{marginnote}
+
+%  \usepackage{epstopdf} 
+%  \usepackage{epsfig}
 
@@
 \hline\hline 
-$T_{a_max}$ &	Max air temperature \\
-$T_{a_min}$ &	Min air temperature \\
+$T_{a_{max}}$ &	Max air temperature \\
+$T_{a_{min}}$ &	Min air temperature \\
 $\overline{T}_{a}$	 & Mean air temperature \\
-$H$	&  Humidity \\
-$H_min$ &	Min humidity \\
-$H_max$	& Max humidity \\
-$R$	& Solar radiation \\
-$\overline{P}$	& Mean precipitation  \\
-$W_max$	& Max speed wind \\
+$\overline{H}$	&  Mean Humidity \\
+$H_{min}$ &	Min humidity \\
+$H_{max}$	& Max humidity \\
+$\overline{R}$	& Mean Solar radiation \\
+$P$	& Sum precipitation  \\
+$W_{max}$	& Max speed wind \\
 $\overline{W}$	& Mean speed wind  \\
@@
 
-Although there are many types of indicators to assess the quality of the prediction, we selected the determination coefficient $(R^2)$.
+Although there are many types of indicators to assess the quality of the prediction, we selected the determination coefficient $(R^2)$ and the Root Mean Square Error $(RMSE)$.
 
@@
 
+$$ RMSE = \sqrt{\frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}}$$
+
 This decision is supported by the widespread use in machine learning and agriculture areas \citep{Soares2014}, \citep{Soares2013}, \citep{Ibrahim2014} and \citep{Demir2014}.
@@
 
-\item Methods: support vector regression with the kernels functions: linear, RBF (Gaussian) and sigmoid; echo state networks; ordinary least squares linear regression, ridge regression and elastic-net regression.
+\item Methods: support vector regression with the kernels functions: linear, gaussian and sigmoid; echo state networks; ordinary least squares linear regression, ridge regression and elastic-net regression.
 
@@
 \item All variables.
-\item from the set $\{ \overline{T}_{a} , H, \overline{P} , \overline{W}  \}$ use the subsets with one, two or four elements.
+\item From the set $\{ \overline{T}_{a} , \overline{H}, P , \overline{W}  \}$ use the subsets with one, two or four elements.
 These variables are according to experts the ones having most impact on the disease development \citep{MarinVargas1995}.
@@
 \section{Results}
+
+In this section we present the main results for phase.
+
+{\bf Phase one }
+
+\figurename $.$\ref{figura4} shows the best $R^2$ for each algorithm in the experiment. Results are group by farm.
+
+\begin{figure}[h] 
+ \centering
+ \includegraphics[scale=.3]{Phase_one_Best_R2_for_each_algortihm}
+ \caption{Phase one - Best $R^2$ for each algorithm} 
+ \label{figura4} 
+\end{figure}
+
+\figurename $.$\ref{figura5} presents, for one, two and three weeks ahead, the best $R^2$. Results are group by farm.
+
+\begin{figure}[h] 
+ \centering
+ \includegraphics[scale=.3]{Phase_one_Best_R2_for_each_number_of_observation_range_in_the_patterm}
+ \caption{Phase one - Best $R^2$ for each algorithm} 
+ \label{figura5} 
+\end{figure}
+
+\figurename $.$\ref{figura6} shows the best $R^2$ for each variables combination. Results are group by farm.
+
+\begin{figure}[h] 
+ \centering
+ \includegraphics[scale=.3]{Phase_one_Best_R2_for_variables}
+ \caption{Phase one - Best $R^2$ for each variable combination} 
+ \label{figura6} 
+\end{figure}
+
+\figurename $.$\ref{figura7} shows the pareto frontier with respect to $R^2$ and $RMSE$. Results are group by farm.
+
+\begin{figure}[h] 
+ \centering
+ \includegraphics[scale=.6]{Phase_one_R2_RMSE}
+ \caption{Phase one - Pareto frontier for $R^2$ and $RMSE$} 
+ \label{figura7} 
+\end{figure}
+
+{\bf Phase two }
+
+In the second phase, the best configurations obtained in phase one are used to validate with the last 52 and 102 weeks.
 

### 2016-02-21 Paper Sigatoka - 2015.tex
2016-02-20

@@
 \usepackage{marginnote}
+\usepackage{float}
 
@@
 
-\begin{figure}[h] 
+\begin{figure}[H] 
  \centering
- \includegraphics[scale=.3]{Phase_one_Best_R2_for_each_algortihm}
+ \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_algortihm}
  \caption{Phase one - Best $R^2$ for each algorithm} 
@@
 
-\begin{figure}[h] 
+\begin{figure}[H] 
  \centering
- \includegraphics[scale=.3]{Phase_one_Best_R2_for_each_number_of_observation_range_in_the_patterm}
+ \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_number_of_observation_range_in_the_patterm}
  \caption{Phase one - Best $R^2$ for each algorithm} 
@@
 
-\begin{figure}[h] 
+\begin{figure}[H] 
  \centering
- \includegraphics[scale=.3]{Phase_one_Best_R2_for_variables}
+ \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
  \caption{Phase one - Best $R^2$ for each variable combination} 
@@
 
-\figurename $.$\ref{figura7} shows the pareto frontier with respect to $R^2$ and $RMSE$. Results are group by farm.
+\figurename $.$\ref{figura7} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
 
-\begin{figure}[h] 
+\begin{figure}[H] 
  \centering
- \includegraphics[scale=.6]{Phase_one_R2_RMSE}
+ \includegraphics[scale=.8]{Phase_one_R2_RMSE}
  \caption{Phase one - Pareto frontier for $R^2$ and $RMSE$} 
@@
 
+The Pareto frontier for the La Rita farm is composed by 96 elements. The \tablename $.$\ref{tabla2} shows the composition about variables and observation ranges.
+
+\begin{table}[h] 
+\caption{Composition of the Pareto frontier - La Rita - Phase one} 
+\label{tabla2} 
+\centering
+\begin{tabular}{c|c|c|c|c} 
+\hline
+\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
+\hline\hline 
+Pair $\overline{T}_{a}$ $\overline{W}$ &	1 to 1  & 36 & 64.25\% & 714.51 \\
+ &	2 to 1  & 6 & 62.97\% & 695.10 \\
+\hline 
+All  & 1 to 1  & 18 & 62.98\% & 701.95 \\
+   & 2 to 1  & 12 & 61.76\% & 679.92 \\
+    & 3 to 1  & 6 & 60.60\% & 676.42 \\
+    & 5 to 1  &  2 & 60.37\% & 672.39 \\
+\hline    
+$\overline{T}_{a}$ & 1 to 1  & 12  & 63.60\% & 708.77 \\
+       &	2 to 1  & 4 & 62.23\% & 689.55 \\
+\hline
+\end{tabular} 
+\end{table}
+
+Similarly, the Pareto frontier for the 28 Millas farm is composed by 75 elements. The \tablename $.$\ref{tabla3} shows the composition about variables and observation ranges.
+
+\begin{table}[h] 
+\caption{Composition of the Pareto frontier - 28 Millas - Phase one} 
+\label{tabla3} 
+\centering
+\begin{tabular}{c|c|c|c|c} 
+\hline
+\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
+\hline\hline 
+Pair $\overline{T}_{a}$ $\overline{W}$ & 1 to 1 & 8 & 57.80\% & 438.09 \\
+\hline 
+All   &	9 to 1 & 2 & 50.93\% & 397.93 \\
+  & 10 to 1	 & 2 & 50.97\% & 398.81 \\
+  &	8 to 1 & 6 & 51.62\% & 398.93 \\
+  &	7 to 1 & 2 & 52.25\% & 400.28 \\
+  &	6 to 1 & 2 & 53.16\% & 404.14 \\
+  &	4 to 1 & 2 & 54.32\% & 407.54 \\
+\hline    
+$\overline{T}_{a}$ & 1 to 1  & 8  & 59.09\% & 439.44 \\
+\hline
+Pair $\overline{T}_{a}$ $\overline{H}$ & 1 to 1	 & 8 & 57.51\% & 428.61 \\
+ &	2 to 1 & 20 & 56.91\% & 414.37 \\
+ &	3 to 1 & 3 & 54.41\% & 411.55 \\
+ &	4 to 1 & 3 & 53.34\% & 406.65 \\
+\hline
+Pair $\overline{T}_{a}$ $P$ & 3 to 1 & 9 & 56.23\% & 422.76 \\
+\hline
+\end{tabular} 
+\end{table}
+
 {\bf Phase two }
 
-In the second phase, the best configurations obtained in phase one are used to validate with the last 52 and 102 weeks.
+In the second phase, the best configurations obtained in phase one were used to validate with the last 50 and 100 weeks.
+
+\figurename $.$\ref{figura8} shows the best $R^2$ for each algorithm in the experiment. Results are group by farm.
+
+\begin{figure}[H] 
+ \centering
+ \includegraphics[scale=.5]{coming}
+ \caption{Phase two - Best $R^2$ for each algorithm} 
+ \label{figura8} 
+\end{figure}
+
+\figurename $.$\ref{figura9} presents, for one, two and three weeks ahead, the best $R^2$. Results are group by farm.
+
+\begin{figure}[H] 
+ \centering
+ \includegraphics[scale=.5]{coming}
+ \caption{Phase two - Best $R^2$ for each algorithm} 
+ \label{figura9} 
+\end{figure}
+
+\figurename $.$\ref{figura10} shows the best $R^2$ for each variables combination. Results are group by farm.
+
+\begin{figure}[H] 
+ \centering
+ \includegraphics[scale=.5]{coming}
+ \caption{Phase two - Best $R^2$ for each variable combination} 
+ \label{figura10} 
+\end{figure}
+
+\figurename $.$\ref{figura11} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
+
+\begin{figure}[H] 
+ \centering
+ \includegraphics[scale=.8]{coming}
+ \caption{Phase two - Pareto frontier for $R^2$ and $RMSE$} 
+ \label{figura11} 
+\end{figure}
 

### 2016-02-21 2016-02-15-Resume archivos de validación.py

@@ -1,0 +1,103 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Mon Nov 23 10:42:31 2015
+
+@author: luis
+"""
+
+import pandas as pd
+import numpy as np
+import os
+import math
+
+
+tTamTotal1 = 55
+tTamUsar1 = 50
+tTamTotal2 = 105
+tTamUsar2 = 100
+tDirectorio = "D:\Cambios\Doctorado\CORRIDAS DEL PROGRAMA\Corbana\Sigatoka - Semanal - EstadoEvolucion\\2016-02-17"
+tArchivos = [   ["2016-02-16-Sigatoka - La Rita=01= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - La Rita=02= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - La Rita=03= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - La Rita=04= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=05= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - 28 Millas=06= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - 28 Millas=07= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=08= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=09= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - 28 Millas=10= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=11= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - 28 Millas=12= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=13= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - 28 Millas=14= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=15= - 55",tTamTotal2, tTamUsar2],
[…]

### 2016-02-21 2016-02-19-Graficar Dispersion.py

@@ -1,0 +1,172 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Feb 04 18:38:48 2016
+
+@author: lcalvo
+"""
+
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as pl
+import operator
+
+
+#########################################################################    
+def guarde_Frente_Pareto(pDatos, pTipos, maxX, maxY, pCategX, pCategY, pCol):
+# maxX : Si True es porque se quiere maximizar, si False se desea minimizar
+# maxY : Si True es porque se quiere maximizar, si False se desea minimizar    
+    global gNombreGlobal  
+    for tL in pTipos:
+        tParte = pDatos[(pDatos[pCol] == tL )]
+        if len(tParte) > 0:    
+            myListX =  tParte.sort([pCategX], ascending=(not maxX)) 
+            myListY =  tParte.sort([pCategY], ascending=(not maxY)) 
+            p_front= pd.DataFrame({})
+            if maxX:
+                p_front = myListY.head(1)
+                orden = myListY
+            else:
+                p_front = myListX.head(1)
+                orden = myListX                
+            tLargo = len(orden)
+            for pair in range(1,tLargo):
+                if maxY:
+                    if maxX:
+                        if (orden.iloc[pair][pCategX] >= p_front.tail(1).iloc[-1][pCategX]) : 
+                            p_front = p_front.append(orden.iloc[pair]) 
+                    else: # if maxX
+                        if (orden.iloc[pair][pCategY] >= p_front.tail(1).iloc[-1][pCategY]):                             
+                            p_front = p_front.append(orden.iloc[pair])                 
+                else: # if maxY
+                    if maxX:
+                        if (orden.iloc[pair][pCategY] <= p_front.tail(1).iloc[-1][pCategY]):
+                            p_front = p_front.append(orden.iloc[pair])                     
+                    else: # if maxX
+                        if (orden.iloc[pair][pCategX] <= p_front.tail(1).iloc[-1][pCategX]): 
+                            p_front = p_front.append(orden.iloc[pair]) 
+        else:
+            print("No se pudo calcular frente de pareto para: ", tL)
[…]

### 2016-02-21 2016-02-19-Graficar Frente de Pareto.py

@@ -1,0 +1,175 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Feb 04 18:38:48 2016
+
+@author: lcalvo
+"""
+
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as pl
+import operator
+import os
+
+
+#########################################################################    
+def guarde_Frente_Pareto(pDatos, pTipos, maxX, maxY, pCategX, pCategY, pCol):
+# maxX : Si True es porque se quiere maximizar, si False se desea minimizar
+# maxY : Si True es porque se quiere maximizar, si False se desea minimizar    
+    global gNombreGlobal  
+    for tL in pTipos:
+        tParte = pDatos[(pDatos[pCol] == tL )]
+        if len(tParte) > 0:    
+            myListX =  tParte.sort([pCategX], ascending=(not maxX)) 
+            myListY =  tParte.sort([pCategY], ascending=(not maxY)) 
+            p_front= pd.DataFrame({})
+            if maxX:
+                p_front = myListY.head(1)
+                orden = myListY
+            else:
+                p_front = myListX.head(1)
+                orden = myListX                
+            tLargo = len(orden)
+            for pair in range(1,tLargo):
+                if maxY:
+                    if maxX:
+                        if (orden.iloc[pair][pCategX] >= p_front.tail(1).iloc[-1][pCategX]) : 
+                            p_front = p_front.append(orden.iloc[pair]) 
+                    else: # if maxX
+                        if (orden.iloc[pair][pCategY] >= p_front.tail(1).iloc[-1][pCategY]):                             
+                            p_front = p_front.append(orden.iloc[pair])                 
+                else: # if maxY
+                    if maxX:
+                        if (orden.iloc[pair][pCategY] <= p_front.tail(1).iloc[-1][pCategY]):
+                            p_front = p_front.append(orden.iloc[pair])                     
+                    else: # if maxX
+                        if (orden.iloc[pair][pCategX] <= p_front.tail(1).iloc[-1][pCategX]): 
+                            p_front = p_front.append(orden.iloc[pair]) 
+        else:
[…]

### 2016-02-21 Resultados - Paper/Fase 2/Prueba graficacion intervalos.py

@@ -1,0 +1,109 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Mon Feb 01 17:14:08 2016
+
+@author: lcalvo
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+from sklearn.ensemble import GradientBoostingRegressor
+from sklearn.cross_validation import cross_val_score
+from sklearn.ensemble import RandomForestClassifier
+from sklearn.ensemble import ExtraTreesClassifier
+from sklearn.tree import DecisionTreeClassifier
+
+###############################################################################
+# Calcula el R2  - Coeficiente de determinación
+def calculeR2(pXorig, pYpred):
+    tn = len(pXorig)
+    ybar = np.sum(pXorig)/ tn
+    sse = np.sum((pXorig - pYpred)**2)  # Suma de cuadrados del error o residual
+    ssr = np.sum((pYpred - ybar)**2)  # Suma de Cuadrados debido a la regresión
+    sst = sse + ssr   # Suma de cuadrados total
+    results =  (ssr / sst)       
+    return results
+
+def demeLista(l):
+    res = []
+    for e in l:
+        res.append(e)
+    return res
+    
+    
+    
+np.random.seed(1)
+nValidar = 52
+
+todos = np.genfromtxt('Sigatoka.csv', delimiter=';', skip_header =1)
+todos = todos.astype(np.float32)
+
+X = todos[1:-nValidar,:-1]
+# Observations
+y = todos[1:-nValidar,-1]
+xx = todos[-nValidar:,:-1]
+realesxx = todos[-nValidar:,-1]
+
+
+alpha = 0.95
+
+clf = GradientBoostingRegressor(loss='quantile', alpha=alpha,
+                                n_estimators=850, max_depth=3,
+                                learning_rate=.1, min_samples_leaf=9,
+                                min_samples_split=9)
+clf.fit(X, y)
+# Make the prediction on the meshed x-axis
+y_upper = clf.predict(xx)
+
+clf.set_params(alpha=1.0 - alpha)
+clf.fit(X, y)
+
+# Make the prediction on the meshed x-axis
+y_lower = clf.predict(xx)
+
+
+clf.set_params(loss='ls')
+clf.fit(X, y)
+
+# Make the prediction on the meshed x-axis
+y_pred = clf.predict(xx)
+
[…]

### 2016-02-21 Paper Sigatoka - 2015.tex
2016-02-21 5:30pm

@@
  \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase one - Best $R^2$ for each algorithm} 
+ \caption{Phase one - Best $R^2$ for each observation range} 
  \label{figura5} 
@@
  \centering
- \includegraphics[scale=.5]{coming}
+ \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_algortihm}
  \caption{Phase two - Best $R^2$ for each algorithm} 
@@
  \centering
- \includegraphics[scale=.5]{coming}
- \caption{Phase two - Best $R^2$ for each algorithm} 
+ \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_number_of_observation_range_in_the_patterm}
+ \caption{Phase two - Best $R^2$ for each observation range} 
  \label{figura9} 
@@
  \centering
- \includegraphics[scale=.5]{coming}
+ \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
  \caption{Phase two - Best $R^2$ for each variable combination} 
@@
  \centering
- \includegraphics[scale=.8]{coming}
+ \includegraphics[scale=.8]{Phase_two_R2_RMSE}
  \caption{Phase two - Pareto frontier for $R^2$ and $RMSE$} 
@@
 \end{figure}
+
+In phase two, the Pareto frontier for the La Rita farm is composed by 2 elements. The \tablename $.$\ref{tabla4} shows the composition about variables and observation ranges.
+
+\begin{table}[h] 
+\caption{Composition of the Pareto frontier - La Rita - Phase two} 
+\label{tabla4} 
+\centering
+\begin{tabular}{c|c|c|c|c} 
+\hline
+\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
+\hline\hline 
+All & 1 to 1  & 2 & 69.33\% & 353.33 \\
+\hline
+\end{tabular} 
+\end{table}
+
+In 28 Millas Farm, the Pareto frontier is composed by 1 element. The \tablename $.$\ref{tabla5} details the result.
+
+\begin{table}[h] 
+\caption{Pareto frontier - 28 Millas - Phase two} 
+\label{tabla5} 
+\centering
+\begin{tabular}{c|c|c|c|c} 
+\hline
+\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
+\hline\hline 
+Pair $\overline{T}_{a}$ $P$ & 1 to 1 & 1 & 61.36\% & 339.89 \\
+\hline
+\end{tabular} 
+\end{table}
 

### 2016-02-21 2016-02-15-Resume archivos de validación.py

@@ -30,8 +30,11 @@ tArchivos = [   ["2016-02-16-Sigatoka - La Rita=01= - 55",tTamTotal1, tTamUsar1]
                 ["2016-02-16-Sigatoka - 28 Millas=12= - 105",tTamTotal2, tTamUsar2],
                 ["2016-02-16-Sigatoka - 28 Millas=13= - 55",tTamTotal1, tTamUsar1],
                 ["2016-02-16-Sigatoka - 28 Millas=14= - 105",tTamTotal2, tTamUsar2],
-                ["2016-02-16-Sigatoka - 28 Millas=15= - 55",tTamTotal2, tTamUsar2],
-                ["2016-02-16-Sigatoka - La Rita=16= - 55",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=15= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - La Rita=16= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - 28 Millas=18= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - La Rita=17= - 105",tTamTotal2, tTamUsar2],
+
                 ["2015-01-27-Sigatoka - 28 Millas - Semanal - Validacion-ESN - 54",54,50],
                 ["2015-01-27-Sigatoka - La Rita - Semanal - Validacion-SVR - 54",54,50],
                 ["2015-01-27-Sigatoka - La Rita - Semanal - Validacion-SVR - 105",105,100],
@@ -50,7 +53,7 @@ tArchivos = [   ["2016-02-16-Sigatoka - La Rita=01= - 55",tTamTotal1, tTamUsar1]
                 ["2015-01-27-Sigatoka - La Rita - Semanal - Validacion-REL - 105",105,100]
         ]
 os.chdir(tDirectorio)
-tArchSalida = "2016-02-19- Resumen - Validacion - semanal - Sigatoka"
+tArchSalida = "2016-02-20- Resumen - Validacion - semanal - Sigatoka"
 
 
 tTempDic = { "tipo": {}, "patron": {}, "algoritmo": {}, "disc_cont": {}, 

### 2016-02-21 2016-02-19-Graficar Frente de Pareto.py

@@ -124,7 +124,7 @@ def genereUnGrafico(pDatos, pColores, pTipos, pCol, pTipoGraf, pX, pY, pLoc, pNo
 os.chdir("D:\Cambios\Doctorado\CORRIDAS DEL PROGRAMA\Graficacion")
 
 tArchivoEntrada = "2016-02-04-datos para dispersion - Fase 1.csv"
-tArchivoEntrada = "2016-02-15- Datos- Validacion - Fase 2.csv"
+tArchivoEntrada = "2016-02-20- Datos- Validacion - Fase 2.csv"
 gNombreGlobal = "Fase 2"
 tCuantosHead = 55000
 tRMSEMenorQue = 1200

### 2016-03-22 Paper Sigatoka - 2015.tex
2016-03-21 11pm

@@
 
-\marginnote{Esto es parte del ``related works''}
-%
 There are efforts to apply machine learning methods for
@@
 Romero concluded that the model to predict the incubation period accounted a $R^2$ of 69\% in his observed data but it was not a good predictor when it was validated against an independent dataset (cross validation).
-For latency, he developed two models that accounted a $R^2$ of 78\% \hl{PONER EL VALOR OBTENIDO} in the observed data, however, when validated against an independent dataset (cross validation), the model was incorrect \hl{PONER EL VALOR OBTENIDO} for Weldeck, and for Rita obtained an adjusted $R^2$ of 82\%.
-\marginnote{Revisar si se puede implementar la propuesta de este autor para comparar}
+For latency, he developed two models that accounted a $R^2$ of 78\% in the observed data, however, when validated against an independent dataset (cross validation), the model was incorrect \hl{PONER EL VALOR OBTENIDO} for Weldeck, and for Rita obtained an adjusted $R^2$ of 82\%.
 
@@
 The value to be predicted in all cases was $E_s$, that is the total measure of the biological warning system.
-\marginnote{Mas detalle aqui?  o ampliar en los conceptos?}
 
@@
 Besides enabling the capture of tendencies, the discretization removes the effect of small variations in the data collection, either by inaccuracies of the instruments (meteorological variables) or by subjective bias introduced by the human who collects the data (biological warning system).
-\hl{ESTO DEBE ESTAR DESCRITO EN ALGUNA PARTE}
 
@@
 
-\section{Results}
+\section{Results and discussion}
 
@@
 
-\figurename $.$\ref{figura4} shows the best $R^2$ for each algorithm in the experiment. Results are group by farm.
+\figurename $.$\ref{figura4} shows the best $R^2$ for each algorithm in the experiment.
+Results are group by farm.
+Though La Rita obtains different results in magnitude than 28 Millas, the trend is similar.
+In both farms, the best results are for linear models, second position is occupied for Echo State Networks and  SVR with gausian and sigmoid kernel are the worst results.
+In linear models, to predict one week ahead is better than two weeks ahead, and this is better than three weeks ahead.
 
@@
 
-\figurename $.$\ref{figura5} presents, for one, two and three weeks ahead, the best $R^2$. Results are group by farm.
+\figurename $.$\ref{figura5} presents, for one, two and three weeks ahead, the best $R^2$.
+Results are group by farm.
+In general, to predict one week ahead is better than two weeks ahead and so on.
+The number of weeks consider in the observation range in the pattern is not the main discriminant factor, but it is clear that we get better $R^2$ for one week ahead than two weeks ahead and so on.
 
@@
 
-\figurename $.$\ref{figura6} shows the best $R^2$ for each variables combination. Results are group by farm.
+\figurename $.$\ref{figura6} shows the best $R^2$ for each variables combination.
+Results are group by farm.
+The better results are obtained with $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms of similarly.
+You can note that the use of all variables in the model or the inclusion of the four variables suggest for expert criteria do not improve significantly the results, then the use of more sensors do not assure a better result.
 
@@
 
-\section{Discussion and conclusions}
+\section{Conclusions}
 

### 2016-03-27 Paper Sigatoka - 2015.tex
2016-03-27

@@
 %% `Elsevier LaTeX' style
-%\bibliographystyle{elsarticle-num}
+\bibliographystyle{elsarticle-num}
 
@@
 \address[afiLuisAlex]{DOCINADE, Instituto Tecnológico de Costa Rica, 
-Computer Research Center, Multidisciplinar program eScience, 
-CNCA/CeNAT, Cartago, Costa Rica}
+Computer Research Center, Multidisciplinar program eScience, Cartago, Costa Rica}
 
@@
 Though La Rita obtains different results in magnitude than 28 Millas, the trend is similar.
-In both farms, the best results are for linear models, second position is occupied for Echo State Networks and  SVR with gausian and sigmoid kernel are the worst results.
+In both farms, the best results are for linear models, second position is occupied for Echo State Networks and  SVR with gausian and sigmoid kernels are the worst results.
 In linear models, to predict one week ahead is better than two weeks ahead, and this is better than three weeks ahead.
@@
 \figurename $.$\ref{figura7} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
+The Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better RMSE than La Rita.
 
@@
 \end{table}
+
+We can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week.
 

### 2016-04-02 Paper Sigatoka - 2015.tex
2016-04-01 noche

@@
 %
-In SVR, the empirical error is measured by the loss function $\epsilon$-insensitive and the generalization ability is measured by the Euclidean norm of $w$ \citep{XXXX201X}.
+In SVR, the empirical error is measured by the loss function $\epsilon$-insensitive and the generalization ability is measured by the Euclidean norm of $w$ \citep{Cristiani2005}.
 %
 Then, the optimization problem to identify the regression model can be
-formulated by \citep{Wei2013}: \marginnote{Falta completar la cita
-  XXXXX con el libro de SVR}
+formulated by \citep{Wei2013}:
 
@@
 Romero concluded that the model to predict the incubation period accounted a $R^2$ of 69\% in his observed data but it was not a good predictor when it was validated against an independent dataset (cross validation).
-For latency, he developed two models that accounted a $R^2$ of 78\% in the observed data, however, when validated against an independent dataset (cross validation), the model was incorrect \hl{PONER EL VALOR OBTENIDO} for Weldeck, and for Rita obtained an adjusted $R^2$ of 82\%.
+For latency, he developed two models that accounted a $R^2$ of 78\% in the observed data, however, when validated against an independent dataset (cross validation), for La Rita obtained an adjusted $R^2$ of 82\%, and for Weldeck, none of the models satisfactorily predicted the latent period; and then those predictions were not shown \citep{Romero1995}.
 
@@
 
-In the second phase, the best configurations obtained in phase one are used to validate with the last 52 and 102 weeks.
+In the second phase, the best configurations obtained in phase one are used to validate with the last 50 and 100 weeks.
 
@@
 $Luko\breve{s} evi \breve{c} ius$ \citep{Lukose2012} on which the necessary were done for adjustments for the experiments of this work.
-The computer used a processor Intel(R) Core™ i7-4800MQ CPU @ 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
+The computer used a processor Intel(R) Core i7-4800MQ CPU 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
 
@@
 \figurename $.$\ref{figura7} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
-The Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better RMSE than La Rita.
+The Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better $RMSE$ than La Rita.
+This situation arise because $RMSE$ considers errors only with respect the prediction and in 28 Millas the average of Stage of Evolution is $4316.16$, unlike, in La Rita the average is $5507.30$.
+So, in La Rita we obtains higher errors in absolute values.
+$R2$ is a relative metric between 0 thru 1 and it is less sensitive to absolute values.
 
@@
 
-\figurename $.$\ref{figura8} shows the best $R^2$ for each algorithm in the experiment. Results are group by farm.
+\figurename $.$\ref{figura8} shows the best $R^2$ for each algorithm in the experiment.
+Results are group by farm.
+Even if linear models continue with good $R^2$, Echo State Networks improve their scores because in $50$ and $100$ last weeks validation, we are in presence of climate change, then the behaviour is less lineal.
 
@@
 
-\figurename $.$\ref{figura9} presents, for one, two and three weeks ahead, the best $R^2$. Results are group by farm.
+\figurename $.$\ref{figura9} presents, for one, two and three weeks ahead, the best $R^2$.
+Results are group by farm.
+The results confirms that, in general, to predict one week ahead is better than two weeks ahead, two than three and so on, this for both farms.
 
@@
 
-\figurename $.$\ref{figura10} shows the best $R^2$ for each variables combination. Results are group by farm.
+\figurename $.$\ref{figura10} shows the best $R^2$ for each variables combination.
+Results are group by farm.
+This results confirm that $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms are the best variables combinations.
 
@@
 
-\figurename $.$\ref{figura11} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
+\figurename $.$\ref{figura11} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.  You can note that the behaviour of $R^2$ and $RMSE$ is similar to the phase one.
 
@@
 
-\section{Conclusions}
+Again, similar to phase one, we can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week, one week to predict one week ahead with combinations of variables listed above.
+
+\section{Acknowledgements}
+
+The authors would like to thank Corbana S.A. for providing the data for this research.
 
@@
 
-\bibitem[Libro SVM XXXX, 201X]{XXXX201X} Libro SVM XXXX.
+\bibitem[Cristiani and Shave-Taylor, 2005]{Cristiani2005} Cristiani, Nello and Shave-Taylor, John.
+An introduction to support vector machines and other kernel-based learning methods.
+Cambridge University Press.
+Ninth printing, United Kingdom, 2005.
 

### 2016-04-25 Paper Sigatoka - 2015.tex
Cambios de referencias en linea a bibtex.

@@
 \documentclass[review]{elsarticle}
+\bibliographystyle{elsarticle-harv}
 
-\usepackage{lineno,hyperref}
+\usepackage{lineno}
 \usepackage[utf8]{inputenc}  %% Permite utilizar tildes si archivo usa utf-8
 \usepackage{amsmath}
+\usepackage{amssymb}
 
-\modulolinenumbers[1]
+\modulolinenumbers[5]
 
@@
 \usepackage{float}
+\usepackage{natbib}
+\usepackage{hyperref}
 
@@
 
-\title{Forecasting the black Sigatoka development rate: A comparison of machine learning techniques 
+\title{Forecasting the black Sigatoka development rate:\\ A comparison of machine learning techniques 
 %\tnoteref{mytitlenote}
@@
   fijiensis Morelet} is the major pathological problem of banana and
-plantain crops in Central America, Panama, Colombia and Ecuador, as in
+plantain crops in Central America, Panama, Colombia and Ecuador, as well as in
 many parts of Africa and Asia \citep{MarinVargas1995}.
@@
 This method fits a linear model with coefficients $w = (w1,..,wp)$ to minimize the residual sum of squares between the observed responses in the dataset, and the responses predicted by the linear approximation.
-Mathematically it solves a problem of the form \citep{ scikit-learn2011}:
+Mathematically it solves a problem of the form \citep{scikitlearn2011}:
 $$\min_{w} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  $$
@@
 According Pedregosa et al.
-\citep{ scikitlearn2011}  the coefficient estimates for Ordinary Least Squares rely on the independence of the model terms.
+\citep{scikitlearn2011}  the coefficient estimates for Ordinary Least Squares rely on the independence of the model terms.
 When terms are correlated and the columns of the design matrix $X$ have an approximate linear dependence, the design matrix becomes close to singular and as a result, the least-squares estimate becomes highly sensitive to random errors in the observed response, producing a large variance.
@@
 
-\begin{thebibliography}{1}
-
-\bibitem[Camargo et al.,2012]{Camargo2012} Camargo, A., Molina, J., Cadena-Torres, J., Jiménez, N., Kim, J.
-2012.
-Intelligent systems for the assessment of crop disorders.
-Computers and Electronics in Agriculture(85), 1-7.
-doi:10.1016/j.compag.2012.02.017.
-
-\bibitem[Chuang and Jeger, 1987]{ChuangJeger1987} Chuang, T., Jeger, M.
-1987.
-Predicting the Rate of Development of Black Sigatoka ( Mycosphaerella fijiensis var.
-difformis ) Disease in Southern Taiwan.
-Phytopathology, 77, 1542-1547.
-
-\bibitem[Huang et al., 2010]{Huang2010} Huang, Y., Lan, Y., Thomson, S., Fang, A., Hoffmann, W., Lacey, R.
-2010.
-Development of soft computing and applications in agricultural and biological engineering.
-Computers and Electronics in Agriculture,(71(2)), 107–127.
-doi:10.1016/j.compag.
-
-\bibitem[Kim et al., 2014]{Kim2014} Kim, Y., Yoo, S., Gu, Y., Lim, J., Han, D.,  Baik, S.
-2014.
-Crop Pests Prediction Method Using Regression and Machine Learning Technology: Survey.
-IERI Procedia(6), 52–56.
-doi:10.1016/j.ieri.2014.03.009.
-
-\bibitem[Marin and Romero, 1995]{MarinVargas1995} Marin Vargas, D., Romero Calderón, R. 1995. El combate de la Sigatoka Negra. Boletín Departamento de Investigaciones, Corbana Costa Rica.
-
-\bibitem[Marin et al., 2003]{Marinetal2003} Marin, D., Romero, R., Guzman, M, Sutton, T. 2003. Black Sigatoka: An increasing threat to banana cultivation. Plant Disease, 87(3), 208-222.
-
-\bibitem[Zhao et al., 2013]{Zhao2013}Zhao, L., He, L., Harry, W., Jin, X.
-2013.
-Intelligent Agricultural Forecasting System Based on Wireless Sensor.
-Journal of Networks(8), 1817–1824.
-doi:10.4304/jnw.8.8.1817-1824.
-
-\bibitem[Wei et al., 2013]{Wei2013} Wei, Z., Tao, T., ZhuoShu, D.,  Zio, E.
-(2013).
-A dynamic particle filter-support vector regression method for reliability prediction.
-Reliability Engineering \& System Safety, 109–116.
-doi:10.1016/j.ress.2013.05.021.
-
-\bibitem[Cristiani and Shave-Taylor, 2005]{Cristiani2005} Cristiani, Nello and Shave-Taylor, John.
-An introduction to support vector machines and other kernel-based learning methods.
-Cambridge University Press.
-Ninth printing, United Kingdom, 2005.
-
-\bibitem[Alonso et al., 2013]{Alonso2013} Alonso, J., Rodríguez Castañón, Á., Bahamonde, A.
-(2013).
-Support Vector Regression to predict carcass weight in beef cattle in advance of the slaughter.
-Computers and Electronics in Agriculture, 116-120.
-
-\bibitem[Pedregosa et al., 2013] {scikitlearn2011} Pedregosa, F.
-and Varoquaux, G.
-and Gramfort, A.
-and Michel, V.
-and Thirion, B.
-and Grisel, O.
-and Blondel, M.
-and Prettenhofer, P.
-and Weiss, R.
-and Dubourg, V.
-and Vanderplas, J.
-and Passos, A.
-and Cournapeau, D.
-and Brucher, M.
-and Perrot, M.
-and Duchesnay, E.
-(2011) Scikit-learn: Machine Learning in Python.
-Journal of Machine Learning Research, vol.
-12, 2825--2830.
-
-\bibitem[Lukosevicius and Jaeger, 2009] {Lukose2009} Lukosevicius, M.
-and Jaeger, H.
-(2009).
-Reservoir computing approaches to recurrent neural network training.
-Computer Science Review(3), 127–149.
-doi:10.1016/j.cosrev.2009.03.005.
-
-\bibitem[Lukosevicius, 2012] {Lukose2012} Lukosevicius, M. (2012). A Practical Guide to Applying Echo State Networks. Neural Networks: Tricks of the Trade. 1-20
-
-\bibitem[Romero, 1995] {Romero1995}  Romero, R.
-(1995).
-Dynamics of fungicide resistant populations of Mycosphaerella fijiensis and Epidemiology of black Sigatoka of banana.
-Costa Rica: North Carolina State University.
-
-\bibitem[Glezakos et al., 2010] {Glezakos2010} Glezakos, T., Moschopoulou, G., Tsiligiridis, T., Kintzios, S., Yialouris, C.
-(2010).
-Plant virus identification based on neural networks with evolutionary preprocessing.
-Computers and Electronics in Agriculture, 70, 263–275.
-
-\bibitem[Alves et al., 2011] {Alves2011} Alves, M., de Carvalho, L., Pozza, E., Sanches, L., Maia, J.
-(2011).
-Ecological zoning of soybean rust, coffee rust and banana black sigatoka based on Brazilian climate changes.
-Procedia Environmental Sciences, 6, 35-49.
-
-\bibitem[Coopersmith et al., 2010] {Coopersmith2014} Coopersmith, E.
-J., Minsker, B.
-S., Wenzel, C.
-E., Gilmore, B.
-J.
-(2014).
-Machine learning assessments of soil drying for agricultural planning.
-Computers and Electronics in Agriculture, 104, 93–104.
-http://doi.org/10.1016/j.compag.2014.04.004
-
-\bibitem[Rumpf et al., 2010] {Rumpf2010} Rumpf, T., Mahlein, a.-K., Steiner, U., Oerke, E.-C., Dehne, H.-W., Plümer, L.
-(2010).
-Early detection and classification of plant diseases with Support Vector Machines based on hyperspectral reflectance.
-Computers and Electronics in Agriculture, 74(1), 91–99.
-http://doi.org/10.1016/j.compag.2010.06.009
-
-\bibitem[Bendini et al., 2010] {Bendini2013} Bendini, H., Moraes, W., da Silva, S., Tezuka, E., Cruvinel, P.
-(2013).
-Análise de risco da ocorrência de Sigatoka-negra baseada em modelos polinomiais: um estudo de caso.
-Tropical Plant Pathology, 38, 035-043.
-
-\bibitem[Soares et al., 2014] {Soares2014} Soares, J., Pasqual, M., Lacerda, W., Silva, S., Donato, S.
-(2014).
-Comparison of techniques used in the prediction of yield in banana plants.
-Scientia Horticulturae journal, 167, 84-90.
-
-\bibitem[Soares et al., 2013] {Soares2013} Soares , J., Pasqual, M., Lacerda, W.
-(2013).
-Utilization of artificial neural networks in the prediction of the bunches’weight in banana plants.
-Scientia Horticulturae(155), 24-29.
-
-\bibitem[Ibrahim and Wibowo, 2014] {Ibrahim2014}  Ibrahim, N.
-and Wibowo, A.
-(2014).
-Time Series Support Vector Regression with Missing Data Treatment Based Variables Selection for Water Level Prediction of Galas River in Kelantan Malaysia.
-International Journal of Applied Research in Engineering and Science, 3, 25-36.
-
-\bibitem[Demir and Bruzzone, 2014] {Demir2014} Demir, B.
-and Bruzzone, L.
-(2014).
-A multiple criteria active learning method for support vector regression.
-Pattern Recognition, 2558–2567.
-doi:10.1016/j.patcog.2014.02.001
-
-\bibitem[Continuum Analitycs, 2015] {Continuum2015} Continuum Analitycs. (2015). Anaconda. Retrieved from https://www.continuum.io/
-
-\bibitem[McKinney, 2010] {mckinneypandas2010} McKinney W. (2010). Data Structures for Statistical Computing in Python, Proceedings of the 9th Python in Science Conference, 51-56
-
-\bibitem[van der Walt et al., 201] {vanderWalt2011} van der Walt S., Colbert C.
-and Varoquaux G.
-(2011).
-The NumPy Array: A Structure for Efficient Numerical Computation, Computing in Science \& Engineering, 13, 22-30 (2011), DOI:10.1109/MCSE.2011.37
-
-\end{thebibliography}
+\bibliography{references}
 

### 2016-04-27 Paper Sigatoka - 2015.tex
Finalmente pude activar el formato autor agno
Text of introduction reviewed.  Some formulations changed

@@
-\documentclass[review]{elsarticle}
+\documentclass[review,authoryear,english]{elsarticle}
+
+\usepackage{babel}
+\usepackage{natbib}
+
 \bibliographystyle{elsarticle-harv}
@@
 \usepackage{float}
-\usepackage{natbib}
-\usepackage{hyperref}
+
+\usepackage[hidelinks]{hyperref}
 
@@
 \journal{Journal of \LaTeX\ Templates}
-
-%%%%%%%%%%%%%%%%%%%%%%%
-%% Elsevier bibliography styles
-%%%%%%%%%%%%%%%%%%%%%%%
-%% To change the style, put a % in front of the second line of the current style and
-%% remove the % from the second line of the style you would like to use.
-%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Numbered
-%\bibliographystyle{model1-num-names}
-
-%% Numbered without titles
-%\bibliographystyle{model1a-num-names}
-
-%% Harvard
-%\bibliographystyle{model2-names.bst}\biboptions{authoryear}
-
-%% Vancouver numbered
-%\usepackage{numcompress}\bibliographystyle{model3-num-names}
-
-%% Vancouver name/year
-%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}
-
-%% AMA style
-%\usepackage{numcompress}\bibliographystyle{model6-num-names}
-
-%% `Elsevier LaTeX' style
-\bibliographystyle{elsarticle-num}
-
-%% APA style
-% \bibliographystyle{model5-names}\biboptions{authoryear}
-
-%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\marginnote{La bibliografia debe ser autor, año}
-
 The black Sigatoka disease caused by the fungus \emph{Mycosphaerella
@@
 of the leaf area, affects the growth and productivity of the plants
-due to the impairment of their photosinthetic ability, causes a
+due to the impairment of the photosynthetic process, causes a
 reduction in the quality of the fruit, and promotes premature
-maturation \marginnote{maturation or ripening?} of bunches, which is
+ripening of bunches, which is
 the major cause of product losses associated with the black
@@
 
-Phytopathological studies point out that precipitation, temperature,
-relative humidity and wind are the main climatic variables that affect
-its development \citep{MarinVargas1995}.
-
-\begin{figure}[h] 
+\begin{figure}[ht] 
 \centering
@@
 
-In Costa Rica the control of black Sigatoka makes use of chemical fungicides.
+Phytopathological studies point out that precipitation, temperature,
+relative humidity and wind are the main climatic variables that affect
+the development of this disease \citep{MarinVargas1995}.
+
+In Costa Rica the black Sigatoka is frequently treated with
+chemical fungicides.
 %
-Depending on the zone of production and the weather conditions, 45--55
-cycles/year of fungicide applications are required to keep this
-disease under control and to produce the fruit quality expected in the
+Depending on the zone of production and the weather conditions,
+45--55~cycles/year of fungicide applications are required to keep this
+disease under control and to produce the expected fruit quality for the
 international markets.
@@
 %
-Overall, this represents 10\%--12\% of the total production cost.
+Overall, this represents 10\%--12\% of the total production costs.
 
@@
 a particular fungicide spray program will be able to effectively and
-economically control the disease \citet{ChuangJeger1987}.
+economically control the disease \citep{ChuangJeger1987}.
 
-There are efforts to apply machine learning methods for
+There are efforts to apply machine learning methods to support
 decision-making in agriculture, including the control of crop
@@
 
-In this work, we compare five machine learning techniques (support
+In this work, we compare five machine learning techniques 
+to predict the development rate of the black Sigatoka disease:
+support
 vector regression (SVR), echo state networks (ESN), ridge regression,
-elastic-net regression and ordinary least squares linear regression)
-to predict the development rate of the black Sigatoka disease.
+elastic-net regression and ordinary least squares linear regression.
 
@@
 This combination allows for learning a sparse model where few of the weights are non-zero like Lasso, while still maintaining the regularization properties of Ridge  \citep{scikitlearn2011}.
-The convex combination of $L1$ and $L2$ is controled by using the $l1_ratio$ parameter.
+The convex combination of $L1$ and $L2$ is controled by using the $l1_{ratio}$ parameter.
 

### 2016-05-01 Paper Sigatoka - 2015.tex
Correcciones de estilo y sugerencias

@@
 
+\usepackage{color}
+\definecolor{dkred}{rgb}{0.5,0,0}       %   dark red
+\definecolor{todocolor}{rgb}{1.0,0.8,0.5}
+
 \usepackage[hidelinks]{hyperref}
+
+\newcommand{\todohdr}{Please review:}
+\newcommand{\TODO}[2][\todohdr]{%
+\smallbreak\noindent\colorbox{todocolor}{%
+\begin{minipage}{1\linewidth}
+\textbf{#1}\par{\footnotesize{#2}}
+\end{minipage}}}
+
+\newcommand{\instead}[2]{\textcolor{dkred}{\st{#1}}#2}
 
@@
 
-\title{Forecasting the black Sigatoka development rate:\\ A comparison of machine learning techniques 
-%\tnoteref{mytitlenote}
-}
-%\tnotetext[mytitlenote]{Fully documented templates are %available in the elsarticle package on \href{http://%www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}%{CTAN}.}
+\title{Forecasting the black Sigatoka development rate:\\ 
+A comparison of machine learning techniques}
 
@@
 
-\author[afiCorbana] {Mauricio Guzmán-Quesada}
-\author[afiCorbana]{José-Antonio Guzmán-Alvarez}
+\author[afiCorbana]{Mauricio Guzmán-Quesada}
+\author[afiCorbana]{José-Antonio~Guzmán-Alvarez}
 \author[afiPablo]{Pablo Alvarado-Moya}
@@
 \end{tabular}
-\caption{Examples of three disease stages of the black Sigatoka. (a)~Initial stage. (b)~Intermediate stage, and (c)~Advanced stage.} 
+\caption{Examples of three disease stages of the black Sigatoka. 
+(a)~Initial stage. (b)~Intermediate stage, and (c)~Advanced stage.} 
 \label{fig:diseasestages} 
@@
 
-Black Sigatoka, disease caused by the fungus Mycosphaerella fijiensis
-Morelet, is the main problem phytopathologic of banana and plantain
-crops in Central America \citep{MarinVargas1995}.
+The black Sigatoka is a disease caused by the fungus
+\emph{Mycosphaerella fijiensis Morelet}.  It is the main
+phytopathologic problem of banana and plantain crops in Central
+America \citep{MarinVargas1995}.
+%
+This disease attacks the plant leaves producing a rapid deterioration
+of the leaf area. It affects the growth and productivity of the
+affected plant by impairing its photosynthetic capacity, which
+consequently causes a reduction in quality of the fruit
+\citep{MarinVargas1995}.
 
-This disease attacks the leaves of plants producing a rapid deterioration of the leaf area.
-It affects the growth and productivity of plants by decreasing photosynthetic capacity.
-Also causes a reduction in quality of the fruit \citep{MarinVargas1995}.
-
-The climate has a major effect on the behavior of the black Sigatoka.
-Precipitation, temperature, relative humidity and wind are the main climatic variables affecting the development of this disease \citep{MarinVargas1995}.
+The climate has a major effect on the development of the black
+Sigatoka.  Precipitation, temperature, relative humidity and wind are
+the main climatic variables affecting the development of this disease
+\citep{MarinVargas1995}.
 
@@
 
-The early warning system for black Sigatoka is an adaptation of the yellow Sigatoka warning system developed by Ganry and Meyer and modified by Ganry and Laville to use for controlling yellow Sigatoka in Cameroon.
-Ternesien and Fouré  later improved Ganry and Laville's system.
+\TODO{Revisar si la nueva redacción de este párrafo coincide con la
+  original.  Eso es lo que creo se quería decir, porque no quedaba
+  claro cuál sistema fue creado por quién...  Las referencias a cada
+  sistema en algo así son absolutamente necesarias.  Favor reemplazar
+  \citep{XXX} por las referencias correspondientes (Pablo, 1/5/16)}
+
+The early warning system developed by Ganry and Meyer \citep{XXX} for
+the control of the yellow Sigatoka in Cameroon was adapted by Ganry
+and Meyer \citep{XXX} for the black Sigatoka. 
 %
-The latter system is based on weekly observations of disease symtoms on young leaves of the plant, according to Fourés symptom (stages) descriptions.
-Arbitrary coefficientes, based on incidence and severity of disease development, are used to calculate two variables: gross sum and state of evolution.
+%The early warning system for black Sigatoka \citep{XXX} is an
+%adaptation of the yellow Sigatoka warning system developed by Ganry
+%and Meyer \citep{XXX}.  It was modified by Ganry and Laville
+%\citep{XXX} for its ito use for controlling yellow Sigatoka in
+%Cameroon.
+Ternesien and Fouré \citep{XXX} later improved Ganry and Laville's system.
+
+The latter system is based on weekly quantified observations of
+the disease progression on young leaves of the plant, according to
+Fouré's scale of the symptom stages.
 %
-Gross sum is based on the stage present and an arbitrary coefficient, which increases with the advance of the symptoms and the juvenility of the leaf.
-The state of evolution is calculated using the gross sum and the foliar emission period.
+The numeric coefficients describe the degree of incidence and the
+severity of the disease development.  These coefficients are used to
+calculate two variables: gross sum and state of evolution.
+\TODO{coefficients??  No suena como coeficientes (un factor que
+  multiplica términos en una función) sino como una calificación o
+  nota o algo así (quizá ``grade''?).  ``Arbitrary coefficient'' no
+  tiene sentido por el ``arbitrary''... No creo que sean tan
+  arbitrarios sino predefinidos.  De hecho, el párrafo que sigue dice
+  que suben con el avance de síntomas, es decir, no es tan
+  ``arbitrario''.  Quizá querés decir ``arbitrarily chosen''?.  No
+  estoy seguro si entiendo lo que se quiere decir aquí.}
+
+The gross sum is based on the present disease progression stage and 
+\instead{an arbitrary coefficients}{the arbitrarily chosen grading},
+which increases with the advance of the symptoms and the
+juvenility of the leaf.
 %
-Although threshold levels were initially suggested as a guide to spray timing, the fluctuation of these two variables was found to better define appropiate times to spray \citep{Marinetal2003}.
+The state of evolution is calculated using the gross sum and the
+foliar emission period.
+%
+Although threshold levels were initially suggested as a guide to spray
+\instead{timing}{schedules}, the fluctuation of these two variables
+was found to better define appropiate times to spray
+\citep{Marinetal2003}.
 
 \subsubsection{Support Vector Regression (SVR)}
+
 From the perspective of Support Vector Regression (SVR) the regression function $y = f(s)$ for a given dataset $D=\{(s_i,y_i)\}_{i=1}^n$ , is represented as a linear function of the form \citep{Wei2013}:

### 2016-05-03 Paper Sigatoka - 2015.tex
d2

--- 2016-04-25 Paper Sigatoka - 2015.tex
@@
 \usepackage{float}
-\usepackage{natbib}
 \usepackage{hyperref}
@@
 reduction in the quality of the fruit, and promotes premature
-maturation \marginnote{maturation or ripening?} of bunches, which is
+ripening of bunches, which is
 the major cause of product losses associated with the black

### 2016-05-20 Paper Sigatoka - 2015.tex
20 mayo

--- 2016-05-01 Paper Sigatoka - 2015.tex
@@
 elastic-net regression and ordinary least squares linear regression.
-
+%
 The main contribution of this work is a comparison between machine
@@
 \marginnote{FALTA COMPLETAR esta parte}
+
+The outline of the paper is as follows.
+In Section 2, we present materials and methods of our research.
+In Section 3, we present the main results and their discussion.
+Our conclusions and future work are presented in Section 4.
 
@@
 
-\TODO{Revisar si la nueva redacción de este párrafo coincide con la
+\TODO{ LISTO  Revisar si la nueva redacción de este párrafo coincide con la
   original.  Eso es lo que creo se quería decir, porque no quedaba
@@
   sistema en algo así son absolutamente necesarias.  Favor reemplazar
-  \citep{XXX} por las referencias correspondientes (Pablo, 1/5/16)}
+  {XXX} por las referencias correspondientes (Pablo, 1/5/16)}
 
-The early warning system developed by Ganry and Meyer \citep{XXX} for
-the control of the yellow Sigatoka in Cameroon was adapted by Ganry
-and Meyer \citep{XXX} for the black Sigatoka. 
-%
-%The early warning system for black Sigatoka \citep{XXX} is an
-%adaptation of the yellow Sigatoka warning system developed by Ganry
-%and Meyer \citep{XXX}.  It was modified by Ganry and Laville
-%\citep{XXX} for its ito use for controlling yellow Sigatoka in
-%Cameroon.
-Ternesien and Fouré \citep{XXX} later improved Ganry and Laville's system.
+The early warning system developed by Ganry and Meyer \citep{ganry1972} and modified by Ganry and Laville \citep{ganry1983} to use for controlling yellow Sigatoka in Cameroon, later was adapted for Ternesien \citep{Ternesien1985} and Fouré \citep{foure1988} for the black Sigatoka.
 
@@
 calculate two variables: gross sum and state of evolution.
-\TODO{coefficients??  No suena como coeficientes (un factor que
+
+\TODO{ EN REALIDAD ASÍ ES USADO EN EL TEXTO FUENTE
+coefficients??  No suena como coeficientes (un factor que
   multiplica términos en una función) sino como una calificación o
@@
 
+\section{Future work}
+
 \section{Acknowledgements}

### 2016-05-22 Paper Sigatoka - 2015.tex
22 mayo 2016

@@
 
+The climate has a major effect on the development of the black
+Sigatoka.  Precipitation, temperature, relative humidity and wind are
+the main climatic variables affecting the development of this disease
+\citep{MarinVargas1995}.
+
+\TODO{ CORREGIDO (LUIS-ALEX)
+revisar si la nueva redacción de este párrafo coincide con la
+  original.  Eso es lo que creo se quería decir, porque no quedaba
+  claro cuál sistema fue creado por quién...  Las referencias a cada
+  sistema en algo así son absolutamente necesarias.  Favor reemplazar
+   por las referencias correspondientes (Pablo, 1/5/16)}
+
+The early warning system developed by Ganry and Meyer \citep{ganry1972} and modified by Ganry and Laville \citep{ganry1983} to use for controlling yellow Sigatoka in Cameroon, later was adapted for Ternesien \citep{Ternesien1985} and Fouré \citep{foure1988} for the black Sigatoka.
+
+This bilogical warning system is based on weekly quantified observations of the disease progression on young leaves of the plant, according to
+Fouré's scale of the symptom stages. The numeric coefficients describe the degree of incidence and the severity of the disease development.  These coefficients are used to
+calculate two variables: gross sum and state of evolution.
+
+\TODO{ EN REALIDAD ASÍ ES USADO EN EL TEXTO FUENTE EN INGLÉS (LUIS ALEX)
+coefficients??  No suena como coeficientes (un factor que
+  multiplica términos en una función) sino como una calificación o
+  nota o algo así (quizá ``grade''?).  ``Arbitrary coefficient'' no
+  tiene sentido por el ``arbitrary''... No creo que sean tan
+  arbitrarios sino predefinidos.  De hecho, el párrafo que sigue dice
+  que suben con el avance de síntomas, es decir, no es tan
+  ``arbitrario''.  Quizá querés decir ``arbitrarily chosen''?.  No
+  estoy seguro si entiendo lo que se quiere decir aquí.}
+
+The gross sum is based on the present disease progression stage and 
+\instead{an arbitrary coefficients}{the arbitrarily chosen grading},
+which increases with the advance of the symptoms and the
+juvenility of the leaf.
+%
+The state of evolution is calculated using the gross sum and the
+foliar emission period.
+%
+Although threshold levels were initially suggested as a guide to spray
+\instead{timing}{schedules}, the fluctuation of these two variables
+was found to better define appropiate times to spray
+\citep{Marinetal2003}.
+
 In Costa Rica the black Sigatoka is frequently treated with
@@
 %
-The main contribution of this work is a comparison between machine
-learning methods to forecast black Sigatoka development rate. 
-\marginnote{FALTA COMPLETAR esta parte}
+The main contribution of this work is the selection of the best machine learning technique to forecast the black Sigatoka development rate according to the metric indicated below.
 
 The outline of the paper is as follows.
-In Section 2, we present materials and methods of our research.
-In Section 3, we present the main results and their discussion.
-Our conclusions and future work are presented in Section 4.
+In Section 2, we introduce the machine learning techniques used in this research and the main related works.
+In section 3, we present the data used and the methodology for the study results.
+In section 4, we present the results and their discussion.
+Our conclusions and future work are presented in Section 5.
 
-\section{Materials and methods}
+\section{Machine learning techniques and Related works}
+In this section we summary the machine learning techniques used and we present the main related works.
 
-\subsection{Concepts}
-
-\subsubsection{Black Sigatoka disease}
-
-The black Sigatoka is a disease caused by the fungus
-\emph{Mycosphaerella fijiensis Morelet}.  It is the main
-phytopathologic problem of banana and plantain crops in Central
-America \citep{MarinVargas1995}.
-%
-This disease attacks the plant leaves producing a rapid deterioration
-of the leaf area. It affects the growth and productivity of the
-affected plant by impairing its photosynthetic capacity, which
-consequently causes a reduction in quality of the fruit
-\citep{MarinVargas1995}.
-
-The climate has a major effect on the development of the black
-Sigatoka.  Precipitation, temperature, relative humidity and wind are
-the main climatic variables affecting the development of this disease
-\citep{MarinVargas1995}.
-
-\subsubsection{Biological warning system}
-
-\TODO{ LISTO  Revisar si la nueva redacción de este párrafo coincide con la
-  original.  Eso es lo que creo se quería decir, porque no quedaba
-  claro cuál sistema fue creado por quién...  Las referencias a cada
-  sistema en algo así son absolutamente necesarias.  Favor reemplazar
-  {XXX} por las referencias correspondientes (Pablo, 1/5/16)}
-
-The early warning system developed by Ganry and Meyer \citep{ganry1972} and modified by Ganry and Laville \citep{ganry1983} to use for controlling yellow Sigatoka in Cameroon, later was adapted for Ternesien \citep{Ternesien1985} and Fouré \citep{foure1988} for the black Sigatoka.
-
-The latter system is based on weekly quantified observations of
-the disease progression on young leaves of the plant, according to
-Fouré's scale of the symptom stages.
-%
-The numeric coefficients describe the degree of incidence and the
-severity of the disease development.  These coefficients are used to
-calculate two variables: gross sum and state of evolution.
-
-\TODO{ EN REALIDAD ASÍ ES USADO EN EL TEXTO FUENTE
-coefficients??  No suena como coeficientes (un factor que
-  multiplica términos en una función) sino como una calificación o
-  nota o algo así (quizá ``grade''?).  ``Arbitrary coefficient'' no
-  tiene sentido por el ``arbitrary''... No creo que sean tan
-  arbitrarios sino predefinidos.  De hecho, el párrafo que sigue dice
-  que suben con el avance de síntomas, es decir, no es tan
-  ``arbitrario''.  Quizá querés decir ``arbitrarily chosen''?.  No
-  estoy seguro si entiendo lo que se quiere decir aquí.}
-
-The gross sum is based on the present disease progression stage and 
-\instead{an arbitrary coefficients}{the arbitrarily chosen grading},
-which increases with the advance of the symptoms and the
-juvenility of the leaf.
-%
-The state of evolution is calculated using the gross sum and the
-foliar emission period.
-%
-Although threshold levels were initially suggested as a guide to spray
-\instead{timing}{schedules}, the fluctuation of these two variables
-was found to better define appropiate times to spray
-\citep{Marinetal2003}.
-
-\subsubsection{Support Vector Regression (SVR)}
+\subsection{Support Vector Regression (SVR)}
 
@@
 
-\subsubsection{Ordinary least squares regression}
+\subsection{Ordinary least squares regression}
 This method fits a linear model with coefficients $w = (w1,..,wp)$ to minimize the residual sum of squares between the observed responses in the dataset, and the responses predicted by the linear approximation.
@@
 
-\subsubsection{Ridge regression}
+\subsection{Ridge regression}
 The ridge regression addresses some of the problems of ordinary least squares regression by imposing a penalty on the size of the coefficients.
@@
 
-\subsubsection{Elastic-Net regression}
+\subsection{Elastic-Net regression}
 Elastic-Net is a linear regression model trained with $L1$ and $L2$ prior as regularizer.
@@
 
-\subsubsection{Echo State Networks (ESN)}
+\subsection{Echo State Networks (ESN)}
 Recurrent Neural Networks (RNN) are useful for temporal patterns, but when they are trained with backpropagation methods, they are very slow.
@@
 
-\subsubsection{Related works}
+\subsection{Related works}
+Now, we present the main and recently works related with our study.
+
 Huang et al.
@@
 
-\subsubsection{Data}
-In this work we use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (previously called Waldeck and located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
+\section{Specification of data and methodology}
+In this section, we present the data and how we worked to obtain our results.
+
+\subsection{Data}
+We use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (previously called Waldeck and located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
 The banana type is Musa AAA, subgroup Cavendish, cv.
@@
 
-\subsubsection{Data preprocessing}
+\subsection{Data preprocessing}
+Usually, data need preprocessing to prepare for use. In this section we explain that process.
 
@@
 
+\subsection{Programming environment}
+
+We use the python programming language with the Integrated Development Environment (IDE) Spyder \citep{Continuum2015}, particularly with the libraries pandas \citep{mckinneypandas2010} and numpy \citep{vanderWalt2011}.
+For SVR, ridge and ordinary least squares regressions, we used sklearn \citep{scikitlearn2011} and for ESN the python-based code of Dr.
+$Luko\breve{s} evi \breve{c} ius$ \citep{Lukose2012} on which the necessary were done for adjustments for the experiments of this work.
+The computer used a processor Intel(R) Core i7-4800MQ CPU 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
+
 \subsection{Methodology}
-The selection of methods and their parametrisation was performed in two stages.
+The selection of the technique and its parametrisation was performed in two stages that they are presented below.
 
@@
 
-This second phase intents to expose how these methods behave on a considerable climate in the years 2014 and 2015.
-\marginnote{Poner una cita que fundamente este ultimo parrafo}
-
-\subsection{Programming environment}
-
-We use the python programming language with the Integrated Development Environment (IDE) Spyder \citep{Continuum2015}, particularly with the libraries pandas \citep{mckinneypandas2010} and numpy \citep{vanderWalt2011}.
-For SVR, ridge and ordinary least squares regressions, we used sklearn \citep{scikitlearn2011} and for ESN the python-based code of Dr.
-$Luko\breve{s} evi \breve{c} ius$ \citep{Lukose2012} on which the necessary were done for adjustments for the experiments of this work.
-The computer used a processor Intel(R) Core i7-4800MQ CPU 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
-
 \section{Results and discussion}
 
-In this section we present the main results for phase.
+In this section we present the main results for phase and their discussion.
 

### 2016-05-22 Paper Sigatoka - 2015.tex
Removing to-do's and check-me's and some minor issues

@@
 
-\modulolinenumbers[5]
+\modulolinenumbers[1]
 
@@
 
-\TODO{ CORREGIDO (LUIS-ALEX)
-revisar si la nueva redacción de este párrafo coincide con la
-  original.  Eso es lo que creo se quería decir, porque no quedaba
-  claro cuál sistema fue creado por quién...  Las referencias a cada
-  sistema en algo así son absolutamente necesarias.  Favor reemplazar
-   por las referencias correspondientes (Pablo, 1/5/16)}
+The early warning system developed by Ganry and Meyer
+\citep{ganry1972} and modified by Ganry and Laville \citep{ganry1983}
+to control yellow Sigatoka in Cameroon, was later adapted by
+Ternesien \citep{Ternesien1985} and Fouré \citep{foure1988} for the
+black Sigatoka.
 
-The early warning system developed by Ganry and Meyer \citep{ganry1972} and modified by Ganry and Laville \citep{ganry1983} to use for controlling yellow Sigatoka in Cameroon, later was adapted for Ternesien \citep{Ternesien1985} and Fouré \citep{foure1988} for the black Sigatoka.
-
-This bilogical warning system is based on weekly quantified observations of the disease progression on young leaves of the plant, according to
-Fouré's scale of the symptom stages. The numeric coefficients describe the degree of incidence and the severity of the disease development.  These coefficients are used to
+This bilogical warning system is based on weekly quantified
+observations of the disease progression on young leaves of the plant,
+according to Fouré's scale of the symptom stages.
+%
+The numeric coefficients describe the degree of incidence and the
+severity of the disease development.  These coefficients are used to
 calculate two variables: gross sum and state of evolution.
 
-\TODO{ EN REALIDAD ASÍ ES USADO EN EL TEXTO FUENTE EN INGLÉS (LUIS ALEX)
-coefficients??  No suena como coeficientes (un factor que
-  multiplica términos en una función) sino como una calificación o
-  nota o algo así (quizá ``grade''?).  ``Arbitrary coefficient'' no
-  tiene sentido por el ``arbitrary''... No creo que sean tan
-  arbitrarios sino predefinidos.  De hecho, el párrafo que sigue dice
-  que suben con el avance de síntomas, es decir, no es tan
-  ``arbitrario''.  Quizá querés decir ``arbitrarily chosen''?.  No
-  estoy seguro si entiendo lo que se quiere decir aquí.}
-
 The gross sum is based on the present disease progression stage and
-\instead{an arbitrary coefficients}{the arbitrarily chosen grading},
-which increases with the advance of the symptoms and the
-juvenility of the leaf.
+the numeric coefficients, which increases with the progression of
+the symptoms and the juvenility of the leaf.
 %
@@
 Although threshold levels were initially suggested as a guide to spray
-\instead{timing}{schedules}, the fluctuation of these two variables
+schedules, the fluctuation of these two variables
 was found to better define appropiate times to spray
@@
 %
-This represent a cost per hectare per year in the range between US\$1600 and
+This represents a cost per hectare per year in the range between US\$1600 and
 US\$2000; about 0.64--0.80 cents of the production costs for a
@@
 a particular fungicide spray program will be able to effectively and
-economically control the disease \citep{ChuangJeger1987}.
+economically control the disease \citep{ChuangJeger1987} or not.
 
@@
 %
-The main contribution of this work is the selection of the best machine learning technique to forecast the black Sigatoka development rate according to the metric indicated below.
+The main contribution of this work is the selection of the best
+machine learning technique to forecast the black Sigatoka development
+rate according to the metric indicated below.
 
-The outline of the paper is as follows.
-In Section 2, we introduce the machine learning techniques used in this research and the main related works.
-In section 3, we present the data used and the methodology for the study results.
-In section 4, we present the results and their discussion.
-Our conclusions and future work are presented in Section 5.
+The outline of the paper is as follows: in Section~\ref{sec:techs}, we
+introduce the machine learning techniques used in this research and
+the main related works. Section~\ref{sec:data} summarizes the data the
+methodology used in this study.  The results and their discussion are
+presented in section~\ref{sec:results}.  We present the main
+conclusions and lines for future works in Section~\ref{sec:concl}.
 
 \section{Machine learning techniques and Related works}
+\label{sec:techs}
+
 In this section we summary the machine learning techniques used and we present the main related works.
@@
 \section{Specification of data and methodology}
+\label{sec:data}
+
 In this section, we present the data and how we worked to obtain our results.
@@
 \section{Results and discussion}
+\label{sec:results}
 
@@
 
-\section{Future work}
+\section{Conclusions}
+\label{sec:concl}
+
+\subsection{Future work}
 

### 2016-05-22 Paper Sigatoka - 2015.tex
Some comments to the introduction and SVR

@@
 \newcommand{\instead}[2]{\textcolor{dkred}{\st{#1}}#2}
+\newcommand{\figref}[1]{\figurename~\ref{#1}}
 
@@
 
-\journal{Journal of \LaTeX\ Templates}
+\journal{Computers and Electronics in Agriculture}
 
@@
 the major cause of product losses associated with the black
-Sigatoka. \figurename~\ref{fig:diseasestages} shows three stages of
+Sigatoka. \figref{fig:diseasestages} shows three stages of
 this disease.
@@
 %
-The numeric coefficients describe the degree of incidence and the
-severity of the disease development.  These coefficients are used to
-calculate two variables: gross sum and state of evolution.
+Numeric coefficients are used to describe the degree of incidence and
+the severity of the disease development.  These coefficients are used
+to calculate two variables: gross sum and state of evolution.
 
 The gross sum is based on the present disease progression stage and
-the numeric coefficients, which increases with the progression of
+the numeric coefficients, which increase with the progression of
 the symptoms and the juvenility of the leaf.
@@
 \citep{Marinetal2003}.
+
+\TODO{No queda claro threshold levels sobre qué variables!?  Tampoco
+  queda claro que quiere decir 'fluctuation' como medida de decisión }
 
@@
 
-In this work, we compare five machine learning techniques 
-to predict the development rate of the black Sigatoka disease:
-support
-vector regression (SVR), echo state networks (ESN), ridge regression,
+In this work, we compare five machine learning techniques to predict
+the development rate of the black Sigatoka disease: support vector
+regression (SVR), echo state networks (ESN), ridge regression,
 elastic-net regression and ordinary least squares linear regression.
@@
 
-The outline of the paper is as follows: in Section~\ref{sec:techs}, we
-introduce the machine learning techniques used in this research and
-the main related works. Section~\ref{sec:data} summarizes the data the
-methodology used in this study.  The results and their discussion are
-presented in section~\ref{sec:results}.  We present the main
-conclusions and lines for future works in Section~\ref{sec:concl}.
+\TODO{Definitivamente ese no es el aporte principal del artículo.
+  Cuando esté listo el resto hay que volver aquí.  Creo que va a ser
+  algo como mostrar la relevancia de la etapa de preprocesamiento, o
+  mostrar que a pesar de ser sistemas complejos, con variables
+  caóticas, lo mejor en este caso son predictores lineales, o algo por
+  el estilo.}
+
+The outline of the paper is as follows: Section~\ref{sec:techs}
+summarizes the machine learning techniques used in this research and
+presents the main related works. In Section~\ref{sec:data} we present
+the methodology used in this study and describe data used for its
+verification.  The results and their discussion are presented in
+section~\ref{sec:results}.  The Section~\ref{sec:concl} concludes this
+article and presents lines for future works.
 
@@
 
-In this section we summary the machine learning techniques used and we present the main related works.
+In this section we summarize the machine learning techniques used and we present the main related works.
 
@@
 
-From the perspective of Support Vector Regression (SVR) the regression function $y = f(s)$ for a given dataset $D=\{(s_i,y_i)\}_{i=1}^n$ , is represented as a linear function of the form \citep{Wei2013}:
-$$f(s)=w^Ts+b$$
-where $w$ and $b$ are respectively the weight vector and the intercept of the model, and they are selected to find an optimal fit to the data available in $D$.
+\TODO{Estoy seguro que los lectores van a solicitar reducir esto y
+  dejar solo la referencia.  No hay nada nuevo en el resto del paper
+  que requiera toda esta información, así que hay que resumirlo a nada
+  más lo esencial. }
 
-For nonlinear cases, one proceeds by mapping the input p-dimensional vectors via a nonlinear function $\phi : R^p\rightarrow F$, onto the feature space $F$.
-After nonlinear mapping, the regression function evolves to a pervasive form:
+From the perspective of Support Vector Regression (SVR) the regression
+function $y = f(s)$ for a given dataset $D=\{(s_i,y_i)\}_{i=1}^n$ is
+represented as a linear function of the form \citep{Wei2013}:
+\begin{equation*}
+  f(s)=w^Ts+b
+\end{equation*}
+where $w$ and $b$ are respectively the weight vector and the intercept
+of the model, and they are selected to find an optimal fit to the data
+available in $D$ in the sense of ...?
+\TODO{optimal respecto a qué?.  Hay muchísimas posibilidades de optimalidad!}
 
-$$f(s)=w^T \phi (s)+b$$
+For nonlinear cases, the $p$-dimensional input vectors are mapped by a
+nonlinear function $\phi : R^p\rightarrow F$ onto a feature space
+$F$.  After the nonlinear mapping, the regression function evolves to a
+pervasive form:
+\begin{equation*}
+f(s)=w^T \phi (s)+b
+\end{equation*}
 
-SVR uses the $\epsilon$-insensitive loss function:
+In SVR the generalization ability is measured by the Euclidean norm of
+$w$, and the empirical error is measured by the $\epsilon$-insensitive
+loss function, given by \citep{Cristiani2005}
+
+\TODO{si $l$ es una función, es función de qué, de $s$? $l(s)$? o de
+  $y$? o de todos $l(y,s;\epsilon)$?}
+
 \begin{equation*}
@@
 \end{equation*}
-which ignores the error if the difference between the prediction value and the actual value is smaller than $\epsilon$.
+which ignores the error if the difference between the prediction value
+and the actual value is smaller than $\epsilon$.
 %
-The $\epsilon$-insensitive loss function allows to find the coefficients $w$ and $b$ by solving a convex optimization problem, which balances the empirical error and the generalization ability.
-%
-In SVR, the empirical error is measured by the loss function $\epsilon$-insensitive and the generalization ability is measured by the Euclidean norm of $w$ \citep{Cristiani2005}.
+The $\epsilon$-insensitive loss function allows to find the
+coefficients $w$ and $b$ by solving a convex optimization problem,
+which balances the empirical error and the generalization ability.
 %
 Then, the optimization problem to identify the regression model can be
-formulated by \citep{Wei2013}:
-
+formulated as \citep{Wei2013}:
+%
 \begin{equation} 
@@
 \end{equation}  
-where $C$ denotes the penalty parameter between empirical and generalization errors, and  $\xi_i , \xi_i^*$ are slack variables. \figurename
-$.$\ref{figura2} shows this situation.
+where $C$ denotes the penalty parameter between empirical and
+generalization errors, and $\xi_i , \xi_i^*$ are slack
+variables. \figref{figura2} shows this situation.
 
@@
 
-The solution of this optimization problem by the Lagrange method is given by:\\
-$$ f(s) = w^T \phi (s) + b = \sum_{i=1}^{n} (\alpha_i - \alpha_i^*) K (s,s_i) + b $$ 
-where $\alpha_i - \alpha_i^*$  are the Lagrange multipliers of the optimization problem’s dual form and $K(s_i,s_j )$ is the kernel function satisfying the Mercer condition, and holds:
-$$K(s_i,s_j ) = \big \langle  \phi(s_i) , \phi(s_j)  \big \rangle $$
-Operations in the kernel function $K(s,s_i )$ are performed in the input space rather than in the potentially high dimensional feature space of $\phi$ \citep{Alonso2013}.
+The solution of this optimization problem by the Lagrange method is given by:
+\begin{equation*}
+f(s) = w^T \phi (s) + b = \sum_{i=1}^{n} (\alpha_i - \alpha_i^*) K (s,s_i) + b
+\end{equation*}
+where $\alpha_i - \alpha_i^*$ are the Lagrange multipliers of the
+optimization problem’s dual form and $K(s_i,s_j )$ is the kernel
+function satisfying the Mercer condition, and holds:
+\begin{equation*}
+K(s_i,s_j ) = \big \langle  \phi(s_i) , \phi(s_j)  \big \rangle
+\end{equation*}
+Operations in the kernel function $K(s,s_i )$ are performed in the
+input space rather than in the potentially high dimensional feature
+space of $\phi$ \citep{Alonso2013}.
 
 \subsection{Ordinary least squares regression}
-This method fits a linear model with coefficients $w = (w1,..,wp)$ to minimize the residual sum of squares between the observed responses in the dataset, and the responses predicted by the linear approximation.
-Mathematically it solves a problem of the form \citep{scikitlearn2011}:
+
+This method fits a linear model with coefficients $w = (w_1,..,w_p)$
+to minimize the residual sum of squares between the observed responses
+in the dataset, and the responses predicted by the linear
+approximation. Mathematically it solves a problem of the form
+\citep{scikitlearn2011}:
 $$\min_{w} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  $$
-where $X$ denotes the features matriz. \\
+where $X$ denotes the features matrix.
+
 According Pedregosa et al.
@@
 When terms are correlated and the columns of the design matrix $X$ have an approximate linear dependence, the design matrix becomes close to singular and as a result, the least-squares estimate becomes highly sensitive to random errors in the observed response, producing a large variance.
-This situation of multicollinearity can arise, for example, when data are collected without an experimental design
+This situation of multicollinearity can arise, for example, when data are collected without an experimental design.
 
@@
 The untrained RNN part of an ESN is called a dynamical reservoir, and the resulting states x(n) are termed echoes of its input history \citep{Lukose2012}.
-Finally, these signals are sent to an output layer as shown in the \figurename
-$.$\ref{figura3}.
+Finally, these signals are sent to an output layer as shown in the \figref{figura3}.
 \begin{figure}[h] 
@@
 
-\figurename $.$\ref{figura4} shows the best $R^2$ for each algorithm in the experiment.
+\figref{figura4} shows the best $R^2$ for each algorithm in the experiment.
 Results are group by farm.
@@
 
-\figurename $.$\ref{figura5} presents, for one, two and three weeks ahead, the best $R^2$.
+\figref{figura5} presents, for one, two and three weeks ahead, the best $R^2$.
 Results are group by farm.
@@
 
-\figurename $.$\ref{figura6} shows the best $R^2$ for each variables combination.
+\figref{figura6} shows the best $R^2$ for each variables combination.
 Results are group by farm.
@@
 
-\figurename $.$\ref{figura7} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
+\figref{figura7} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
 The Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better $RMSE$ than La Rita.
@@
 
-\figurename $.$\ref{figura8} shows the best $R^2$ for each algorithm in the experiment.
+\figref{figura8} shows the best $R^2$ for each algorithm in the experiment.
 Results are group by farm.
@@
 
-\figurename $.$\ref{figura9} presents, for one, two and three weeks ahead, the best $R^2$.
+\figref{figura9} presents, for one, two and three weeks ahead, the best $R^2$.
 Results are group by farm.
@@
 
-\figurename $.$\ref{figura10} shows the best $R^2$ for each variables combination.
+\figref{figura10} shows the best $R^2$ for each variables combination.
 Results are group by farm.
@@
 
-\figurename $.$\ref{figura11} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.  You can note that the behaviour of $R^2$ and $RMSE$ is similar to the phase one.
+\figref{figura11} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.  You can note that the behaviour of $R^2$ and $RMSE$ is similar to the phase one.
 

### 2016-06-11 conclusions.tex
Nueva organizacion de los archivos.  Partes del paper en archivos separados y todas las imagenes en subdirectorio fig

--- 2016-05-22 Paper Sigatoka - 2015.tex
@@
-\documentclass[review,authoryear,english]{elsarticle}
-
-\usepackage{babel}
-\usepackage{natbib}
-
-\bibliographystyle{elsarticle-harv}
-
-\usepackage{lineno}
-\usepackage[utf8]{inputenc}  %% Permite utilizar tildes si archivo usa utf-8
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\modulolinenumbers[1]
-
-\usepackage{soul}
-\usepackage{marginnote}
-\usepackage{float}
-
-\usepackage{color}
-\definecolor{dkred}{rgb}{0.5,0,0}       %   dark red
-\definecolor{todocolor}{rgb}{1.0,0.8,0.5}
-
-\usepackage[hidelinks]{hyperref}
-
-\newcommand{\todohdr}{Please review:}
-\newcommand{\TODO}[2][\todohdr]{%
-\smallbreak\noindent\colorbox{todocolor}{%
-\begin{minipage}{1\linewidth}
-\textbf{#1}\par{\footnotesize{#2}}
-\end{minipage}}}
-
-\newcommand{\instead}[2]{\textcolor{dkred}{\st{#1}}#2}
-
-%  \usepackage{epstopdf} 
-%  \usepackage{epsfig}
-
-\journal{Journal of \LaTeX\ Templates}
-
-\usepackage{subcaption}
-
-\begin{document}
-
-\begin{frontmatter}
-
-\title{Forecasting the black Sigatoka development rate:\\ 
-A comparison of machine learning techniques}
-
-%% Group authors per affiliation:
-\author[afiLuisAlex]{Luis-Alexander Calvo-Valverde\fnref{myfootnote}}
-\ead{lualcava.sa@gmail.com}
-\fntext[myfootnote]{Corresponding author. (506)70104420}
-
-\author[afiCorbana]{Mauricio Guzmán-Quesada}
-\author[afiCorbana]{José-Antonio~Guzmán-Alvarez}
-\author[afiPablo]{Pablo Alvarado-Moya}
-
-\address[afiLuisAlex]{DOCINADE, Instituto Tecnológico de Costa Rica, 
-Computer Research Center, Multidisciplinar program eScience, Cartago, Costa Rica}
-
-\address[afiCorbana]{Dirección de Investigaciones, Corporación Bananera Nacional S.A., Guápiles, Costa Rica}
-
-\address[afiPablo]{DOCINADE, Instituto Tecnológico de Costa Rica, Cartago, Costa Rica}
-
-\begin{abstract}
-Pending.
-\end{abstract}
-
-\begin{keyword}
-Machine learning \sep Black Sigatoka \sep Support vector regression \sep
-Banana disease prediction \sep Biological warning system 
-\end{keyword}
-
-\end{frontmatter}
-
-\linenumbers
-
-\section{Introduction}
-
-The black Sigatoka disease caused by the fungus \emph{Mycosphaerella
-  fijiensis Morelet} is the major pathological problem of banana and
-plantain crops in Central America, Panama, Colombia and Ecuador, as well as in
-many parts of Africa and Asia \citep{MarinVargas1995}.
-
-This disease attacks the plant leaves producing a rapid deterioration
-of the leaf area, affects the growth and productivity of the plants
-due to the impairment of the photosynthetic process, causes a
-reduction in the quality of the fruit, and promotes premature
-ripening of bunches, which is
-the major cause of product losses associated with the black
-Sigatoka. \figurename~\ref{fig:diseasestages} shows three stages of
-this disease.
-
-\begin{figure}[ht] 
-\centering
-\begin{tabular}{c@{\;}c@{\;}c}
-  \includegraphics[width=.32\linewidth]{Roya_a} &
-  \includegraphics[width=.32\linewidth]{Roya_b} &
-  \includegraphics[width=.32\linewidth]{Roya_c} \\
-  (a) & (b) & (c) 
-\end{tabular}
-\caption{Examples of three disease stages of the black Sigatoka. 
-(a)~Initial stage. (b)~Intermediate stage, and (c)~Advanced stage.} 
-\label{fig:diseasestages} 
-\end{figure}
-
-Phytopathological studies point out that precipitation, temperature,
-relative humidity and wind are the main climatic variables that affect
-the development of this disease \citep{MarinVargas1995}.
-
-The climate has a major effect on the development of the black
-Sigatoka.  Precipitation, temperature, relative humidity and wind are
-the main climatic variables affecting the development of this disease
-\citep{MarinVargas1995}.
-
-The early warning system developed by Ganry and Meyer
-\citep{ganry1972} and modified by Ganry and Laville \citep{ganry1983}
-to control yellow Sigatoka in Cameroon, was later adapted by
-Ternesien \citep{Ternesien1985} and Fouré \citep{foure1988} for the
-black Sigatoka.
-
-This bilogical warning system is based on weekly quantified
-observations of the disease progression on young leaves of the plant,
-according to Fouré's scale of the symptom stages.
-%
-The numeric coefficients describe the degree of incidence and the
-severity of the disease development.  These coefficients are used to
-calculate two variables: gross sum and state of evolution.
-
-The gross sum is based on the present disease progression stage and
-the numeric coefficients, which increases with the progression of
-the symptoms and the juvenility of the leaf.
-%
-The state of evolution is calculated using the gross sum and the
-foliar emission period.
-%
-Although threshold levels were initially suggested as a guide to spray
-schedules, the fluctuation of these two variables
-was found to better define appropiate times to spray
-\citep{Marinetal2003}.
-
-In Costa Rica the black Sigatoka is frequently treated with
-chemical fungicides.
-%
-Depending on the zone of production and the weather conditions,
-45--55~cycles/year of fungicide applications are required to keep this
-disease under control and to produce the expected fruit quality for the
-international markets.
-%
-This represents a cost per hectare per year in the range between US\$1600 and
-US\$2000; about 0.64--0.80 cents of the production costs for a
-18.14\,kg box.
-%
-Overall, this represents 10\%--12\% of the total production costs.
-
-The past and present rates of disease development can in principle be used
-to predict its future behavior and to determine whether
-a particular fungicide spray program will be able to effectively and
-economically control the disease \citep{ChuangJeger1987} or not.
-
-There are efforts to apply machine learning methods to support
-decision-making in agriculture, including the control of crop
-diseases. For example, \cite{Camargo2012} present an intelligent
-system for the assessment of crop disorders, \cite{Huang2010}
-introduce a plant virus identification method based on neural networks
-with an evolutionary preprocessing stage, \cite{Kim2014} summarize in
-their survey crop pests prediction methods using regression and
-machine learning approaches, while \cite{Zhao2013} present an
-intelligent agricultural forecasting system based on wireless sensor
-networks.
-
-In this work, we compare five machine learning techniques 
-to predict the development rate of the black Sigatoka disease:
-support
-vector regression (SVR), echo state networks (ESN), ridge regression,
-elastic-net regression and ordinary least squares linear regression.
-%
-The main contribution of this work is the selection of the best
-machine learning technique to forecast the black Sigatoka development
-rate according to the metric indicated below.
-
-The outline of the paper is as follows: in Section~\ref{sec:techs}, we
-introduce the machine learning techniques used in this research and
-the main related works. Section~\ref{sec:data} summarizes the data the
-methodology used in this study.  The results and their discussion are
-presented in section~\ref{sec:results}.  We present the main
-conclusions and lines for future works in Section~\ref{sec:concl}.
-
-\section{Machine learning techniques and Related works}
-\label{sec:techs}
-
-In this section we summary the machine learning techniques used and we present the main related works.
-
-\subsection{Support Vector Regression (SVR)}
-
-From the perspective of Support Vector Regression (SVR) the regression function $y = f(s)$ for a given dataset $D=\{(s_i,y_i)\}_{i=1}^n$ , is represented as a linear function of the form \citep{Wei2013}:
-$$f(s)=w^Ts+b$$
-where $w$ and $b$ are respectively the weight vector and the intercept of the model, and they are selected to find an optimal fit to the data available in $D$.
-
-For nonlinear cases, one proceeds by mapping the input p-dimensional vectors via a nonlinear function $\phi : R^p\rightarrow F$, onto the feature space $F$.
-After nonlinear mapping, the regression function evolves to a pervasive form:
-
-$$f(s)=w^T \phi (s)+b$$
-
-SVR uses the $\epsilon$-insensitive loss function:
-\begin{equation*}
-l= {\left| y - f(s) \right| }_{\epsilon } = 
-\begin{cases}
-   0 & {\left| y - f(s) \right| } \leq \epsilon \\
-   \left| y - f(s) \right| - \epsilon & \text{otherwise}
-\end{cases}
-\end{equation*}
-which ignores the error if the difference between the prediction value and the actual value is smaller than $\epsilon$.
-%
-The $\epsilon$-insensitive loss function allows to find the coefficients $w$ and $b$ by solving a convex optimization problem, which balances the empirical error and the generalization ability.
-%
-In SVR, the empirical error is measured by the loss function $\epsilon$-insensitive and the generalization ability is measured by the Euclidean norm of $w$ \citep{Cristiani2005}.
-%
-Then, the optimization problem to identify the regression model can be
-formulated by \citep{Wei2013}:
-
-\begin{equation} 
-\begin{aligned}
-& \underset{}{\text{minimize}}
-& & J(w,\xi_i , \xi_i^* ) = \frac{1}{2}   \Bigr| \Bigr| w \Bigr| \Bigr|^2 + C \sum_{i=1}^{n} (\xi_i , \xi_i^* )    \\
-& \text{subject to}
-& & \begin{array}{lcl} 
-y_i - w^T \phi (s) - b  \leq \epsilon + \xi_i   \\
-w^T \phi (s) + b - y_i \leq \epsilon + \xi_i^* & i= 1,2,...,n \\
-\xi_i , \xi_i^* \geq 0  \\
-\end{array}
-\end{aligned}
-\end{equation}  
-where $C$ denotes the penalty parameter between empirical and generalization errors, and  $\xi_i , \xi_i^*$ are slack variables. \figurename
-$.$\ref{figura2} shows this situation.
-
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{SVR}
- \caption{$\epsilon$-insensitive loss function \citep{Wei2013}.} 
- \label{figura2} 
-\end{figure}
-
-The solution of this optimization problem by the Lagrange method is given by:\\
-$$ f(s) = w^T \phi (s) + b = \sum_{i=1}^{n} (\alpha_i - \alpha_i^*) K (s,s_i) + b $$ 
-where $\alpha_i - \alpha_i^*$  are the Lagrange multipliers of the optimization problem’s dual form and $K(s_i,s_j )$ is the kernel function satisfying the Mercer condition, and holds:
-$$K(s_i,s_j ) = \big \langle  \phi(s_i) , \phi(s_j)  \big \rangle $$
-Operations in the kernel function $K(s,s_i )$ are performed in the input space rather than in the potentially high dimensional feature space of $\phi$ \citep{Alonso2013}.
-
-\subsection{Ordinary least squares regression}
-This method fits a linear model with coefficients $w = (w1,..,wp)$ to minimize the residual sum of squares between the observed responses in the dataset, and the responses predicted by the linear approximation.
-Mathematically it solves a problem of the form \citep{scikitlearn2011}:
-$$\min_{w} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  $$
-where $X$ denotes the features matriz. \\
-According Pedregosa et al.
-\citep{scikitlearn2011}  the coefficient estimates for Ordinary Least Squares rely on the independence of the model terms.
-When terms are correlated and the columns of the design matrix $X$ have an approximate linear dependence, the design matrix becomes close to singular and as a result, the least-squares estimate becomes highly sensitive to random errors in the observed response, producing a large variance.
-This situation of multicollinearity can arise, for example, when data are collected without an experimental design
-
-\subsection{Ridge regression}
-The ridge regression addresses some of the problems of ordinary least squares regression by imposing a penalty on the size of the coefficients.
-The ridge coefficients minimize a penalized residual sum of squares  \citep{scikitlearn2011}:
-
-$$\min_{w} { \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-Here, $\alpha \ \textgreater \  0$ is a complexity parameter that controls the amount of shrinkage: the larger the value of $\alpha$, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
-
-\subsection{Elastic-Net regression}
-Elastic-Net is a linear regression model trained with $L1$ and $L2$ prior as regularizer.
-This combination allows for learning a sparse model where few of the weights are non-zero like Lasso, while still maintaining the regularization properties of Ridge  \citep{scikitlearn2011}.
-The convex combination of $L1$ and $L2$ is controled by using the $l1_{ratio}$ parameter.
-
-Elastic-Net is useful when there are multiple features which are correlated with one another. Lasso is likely to pick one of these at random, while elastic-net is likely to pick both.
-A practical advantage of trading-off between Lasso and Ridge is it allows Elastic-Net to inherit some of Ridge’s stability under rotation.
-The objective function to minimize is \citep{scikitlearn2011}:
-
-$$\min_{w} { \frac{1}{2n_{samples}} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \rho \Bigr| \Bigr| w \Bigr| \Bigr|_1 + \frac{\alpha (1- \rho)}{2} \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-\subsection{Echo State Networks (ESN)}
-Recurrent Neural Networks (RNN) are useful for temporal patterns, but when they are trained with backpropagation methods, they are very slow.
-Echo State Network (ESN) is an alternative training method to solve that problem.
-ESN is based on the observation that if a random RNN possesses certain algebraic properties, training only a linear readout from it is often sufficient to achieve excellent performance in practical applications \citep{Lukose2009}.
-For a given training input signal $u(n)  \in R^{N_u}$ a desired target output signal $y^{target}(n) \in R^{N_y}$
-is known.
-Here $n = 1, .
-.
-.
-,T$ is the discrete time and $T$ is the number of data points in the training dataset.
-The task is to learn a model with output $y(n) \in R^{N_y}$, where $y(n)$ matches $y^target(n)$ as well as possible, minimizing an error measure $E(y,y^target)$, and, more importantly, generalizes well to unseen data.
-The untrained RNN part of an ESN is called a dynamical reservoir, and the resulting states x(n) are termed echoes of its input history \citep{Lukose2012}.
-Finally, these signals are sent to an output layer as shown in the \figurename
-$.$\ref{figura3}.
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{Reservorio}
- \caption{An echo state network \citep{Lukose2012}.} 
- \label{figura3} 
-\end{figure}
-
-The connections between the different elements of an Echo State Network have weights randomly generated.
-The weights of the internal connections of the reservoir $(W)$ as well as the weights of the input layer $(W_in)$, after being generated are set statically during all stages of implementation of the algorithm.
-The weights between the reservoir and the output layer $(W_out)$ are subject to changes of a supervised learning algorithm to correct the degree of error generated by the entire system \citep{Lukose2012}.
-
-\subsection{Related works}
-Now, we present the main and recently works related with our study.
-
-Huang et al.
-\citep{Huang2010}  surveyed the development of soft computing techniques in agricultural and biological engineering, including fuzzy logic, artificial neural networks, genetic algorithms, bayesian inference and decision trees.
-
-A related work, proposed by Romero \citep{Romero1995} relies on regression models using a stepwise procedure to predict incubation and latency times of black Sigatoka.
-The author performed experiments on two farms located in Costa Rica (La Rita and Waldeck, the same as those used in this study but with different names).
-The study used data from: December 1993 to August 1995.
-Romero concluded that the model to predict the incubation period accounted a $R^2$ of 69\% in his observed data but it was not a good predictor when it was validated against an independent dataset (cross validation).
-For latency, he developed two models that accounted a $R^2$ of 78\% in the observed data, however, when validated against an independent dataset (cross validation), for La Rita obtained an adjusted $R^2$ of 82\%, and for Weldeck, none of the models satisfactorily predicted the latent period; and then those predictions were not shown \citep{Romero1995}.
-
-Glezakos et al.
-\citep{Glezakos2010} proposed to use Genetic Algorithms (GA) and Neural Networks (NN) to identify plant virus (Tobacco Rattle Virus (TRV) and the Cucumber Green Mottle Mosaic Virus (CGMMV)).
-This is achieved by the development of ana- lytical tools of evolutionary adaptive width, propelled by Genetic Algorithms (GAs) and Neural Networks (NNs).
-The method was tested against some of the most commonly used classifiers in machine learning (Bayes, Trees and k-NN) via cross-validation and proved its potential towards the identification.
-
-In the agricultural context, Alves et al.
-\citep{Alves2011} used  geoinformation techniques to develop predictive models to study the areas of risk to soybean rust in soybean, coffee leaf rust in coffee, and black Sigatoka in banana, considering Brazil’s climatic characterization and the distribution of soybean, coffee and banana crops.
-Temperature and rainfall data were obtained for the period from 1950 to 2000, and of simulations for 2020, 2050 and 2080 using the SRES A2 climate change scenarios.
-Using principal components analysis, a single variable was generated based on 57 variables, in order to determine an index explaining 87\%, 88\% and 90\% of the variability of soybean, coffee and banana crops, respectively, in municipal districts across Brazil.
-The climatic model was used to generate the zoning of the three plant diseases, using temperature and leaf wetness as input.
-Areas of favorability for the diseases were plotted against the main coffee, soybean and banana growing areas in Brazil.
-This methodology enabled the visualization of the changes in areas favorable for epidemics under possible future scenarios of climate change.
-
-Other applications of machine learning methods in precision agriculture include the use of support vector regression to predict carcass weight in beef cattle in advance to the slaughter  \citep{Alonso2013}, machine learning assessments of soil drying for agricultural planning \citep{Coopersmith2014}, and early detection and classification of plant diseases with support vector machines based on hyperspectral reflectance \citep{Rumpf2010}.
-
-Furthermore, there have been attempts to generate software tools.
-Camargo et al.
-\citep{Camargo2012} presented an information system for the assessment of plant disorders (Isacrodi).
-They proposed that experts will attain a much better accuracy than the Isacrodi classifier, particularly when provided with samples from the affected crop.
-However, those cases where such expertise is not available, they suggest that Isacrodi can provide valuable support to farmers.
-Isacordi includes 15 crop disorders, but the black Sigatoka no is one of them.
-The prediction process is based on multi-class support vector machines.
-
-Regarding the prediction of the develpment of the black Sigatoka with machine learning methods, Bendini et al.
-\citep{Bendini2013}  presented a study about the risk analysis of black Sigatoka occurrence based on polynomial models.
-A case study was developed in a commercial banana plantation located in Jacupiranga, Brazil.
-It was monitored weekly during the period from February to December 2005.
-Data included the weekly monitoring of the disease’s evolution stage, time series of meteorological data and remote sensing data.
-They obtained a model to estimate the evolution of the disease from satellite imagery.
-This model relates gray levels (NC) of the band 2 images of the Landsat-5 satellite, with the progress status or disease severity (EE).
-The authors claim to reach an $R^2$ of 90\%.
-
-Also there are works related to banana fruit.
-Soares et al.
-\citep{Soares2014} apply two techniques: artificial neural networks (ANNs) and multiple linear regression (MLR) in banana plant to predict the yield, their results show that the neural network proved to be more accurate in forecasting the weight of the bunch in comparison to the multiple linear regressions in terms of the mean prediction-error $(MPE = 1.40)$, mean square deviation $(MSD = 2.29)$ and coefficient of determination $(R^2 = 91\%)$.
-
-In general, the machine learning methods applied to predict the evolution of plant diseases, can be classified in two main approaches: 1) Those whose main inputs are images, and 2) Those whose main inputs are environmental and biological variables.
-Our study focuses in the second case.
-
-\section{Specification of data and methodology}
-\label{sec:data}
-
-In this section, we present the data and how we worked to obtain our results.
-
-\subsection{Data}
-We use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (previously called Waldeck and located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
-The banana type is Musa AAA, subgroup Cavendish, cv.
-Grande Naine.
-The \tablename $.$\ref{tabla1} shows the variables available.
-
-\begin{table}[h] 
-\caption{Variables used in the study} 
-\label{tabla1} 
-\centering
-\begin{tabular}{c|c} 
-\hline
-\bfseries Variable & \bfseries Meaning0'0' \\ 
-\hline\hline 
-$T_{a_{max}}$ &	Max air temperature \\
-$T_{a_{min}}$ &	Min air temperature \\
-$\overline{T}_{a}$	 & Mean air temperature \\
-$\overline{H}$	&  Mean Humidity \\
-$H_{min}$ &	Min humidity \\
-$H_{max}$	& Max humidity \\
-$\overline{R}$	& Mean Solar radiation \\
-$P$	& Sum precipitation  \\
-$W_{max}$	& Max speed wind \\
-$\overline{W}$	& Mean speed wind  \\
-$L_2$	& Biological warning system – Leaf 2 \\
-$L_3$	& Biological warning system – Leaf 3 \\
-$L_4$	& Biological warning system – Leaf 4 \\
-$E_s$	& Biological warning system – Evolution Stage \\
-\hline
-\end{tabular} 
-\end{table}
-
-The value to be predicted in all cases was $E_s$, that is the total measure of the biological warning system.
-
-The data on the biological warning system are collected once a week.
-Although Corbana has meteorological stations that take data every five minutes, for these experiments, weekly averages generated by nearby stations to each of the farms were used.
-
-The time intervals used for this study were: La Rita, week 48 of 2002 to week 17 of the 2015 (647 weeks) and for 28 Miles, week 37 of 2003 to week 18 of 2015 (605 weeks).
-
-\subsection{Data preprocessing}
-Usually, data need preprocessing to prepare for use. In this section we explain that process.
-
-In 28 Miles farm, 1\% of the data were missing, while in La Rita was 2.25\%. To fill-in the missing values we use spline interpolation. The data collected did not exhibit outliers.
-
-Due the fact that the variables measure meteorological or biological process, they are discretized in order to reflect trends in the data, i.e.
-the continuous values are not directly used.
-The coefficient of variation $C_v(x)$ of each variable x was used to determine the number $n$ of discretization levels.
-$$n= \lfloor 100 \ C_v(x) \rfloor$$
-where $\lfloor \ \rfloor$  is the round operator.
-
-Each discretization range was uniformly partitioned.
-Besides enabling the capture of tendencies, the discretization removes the effect of small variations in the data collection, either by inaccuracies of the instruments (meteorological variables) or by subjective bias introduced by the human who collects the data (biological warning system).
-
-Each feature was scaled to fit in a range between 0 and 1. The variable to be predicted was not scaled.
-
-\subsection{Evaluation criteria}
-
-Although there are many types of indicators to assess the quality of the prediction, we selected the determination coefficient $(R^2)$ and the Root Mean Square Error $(RMSE)$.
-
-Given $n$ records, let be $y$ the actual value of the series, $\hat{y}$ the predicted value and $\acute{y}$ the mean of the observed data.
-
-$$ \bar{y} = \frac{1}{n} \sum_{i=1}^{n} y_i $$
-
-$$ S_e^2 = \frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}$$
-
-$$ S_R^2 = \frac{\sum_{i=1}^{n} {(\hat{y}_i-\bar{y}_i)}^2 }{n}$$
-
-$$ S_y^2 = S_R^2 + S_e^2$$
-
-$$ R^2 = \frac{S_R^2}{S_y^2}$$
-
-$$ RMSE = \sqrt{\frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}}$$
-
-This decision is supported by the widespread use in machine learning and agriculture areas \citep{Soares2014}, \citep{Soares2013}, \citep{Ibrahim2014} and \citep{Demir2014}.
-
-\subsection{Programming environment}
-
-We use the python programming language with the Integrated Development Environment (IDE) Spyder \citep{Continuum2015}, particularly with the libraries pandas \citep{mckinneypandas2010} and numpy \citep{vanderWalt2011}.
-For SVR, ridge and ordinary least squares regressions, we used sklearn \citep{scikitlearn2011} and for ESN the python-based code of Dr.
-$Luko\breve{s} evi \breve{c} ius$ \citep{Lukose2012} on which the necessary were done for adjustments for the experiments of this work.
-The computer used a processor Intel(R) Core i7-4800MQ CPU 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
-
-\subsection{Methodology}
-The selection of the technique and its parametrisation was performed in two stages that they are presented below.
-
-{\bf Phase one }
-
-In the phase one, we did ten-fold-cross-validation on a set of  machine learning methods and different configurations:
-
-\begin{itemize}
-\item 	Patterns: n by m, where n from 1 to 8 and m from 1 to 2.
-
-\item Methods: support vector regression with the kernels functions: linear, gaussian and sigmoid; echo state networks; ordinary least squares linear regression, ridge regression and elastic-net regression.
-
-\item Variables included in the model:
-\begin{itemize}
-\item All variables.
-\item From the set $\{ \overline{T}_{a} , \overline{H}, P , \overline{W}  \}$ use the subsets with one, two or four elements.
-These variables are according to experts the ones having most impact on the disease development \citep{MarinVargas1995}.
-\end{itemize}
-
-\end{itemize}
-
-{\bf Phase two }
-
-In the second phase, the best configurations obtained in phase one are used to validate with the last 50 and 100 weeks.
-
-\section{Results and discussion}
-\label{sec:results}
-
-In this section we present the main results for phase and their discussion.
-
-{\bf Phase one }
-
-\figurename $.$\ref{figura4} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Though La Rita obtains different results in magnitude than 28 Millas, the trend is similar.
-In both farms, the best results are for linear models, second position is occupied for Echo State Networks and  SVR with gausian and sigmoid kernels are the worst results.
-In linear models, to predict one week ahead is better than two weeks ahead, and this is better than three weeks ahead.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_algortihm}
- \caption{Phase one - Best $R^2$ for each algorithm} 
- \label{figura4} 
-\end{figure}
-
-\figurename $.$\ref{figura5} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-In general, to predict one week ahead is better than two weeks ahead and so on.
-The number of weeks consider in the observation range in the pattern is not the main discriminant factor, but it is clear that we get better $R^2$ for one week ahead than two weeks ahead and so on.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase one - Best $R^2$ for each observation range} 
- \label{figura5} 
-\end{figure}
-
-\figurename $.$\ref{figura6} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-The better results are obtained with $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms of similarly.
-You can note that the use of all variables in the model or the inclusion of the four variables suggest for expert criteria do not improve significantly the results, then the use of more sensors do not assure a better result.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
- \caption{Phase one - Best $R^2$ for each variable combination} 
- \label{figura6} 
-\end{figure}
-
-\figurename $.$\ref{figura7} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
-The Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better $RMSE$ than La Rita.
-This situation arise because $RMSE$ considers errors only with respect the prediction and in 28 Millas the average of Stage of Evolution is $4316.16$, unlike, in La Rita the average is $5507.30$.
-So, in La Rita we obtains higher errors in absolute values.
-$R2$ is a relative metric between 0 thru 1 and it is less sensitive to absolute values.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.8]{Phase_one_R2_RMSE}
- \caption{Phase one - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura7} 
-\end{figure}
-
-The Pareto frontier for the La Rita farm is composed by 96 elements. The \tablename $.$\ref{tabla2} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase one} 
-\label{tabla2} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $\overline{W}$ &	1 to 1  & 36 & 64.25\% & 714.51 \\
- &	2 to 1  & 6 & 62.97\% & 695.10 \\
-\hline 
-All  & 1 to 1  & 18 & 62.98\% & 701.95 \\
-   & 2 to 1  & 12 & 61.76\% & 679.92 \\
-    & 3 to 1  & 6 & 60.60\% & 676.42 \\
-    & 5 to 1  &  2 & 60.37\% & 672.39 \\
-\hline    
-$\overline{T}_{a}$ & 1 to 1  & 12  & 63.60\% & 708.77 \\
-       &	2 to 1  & 4 & 62.23\% & 689.55 \\
-\hline
-\end{tabular} 
-\end{table}
-
-Similarly, the Pareto frontier for the 28 Millas farm is composed by 75 elements. The \tablename $.$\ref{tabla3} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - 28 Millas - Phase one} 
-\label{tabla3} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $\overline{W}$ & 1 to 1 & 8 & 57.80\% & 438.09 \\
-\hline 
-All   &	9 to 1 & 2 & 50.93\% & 397.93 \\
-  & 10 to 1	 & 2 & 50.97\% & 398.81 \\
-  &	8 to 1 & 6 & 51.62\% & 398.93 \\
-  &	7 to 1 & 2 & 52.25\% & 400.28 \\
-  &	6 to 1 & 2 & 53.16\% & 404.14 \\
-  &	4 to 1 & 2 & 54.32\% & 407.54 \\
-\hline    
-$\overline{T}_{a}$ & 1 to 1  & 8  & 59.09\% & 439.44 \\
-\hline
-Pair $\overline{T}_{a}$ $\overline{H}$ & 1 to 1	 & 8 & 57.51\% & 428.61 \\
- &	2 to 1 & 20 & 56.91\% & 414.37 \\
- &	3 to 1 & 3 & 54.41\% & 411.55 \\
- &	4 to 1 & 3 & 53.34\% & 406.65 \\
-\hline
-Pair $\overline{T}_{a}$ $P$ & 3 to 1 & 9 & 56.23\% & 422.76 \\
-\hline
-\end{tabular} 
-\end{table}
-
-We can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week.
-
-{\bf Phase two }
-
-In the second phase, the best configurations obtained in phase one were used to validate with the last 50 and 100 weeks.
-
-\figurename $.$\ref{figura8} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Even if linear models continue with good $R^2$, Echo State Networks improve their scores because in $50$ and $100$ last weeks validation, we are in presence of climate change, then the behaviour is less lineal.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_algortihm}
- \caption{Phase two - Best $R^2$ for each algorithm} 
- \label{figura8} 
-\end{figure}
-
-\figurename $.$\ref{figura9} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-The results confirms that, in general, to predict one week ahead is better than two weeks ahead, two than three and so on, this for both farms.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase two - Best $R^2$ for each observation range} 
- \label{figura9} 
-\end{figure}
-
-\figurename $.$\ref{figura10} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-This results confirm that $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms are the best variables combinations.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
- \caption{Phase two - Best $R^2$ for each variable combination} 
- \label{figura10} 
-\end{figure}
-
-\figurename $.$\ref{figura11} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.  You can note that the behaviour of $R^2$ and $RMSE$ is similar to the phase one.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.8]{Phase_two_R2_RMSE}
- \caption{Phase two - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura11} 
-\end{figure}
-
-In phase two, the Pareto frontier for the La Rita farm is composed by 2 elements. The \tablename $.$\ref{tabla4} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase two} 
-\label{tabla4} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-All & 1 to 1  & 2 & 69.33\% & 353.33 \\
-\hline
-\end{tabular} 
-\end{table}
-
-In 28 Millas Farm, the Pareto frontier is composed by 1 element. The \tablename $.$\ref{tabla5} details the result.
-
-\begin{table}[h] 
-\caption{Pareto frontier - 28 Millas - Phase two} 
-\label{tabla5} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $P$ & 1 to 1 & 1 & 61.36\% & 339.89 \\
-\hline
-\end{tabular} 
-\end{table}
-
-Again, similar to phase one, we can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week, one week to predict one week ahead with combinations of variables listed above.
-
 \section{Conclusions}
@@
 \subsection{Future work}
-
-\section{Acknowledgements}
-
-The authors would like to thank Corbana S.A. for providing the data for this research.
-
-\section{References}
-
-\bibliography{references}
-
-\end{document}

### 2016-06-11 intro.tex
Nueva organizacion de los archivos.  Partes del paper en archivos separados y todas las imagenes en subdirectorio fig

--- 2016-05-22 Paper Sigatoka - 2015.tex
@@
-\documentclass[review,authoryear,english]{elsarticle}
-
-\usepackage{babel}
-\usepackage{natbib}
-
-\bibliographystyle{elsarticle-harv}
-
-\usepackage{lineno}
-\usepackage[utf8]{inputenc}  %% Permite utilizar tildes si archivo usa utf-8
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\modulolinenumbers[1]
-
-\usepackage{soul}
-\usepackage{marginnote}
-\usepackage{float}
-
-\usepackage{color}
-\definecolor{dkred}{rgb}{0.5,0,0}       %   dark red
-\definecolor{todocolor}{rgb}{1.0,0.8,0.5}
-
-\usepackage[hidelinks]{hyperref}
-
-\newcommand{\todohdr}{Please review:}
-\newcommand{\TODO}[2][\todohdr]{%
-\smallbreak\noindent\colorbox{todocolor}{%
-\begin{minipage}{1\linewidth}
-\textbf{#1}\par{\footnotesize{#2}}
-\end{minipage}}}
-
-\newcommand{\instead}[2]{\textcolor{dkred}{\st{#1}}#2}
-\newcommand{\figref}[1]{\figurename~\ref{#1}}
-
-%  \usepackage{epstopdf} 
-%  \usepackage{epsfig}
-
-\journal{Computers and Electronics in Agriculture}
-
-\usepackage{subcaption}
-
-\begin{document}
-
-\begin{frontmatter}
-
-\title{Forecasting the black Sigatoka development rate:\\ 
-A comparison of machine learning techniques}
-
-%% Group authors per affiliation:
-\author[afiLuisAlex]{Luis-Alexander Calvo-Valverde\fnref{myfootnote}}
-\ead{lualcava.sa@gmail.com}
-\fntext[myfootnote]{Corresponding author. (506)70104420}
-
-\author[afiCorbana]{Mauricio Guzmán-Quesada}
-\author[afiCorbana]{José-Antonio~Guzmán-Alvarez}
-\author[afiPablo]{Pablo Alvarado-Moya}
-
-\address[afiLuisAlex]{DOCINADE, Instituto Tecnológico de Costa Rica, 
-Computer Research Center, Multidisciplinar program eScience, Cartago, Costa Rica}
-
-\address[afiCorbana]{Dirección de Investigaciones, Corporación Bananera Nacional S.A., Guápiles, Costa Rica}
-
-\address[afiPablo]{DOCINADE, Instituto Tecnológico de Costa Rica, Cartago, Costa Rica}
-
-\begin{abstract}
-Pending.
-\end{abstract}
-
-\begin{keyword}
-Machine learning \sep Black Sigatoka \sep Support vector regression \sep
-Banana disease prediction \sep Biological warning system 
-\end{keyword}
-
-\end{frontmatter}
-
-\linenumbers
-
 \section{Introduction}
@@
 %
-Although threshold levels were initially suggested as a guide to spray
+Although \todofn{threshold levels}% 
+{No queda claro threshold levels sobre qué variables!?  Tampoco
+  queda claro que quiere decir 'fluctuation' como medida de decisión}
+%
+were initially suggested as a guide to spray
 schedules, the fluctuation of these two variables
@@
 \citep{Marinetal2003}.
-
-\TODO{No queda claro threshold levels sobre qué variables!?  Tampoco
-  queda claro que quiere decir 'fluctuation' como medida de decisión }
 
@@
 %
-The main contribution of this work is the selection of the best
-machine learning technique to forecast the black Sigatoka development
-rate according to the metric indicated below.
-
-\TODO{Definitivamente ese no es el aporte principal del artículo.
-  Cuando esté listo el resto hay que volver aquí.  Creo que va a ser
-  algo como mostrar la relevancia de la etapa de preprocesamiento, o
+\todofn{The main contribution of this work}%
+{Definitivamente ese no es el aporte principal del artículo.  Cuando
+  esté listo el resto hay que volver aquí.  Creo que va a ser algo
+  como mostrar la relevancia de la etapa de preprocesamiento, o
   mostrar que a pesar de ser sistemas complejos, con variables
   caóticas, lo mejor en este caso son predictores lineales, o algo por
-  el estilo.}
+  el estilo.} %
+% 
+ is the selection of the best machine learning technique
+to forecast the black Sigatoka development rate according to the
+metric indicated below.
 
@@
 article and presents lines for future works.
-
-\section{Machine learning techniques and Related works}
-\label{sec:techs}
-
-In this section we summarize the machine learning techniques used and we present the main related works.
-
-\subsection{Support Vector Regression (SVR)}
-
-\TODO{Estoy seguro que los lectores van a solicitar reducir esto y
-  dejar solo la referencia.  No hay nada nuevo en el resto del paper
-  que requiera toda esta información, así que hay que resumirlo a nada
-  más lo esencial. }
-
-From the perspective of Support Vector Regression (SVR) the regression
-function $y = f(s)$ for a given dataset $D=\{(s_i,y_i)\}_{i=1}^n$ is
-represented as a linear function of the form \citep{Wei2013}:
-\begin{equation*}
-  f(s)=w^Ts+b
-\end{equation*}
-where $w$ and $b$ are respectively the weight vector and the intercept
-of the model, and they are selected to find an optimal fit to the data
-available in $D$ in the sense of ...?
-\TODO{optimal respecto a qué?.  Hay muchísimas posibilidades de optimalidad!}
-
-For nonlinear cases, the $p$-dimensional input vectors are mapped by a
-nonlinear function $\phi : R^p\rightarrow F$ onto a feature space
-$F$.  After the nonlinear mapping, the regression function evolves to a
-pervasive form:
-\begin{equation*}
-f(s)=w^T \phi (s)+b
-\end{equation*}
-
-In SVR the generalization ability is measured by the Euclidean norm of
-$w$, and the empirical error is measured by the $\epsilon$-insensitive
-loss function, given by \citep{Cristiani2005}
-
-\TODO{si $l$ es una función, es función de qué, de $s$? $l(s)$? o de
-  $y$? o de todos $l(y,s;\epsilon)$?}
-
-\begin{equation*}
-l= {\left| y - f(s) \right| }_{\epsilon } = 
-\begin{cases}
-   0 & {\left| y - f(s) \right| } \leq \epsilon \\
-   \left| y - f(s) \right| - \epsilon & \text{otherwise}
-\end{cases}
-\end{equation*}
-which ignores the error if the difference between the prediction value
-and the actual value is smaller than $\epsilon$.
-%
-The $\epsilon$-insensitive loss function allows to find the
-coefficients $w$ and $b$ by solving a convex optimization problem,
-which balances the empirical error and the generalization ability.
-%
-Then, the optimization problem to identify the regression model can be
-formulated as \citep{Wei2013}:
-%
-\begin{equation} 
-\begin{aligned}
-& \underset{}{\text{minimize}}
-& & J(w,\xi_i , \xi_i^* ) = \frac{1}{2}   \Bigr| \Bigr| w \Bigr| \Bigr|^2 + C \sum_{i=1}^{n} (\xi_i , \xi_i^* )    \\
-& \text{subject to}
-& & \begin{array}{lcl} 
-y_i - w^T \phi (s) - b  \leq \epsilon + \xi_i   \\
-w^T \phi (s) + b - y_i \leq \epsilon + \xi_i^* & i= 1,2,...,n \\
-\xi_i , \xi_i^* \geq 0  \\
-\end{array}
-\end{aligned}
-\end{equation}  
-where $C$ denotes the penalty parameter between empirical and
-generalization errors, and $\xi_i , \xi_i^*$ are slack
-variables. \figref{figura2} shows this situation.
-
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{SVR}
- \caption{$\epsilon$-insensitive loss function \citep{Wei2013}.} 
- \label{figura2} 
-\end{figure}
-
-The solution of this optimization problem by the Lagrange method is given by:
-\begin{equation*}
-f(s) = w^T \phi (s) + b = \sum_{i=1}^{n} (\alpha_i - \alpha_i^*) K (s,s_i) + b
-\end{equation*}
-where $\alpha_i - \alpha_i^*$ are the Lagrange multipliers of the
-optimization problem’s dual form and $K(s_i,s_j )$ is the kernel
-function satisfying the Mercer condition, and holds:
-\begin{equation*}
-K(s_i,s_j ) = \big \langle  \phi(s_i) , \phi(s_j)  \big \rangle
-\end{equation*}
-Operations in the kernel function $K(s,s_i )$ are performed in the
-input space rather than in the potentially high dimensional feature
-space of $\phi$ \citep{Alonso2013}.
-
-\subsection{Ordinary least squares regression}
-
-This method fits a linear model with coefficients $w = (w_1,..,w_p)$
-to minimize the residual sum of squares between the observed responses
-in the dataset, and the responses predicted by the linear
-approximation. Mathematically it solves a problem of the form
-\citep{scikitlearn2011}:
-$$\min_{w} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  $$
-where $X$ denotes the features matrix.
-
-According Pedregosa et al.
-\citep{scikitlearn2011}  the coefficient estimates for Ordinary Least Squares rely on the independence of the model terms.
-When terms are correlated and the columns of the design matrix $X$ have an approximate linear dependence, the design matrix becomes close to singular and as a result, the least-squares estimate becomes highly sensitive to random errors in the observed response, producing a large variance.
-This situation of multicollinearity can arise, for example, when data are collected without an experimental design.
-
-\subsection{Ridge regression}
-The ridge regression addresses some of the problems of ordinary least squares regression by imposing a penalty on the size of the coefficients.
-The ridge coefficients minimize a penalized residual sum of squares  \citep{scikitlearn2011}:
-
-$$\min_{w} { \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-Here, $\alpha \ \textgreater \  0$ is a complexity parameter that controls the amount of shrinkage: the larger the value of $\alpha$, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
-
-\subsection{Elastic-Net regression}
-Elastic-Net is a linear regression model trained with $L1$ and $L2$ prior as regularizer.
-This combination allows for learning a sparse model where few of the weights are non-zero like Lasso, while still maintaining the regularization properties of Ridge  \citep{scikitlearn2011}.
-The convex combination of $L1$ and $L2$ is controled by using the $l1_{ratio}$ parameter.
-
-Elastic-Net is useful when there are multiple features which are correlated with one another. Lasso is likely to pick one of these at random, while elastic-net is likely to pick both.
-A practical advantage of trading-off between Lasso and Ridge is it allows Elastic-Net to inherit some of Ridge’s stability under rotation.
-The objective function to minimize is \citep{scikitlearn2011}:
-
-$$\min_{w} { \frac{1}{2n_{samples}} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \rho \Bigr| \Bigr| w \Bigr| \Bigr|_1 + \frac{\alpha (1- \rho)}{2} \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-\subsection{Echo State Networks (ESN)}
-Recurrent Neural Networks (RNN) are useful for temporal patterns, but when they are trained with backpropagation methods, they are very slow.
-Echo State Network (ESN) is an alternative training method to solve that problem.
-ESN is based on the observation that if a random RNN possesses certain algebraic properties, training only a linear readout from it is often sufficient to achieve excellent performance in practical applications \citep{Lukose2009}.
-For a given training input signal $u(n)  \in R^{N_u}$ a desired target output signal $y^{target}(n) \in R^{N_y}$
-is known.
-Here $n = 1, .
-.
-.
-,T$ is the discrete time and $T$ is the number of data points in the training dataset.
-The task is to learn a model with output $y(n) \in R^{N_y}$, where $y(n)$ matches $y^target(n)$ as well as possible, minimizing an error measure $E(y,y^target)$, and, more importantly, generalizes well to unseen data.
-The untrained RNN part of an ESN is called a dynamical reservoir, and the resulting states x(n) are termed echoes of its input history \citep{Lukose2012}.
-Finally, these signals are sent to an output layer as shown in the \figref{figura3}.
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{Reservorio}
- \caption{An echo state network \citep{Lukose2012}.} 
- \label{figura3} 
-\end{figure}
-
-The connections between the different elements of an Echo State Network have weights randomly generated.
-The weights of the internal connections of the reservoir $(W)$ as well as the weights of the input layer $(W_in)$, after being generated are set statically during all stages of implementation of the algorithm.
-The weights between the reservoir and the output layer $(W_out)$ are subject to changes of a supervised learning algorithm to correct the degree of error generated by the entire system \citep{Lukose2012}.
-
-\subsection{Related works}
-Now, we present the main and recently works related with our study.
-
-Huang et al.
-\citep{Huang2010}  surveyed the development of soft computing techniques in agricultural and biological engineering, including fuzzy logic, artificial neural networks, genetic algorithms, bayesian inference and decision trees.
-
-A related work, proposed by Romero \citep{Romero1995} relies on regression models using a stepwise procedure to predict incubation and latency times of black Sigatoka.
-The author performed experiments on two farms located in Costa Rica (La Rita and Waldeck, the same as those used in this study but with different names).
-The study used data from: December 1993 to August 1995.
-Romero concluded that the model to predict the incubation period accounted a $R^2$ of 69\% in his observed data but it was not a good predictor when it was validated against an independent dataset (cross validation).
-For latency, he developed two models that accounted a $R^2$ of 78\% in the observed data, however, when validated against an independent dataset (cross validation), for La Rita obtained an adjusted $R^2$ of 82\%, and for Weldeck, none of the models satisfactorily predicted the latent period; and then those predictions were not shown \citep{Romero1995}.
-
-Glezakos et al.
-\citep{Glezakos2010} proposed to use Genetic Algorithms (GA) and Neural Networks (NN) to identify plant virus (Tobacco Rattle Virus (TRV) and the Cucumber Green Mottle Mosaic Virus (CGMMV)).
-This is achieved by the development of ana- lytical tools of evolutionary adaptive width, propelled by Genetic Algorithms (GAs) and Neural Networks (NNs).
-The method was tested against some of the most commonly used classifiers in machine learning (Bayes, Trees and k-NN) via cross-validation and proved its potential towards the identification.
-
-In the agricultural context, Alves et al.
-\citep{Alves2011} used  geoinformation techniques to develop predictive models to study the areas of risk to soybean rust in soybean, coffee leaf rust in coffee, and black Sigatoka in banana, considering Brazil’s climatic characterization and the distribution of soybean, coffee and banana crops.
-Temperature and rainfall data were obtained for the period from 1950 to 2000, and of simulations for 2020, 2050 and 2080 using the SRES A2 climate change scenarios.
-Using principal components analysis, a single variable was generated based on 57 variables, in order to determine an index explaining 87\%, 88\% and 90\% of the variability of soybean, coffee and banana crops, respectively, in municipal districts across Brazil.
-The climatic model was used to generate the zoning of the three plant diseases, using temperature and leaf wetness as input.
-Areas of favorability for the diseases were plotted against the main coffee, soybean and banana growing areas in Brazil.
-This methodology enabled the visualization of the changes in areas favorable for epidemics under possible future scenarios of climate change.
-
-Other applications of machine learning methods in precision agriculture include the use of support vector regression to predict carcass weight in beef cattle in advance to the slaughter  \citep{Alonso2013}, machine learning assessments of soil drying for agricultural planning \citep{Coopersmith2014}, and early detection and classification of plant diseases with support vector machines based on hyperspectral reflectance \citep{Rumpf2010}.
-
-Furthermore, there have been attempts to generate software tools.
-Camargo et al.
-\citep{Camargo2012} presented an information system for the assessment of plant disorders (Isacrodi).
-They proposed that experts will attain a much better accuracy than the Isacrodi classifier, particularly when provided with samples from the affected crop.
-However, those cases where such expertise is not available, they suggest that Isacrodi can provide valuable support to farmers.
-Isacordi includes 15 crop disorders, but the black Sigatoka no is one of them.
-The prediction process is based on multi-class support vector machines.
-
-Regarding the prediction of the develpment of the black Sigatoka with machine learning methods, Bendini et al.
-\citep{Bendini2013}  presented a study about the risk analysis of black Sigatoka occurrence based on polynomial models.
-A case study was developed in a commercial banana plantation located in Jacupiranga, Brazil.
-It was monitored weekly during the period from February to December 2005.
-Data included the weekly monitoring of the disease’s evolution stage, time series of meteorological data and remote sensing data.
-They obtained a model to estimate the evolution of the disease from satellite imagery.
-This model relates gray levels (NC) of the band 2 images of the Landsat-5 satellite, with the progress status or disease severity (EE).
-The authors claim to reach an $R^2$ of 90\%.
-
-Also there are works related to banana fruit.
-Soares et al.
-\citep{Soares2014} apply two techniques: artificial neural networks (ANNs) and multiple linear regression (MLR) in banana plant to predict the yield, their results show that the neural network proved to be more accurate in forecasting the weight of the bunch in comparison to the multiple linear regressions in terms of the mean prediction-error $(MPE = 1.40)$, mean square deviation $(MSD = 2.29)$ and coefficient of determination $(R^2 = 91\%)$.
-
-In general, the machine learning methods applied to predict the evolution of plant diseases, can be classified in two main approaches: 1) Those whose main inputs are images, and 2) Those whose main inputs are environmental and biological variables.
-Our study focuses in the second case.
-
-\section{Specification of data and methodology}
-\label{sec:data}
-
-In this section, we present the data and how we worked to obtain our results.
-
-\subsection{Data}
-We use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (previously called Waldeck and located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
-The banana type is Musa AAA, subgroup Cavendish, cv.
-Grande Naine.
-The \tablename $.$\ref{tabla1} shows the variables available.
-
-\begin{table}[h] 
-\caption{Variables used in the study} 
-\label{tabla1} 
-\centering
-\begin{tabular}{c|c} 
-\hline
-\bfseries Variable & \bfseries Meaning0'0' \\ 
-\hline\hline 
-$T_{a_{max}}$ &	Max air temperature \\
-$T_{a_{min}}$ &	Min air temperature \\
-$\overline{T}_{a}$	 & Mean air temperature \\
-$\overline{H}$	&  Mean Humidity \\
-$H_{min}$ &	Min humidity \\
-$H_{max}$	& Max humidity \\
-$\overline{R}$	& Mean Solar radiation \\
-$P$	& Sum precipitation  \\
-$W_{max}$	& Max speed wind \\
-$\overline{W}$	& Mean speed wind  \\
-$L_2$	& Biological warning system – Leaf 2 \\
-$L_3$	& Biological warning system – Leaf 3 \\
-$L_4$	& Biological warning system – Leaf 4 \\
-$E_s$	& Biological warning system – Evolution Stage \\
-\hline
-\end{tabular} 
-\end{table}
-
-The value to be predicted in all cases was $E_s$, that is the total measure of the biological warning system.
-
-The data on the biological warning system are collected once a week.
-Although Corbana has meteorological stations that take data every five minutes, for these experiments, weekly averages generated by nearby stations to each of the farms were used.
-
-The time intervals used for this study were: La Rita, week 48 of 2002 to week 17 of the 2015 (647 weeks) and for 28 Miles, week 37 of 2003 to week 18 of 2015 (605 weeks).
-
-\subsection{Data preprocessing}
-Usually, data need preprocessing to prepare for use. In this section we explain that process.
-
-In 28 Miles farm, 1\% of the data were missing, while in La Rita was 2.25\%. To fill-in the missing values we use spline interpolation. The data collected did not exhibit outliers.
-
-Due the fact that the variables measure meteorological or biological process, they are discretized in order to reflect trends in the data, i.e.
-the continuous values are not directly used.
-The coefficient of variation $C_v(x)$ of each variable x was used to determine the number $n$ of discretization levels.
-$$n= \lfloor 100 \ C_v(x) \rfloor$$
-where $\lfloor \ \rfloor$  is the round operator.
-
-Each discretization range was uniformly partitioned.
-Besides enabling the capture of tendencies, the discretization removes the effect of small variations in the data collection, either by inaccuracies of the instruments (meteorological variables) or by subjective bias introduced by the human who collects the data (biological warning system).
-
-Each feature was scaled to fit in a range between 0 and 1. The variable to be predicted was not scaled.
-
-\subsection{Evaluation criteria}
-
-Although there are many types of indicators to assess the quality of the prediction, we selected the determination coefficient $(R^2)$ and the Root Mean Square Error $(RMSE)$.
-
-Given $n$ records, let be $y$ the actual value of the series, $\hat{y}$ the predicted value and $\acute{y}$ the mean of the observed data.
-
-$$ \bar{y} = \frac{1}{n} \sum_{i=1}^{n} y_i $$
-
-$$ S_e^2 = \frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}$$
-
-$$ S_R^2 = \frac{\sum_{i=1}^{n} {(\hat{y}_i-\bar{y}_i)}^2 }{n}$$
-
-$$ S_y^2 = S_R^2 + S_e^2$$
-
-$$ R^2 = \frac{S_R^2}{S_y^2}$$
-
-$$ RMSE = \sqrt{\frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}}$$
-
-This decision is supported by the widespread use in machine learning and agriculture areas \citep{Soares2014}, \citep{Soares2013}, \citep{Ibrahim2014} and \citep{Demir2014}.
-
-\subsection{Programming environment}
-
-We use the python programming language with the Integrated Development Environment (IDE) Spyder \citep{Continuum2015}, particularly with the libraries pandas \citep{mckinneypandas2010} and numpy \citep{vanderWalt2011}.
-For SVR, ridge and ordinary least squares regressions, we used sklearn \citep{scikitlearn2011} and for ESN the python-based code of Dr.
-$Luko\breve{s} evi \breve{c} ius$ \citep{Lukose2012} on which the necessary were done for adjustments for the experiments of this work.
-The computer used a processor Intel(R) Core i7-4800MQ CPU 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
-
-\subsection{Methodology}
-The selection of the technique and its parametrisation was performed in two stages that they are presented below.
-
-{\bf Phase one }
-
-In the phase one, we did ten-fold-cross-validation on a set of  machine learning methods and different configurations:
-
-\begin{itemize}
-\item 	Patterns: n by m, where n from 1 to 8 and m from 1 to 2.
-
-\item Methods: support vector regression with the kernels functions: linear, gaussian and sigmoid; echo state networks; ordinary least squares linear regression, ridge regression and elastic-net regression.
-
-\item Variables included in the model:
-\begin{itemize}
-\item All variables.
-\item From the set $\{ \overline{T}_{a} , \overline{H}, P , \overline{W}  \}$ use the subsets with one, two or four elements.
-These variables are according to experts the ones having most impact on the disease development \citep{MarinVargas1995}.
-\end{itemize}
-
-\end{itemize}
-
-{\bf Phase two }
-
-In the second phase, the best configurations obtained in phase one are used to validate with the last 50 and 100 weeks.
-
-\section{Results and discussion}
-\label{sec:results}
-
-In this section we present the main results for phase and their discussion.
-
-{\bf Phase one }
-
-\figref{figura4} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Though La Rita obtains different results in magnitude than 28 Millas, the trend is similar.
-In both farms, the best results are for linear models, second position is occupied for Echo State Networks and  SVR with gausian and sigmoid kernels are the worst results.
-In linear models, to predict one week ahead is better than two weeks ahead, and this is better than three weeks ahead.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_algortihm}
- \caption{Phase one - Best $R^2$ for each algorithm} 
- \label{figura4} 
-\end{figure}
-
-\figref{figura5} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-In general, to predict one week ahead is better than two weeks ahead and so on.
-The number of weeks consider in the observation range in the pattern is not the main discriminant factor, but it is clear that we get better $R^2$ for one week ahead than two weeks ahead and so on.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase one - Best $R^2$ for each observation range} 
- \label{figura5} 
-\end{figure}
-
-\figref{figura6} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-The better results are obtained with $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms of similarly.
-You can note that the use of all variables in the model or the inclusion of the four variables suggest for expert criteria do not improve significantly the results, then the use of more sensors do not assure a better result.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
- \caption{Phase one - Best $R^2$ for each variable combination} 
- \label{figura6} 
-\end{figure}
-
-\figref{figura7} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
-The Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better $RMSE$ than La Rita.
-This situation arise because $RMSE$ considers errors only with respect the prediction and in 28 Millas the average of Stage of Evolution is $4316.16$, unlike, in La Rita the average is $5507.30$.
-So, in La Rita we obtains higher errors in absolute values.
-$R2$ is a relative metric between 0 thru 1 and it is less sensitive to absolute values.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.8]{Phase_one_R2_RMSE}
- \caption{Phase one - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura7} 
-\end{figure}
-
-The Pareto frontier for the La Rita farm is composed by 96 elements. The \tablename $.$\ref{tabla2} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase one} 
-\label{tabla2} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $\overline{W}$ &	1 to 1  & 36 & 64.25\% & 714.51 \\
- &	2 to 1  & 6 & 62.97\% & 695.10 \\
-\hline 
-All  & 1 to 1  & 18 & 62.98\% & 701.95 \\
-   & 2 to 1  & 12 & 61.76\% & 679.92 \\
-    & 3 to 1  & 6 & 60.60\% & 676.42 \\
-    & 5 to 1  &  2 & 60.37\% & 672.39 \\
-\hline    
-$\overline{T}_{a}$ & 1 to 1  & 12  & 63.60\% & 708.77 \\
-       &	2 to 1  & 4 & 62.23\% & 689.55 \\
-\hline
-\end{tabular} 
-\end{table}
-
-Similarly, the Pareto frontier for the 28 Millas farm is composed by 75 elements. The \tablename $.$\ref{tabla3} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - 28 Millas - Phase one} 
-\label{tabla3} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $\overline{W}$ & 1 to 1 & 8 & 57.80\% & 438.09 \\
-\hline 
-All   &	9 to 1 & 2 & 50.93\% & 397.93 \\
-  & 10 to 1	 & 2 & 50.97\% & 398.81 \\
-  &	8 to 1 & 6 & 51.62\% & 398.93 \\
-  &	7 to 1 & 2 & 52.25\% & 400.28 \\
-  &	6 to 1 & 2 & 53.16\% & 404.14 \\
-  &	4 to 1 & 2 & 54.32\% & 407.54 \\
-\hline    
-$\overline{T}_{a}$ & 1 to 1  & 8  & 59.09\% & 439.44 \\
-\hline
-Pair $\overline{T}_{a}$ $\overline{H}$ & 1 to 1	 & 8 & 57.51\% & 428.61 \\
- &	2 to 1 & 20 & 56.91\% & 414.37 \\
- &	3 to 1 & 3 & 54.41\% & 411.55 \\
- &	4 to 1 & 3 & 53.34\% & 406.65 \\
-\hline
-Pair $\overline{T}_{a}$ $P$ & 3 to 1 & 9 & 56.23\% & 422.76 \\
-\hline
-\end{tabular} 
-\end{table}
-
-We can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week.
-
-{\bf Phase two }
-
-In the second phase, the best configurations obtained in phase one were used to validate with the last 50 and 100 weeks.
-
-\figref{figura8} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Even if linear models continue with good $R^2$, Echo State Networks improve their scores because in $50$ and $100$ last weeks validation, we are in presence of climate change, then the behaviour is less lineal.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_algortihm}
- \caption{Phase two - Best $R^2$ for each algorithm} 
- \label{figura8} 
-\end{figure}
-
-\figref{figura9} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-The results confirms that, in general, to predict one week ahead is better than two weeks ahead, two than three and so on, this for both farms.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase two - Best $R^2$ for each observation range} 
- \label{figura9} 
-\end{figure}
-
-\figref{figura10} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-This results confirm that $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms are the best variables combinations.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
- \caption{Phase two - Best $R^2$ for each variable combination} 
- \label{figura10} 
-\end{figure}
-
-\figref{figura11} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.  You can note that the behaviour of $R^2$ and $RMSE$ is similar to the phase one.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.8]{Phase_two_R2_RMSE}
- \caption{Phase two - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura11} 
-\end{figure}
-
-In phase two, the Pareto frontier for the La Rita farm is composed by 2 elements. The \tablename $.$\ref{tabla4} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase two} 
-\label{tabla4} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-All & 1 to 1  & 2 & 69.33\% & 353.33 \\
-\hline
-\end{tabular} 
-\end{table}
-
-In 28 Millas Farm, the Pareto frontier is composed by 1 element. The \tablename $.$\ref{tabla5} details the result.
-
-\begin{table}[h] 
-\caption{Pareto frontier - 28 Millas - Phase two} 
-\label{tabla5} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $P$ & 1 to 1 & 1 & 61.36\% & 339.89 \\
-\hline
-\end{tabular} 
-\end{table}
-
-Again, similar to phase one, we can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week, one week to predict one week ahead with combinations of variables listed above.
-
-\section{Conclusions}
-\label{sec:concl}
-
-\subsection{Future work}
-
-\section{Acknowledgements}
-
-The authors would like to thank Corbana S.A. for providing the data for this research.
-
-\section{References}
-
-\bibliography{references}
-
-\end{document}

### 2016-06-11 methodology.tex
Nueva organizacion de los archivos.  Partes del paper en archivos separados y todas las imagenes en subdirectorio fig

--- 2016-05-22 Paper Sigatoka - 2015.tex
@@
-\documentclass[review,authoryear,english]{elsarticle}
-
-\usepackage{babel}
-\usepackage{natbib}
-
-\bibliographystyle{elsarticle-harv}
-
-\usepackage{lineno}
-\usepackage[utf8]{inputenc}  %% Permite utilizar tildes si archivo usa utf-8
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\modulolinenumbers[1]
-
-\usepackage{soul}
-\usepackage{marginnote}
-\usepackage{float}
-
-\usepackage{color}
-\definecolor{dkred}{rgb}{0.5,0,0}       %   dark red
-\definecolor{todocolor}{rgb}{1.0,0.8,0.5}
-
-\usepackage[hidelinks]{hyperref}
-
-\newcommand{\todohdr}{Please review:}
-\newcommand{\TODO}[2][\todohdr]{%
-\smallbreak\noindent\colorbox{todocolor}{%
-\begin{minipage}{1\linewidth}
-\textbf{#1}\par{\footnotesize{#2}}
-\end{minipage}}}
-
-\newcommand{\instead}[2]{\textcolor{dkred}{\st{#1}}#2}
-
-%  \usepackage{epstopdf} 
-%  \usepackage{epsfig}
-
-\journal{Journal of \LaTeX\ Templates}
-
-\usepackage{subcaption}
-
-\begin{document}
-
-\begin{frontmatter}
-
-\title{Forecasting the black Sigatoka development rate:\\ 
-A comparison of machine learning techniques}
-
-%% Group authors per affiliation:
-\author[afiLuisAlex]{Luis-Alexander Calvo-Valverde\fnref{myfootnote}}
-\ead{lualcava.sa@gmail.com}
-\fntext[myfootnote]{Corresponding author. (506)70104420}
-
-\author[afiCorbana]{Mauricio Guzmán-Quesada}
-\author[afiCorbana]{José-Antonio~Guzmán-Alvarez}
-\author[afiPablo]{Pablo Alvarado-Moya}
-
-\address[afiLuisAlex]{DOCINADE, Instituto Tecnológico de Costa Rica, 
-Computer Research Center, Multidisciplinar program eScience, Cartago, Costa Rica}
-
-\address[afiCorbana]{Dirección de Investigaciones, Corporación Bananera Nacional S.A., Guápiles, Costa Rica}
-
-\address[afiPablo]{DOCINADE, Instituto Tecnológico de Costa Rica, Cartago, Costa Rica}
-
-\begin{abstract}
-Pending.
-\end{abstract}
-
-\begin{keyword}
-Machine learning \sep Black Sigatoka \sep Support vector regression \sep
-Banana disease prediction \sep Biological warning system 
-\end{keyword}
-
-\end{frontmatter}
-
-\linenumbers
-
-\section{Introduction}
-
-The black Sigatoka disease caused by the fungus \emph{Mycosphaerella
-  fijiensis Morelet} is the major pathological problem of banana and
-plantain crops in Central America, Panama, Colombia and Ecuador, as well as in
-many parts of Africa and Asia \citep{MarinVargas1995}.
-
-This disease attacks the plant leaves producing a rapid deterioration
-of the leaf area, affects the growth and productivity of the plants
-due to the impairment of the photosynthetic process, causes a
-reduction in the quality of the fruit, and promotes premature
-ripening of bunches, which is
-the major cause of product losses associated with the black
-Sigatoka. \figurename~\ref{fig:diseasestages} shows three stages of
-this disease.
-
-\begin{figure}[ht] 
-\centering
-\begin{tabular}{c@{\;}c@{\;}c}
-  \includegraphics[width=.32\linewidth]{Roya_a} &
-  \includegraphics[width=.32\linewidth]{Roya_b} &
-  \includegraphics[width=.32\linewidth]{Roya_c} \\
-  (a) & (b) & (c) 
-\end{tabular}
-\caption{Examples of three disease stages of the black Sigatoka. 
-(a)~Initial stage. (b)~Intermediate stage, and (c)~Advanced stage.} 
-\label{fig:diseasestages} 
-\end{figure}
-
-Phytopathological studies point out that precipitation, temperature,
-relative humidity and wind are the main climatic variables that affect
-the development of this disease \citep{MarinVargas1995}.
-
-The climate has a major effect on the development of the black
-Sigatoka.  Precipitation, temperature, relative humidity and wind are
-the main climatic variables affecting the development of this disease
-\citep{MarinVargas1995}.
-
-The early warning system developed by Ganry and Meyer
-\citep{ganry1972} and modified by Ganry and Laville \citep{ganry1983}
-to control yellow Sigatoka in Cameroon, was later adapted by
-Ternesien \citep{Ternesien1985} and Fouré \citep{foure1988} for the
-black Sigatoka.
-
-This bilogical warning system is based on weekly quantified
-observations of the disease progression on young leaves of the plant,
-according to Fouré's scale of the symptom stages.
-%
-The numeric coefficients describe the degree of incidence and the
-severity of the disease development.  These coefficients are used to
-calculate two variables: gross sum and state of evolution.
-
-The gross sum is based on the present disease progression stage and
-the numeric coefficients, which increases with the progression of
-the symptoms and the juvenility of the leaf.
-%
-The state of evolution is calculated using the gross sum and the
-foliar emission period.
-%
-Although threshold levels were initially suggested as a guide to spray
-schedules, the fluctuation of these two variables
-was found to better define appropiate times to spray
-\citep{Marinetal2003}.
-
-In Costa Rica the black Sigatoka is frequently treated with
-chemical fungicides.
-%
-Depending on the zone of production and the weather conditions,
-45--55~cycles/year of fungicide applications are required to keep this
-disease under control and to produce the expected fruit quality for the
-international markets.
-%
-This represents a cost per hectare per year in the range between US\$1600 and
-US\$2000; about 0.64--0.80 cents of the production costs for a
-18.14\,kg box.
-%
-Overall, this represents 10\%--12\% of the total production costs.
-
-The past and present rates of disease development can in principle be used
-to predict its future behavior and to determine whether
-a particular fungicide spray program will be able to effectively and
-economically control the disease \citep{ChuangJeger1987} or not.
-
-There are efforts to apply machine learning methods to support
-decision-making in agriculture, including the control of crop
-diseases. For example, \cite{Camargo2012} present an intelligent
-system for the assessment of crop disorders, \cite{Huang2010}
-introduce a plant virus identification method based on neural networks
-with an evolutionary preprocessing stage, \cite{Kim2014} summarize in
-their survey crop pests prediction methods using regression and
-machine learning approaches, while \cite{Zhao2013} present an
-intelligent agricultural forecasting system based on wireless sensor
-networks.
-
-In this work, we compare five machine learning techniques 
-to predict the development rate of the black Sigatoka disease:
-support
-vector regression (SVR), echo state networks (ESN), ridge regression,
-elastic-net regression and ordinary least squares linear regression.
-%
-The main contribution of this work is the selection of the best
-machine learning technique to forecast the black Sigatoka development
-rate according to the metric indicated below.
-
-The outline of the paper is as follows: in Section~\ref{sec:techs}, we
-introduce the machine learning techniques used in this research and
-the main related works. Section~\ref{sec:data} summarizes the data the
-methodology used in this study.  The results and their discussion are
-presented in section~\ref{sec:results}.  We present the main
-conclusions and lines for future works in Section~\ref{sec:concl}.
-
-\section{Machine learning techniques and Related works}
-\label{sec:techs}
-
-In this section we summary the machine learning techniques used and we present the main related works.
-
-\subsection{Support Vector Regression (SVR)}
-
-From the perspective of Support Vector Regression (SVR) the regression function $y = f(s)$ for a given dataset $D=\{(s_i,y_i)\}_{i=1}^n$ , is represented as a linear function of the form \citep{Wei2013}:
-$$f(s)=w^Ts+b$$
-where $w$ and $b$ are respectively the weight vector and the intercept of the model, and they are selected to find an optimal fit to the data available in $D$.
-
-For nonlinear cases, one proceeds by mapping the input p-dimensional vectors via a nonlinear function $\phi : R^p\rightarrow F$, onto the feature space $F$.
-After nonlinear mapping, the regression function evolves to a pervasive form:
-
-$$f(s)=w^T \phi (s)+b$$
-
-SVR uses the $\epsilon$-insensitive loss function:
-\begin{equation*}
-l= {\left| y - f(s) \right| }_{\epsilon } = 
-\begin{cases}
-   0 & {\left| y - f(s) \right| } \leq \epsilon \\
-   \left| y - f(s) \right| - \epsilon & \text{otherwise}
-\end{cases}
-\end{equation*}
-which ignores the error if the difference between the prediction value and the actual value is smaller than $\epsilon$.
-%
-The $\epsilon$-insensitive loss function allows to find the coefficients $w$ and $b$ by solving a convex optimization problem, which balances the empirical error and the generalization ability.
-%
-In SVR, the empirical error is measured by the loss function $\epsilon$-insensitive and the generalization ability is measured by the Euclidean norm of $w$ \citep{Cristiani2005}.
-%
-Then, the optimization problem to identify the regression model can be
-formulated by \citep{Wei2013}:
-
-\begin{equation} 
-\begin{aligned}
-& \underset{}{\text{minimize}}
-& & J(w,\xi_i , \xi_i^* ) = \frac{1}{2}   \Bigr| \Bigr| w \Bigr| \Bigr|^2 + C \sum_{i=1}^{n} (\xi_i , \xi_i^* )    \\
-& \text{subject to}
-& & \begin{array}{lcl} 
-y_i - w^T \phi (s) - b  \leq \epsilon + \xi_i   \\
-w^T \phi (s) + b - y_i \leq \epsilon + \xi_i^* & i= 1,2,...,n \\
-\xi_i , \xi_i^* \geq 0  \\
-\end{array}
-\end{aligned}
-\end{equation}  
-where $C$ denotes the penalty parameter between empirical and generalization errors, and  $\xi_i , \xi_i^*$ are slack variables. \figurename
-$.$\ref{figura2} shows this situation.
-
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{SVR}
- \caption{$\epsilon$-insensitive loss function \citep{Wei2013}.} 
- \label{figura2} 
-\end{figure}
-
-The solution of this optimization problem by the Lagrange method is given by:\\
-$$ f(s) = w^T \phi (s) + b = \sum_{i=1}^{n} (\alpha_i - \alpha_i^*) K (s,s_i) + b $$ 
-where $\alpha_i - \alpha_i^*$  are the Lagrange multipliers of the optimization problem’s dual form and $K(s_i,s_j )$ is the kernel function satisfying the Mercer condition, and holds:
-$$K(s_i,s_j ) = \big \langle  \phi(s_i) , \phi(s_j)  \big \rangle $$
-Operations in the kernel function $K(s,s_i )$ are performed in the input space rather than in the potentially high dimensional feature space of $\phi$ \citep{Alonso2013}.
-
-\subsection{Ordinary least squares regression}
-This method fits a linear model with coefficients $w = (w1,..,wp)$ to minimize the residual sum of squares between the observed responses in the dataset, and the responses predicted by the linear approximation.
-Mathematically it solves a problem of the form \citep{scikitlearn2011}:
-$$\min_{w} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  $$
-where $X$ denotes the features matriz. \\
-According Pedregosa et al.
-\citep{scikitlearn2011}  the coefficient estimates for Ordinary Least Squares rely on the independence of the model terms.
-When terms are correlated and the columns of the design matrix $X$ have an approximate linear dependence, the design matrix becomes close to singular and as a result, the least-squares estimate becomes highly sensitive to random errors in the observed response, producing a large variance.
-This situation of multicollinearity can arise, for example, when data are collected without an experimental design
-
-\subsection{Ridge regression}
-The ridge regression addresses some of the problems of ordinary least squares regression by imposing a penalty on the size of the coefficients.
-The ridge coefficients minimize a penalized residual sum of squares  \citep{scikitlearn2011}:
-
-$$\min_{w} { \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-Here, $\alpha \ \textgreater \  0$ is a complexity parameter that controls the amount of shrinkage: the larger the value of $\alpha$, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
-
-\subsection{Elastic-Net regression}
-Elastic-Net is a linear regression model trained with $L1$ and $L2$ prior as regularizer.
-This combination allows for learning a sparse model where few of the weights are non-zero like Lasso, while still maintaining the regularization properties of Ridge  \citep{scikitlearn2011}.
-The convex combination of $L1$ and $L2$ is controled by using the $l1_{ratio}$ parameter.
-
-Elastic-Net is useful when there are multiple features which are correlated with one another. Lasso is likely to pick one of these at random, while elastic-net is likely to pick both.
-A practical advantage of trading-off between Lasso and Ridge is it allows Elastic-Net to inherit some of Ridge’s stability under rotation.
-The objective function to minimize is \citep{scikitlearn2011}:
-
-$$\min_{w} { \frac{1}{2n_{samples}} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \rho \Bigr| \Bigr| w \Bigr| \Bigr|_1 + \frac{\alpha (1- \rho)}{2} \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-\subsection{Echo State Networks (ESN)}
-Recurrent Neural Networks (RNN) are useful for temporal patterns, but when they are trained with backpropagation methods, they are very slow.
-Echo State Network (ESN) is an alternative training method to solve that problem.
-ESN is based on the observation that if a random RNN possesses certain algebraic properties, training only a linear readout from it is often sufficient to achieve excellent performance in practical applications \citep{Lukose2009}.
-For a given training input signal $u(n)  \in R^{N_u}$ a desired target output signal $y^{target}(n) \in R^{N_y}$
-is known.
-Here $n = 1, .
-.
-.
-,T$ is the discrete time and $T$ is the number of data points in the training dataset.
-The task is to learn a model with output $y(n) \in R^{N_y}$, where $y(n)$ matches $y^target(n)$ as well as possible, minimizing an error measure $E(y,y^target)$, and, more importantly, generalizes well to unseen data.
-The untrained RNN part of an ESN is called a dynamical reservoir, and the resulting states x(n) are termed echoes of its input history \citep{Lukose2012}.
-Finally, these signals are sent to an output layer as shown in the \figurename
-$.$\ref{figura3}.
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{Reservorio}
- \caption{An echo state network \citep{Lukose2012}.} 
- \label{figura3} 
-\end{figure}
-
-The connections between the different elements of an Echo State Network have weights randomly generated.
-The weights of the internal connections of the reservoir $(W)$ as well as the weights of the input layer $(W_in)$, after being generated are set statically during all stages of implementation of the algorithm.
-The weights between the reservoir and the output layer $(W_out)$ are subject to changes of a supervised learning algorithm to correct the degree of error generated by the entire system \citep{Lukose2012}.
-
-\subsection{Related works}
-Now, we present the main and recently works related with our study.
-
-Huang et al.
-\citep{Huang2010}  surveyed the development of soft computing techniques in agricultural and biological engineering, including fuzzy logic, artificial neural networks, genetic algorithms, bayesian inference and decision trees.
-
-A related work, proposed by Romero \citep{Romero1995} relies on regression models using a stepwise procedure to predict incubation and latency times of black Sigatoka.
-The author performed experiments on two farms located in Costa Rica (La Rita and Waldeck, the same as those used in this study but with different names).
-The study used data from: December 1993 to August 1995.
-Romero concluded that the model to predict the incubation period accounted a $R^2$ of 69\% in his observed data but it was not a good predictor when it was validated against an independent dataset (cross validation).
-For latency, he developed two models that accounted a $R^2$ of 78\% in the observed data, however, when validated against an independent dataset (cross validation), for La Rita obtained an adjusted $R^2$ of 82\%, and for Weldeck, none of the models satisfactorily predicted the latent period; and then those predictions were not shown \citep{Romero1995}.
-
-Glezakos et al.
-\citep{Glezakos2010} proposed to use Genetic Algorithms (GA) and Neural Networks (NN) to identify plant virus (Tobacco Rattle Virus (TRV) and the Cucumber Green Mottle Mosaic Virus (CGMMV)).
-This is achieved by the development of ana- lytical tools of evolutionary adaptive width, propelled by Genetic Algorithms (GAs) and Neural Networks (NNs).
-The method was tested against some of the most commonly used classifiers in machine learning (Bayes, Trees and k-NN) via cross-validation and proved its potential towards the identification.
-
-In the agricultural context, Alves et al.
-\citep{Alves2011} used  geoinformation techniques to develop predictive models to study the areas of risk to soybean rust in soybean, coffee leaf rust in coffee, and black Sigatoka in banana, considering Brazil’s climatic characterization and the distribution of soybean, coffee and banana crops.
-Temperature and rainfall data were obtained for the period from 1950 to 2000, and of simulations for 2020, 2050 and 2080 using the SRES A2 climate change scenarios.
-Using principal components analysis, a single variable was generated based on 57 variables, in order to determine an index explaining 87\%, 88\% and 90\% of the variability of soybean, coffee and banana crops, respectively, in municipal districts across Brazil.
-The climatic model was used to generate the zoning of the three plant diseases, using temperature and leaf wetness as input.
-Areas of favorability for the diseases were plotted against the main coffee, soybean and banana growing areas in Brazil.
-This methodology enabled the visualization of the changes in areas favorable for epidemics under possible future scenarios of climate change.
-
-Other applications of machine learning methods in precision agriculture include the use of support vector regression to predict carcass weight in beef cattle in advance to the slaughter  \citep{Alonso2013}, machine learning assessments of soil drying for agricultural planning \citep{Coopersmith2014}, and early detection and classification of plant diseases with support vector machines based on hyperspectral reflectance \citep{Rumpf2010}.
-
-Furthermore, there have been attempts to generate software tools.
-Camargo et al.
-\citep{Camargo2012} presented an information system for the assessment of plant disorders (Isacrodi).
-They proposed that experts will attain a much better accuracy than the Isacrodi classifier, particularly when provided with samples from the affected crop.
-However, those cases where such expertise is not available, they suggest that Isacrodi can provide valuable support to farmers.
-Isacordi includes 15 crop disorders, but the black Sigatoka no is one of them.
-The prediction process is based on multi-class support vector machines.
-
-Regarding the prediction of the develpment of the black Sigatoka with machine learning methods, Bendini et al.
-\citep{Bendini2013}  presented a study about the risk analysis of black Sigatoka occurrence based on polynomial models.
-A case study was developed in a commercial banana plantation located in Jacupiranga, Brazil.
-It was monitored weekly during the period from February to December 2005.
-Data included the weekly monitoring of the disease’s evolution stage, time series of meteorological data and remote sensing data.
-They obtained a model to estimate the evolution of the disease from satellite imagery.
-This model relates gray levels (NC) of the band 2 images of the Landsat-5 satellite, with the progress status or disease severity (EE).
-The authors claim to reach an $R^2$ of 90\%.
-
-Also there are works related to banana fruit.
-Soares et al.
-\citep{Soares2014} apply two techniques: artificial neural networks (ANNs) and multiple linear regression (MLR) in banana plant to predict the yield, their results show that the neural network proved to be more accurate in forecasting the weight of the bunch in comparison to the multiple linear regressions in terms of the mean prediction-error $(MPE = 1.40)$, mean square deviation $(MSD = 2.29)$ and coefficient of determination $(R^2 = 91\%)$.
-
-In general, the machine learning methods applied to predict the evolution of plant diseases, can be classified in two main approaches: 1) Those whose main inputs are images, and 2) Those whose main inputs are environmental and biological variables.
-Our study focuses in the second case.
-
 \section{Specification of data and methodology}
@@
 In the second phase, the best configurations obtained in phase one are used to validate with the last 50 and 100 weeks.
-
-\section{Results and discussion}
-\label{sec:results}
-
-In this section we present the main results for phase and their discussion.
-
-{\bf Phase one }
-
-\figurename $.$\ref{figura4} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Though La Rita obtains different results in magnitude than 28 Millas, the trend is similar.
-In both farms, the best results are for linear models, second position is occupied for Echo State Networks and  SVR with gausian and sigmoid kernels are the worst results.
-In linear models, to predict one week ahead is better than two weeks ahead, and this is better than three weeks ahead.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_algortihm}
- \caption{Phase one - Best $R^2$ for each algorithm} 
- \label{figura4} 
-\end{figure}
-
-\figurename $.$\ref{figura5} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-In general, to predict one week ahead is better than two weeks ahead and so on.
-The number of weeks consider in the observation range in the pattern is not the main discriminant factor, but it is clear that we get better $R^2$ for one week ahead than two weeks ahead and so on.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase one - Best $R^2$ for each observation range} 
- \label{figura5} 
-\end{figure}
-
-\figurename $.$\ref{figura6} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-The better results are obtained with $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms of similarly.
-You can note that the use of all variables in the model or the inclusion of the four variables suggest for expert criteria do not improve significantly the results, then the use of more sensors do not assure a better result.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
- \caption{Phase one - Best $R^2$ for each variable combination} 
- \label{figura6} 
-\end{figure}
-
-\figurename $.$\ref{figura7} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
-The Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better $RMSE$ than La Rita.
-This situation arise because $RMSE$ considers errors only with respect the prediction and in 28 Millas the average of Stage of Evolution is $4316.16$, unlike, in La Rita the average is $5507.30$.
-So, in La Rita we obtains higher errors in absolute values.
-$R2$ is a relative metric between 0 thru 1 and it is less sensitive to absolute values.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.8]{Phase_one_R2_RMSE}
- \caption{Phase one - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura7} 
-\end{figure}
-
-The Pareto frontier for the La Rita farm is composed by 96 elements. The \tablename $.$\ref{tabla2} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase one} 
-\label{tabla2} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $\overline{W}$ &	1 to 1  & 36 & 64.25\% & 714.51 \\
- &	2 to 1  & 6 & 62.97\% & 695.10 \\
-\hline 
-All  & 1 to 1  & 18 & 62.98\% & 701.95 \\
-   & 2 to 1  & 12 & 61.76\% & 679.92 \\
-    & 3 to 1  & 6 & 60.60\% & 676.42 \\
-    & 5 to 1  &  2 & 60.37\% & 672.39 \\
-\hline    
-$\overline{T}_{a}$ & 1 to 1  & 12  & 63.60\% & 708.77 \\
-       &	2 to 1  & 4 & 62.23\% & 689.55 \\
-\hline
-\end{tabular} 
-\end{table}
-
-Similarly, the Pareto frontier for the 28 Millas farm is composed by 75 elements. The \tablename $.$\ref{tabla3} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - 28 Millas - Phase one} 
-\label{tabla3} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $\overline{W}$ & 1 to 1 & 8 & 57.80\% & 438.09 \\
-\hline 
-All   &	9 to 1 & 2 & 50.93\% & 397.93 \\
-  & 10 to 1	 & 2 & 50.97\% & 398.81 \\
-  &	8 to 1 & 6 & 51.62\% & 398.93 \\
-  &	7 to 1 & 2 & 52.25\% & 400.28 \\
-  &	6 to 1 & 2 & 53.16\% & 404.14 \\
-  &	4 to 1 & 2 & 54.32\% & 407.54 \\
-\hline    
-$\overline{T}_{a}$ & 1 to 1  & 8  & 59.09\% & 439.44 \\
-\hline
-Pair $\overline{T}_{a}$ $\overline{H}$ & 1 to 1	 & 8 & 57.51\% & 428.61 \\
- &	2 to 1 & 20 & 56.91\% & 414.37 \\
- &	3 to 1 & 3 & 54.41\% & 411.55 \\
- &	4 to 1 & 3 & 53.34\% & 406.65 \\
-\hline
-Pair $\overline{T}_{a}$ $P$ & 3 to 1 & 9 & 56.23\% & 422.76 \\
-\hline
-\end{tabular} 
-\end{table}
-
-We can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week.
-
-{\bf Phase two }
-
-In the second phase, the best configurations obtained in phase one were used to validate with the last 50 and 100 weeks.
-
-\figurename $.$\ref{figura8} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Even if linear models continue with good $R^2$, Echo State Networks improve their scores because in $50$ and $100$ last weeks validation, we are in presence of climate change, then the behaviour is less lineal.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_algortihm}
- \caption{Phase two - Best $R^2$ for each algorithm} 
- \label{figura8} 
-\end{figure}
-
-\figurename $.$\ref{figura9} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-The results confirms that, in general, to predict one week ahead is better than two weeks ahead, two than three and so on, this for both farms.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase two - Best $R^2$ for each observation range} 
- \label{figura9} 
-\end{figure}
-
-\figurename $.$\ref{figura10} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-This results confirm that $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms are the best variables combinations.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
- \caption{Phase two - Best $R^2$ for each variable combination} 
- \label{figura10} 
-\end{figure}
-
-\figurename $.$\ref{figura11} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.  You can note that the behaviour of $R^2$ and $RMSE$ is similar to the phase one.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.8]{Phase_two_R2_RMSE}
- \caption{Phase two - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura11} 
-\end{figure}
-
-In phase two, the Pareto frontier for the La Rita farm is composed by 2 elements. The \tablename $.$\ref{tabla4} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase two} 
-\label{tabla4} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-All & 1 to 1  & 2 & 69.33\% & 353.33 \\
-\hline
-\end{tabular} 
-\end{table}
-
-In 28 Millas Farm, the Pareto frontier is composed by 1 element. The \tablename $.$\ref{tabla5} details the result.
-
-\begin{table}[h] 
-\caption{Pareto frontier - 28 Millas - Phase two} 
-\label{tabla5} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $P$ & 1 to 1 & 1 & 61.36\% & 339.89 \\
-\hline
-\end{tabular} 
-\end{table}
-
-Again, similar to phase one, we can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week, one week to predict one week ahead with combinations of variables listed above.
-
-\section{Conclusions}
-\label{sec:concl}
-
-\subsection{Future work}
-
-\section{Acknowledgements}
-
-The authors would like to thank Corbana S.A. for providing the data for this research.
-
-\section{References}
-
-\bibliography{references}
-
-\end{document}

### 2016-06-11 related.tex
Nueva organizacion de los archivos.  Partes del paper en archivos separados y todas las imagenes en subdirectorio fig

--- 2016-05-22 Paper Sigatoka - 2015.tex
@@
-\documentclass[review,authoryear,english]{elsarticle}
-
-\usepackage{babel}
-\usepackage{natbib}
-
-\bibliographystyle{elsarticle-harv}
-
-\usepackage{lineno}
-\usepackage[utf8]{inputenc}  %% Permite utilizar tildes si archivo usa utf-8
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\modulolinenumbers[1]
-
-\usepackage{soul}
-\usepackage{marginnote}
-\usepackage{float}
-
-\usepackage{color}
-\definecolor{dkred}{rgb}{0.5,0,0}       %   dark red
-\definecolor{todocolor}{rgb}{1.0,0.8,0.5}
-
-\usepackage[hidelinks]{hyperref}
-
-\newcommand{\todohdr}{Please review:}
-\newcommand{\TODO}[2][\todohdr]{%
-\smallbreak\noindent\colorbox{todocolor}{%
-\begin{minipage}{1\linewidth}
-\textbf{#1}\par{\footnotesize{#2}}
-\end{minipage}}}
-
-\newcommand{\instead}[2]{\textcolor{dkred}{\st{#1}}#2}
-
-%  \usepackage{epstopdf} 
-%  \usepackage{epsfig}
-
-\journal{Journal of \LaTeX\ Templates}
-
-\usepackage{subcaption}
-
-\begin{document}
-
-\begin{frontmatter}
-
-\title{Forecasting the black Sigatoka development rate:\\ 
-A comparison of machine learning techniques}
-
-%% Group authors per affiliation:
-\author[afiLuisAlex]{Luis-Alexander Calvo-Valverde\fnref{myfootnote}}
-\ead{lualcava.sa@gmail.com}
-\fntext[myfootnote]{Corresponding author. (506)70104420}
-
-\author[afiCorbana]{Mauricio Guzmán-Quesada}
-\author[afiCorbana]{José-Antonio~Guzmán-Alvarez}
-\author[afiPablo]{Pablo Alvarado-Moya}
-
-\address[afiLuisAlex]{DOCINADE, Instituto Tecnológico de Costa Rica, 
-Computer Research Center, Multidisciplinar program eScience, Cartago, Costa Rica}
-
-\address[afiCorbana]{Dirección de Investigaciones, Corporación Bananera Nacional S.A., Guápiles, Costa Rica}
-
-\address[afiPablo]{DOCINADE, Instituto Tecnológico de Costa Rica, Cartago, Costa Rica}
-
-\begin{abstract}
-Pending.
-\end{abstract}
-
-\begin{keyword}
-Machine learning \sep Black Sigatoka \sep Support vector regression \sep
-Banana disease prediction \sep Biological warning system 
-\end{keyword}
-
-\end{frontmatter}
-
-\linenumbers
-
-\section{Introduction}
-
-The black Sigatoka disease caused by the fungus \emph{Mycosphaerella
-  fijiensis Morelet} is the major pathological problem of banana and
-plantain crops in Central America, Panama, Colombia and Ecuador, as well as in
-many parts of Africa and Asia \citep{MarinVargas1995}.
-
-This disease attacks the plant leaves producing a rapid deterioration
-of the leaf area, affects the growth and productivity of the plants
-due to the impairment of the photosynthetic process, causes a
-reduction in the quality of the fruit, and promotes premature
-ripening of bunches, which is
-the major cause of product losses associated with the black
-Sigatoka. \figurename~\ref{fig:diseasestages} shows three stages of
-this disease.
-
-\begin{figure}[ht] 
-\centering
-\begin{tabular}{c@{\;}c@{\;}c}
-  \includegraphics[width=.32\linewidth]{Roya_a} &
-  \includegraphics[width=.32\linewidth]{Roya_b} &
-  \includegraphics[width=.32\linewidth]{Roya_c} \\
-  (a) & (b) & (c) 
-\end{tabular}
-\caption{Examples of three disease stages of the black Sigatoka. 
-(a)~Initial stage. (b)~Intermediate stage, and (c)~Advanced stage.} 
-\label{fig:diseasestages} 
-\end{figure}
-
-Phytopathological studies point out that precipitation, temperature,
-relative humidity and wind are the main climatic variables that affect
-the development of this disease \citep{MarinVargas1995}.
-
-The climate has a major effect on the development of the black
-Sigatoka.  Precipitation, temperature, relative humidity and wind are
-the main climatic variables affecting the development of this disease
-\citep{MarinVargas1995}.
-
-The early warning system developed by Ganry and Meyer
-\citep{ganry1972} and modified by Ganry and Laville \citep{ganry1983}
-to control yellow Sigatoka in Cameroon, was later adapted by
-Ternesien \citep{Ternesien1985} and Fouré \citep{foure1988} for the
-black Sigatoka.
-
-This bilogical warning system is based on weekly quantified
-observations of the disease progression on young leaves of the plant,
-according to Fouré's scale of the symptom stages.
-%
-The numeric coefficients describe the degree of incidence and the
-severity of the disease development.  These coefficients are used to
-calculate two variables: gross sum and state of evolution.
-
-The gross sum is based on the present disease progression stage and
-the numeric coefficients, which increases with the progression of
-the symptoms and the juvenility of the leaf.
-%
-The state of evolution is calculated using the gross sum and the
-foliar emission period.
-%
-Although threshold levels were initially suggested as a guide to spray
-schedules, the fluctuation of these two variables
-was found to better define appropiate times to spray
-\citep{Marinetal2003}.
-
-In Costa Rica the black Sigatoka is frequently treated with
-chemical fungicides.
-%
-Depending on the zone of production and the weather conditions,
-45--55~cycles/year of fungicide applications are required to keep this
-disease under control and to produce the expected fruit quality for the
-international markets.
-%
-This represents a cost per hectare per year in the range between US\$1600 and
-US\$2000; about 0.64--0.80 cents of the production costs for a
-18.14\,kg box.
-%
-Overall, this represents 10\%--12\% of the total production costs.
-
-The past and present rates of disease development can in principle be used
-to predict its future behavior and to determine whether
-a particular fungicide spray program will be able to effectively and
-economically control the disease \citep{ChuangJeger1987} or not.
-
-There are efforts to apply machine learning methods to support
-decision-making in agriculture, including the control of crop
-diseases. For example, \cite{Camargo2012} present an intelligent
-system for the assessment of crop disorders, \cite{Huang2010}
-introduce a plant virus identification method based on neural networks
-with an evolutionary preprocessing stage, \cite{Kim2014} summarize in
-their survey crop pests prediction methods using regression and
-machine learning approaches, while \cite{Zhao2013} present an
-intelligent agricultural forecasting system based on wireless sensor
-networks.
-
-In this work, we compare five machine learning techniques 
-to predict the development rate of the black Sigatoka disease:
-support
-vector regression (SVR), echo state networks (ESN), ridge regression,
-elastic-net regression and ordinary least squares linear regression.
-%
-The main contribution of this work is the selection of the best
-machine learning technique to forecast the black Sigatoka development
-rate according to the metric indicated below.
-
-The outline of the paper is as follows: in Section~\ref{sec:techs}, we
-introduce the machine learning techniques used in this research and
-the main related works. Section~\ref{sec:data} summarizes the data the
-methodology used in this study.  The results and their discussion are
-presented in section~\ref{sec:results}.  We present the main
-conclusions and lines for future works in Section~\ref{sec:concl}.
-
-\section{Machine learning techniques and Related works}
-\label{sec:techs}
-
-In this section we summary the machine learning techniques used and we present the main related works.
-
-\subsection{Support Vector Regression (SVR)}
-
-From the perspective of Support Vector Regression (SVR) the regression function $y = f(s)$ for a given dataset $D=\{(s_i,y_i)\}_{i=1}^n$ , is represented as a linear function of the form \citep{Wei2013}:
-$$f(s)=w^Ts+b$$
-where $w$ and $b$ are respectively the weight vector and the intercept of the model, and they are selected to find an optimal fit to the data available in $D$.
-
-For nonlinear cases, one proceeds by mapping the input p-dimensional vectors via a nonlinear function $\phi : R^p\rightarrow F$, onto the feature space $F$.
-After nonlinear mapping, the regression function evolves to a pervasive form:
-
-$$f(s)=w^T \phi (s)+b$$
-
-SVR uses the $\epsilon$-insensitive loss function:
-\begin{equation*}
-l= {\left| y - f(s) \right| }_{\epsilon } = 
-\begin{cases}
-   0 & {\left| y - f(s) \right| } \leq \epsilon \\
-   \left| y - f(s) \right| - \epsilon & \text{otherwise}
-\end{cases}
-\end{equation*}
-which ignores the error if the difference between the prediction value and the actual value is smaller than $\epsilon$.
-%
-The $\epsilon$-insensitive loss function allows to find the coefficients $w$ and $b$ by solving a convex optimization problem, which balances the empirical error and the generalization ability.
-%
-In SVR, the empirical error is measured by the loss function $\epsilon$-insensitive and the generalization ability is measured by the Euclidean norm of $w$ \citep{Cristiani2005}.
-%
-Then, the optimization problem to identify the regression model can be
-formulated by \citep{Wei2013}:
-
-\begin{equation} 
-\begin{aligned}
-& \underset{}{\text{minimize}}
-& & J(w,\xi_i , \xi_i^* ) = \frac{1}{2}   \Bigr| \Bigr| w \Bigr| \Bigr|^2 + C \sum_{i=1}^{n} (\xi_i , \xi_i^* )    \\
-& \text{subject to}
-& & \begin{array}{lcl} 
-y_i - w^T \phi (s) - b  \leq \epsilon + \xi_i   \\
-w^T \phi (s) + b - y_i \leq \epsilon + \xi_i^* & i= 1,2,...,n \\
-\xi_i , \xi_i^* \geq 0  \\
-\end{array}
-\end{aligned}
-\end{equation}  
-where $C$ denotes the penalty parameter between empirical and generalization errors, and  $\xi_i , \xi_i^*$ are slack variables. \figurename
-$.$\ref{figura2} shows this situation.
-
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{SVR}
- \caption{$\epsilon$-insensitive loss function \citep{Wei2013}.} 
- \label{figura2} 
-\end{figure}
-
-The solution of this optimization problem by the Lagrange method is given by:\\
-$$ f(s) = w^T \phi (s) + b = \sum_{i=1}^{n} (\alpha_i - \alpha_i^*) K (s,s_i) + b $$ 
-where $\alpha_i - \alpha_i^*$  are the Lagrange multipliers of the optimization problem’s dual form and $K(s_i,s_j )$ is the kernel function satisfying the Mercer condition, and holds:
-$$K(s_i,s_j ) = \big \langle  \phi(s_i) , \phi(s_j)  \big \rangle $$
-Operations in the kernel function $K(s,s_i )$ are performed in the input space rather than in the potentially high dimensional feature space of $\phi$ \citep{Alonso2013}.
-
-\subsection{Ordinary least squares regression}
-This method fits a linear model with coefficients $w = (w1,..,wp)$ to minimize the residual sum of squares between the observed responses in the dataset, and the responses predicted by the linear approximation.
-Mathematically it solves a problem of the form \citep{scikitlearn2011}:
-$$\min_{w} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  $$
-where $X$ denotes the features matriz. \\
-According Pedregosa et al.
-\citep{scikitlearn2011}  the coefficient estimates for Ordinary Least Squares rely on the independence of the model terms.
-When terms are correlated and the columns of the design matrix $X$ have an approximate linear dependence, the design matrix becomes close to singular and as a result, the least-squares estimate becomes highly sensitive to random errors in the observed response, producing a large variance.
-This situation of multicollinearity can arise, for example, when data are collected without an experimental design
-
-\subsection{Ridge regression}
-The ridge regression addresses some of the problems of ordinary least squares regression by imposing a penalty on the size of the coefficients.
-The ridge coefficients minimize a penalized residual sum of squares  \citep{scikitlearn2011}:
-
-$$\min_{w} { \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-Here, $\alpha \ \textgreater \  0$ is a complexity parameter that controls the amount of shrinkage: the larger the value of $\alpha$, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
-
-\subsection{Elastic-Net regression}
-Elastic-Net is a linear regression model trained with $L1$ and $L2$ prior as regularizer.
-This combination allows for learning a sparse model where few of the weights are non-zero like Lasso, while still maintaining the regularization properties of Ridge  \citep{scikitlearn2011}.
-The convex combination of $L1$ and $L2$ is controled by using the $l1_{ratio}$ parameter.
-
-Elastic-Net is useful when there are multiple features which are correlated with one another. Lasso is likely to pick one of these at random, while elastic-net is likely to pick both.
-A practical advantage of trading-off between Lasso and Ridge is it allows Elastic-Net to inherit some of Ridge’s stability under rotation.
-The objective function to minimize is \citep{scikitlearn2011}:
-
-$$\min_{w} { \frac{1}{2n_{samples}} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \rho \Bigr| \Bigr| w \Bigr| \Bigr|_1 + \frac{\alpha (1- \rho)}{2} \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-\subsection{Echo State Networks (ESN)}
-Recurrent Neural Networks (RNN) are useful for temporal patterns, but when they are trained with backpropagation methods, they are very slow.
-Echo State Network (ESN) is an alternative training method to solve that problem.
-ESN is based on the observation that if a random RNN possesses certain algebraic properties, training only a linear readout from it is often sufficient to achieve excellent performance in practical applications \citep{Lukose2009}.
-For a given training input signal $u(n)  \in R^{N_u}$ a desired target output signal $y^{target}(n) \in R^{N_y}$
-is known.
-Here $n = 1, .
-.
-.
-,T$ is the discrete time and $T$ is the number of data points in the training dataset.
-The task is to learn a model with output $y(n) \in R^{N_y}$, where $y(n)$ matches $y^target(n)$ as well as possible, minimizing an error measure $E(y,y^target)$, and, more importantly, generalizes well to unseen data.
-The untrained RNN part of an ESN is called a dynamical reservoir, and the resulting states x(n) are termed echoes of its input history \citep{Lukose2012}.
-Finally, these signals are sent to an output layer as shown in the \figurename
-$.$\ref{figura3}.
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{Reservorio}
- \caption{An echo state network \citep{Lukose2012}.} 
- \label{figura3} 
-\end{figure}
-
-The connections between the different elements of an Echo State Network have weights randomly generated.
-The weights of the internal connections of the reservoir $(W)$ as well as the weights of the input layer $(W_in)$, after being generated are set statically during all stages of implementation of the algorithm.
-The weights between the reservoir and the output layer $(W_out)$ are subject to changes of a supervised learning algorithm to correct the degree of error generated by the entire system \citep{Lukose2012}.
-
-\subsection{Related works}
-Now, we present the main and recently works related with our study.
+\section{Related works}
+\label{sec:related}
 
@@
 Our study focuses in the second case.
-
-\section{Specification of data and methodology}
-\label{sec:data}
-
-In this section, we present the data and how we worked to obtain our results.
-
-\subsection{Data}
-We use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (previously called Waldeck and located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
-The banana type is Musa AAA, subgroup Cavendish, cv.
-Grande Naine.
-The \tablename $.$\ref{tabla1} shows the variables available.
-
-\begin{table}[h] 
-\caption{Variables used in the study} 
-\label{tabla1} 
-\centering
-\begin{tabular}{c|c} 
-\hline
-\bfseries Variable & \bfseries Meaning0'0' \\ 
-\hline\hline 
-$T_{a_{max}}$ &	Max air temperature \\
-$T_{a_{min}}$ &	Min air temperature \\
-$\overline{T}_{a}$	 & Mean air temperature \\
-$\overline{H}$	&  Mean Humidity \\
-$H_{min}$ &	Min humidity \\
-$H_{max}$	& Max humidity \\
-$\overline{R}$	& Mean Solar radiation \\
-$P$	& Sum precipitation  \\
-$W_{max}$	& Max speed wind \\
-$\overline{W}$	& Mean speed wind  \\
-$L_2$	& Biological warning system – Leaf 2 \\
-$L_3$	& Biological warning system – Leaf 3 \\
-$L_4$	& Biological warning system – Leaf 4 \\
-$E_s$	& Biological warning system – Evolution Stage \\
-\hline
-\end{tabular} 
-\end{table}
-
-The value to be predicted in all cases was $E_s$, that is the total measure of the biological warning system.
-
-The data on the biological warning system are collected once a week.
-Although Corbana has meteorological stations that take data every five minutes, for these experiments, weekly averages generated by nearby stations to each of the farms were used.
-
-The time intervals used for this study were: La Rita, week 48 of 2002 to week 17 of the 2015 (647 weeks) and for 28 Miles, week 37 of 2003 to week 18 of 2015 (605 weeks).
-
-\subsection{Data preprocessing}
-Usually, data need preprocessing to prepare for use. In this section we explain that process.
-
-In 28 Miles farm, 1\% of the data were missing, while in La Rita was 2.25\%. To fill-in the missing values we use spline interpolation. The data collected did not exhibit outliers.
-
-Due the fact that the variables measure meteorological or biological process, they are discretized in order to reflect trends in the data, i.e.
-the continuous values are not directly used.
-The coefficient of variation $C_v(x)$ of each variable x was used to determine the number $n$ of discretization levels.
-$$n= \lfloor 100 \ C_v(x) \rfloor$$
-where $\lfloor \ \rfloor$  is the round operator.
-
-Each discretization range was uniformly partitioned.
-Besides enabling the capture of tendencies, the discretization removes the effect of small variations in the data collection, either by inaccuracies of the instruments (meteorological variables) or by subjective bias introduced by the human who collects the data (biological warning system).
-
-Each feature was scaled to fit in a range between 0 and 1. The variable to be predicted was not scaled.
-
-\subsection{Evaluation criteria}
-
-Although there are many types of indicators to assess the quality of the prediction, we selected the determination coefficient $(R^2)$ and the Root Mean Square Error $(RMSE)$.
-
-Given $n$ records, let be $y$ the actual value of the series, $\hat{y}$ the predicted value and $\acute{y}$ the mean of the observed data.
-
-$$ \bar{y} = \frac{1}{n} \sum_{i=1}^{n} y_i $$
-
-$$ S_e^2 = \frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}$$
-
-$$ S_R^2 = \frac{\sum_{i=1}^{n} {(\hat{y}_i-\bar{y}_i)}^2 }{n}$$
-
-$$ S_y^2 = S_R^2 + S_e^2$$
-
-$$ R^2 = \frac{S_R^2}{S_y^2}$$
-
-$$ RMSE = \sqrt{\frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}}$$
-
-This decision is supported by the widespread use in machine learning and agriculture areas \citep{Soares2014}, \citep{Soares2013}, \citep{Ibrahim2014} and \citep{Demir2014}.
-
-\subsection{Programming environment}
-
-We use the python programming language with the Integrated Development Environment (IDE) Spyder \citep{Continuum2015}, particularly with the libraries pandas \citep{mckinneypandas2010} and numpy \citep{vanderWalt2011}.
-For SVR, ridge and ordinary least squares regressions, we used sklearn \citep{scikitlearn2011} and for ESN the python-based code of Dr.
-$Luko\breve{s} evi \breve{c} ius$ \citep{Lukose2012} on which the necessary were done for adjustments for the experiments of this work.
-The computer used a processor Intel(R) Core i7-4800MQ CPU 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
-
-\subsection{Methodology}
-The selection of the technique and its parametrisation was performed in two stages that they are presented below.
-
-{\bf Phase one }
-
-In the phase one, we did ten-fold-cross-validation on a set of  machine learning methods and different configurations:
-
-\begin{itemize}
-\item 	Patterns: n by m, where n from 1 to 8 and m from 1 to 2.
-
-\item Methods: support vector regression with the kernels functions: linear, gaussian and sigmoid; echo state networks; ordinary least squares linear regression, ridge regression and elastic-net regression.
-
-\item Variables included in the model:
-\begin{itemize}
-\item All variables.
-\item From the set $\{ \overline{T}_{a} , \overline{H}, P , \overline{W}  \}$ use the subsets with one, two or four elements.
-These variables are according to experts the ones having most impact on the disease development \citep{MarinVargas1995}.
-\end{itemize}
-
-\end{itemize}
-
-{\bf Phase two }
-
-In the second phase, the best configurations obtained in phase one are used to validate with the last 50 and 100 weeks.
-
-\section{Results and discussion}
-\label{sec:results}
-
-In this section we present the main results for phase and their discussion.
-
-{\bf Phase one }
-
-\figurename $.$\ref{figura4} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Though La Rita obtains different results in magnitude than 28 Millas, the trend is similar.
-In both farms, the best results are for linear models, second position is occupied for Echo State Networks and  SVR with gausian and sigmoid kernels are the worst results.
-In linear models, to predict one week ahead is better than two weeks ahead, and this is better than three weeks ahead.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_algortihm}
- \caption{Phase one - Best $R^2$ for each algorithm} 
- \label{figura4} 
-\end{figure}
-
-\figurename $.$\ref{figura5} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-In general, to predict one week ahead is better than two weeks ahead and so on.
-The number of weeks consider in the observation range in the pattern is not the main discriminant factor, but it is clear that we get better $R^2$ for one week ahead than two weeks ahead and so on.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase one - Best $R^2$ for each observation range} 
- \label{figura5} 
-\end{figure}
-
-\figurename $.$\ref{figura6} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-The better results are obtained with $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms of similarly.
-You can note that the use of all variables in the model or the inclusion of the four variables suggest for expert criteria do not improve significantly the results, then the use of more sensors do not assure a better result.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
- \caption{Phase one - Best $R^2$ for each variable combination} 
- \label{figura6} 
-\end{figure}
-
-\figurename $.$\ref{figura7} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
-The Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better $RMSE$ than La Rita.
-This situation arise because $RMSE$ considers errors only with respect the prediction and in 28 Millas the average of Stage of Evolution is $4316.16$, unlike, in La Rita the average is $5507.30$.
-So, in La Rita we obtains higher errors in absolute values.
-$R2$ is a relative metric between 0 thru 1 and it is less sensitive to absolute values.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.8]{Phase_one_R2_RMSE}
- \caption{Phase one - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura7} 
-\end{figure}
-
-The Pareto frontier for the La Rita farm is composed by 96 elements. The \tablename $.$\ref{tabla2} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase one} 
-\label{tabla2} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $\overline{W}$ &	1 to 1  & 36 & 64.25\% & 714.51 \\
- &	2 to 1  & 6 & 62.97\% & 695.10 \\
-\hline 
-All  & 1 to 1  & 18 & 62.98\% & 701.95 \\
-   & 2 to 1  & 12 & 61.76\% & 679.92 \\
-    & 3 to 1  & 6 & 60.60\% & 676.42 \\
-    & 5 to 1  &  2 & 60.37\% & 672.39 \\
-\hline    
-$\overline{T}_{a}$ & 1 to 1  & 12  & 63.60\% & 708.77 \\
-       &	2 to 1  & 4 & 62.23\% & 689.55 \\
-\hline
-\end{tabular} 
-\end{table}
-
-Similarly, the Pareto frontier for the 28 Millas farm is composed by 75 elements. The \tablename $.$\ref{tabla3} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - 28 Millas - Phase one} 
-\label{tabla3} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $\overline{W}$ & 1 to 1 & 8 & 57.80\% & 438.09 \\
-\hline 
-All   &	9 to 1 & 2 & 50.93\% & 397.93 \\
-  & 10 to 1	 & 2 & 50.97\% & 398.81 \\
-  &	8 to 1 & 6 & 51.62\% & 398.93 \\
-  &	7 to 1 & 2 & 52.25\% & 400.28 \\
-  &	6 to 1 & 2 & 53.16\% & 404.14 \\
-  &	4 to 1 & 2 & 54.32\% & 407.54 \\
-\hline    
-$\overline{T}_{a}$ & 1 to 1  & 8  & 59.09\% & 439.44 \\
-\hline
-Pair $\overline{T}_{a}$ $\overline{H}$ & 1 to 1	 & 8 & 57.51\% & 428.61 \\
- &	2 to 1 & 20 & 56.91\% & 414.37 \\
- &	3 to 1 & 3 & 54.41\% & 411.55 \\
- &	4 to 1 & 3 & 53.34\% & 406.65 \\
-\hline
-Pair $\overline{T}_{a}$ $P$ & 3 to 1 & 9 & 56.23\% & 422.76 \\
-\hline
-\end{tabular} 
-\end{table}
-
-We can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week.
-
-{\bf Phase two }
-
-In the second phase, the best configurations obtained in phase one were used to validate with the last 50 and 100 weeks.
-
-\figurename $.$\ref{figura8} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Even if linear models continue with good $R^2$, Echo State Networks improve their scores because in $50$ and $100$ last weeks validation, we are in presence of climate change, then the behaviour is less lineal.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_algortihm}
- \caption{Phase two - Best $R^2$ for each algorithm} 
- \label{figura8} 
-\end{figure}
-
-\figurename $.$\ref{figura9} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-The results confirms that, in general, to predict one week ahead is better than two weeks ahead, two than three and so on, this for both farms.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase two - Best $R^2$ for each observation range} 
- \label{figura9} 
-\end{figure}
-
-\figurename $.$\ref{figura10} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-This results confirm that $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms are the best variables combinations.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
- \caption{Phase two - Best $R^2$ for each variable combination} 
- \label{figura10} 
-\end{figure}
-
-\figurename $.$\ref{figura11} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.  You can note that the behaviour of $R^2$ and $RMSE$ is similar to the phase one.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.8]{Phase_two_R2_RMSE}
- \caption{Phase two - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura11} 
-\end{figure}
-
-In phase two, the Pareto frontier for the La Rita farm is composed by 2 elements. The \tablename $.$\ref{tabla4} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase two} 
-\label{tabla4} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-All & 1 to 1  & 2 & 69.33\% & 353.33 \\
-\hline
-\end{tabular} 
-\end{table}
-
-In 28 Millas Farm, the Pareto frontier is composed by 1 element. The \tablename $.$\ref{tabla5} details the result.
-
-\begin{table}[h] 
-\caption{Pareto frontier - 28 Millas - Phase two} 
-\label{tabla5} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $P$ & 1 to 1 & 1 & 61.36\% & 339.89 \\
-\hline
-\end{tabular} 
-\end{table}
-
-Again, similar to phase one, we can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week, one week to predict one week ahead with combinations of variables listed above.
-
-\section{Conclusions}
-\label{sec:concl}
-
-\subsection{Future work}
-
-\section{Acknowledgements}
-
-The authors would like to thank Corbana S.A. for providing the data for this research.
-
-\section{References}
-
-\bibliography{references}
-
-\end{document}

### 2016-06-11 results.tex
Nueva organizacion de los archivos.  Partes del paper en archivos separados y todas las imagenes en subdirectorio fig

--- 2016-05-22 Paper Sigatoka - 2015.tex
@@
-\documentclass[review,authoryear,english]{elsarticle}
-
-\usepackage{babel}
-\usepackage{natbib}
-
-\bibliographystyle{elsarticle-harv}
-
-\usepackage{lineno}
-\usepackage[utf8]{inputenc}  %% Permite utilizar tildes si archivo usa utf-8
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\modulolinenumbers[1]
-
-\usepackage{soul}
-\usepackage{marginnote}
-\usepackage{float}
-
-\usepackage{color}
-\definecolor{dkred}{rgb}{0.5,0,0}       %   dark red
-\definecolor{todocolor}{rgb}{1.0,0.8,0.5}
-
-\usepackage[hidelinks]{hyperref}
-
-\newcommand{\todohdr}{Please review:}
-\newcommand{\TODO}[2][\todohdr]{%
-\smallbreak\noindent\colorbox{todocolor}{%
-\begin{minipage}{1\linewidth}
-\textbf{#1}\par{\footnotesize{#2}}
-\end{minipage}}}
-
-\newcommand{\instead}[2]{\textcolor{dkred}{\st{#1}}#2}
-\newcommand{\figref}[1]{\figurename~\ref{#1}}
-
-%  \usepackage{epstopdf} 
-%  \usepackage{epsfig}
-
-\journal{Computers and Electronics in Agriculture}
-
-\usepackage{subcaption}
-
-\begin{document}
-
-\begin{frontmatter}
-
-\title{Forecasting the black Sigatoka development rate:\\ 
-A comparison of machine learning techniques}
-
-%% Group authors per affiliation:
-\author[afiLuisAlex]{Luis-Alexander Calvo-Valverde\fnref{myfootnote}}
-\ead{lualcava.sa@gmail.com}
-\fntext[myfootnote]{Corresponding author. (506)70104420}
-
-\author[afiCorbana]{Mauricio Guzmán-Quesada}
-\author[afiCorbana]{José-Antonio~Guzmán-Alvarez}
-\author[afiPablo]{Pablo Alvarado-Moya}
-
-\address[afiLuisAlex]{DOCINADE, Instituto Tecnológico de Costa Rica, 
-Computer Research Center, Multidisciplinar program eScience, Cartago, Costa Rica}
-
-\address[afiCorbana]{Dirección de Investigaciones, Corporación Bananera Nacional S.A., Guápiles, Costa Rica}
-
-\address[afiPablo]{DOCINADE, Instituto Tecnológico de Costa Rica, Cartago, Costa Rica}
-
-\begin{abstract}
-Pending.
-\end{abstract}
-
-\begin{keyword}
-Machine learning \sep Black Sigatoka \sep Support vector regression \sep
-Banana disease prediction \sep Biological warning system 
-\end{keyword}
-
-\end{frontmatter}
-
-\linenumbers
-
-\section{Introduction}
-
-The black Sigatoka disease caused by the fungus \emph{Mycosphaerella
-  fijiensis Morelet} is the major pathological problem of banana and
-plantain crops in Central America, Panama, Colombia and Ecuador, as well as in
-many parts of Africa and Asia \citep{MarinVargas1995}.
-
-This disease attacks the plant leaves producing a rapid deterioration
-of the leaf area, affects the growth and productivity of the plants
-due to the impairment of the photosynthetic process, causes a
-reduction in the quality of the fruit, and promotes premature
-ripening of bunches, which is
-the major cause of product losses associated with the black
-Sigatoka. \figref{fig:diseasestages} shows three stages of
-this disease.
-
-\begin{figure}[ht] 
-\centering
-\begin{tabular}{c@{\;}c@{\;}c}
-  \includegraphics[width=.32\linewidth]{Roya_a} &
-  \includegraphics[width=.32\linewidth]{Roya_b} &
-  \includegraphics[width=.32\linewidth]{Roya_c} \\
-  (a) & (b) & (c) 
-\end{tabular}
-\caption{Examples of three disease stages of the black Sigatoka. 
-(a)~Initial stage. (b)~Intermediate stage, and (c)~Advanced stage.} 
-\label{fig:diseasestages} 
-\end{figure}
-
-Phytopathological studies point out that precipitation, temperature,
-relative humidity and wind are the main climatic variables that affect
-the development of this disease \citep{MarinVargas1995}.
-
-The climate has a major effect on the development of the black
-Sigatoka.  Precipitation, temperature, relative humidity and wind are
-the main climatic variables affecting the development of this disease
-\citep{MarinVargas1995}.
-
-The early warning system developed by Ganry and Meyer
-\citep{ganry1972} and modified by Ganry and Laville \citep{ganry1983}
-to control yellow Sigatoka in Cameroon, was later adapted by
-Ternesien \citep{Ternesien1985} and Fouré \citep{foure1988} for the
-black Sigatoka.
-
-This bilogical warning system is based on weekly quantified
-observations of the disease progression on young leaves of the plant,
-according to Fouré's scale of the symptom stages.
-%
-Numeric coefficients are used to describe the degree of incidence and
-the severity of the disease development.  These coefficients are used
-to calculate two variables: gross sum and state of evolution.
-
-The gross sum is based on the present disease progression stage and
-the numeric coefficients, which increase with the progression of
-the symptoms and the juvenility of the leaf.
-%
-The state of evolution is calculated using the gross sum and the
-foliar emission period.
-%
-Although threshold levels were initially suggested as a guide to spray
-schedules, the fluctuation of these two variables
-was found to better define appropiate times to spray
-\citep{Marinetal2003}.
-
-\TODO{No queda claro threshold levels sobre qué variables!?  Tampoco
-  queda claro que quiere decir 'fluctuation' como medida de decisión }
-
-In Costa Rica the black Sigatoka is frequently treated with
-chemical fungicides.
-%
-Depending on the zone of production and the weather conditions,
-45--55~cycles/year of fungicide applications are required to keep this
-disease under control and to produce the expected fruit quality for the
-international markets.
-%
-This represents a cost per hectare per year in the range between US\$1600 and
-US\$2000; about 0.64--0.80 cents of the production costs for a
-18.14\,kg box.
-%
-Overall, this represents 10\%--12\% of the total production costs.
-
-The past and present rates of disease development can in principle be used
-to predict its future behavior and to determine whether
-a particular fungicide spray program will be able to effectively and
-economically control the disease \citep{ChuangJeger1987} or not.
-
-There are efforts to apply machine learning methods to support
-decision-making in agriculture, including the control of crop
-diseases. For example, \cite{Camargo2012} present an intelligent
-system for the assessment of crop disorders, \cite{Huang2010}
-introduce a plant virus identification method based on neural networks
-with an evolutionary preprocessing stage, \cite{Kim2014} summarize in
-their survey crop pests prediction methods using regression and
-machine learning approaches, while \cite{Zhao2013} present an
-intelligent agricultural forecasting system based on wireless sensor
-networks.
-
-In this work, we compare five machine learning techniques to predict
-the development rate of the black Sigatoka disease: support vector
-regression (SVR), echo state networks (ESN), ridge regression,
-elastic-net regression and ordinary least squares linear regression.
-%
-The main contribution of this work is the selection of the best
-machine learning technique to forecast the black Sigatoka development
-rate according to the metric indicated below.
-
-\TODO{Definitivamente ese no es el aporte principal del artículo.
-  Cuando esté listo el resto hay que volver aquí.  Creo que va a ser
-  algo como mostrar la relevancia de la etapa de preprocesamiento, o
-  mostrar que a pesar de ser sistemas complejos, con variables
-  caóticas, lo mejor en este caso son predictores lineales, o algo por
-  el estilo.}
-
-The outline of the paper is as follows: Section~\ref{sec:techs}
-summarizes the machine learning techniques used in this research and
-presents the main related works. In Section~\ref{sec:data} we present
-the methodology used in this study and describe data used for its
-verification.  The results and their discussion are presented in
-section~\ref{sec:results}.  The Section~\ref{sec:concl} concludes this
-article and presents lines for future works.
-
-\section{Machine learning techniques and Related works}
-\label{sec:techs}
-
-In this section we summarize the machine learning techniques used and we present the main related works.
-
-\subsection{Support Vector Regression (SVR)}
-
-\TODO{Estoy seguro que los lectores van a solicitar reducir esto y
-  dejar solo la referencia.  No hay nada nuevo en el resto del paper
-  que requiera toda esta información, así que hay que resumirlo a nada
-  más lo esencial. }
-
-From the perspective of Support Vector Regression (SVR) the regression
-function $y = f(s)$ for a given dataset $D=\{(s_i,y_i)\}_{i=1}^n$ is
-represented as a linear function of the form \citep{Wei2013}:
-\begin{equation*}
-  f(s)=w^Ts+b
-\end{equation*}
-where $w$ and $b$ are respectively the weight vector and the intercept
-of the model, and they are selected to find an optimal fit to the data
-available in $D$ in the sense of ...?
-\TODO{optimal respecto a qué?.  Hay muchísimas posibilidades de optimalidad!}
-
-For nonlinear cases, the $p$-dimensional input vectors are mapped by a
-nonlinear function $\phi : R^p\rightarrow F$ onto a feature space
-$F$.  After the nonlinear mapping, the regression function evolves to a
-pervasive form:
-\begin{equation*}
-f(s)=w^T \phi (s)+b
-\end{equation*}
-
-In SVR the generalization ability is measured by the Euclidean norm of
-$w$, and the empirical error is measured by the $\epsilon$-insensitive
-loss function, given by \citep{Cristiani2005}
-
-\TODO{si $l$ es una función, es función de qué, de $s$? $l(s)$? o de
-  $y$? o de todos $l(y,s;\epsilon)$?}
-
-\begin{equation*}
-l= {\left| y - f(s) \right| }_{\epsilon } = 
-\begin{cases}
-   0 & {\left| y - f(s) \right| } \leq \epsilon \\
-   \left| y - f(s) \right| - \epsilon & \text{otherwise}
-\end{cases}
-\end{equation*}
-which ignores the error if the difference between the prediction value
-and the actual value is smaller than $\epsilon$.
-%
-The $\epsilon$-insensitive loss function allows to find the
-coefficients $w$ and $b$ by solving a convex optimization problem,
-which balances the empirical error and the generalization ability.
-%
-Then, the optimization problem to identify the regression model can be
-formulated as \citep{Wei2013}:
-%
-\begin{equation} 
-\begin{aligned}
-& \underset{}{\text{minimize}}
-& & J(w,\xi_i , \xi_i^* ) = \frac{1}{2}   \Bigr| \Bigr| w \Bigr| \Bigr|^2 + C \sum_{i=1}^{n} (\xi_i , \xi_i^* )    \\
-& \text{subject to}
-& & \begin{array}{lcl} 
-y_i - w^T \phi (s) - b  \leq \epsilon + \xi_i   \\
-w^T \phi (s) + b - y_i \leq \epsilon + \xi_i^* & i= 1,2,...,n \\
-\xi_i , \xi_i^* \geq 0  \\
-\end{array}
-\end{aligned}
-\end{equation}  
-where $C$ denotes the penalty parameter between empirical and
-generalization errors, and $\xi_i , \xi_i^*$ are slack
-variables. \figref{figura2} shows this situation.
-
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{SVR}
- \caption{$\epsilon$-insensitive loss function \citep{Wei2013}.} 
- \label{figura2} 
-\end{figure}
-
-The solution of this optimization problem by the Lagrange method is given by:
-\begin{equation*}
-f(s) = w^T \phi (s) + b = \sum_{i=1}^{n} (\alpha_i - \alpha_i^*) K (s,s_i) + b
-\end{equation*}
-where $\alpha_i - \alpha_i^*$ are the Lagrange multipliers of the
-optimization problem’s dual form and $K(s_i,s_j )$ is the kernel
-function satisfying the Mercer condition, and holds:
-\begin{equation*}
-K(s_i,s_j ) = \big \langle  \phi(s_i) , \phi(s_j)  \big \rangle
-\end{equation*}
-Operations in the kernel function $K(s,s_i )$ are performed in the
-input space rather than in the potentially high dimensional feature
-space of $\phi$ \citep{Alonso2013}.
-
-\subsection{Ordinary least squares regression}
-
-This method fits a linear model with coefficients $w = (w_1,..,w_p)$
-to minimize the residual sum of squares between the observed responses
-in the dataset, and the responses predicted by the linear
-approximation. Mathematically it solves a problem of the form
-\citep{scikitlearn2011}:
-$$\min_{w} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  $$
-where $X$ denotes the features matrix.
-
-According Pedregosa et al.
-\citep{scikitlearn2011}  the coefficient estimates for Ordinary Least Squares rely on the independence of the model terms.
-When terms are correlated and the columns of the design matrix $X$ have an approximate linear dependence, the design matrix becomes close to singular and as a result, the least-squares estimate becomes highly sensitive to random errors in the observed response, producing a large variance.
-This situation of multicollinearity can arise, for example, when data are collected without an experimental design.
-
-\subsection{Ridge regression}
-The ridge regression addresses some of the problems of ordinary least squares regression by imposing a penalty on the size of the coefficients.
-The ridge coefficients minimize a penalized residual sum of squares  \citep{scikitlearn2011}:
-
-$$\min_{w} { \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-Here, $\alpha \ \textgreater \  0$ is a complexity parameter that controls the amount of shrinkage: the larger the value of $\alpha$, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
-
-\subsection{Elastic-Net regression}
-Elastic-Net is a linear regression model trained with $L1$ and $L2$ prior as regularizer.
-This combination allows for learning a sparse model where few of the weights are non-zero like Lasso, while still maintaining the regularization properties of Ridge  \citep{scikitlearn2011}.
-The convex combination of $L1$ and $L2$ is controled by using the $l1_{ratio}$ parameter.
-
-Elastic-Net is useful when there are multiple features which are correlated with one another. Lasso is likely to pick one of these at random, while elastic-net is likely to pick both.
-A practical advantage of trading-off between Lasso and Ridge is it allows Elastic-Net to inherit some of Ridge’s stability under rotation.
-The objective function to minimize is \citep{scikitlearn2011}:
-
-$$\min_{w} { \frac{1}{2n_{samples}} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \rho \Bigr| \Bigr| w \Bigr| \Bigr|_1 + \frac{\alpha (1- \rho)}{2} \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-\subsection{Echo State Networks (ESN)}
-Recurrent Neural Networks (RNN) are useful for temporal patterns, but when they are trained with backpropagation methods, they are very slow.
-Echo State Network (ESN) is an alternative training method to solve that problem.
-ESN is based on the observation that if a random RNN possesses certain algebraic properties, training only a linear readout from it is often sufficient to achieve excellent performance in practical applications \citep{Lukose2009}.
-For a given training input signal $u(n)  \in R^{N_u}$ a desired target output signal $y^{target}(n) \in R^{N_y}$
-is known.
-Here $n = 1, .
-.
-.
-,T$ is the discrete time and $T$ is the number of data points in the training dataset.
-The task is to learn a model with output $y(n) \in R^{N_y}$, where $y(n)$ matches $y^target(n)$ as well as possible, minimizing an error measure $E(y,y^target)$, and, more importantly, generalizes well to unseen data.
-The untrained RNN part of an ESN is called a dynamical reservoir, and the resulting states x(n) are termed echoes of its input history \citep{Lukose2012}.
-Finally, these signals are sent to an output layer as shown in the \figref{figura3}.
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{Reservorio}
- \caption{An echo state network \citep{Lukose2012}.} 
- \label{figura3} 
-\end{figure}
-
-The connections between the different elements of an Echo State Network have weights randomly generated.
-The weights of the internal connections of the reservoir $(W)$ as well as the weights of the input layer $(W_in)$, after being generated are set statically during all stages of implementation of the algorithm.
-The weights between the reservoir and the output layer $(W_out)$ are subject to changes of a supervised learning algorithm to correct the degree of error generated by the entire system \citep{Lukose2012}.
-
-\subsection{Related works}
-Now, we present the main and recently works related with our study.
-
-Huang et al.
-\citep{Huang2010}  surveyed the development of soft computing techniques in agricultural and biological engineering, including fuzzy logic, artificial neural networks, genetic algorithms, bayesian inference and decision trees.
-
-A related work, proposed by Romero \citep{Romero1995} relies on regression models using a stepwise procedure to predict incubation and latency times of black Sigatoka.
-The author performed experiments on two farms located in Costa Rica (La Rita and Waldeck, the same as those used in this study but with different names).
-The study used data from: December 1993 to August 1995.
-Romero concluded that the model to predict the incubation period accounted a $R^2$ of 69\% in his observed data but it was not a good predictor when it was validated against an independent dataset (cross validation).
-For latency, he developed two models that accounted a $R^2$ of 78\% in the observed data, however, when validated against an independent dataset (cross validation), for La Rita obtained an adjusted $R^2$ of 82\%, and for Weldeck, none of the models satisfactorily predicted the latent period; and then those predictions were not shown \citep{Romero1995}.
-
-Glezakos et al.
-\citep{Glezakos2010} proposed to use Genetic Algorithms (GA) and Neural Networks (NN) to identify plant virus (Tobacco Rattle Virus (TRV) and the Cucumber Green Mottle Mosaic Virus (CGMMV)).
-This is achieved by the development of ana- lytical tools of evolutionary adaptive width, propelled by Genetic Algorithms (GAs) and Neural Networks (NNs).
-The method was tested against some of the most commonly used classifiers in machine learning (Bayes, Trees and k-NN) via cross-validation and proved its potential towards the identification.
-
-In the agricultural context, Alves et al.
-\citep{Alves2011} used  geoinformation techniques to develop predictive models to study the areas of risk to soybean rust in soybean, coffee leaf rust in coffee, and black Sigatoka in banana, considering Brazil’s climatic characterization and the distribution of soybean, coffee and banana crops.
-Temperature and rainfall data were obtained for the period from 1950 to 2000, and of simulations for 2020, 2050 and 2080 using the SRES A2 climate change scenarios.
-Using principal components analysis, a single variable was generated based on 57 variables, in order to determine an index explaining 87\%, 88\% and 90\% of the variability of soybean, coffee and banana crops, respectively, in municipal districts across Brazil.
-The climatic model was used to generate the zoning of the three plant diseases, using temperature and leaf wetness as input.
-Areas of favorability for the diseases were plotted against the main coffee, soybean and banana growing areas in Brazil.
-This methodology enabled the visualization of the changes in areas favorable for epidemics under possible future scenarios of climate change.
-
-Other applications of machine learning methods in precision agriculture include the use of support vector regression to predict carcass weight in beef cattle in advance to the slaughter  \citep{Alonso2013}, machine learning assessments of soil drying for agricultural planning \citep{Coopersmith2014}, and early detection and classification of plant diseases with support vector machines based on hyperspectral reflectance \citep{Rumpf2010}.
-
-Furthermore, there have been attempts to generate software tools.
-Camargo et al.
-\citep{Camargo2012} presented an information system for the assessment of plant disorders (Isacrodi).
-They proposed that experts will attain a much better accuracy than the Isacrodi classifier, particularly when provided with samples from the affected crop.
-However, those cases where such expertise is not available, they suggest that Isacrodi can provide valuable support to farmers.
-Isacordi includes 15 crop disorders, but the black Sigatoka no is one of them.
-The prediction process is based on multi-class support vector machines.
-
-Regarding the prediction of the develpment of the black Sigatoka with machine learning methods, Bendini et al.
-\citep{Bendini2013}  presented a study about the risk analysis of black Sigatoka occurrence based on polynomial models.
-A case study was developed in a commercial banana plantation located in Jacupiranga, Brazil.
-It was monitored weekly during the period from February to December 2005.
-Data included the weekly monitoring of the disease’s evolution stage, time series of meteorological data and remote sensing data.
-They obtained a model to estimate the evolution of the disease from satellite imagery.
-This model relates gray levels (NC) of the band 2 images of the Landsat-5 satellite, with the progress status or disease severity (EE).
-The authors claim to reach an $R^2$ of 90\%.
-
-Also there are works related to banana fruit.
-Soares et al.
-\citep{Soares2014} apply two techniques: artificial neural networks (ANNs) and multiple linear regression (MLR) in banana plant to predict the yield, their results show that the neural network proved to be more accurate in forecasting the weight of the bunch in comparison to the multiple linear regressions in terms of the mean prediction-error $(MPE = 1.40)$, mean square deviation $(MSD = 2.29)$ and coefficient of determination $(R^2 = 91\%)$.
-
-In general, the machine learning methods applied to predict the evolution of plant diseases, can be classified in two main approaches: 1) Those whose main inputs are images, and 2) Those whose main inputs are environmental and biological variables.
-Our study focuses in the second case.
-
-\section{Specification of data and methodology}
-\label{sec:data}
-
-In this section, we present the data and how we worked to obtain our results.
-
-\subsection{Data}
-We use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (previously called Waldeck and located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
-The banana type is Musa AAA, subgroup Cavendish, cv.
-Grande Naine.
-The \tablename $.$\ref{tabla1} shows the variables available.
-
-\begin{table}[h] 
-\caption{Variables used in the study} 
-\label{tabla1} 
-\centering
-\begin{tabular}{c|c} 
-\hline
-\bfseries Variable & \bfseries Meaning0'0' \\ 
-\hline\hline 
-$T_{a_{max}}$ &	Max air temperature \\
-$T_{a_{min}}$ &	Min air temperature \\
-$\overline{T}_{a}$	 & Mean air temperature \\
-$\overline{H}$	&  Mean Humidity \\
-$H_{min}$ &	Min humidity \\
-$H_{max}$	& Max humidity \\
-$\overline{R}$	& Mean Solar radiation \\
-$P$	& Sum precipitation  \\
-$W_{max}$	& Max speed wind \\
-$\overline{W}$	& Mean speed wind  \\
-$L_2$	& Biological warning system – Leaf 2 \\
-$L_3$	& Biological warning system – Leaf 3 \\
-$L_4$	& Biological warning system – Leaf 4 \\
-$E_s$	& Biological warning system – Evolution Stage \\
-\hline
-\end{tabular} 
-\end{table}
-
-The value to be predicted in all cases was $E_s$, that is the total measure of the biological warning system.
-
-The data on the biological warning system are collected once a week.
-Although Corbana has meteorological stations that take data every five minutes, for these experiments, weekly averages generated by nearby stations to each of the farms were used.
-
-The time intervals used for this study were: La Rita, week 48 of 2002 to week 17 of the 2015 (647 weeks) and for 28 Miles, week 37 of 2003 to week 18 of 2015 (605 weeks).
-
-\subsection{Data preprocessing}
-Usually, data need preprocessing to prepare for use. In this section we explain that process.
-
-In 28 Miles farm, 1\% of the data were missing, while in La Rita was 2.25\%. To fill-in the missing values we use spline interpolation. The data collected did not exhibit outliers.
-
-Due the fact that the variables measure meteorological or biological process, they are discretized in order to reflect trends in the data, i.e.
-the continuous values are not directly used.
-The coefficient of variation $C_v(x)$ of each variable x was used to determine the number $n$ of discretization levels.
-$$n= \lfloor 100 \ C_v(x) \rfloor$$
-where $\lfloor \ \rfloor$  is the round operator.
-
-Each discretization range was uniformly partitioned.
-Besides enabling the capture of tendencies, the discretization removes the effect of small variations in the data collection, either by inaccuracies of the instruments (meteorological variables) or by subjective bias introduced by the human who collects the data (biological warning system).
-
-Each feature was scaled to fit in a range between 0 and 1. The variable to be predicted was not scaled.
-
-\subsection{Evaluation criteria}
-
-Although there are many types of indicators to assess the quality of the prediction, we selected the determination coefficient $(R^2)$ and the Root Mean Square Error $(RMSE)$.
-
-Given $n$ records, let be $y$ the actual value of the series, $\hat{y}$ the predicted value and $\acute{y}$ the mean of the observed data.
-
-$$ \bar{y} = \frac{1}{n} \sum_{i=1}^{n} y_i $$
-
-$$ S_e^2 = \frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}$$
-
-$$ S_R^2 = \frac{\sum_{i=1}^{n} {(\hat{y}_i-\bar{y}_i)}^2 }{n}$$
-
-$$ S_y^2 = S_R^2 + S_e^2$$
-
-$$ R^2 = \frac{S_R^2}{S_y^2}$$
-
-$$ RMSE = \sqrt{\frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}}$$
-
-This decision is supported by the widespread use in machine learning and agriculture areas \citep{Soares2014}, \citep{Soares2013}, \citep{Ibrahim2014} and \citep{Demir2014}.
-
-\subsection{Programming environment}
-
-We use the python programming language with the Integrated Development Environment (IDE) Spyder \citep{Continuum2015}, particularly with the libraries pandas \citep{mckinneypandas2010} and numpy \citep{vanderWalt2011}.
-For SVR, ridge and ordinary least squares regressions, we used sklearn \citep{scikitlearn2011} and for ESN the python-based code of Dr.
-$Luko\breve{s} evi \breve{c} ius$ \citep{Lukose2012} on which the necessary were done for adjustments for the experiments of this work.
-The computer used a processor Intel(R) Core i7-4800MQ CPU 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
-
-\subsection{Methodology}
-The selection of the technique and its parametrisation was performed in two stages that they are presented below.
-
-{\bf Phase one }
-
-In the phase one, we did ten-fold-cross-validation on a set of  machine learning methods and different configurations:
-
-\begin{itemize}
-\item 	Patterns: n by m, where n from 1 to 8 and m from 1 to 2.
-
-\item Methods: support vector regression with the kernels functions: linear, gaussian and sigmoid; echo state networks; ordinary least squares linear regression, ridge regression and elastic-net regression.
-
-\item Variables included in the model:
-\begin{itemize}
-\item All variables.
-\item From the set $\{ \overline{T}_{a} , \overline{H}, P , \overline{W}  \}$ use the subsets with one, two or four elements.
-These variables are according to experts the ones having most impact on the disease development \citep{MarinVargas1995}.
-\end{itemize}
-
-\end{itemize}
-
-{\bf Phase two }
-
-In the second phase, the best configurations obtained in phase one are used to validate with the last 50 and 100 weeks.
-
 \section{Results and discussion}
@@
 Again, similar to phase one, we can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week, one week to predict one week ahead with combinations of variables listed above.
-
-\section{Conclusions}
-\label{sec:concl}
-
-\subsection{Future work}
-
-\section{Acknowledgements}
-
-The authors would like to thank Corbana S.A. for providing the data for this research.
-
-\section{References}
-
-\bibliography{references}
-
-\end{document}

### 2016-06-11 techniques.tex
Nueva organizacion de los archivos.  Partes del paper en archivos separados y todas las imagenes en subdirectorio fig

--- 2016-05-22 Paper Sigatoka - 2015.tex
@@
-\documentclass[review,authoryear,english]{elsarticle}
-
-\usepackage{babel}
-\usepackage{natbib}
-
-\bibliographystyle{elsarticle-harv}
-
-\usepackage{lineno}
-\usepackage[utf8]{inputenc}  %% Permite utilizar tildes si archivo usa utf-8
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\modulolinenumbers[1]
-
-\usepackage{soul}
-\usepackage{marginnote}
-\usepackage{float}
-
-\usepackage{color}
-\definecolor{dkred}{rgb}{0.5,0,0}       %   dark red
-\definecolor{todocolor}{rgb}{1.0,0.8,0.5}
-
-\usepackage[hidelinks]{hyperref}
-
-\newcommand{\todohdr}{Please review:}
-\newcommand{\TODO}[2][\todohdr]{%
-\smallbreak\noindent\colorbox{todocolor}{%
-\begin{minipage}{1\linewidth}
-\textbf{#1}\par{\footnotesize{#2}}
-\end{minipage}}}
-
-\newcommand{\instead}[2]{\textcolor{dkred}{\st{#1}}#2}
-\newcommand{\figref}[1]{\figurename~\ref{#1}}
-
-%  \usepackage{epstopdf} 
-%  \usepackage{epsfig}
-
-\journal{Computers and Electronics in Agriculture}
-
-\usepackage{subcaption}
-
-\begin{document}
-
-\begin{frontmatter}
-
-\title{Forecasting the black Sigatoka development rate:\\ 
-A comparison of machine learning techniques}
-
-%% Group authors per affiliation:
-\author[afiLuisAlex]{Luis-Alexander Calvo-Valverde\fnref{myfootnote}}
-\ead{lualcava.sa@gmail.com}
-\fntext[myfootnote]{Corresponding author. (506)70104420}
-
-\author[afiCorbana]{Mauricio Guzmán-Quesada}
-\author[afiCorbana]{José-Antonio~Guzmán-Alvarez}
-\author[afiPablo]{Pablo Alvarado-Moya}
-
-\address[afiLuisAlex]{DOCINADE, Instituto Tecnológico de Costa Rica, 
-Computer Research Center, Multidisciplinar program eScience, Cartago, Costa Rica}
-
-\address[afiCorbana]{Dirección de Investigaciones, Corporación Bananera Nacional S.A., Guápiles, Costa Rica}
-
-\address[afiPablo]{DOCINADE, Instituto Tecnológico de Costa Rica, Cartago, Costa Rica}
-
-\begin{abstract}
-Pending.
-\end{abstract}
-
-\begin{keyword}
-Machine learning \sep Black Sigatoka \sep Support vector regression \sep
-Banana disease prediction \sep Biological warning system 
-\end{keyword}
-
-\end{frontmatter}
-
-\linenumbers
-
-\section{Introduction}
-
-The black Sigatoka disease caused by the fungus \emph{Mycosphaerella
-  fijiensis Morelet} is the major pathological problem of banana and
-plantain crops in Central America, Panama, Colombia and Ecuador, as well as in
-many parts of Africa and Asia \citep{MarinVargas1995}.
-
-This disease attacks the plant leaves producing a rapid deterioration
-of the leaf area, affects the growth and productivity of the plants
-due to the impairment of the photosynthetic process, causes a
-reduction in the quality of the fruit, and promotes premature
-ripening of bunches, which is
-the major cause of product losses associated with the black
-Sigatoka. \figref{fig:diseasestages} shows three stages of
-this disease.
-
-\begin{figure}[ht] 
-\centering
-\begin{tabular}{c@{\;}c@{\;}c}
-  \includegraphics[width=.32\linewidth]{Roya_a} &
-  \includegraphics[width=.32\linewidth]{Roya_b} &
-  \includegraphics[width=.32\linewidth]{Roya_c} \\
-  (a) & (b) & (c) 
-\end{tabular}
-\caption{Examples of three disease stages of the black Sigatoka. 
-(a)~Initial stage. (b)~Intermediate stage, and (c)~Advanced stage.} 
-\label{fig:diseasestages} 
-\end{figure}
-
-Phytopathological studies point out that precipitation, temperature,
-relative humidity and wind are the main climatic variables that affect
-the development of this disease \citep{MarinVargas1995}.
-
-The climate has a major effect on the development of the black
-Sigatoka.  Precipitation, temperature, relative humidity and wind are
-the main climatic variables affecting the development of this disease
-\citep{MarinVargas1995}.
-
-The early warning system developed by Ganry and Meyer
-\citep{ganry1972} and modified by Ganry and Laville \citep{ganry1983}
-to control yellow Sigatoka in Cameroon, was later adapted by
-Ternesien \citep{Ternesien1985} and Fouré \citep{foure1988} for the
-black Sigatoka.
-
-This bilogical warning system is based on weekly quantified
-observations of the disease progression on young leaves of the plant,
-according to Fouré's scale of the symptom stages.
-%
-Numeric coefficients are used to describe the degree of incidence and
-the severity of the disease development.  These coefficients are used
-to calculate two variables: gross sum and state of evolution.
-
-The gross sum is based on the present disease progression stage and
-the numeric coefficients, which increase with the progression of
-the symptoms and the juvenility of the leaf.
-%
-The state of evolution is calculated using the gross sum and the
-foliar emission period.
-%
-Although threshold levels were initially suggested as a guide to spray
-schedules, the fluctuation of these two variables
-was found to better define appropiate times to spray
-\citep{Marinetal2003}.
-
-\TODO{No queda claro threshold levels sobre qué variables!?  Tampoco
-  queda claro que quiere decir 'fluctuation' como medida de decisión }
-
-In Costa Rica the black Sigatoka is frequently treated with
-chemical fungicides.
-%
-Depending on the zone of production and the weather conditions,
-45--55~cycles/year of fungicide applications are required to keep this
-disease under control and to produce the expected fruit quality for the
-international markets.
-%
-This represents a cost per hectare per year in the range between US\$1600 and
-US\$2000; about 0.64--0.80 cents of the production costs for a
-18.14\,kg box.
-%
-Overall, this represents 10\%--12\% of the total production costs.
-
-The past and present rates of disease development can in principle be used
-to predict its future behavior and to determine whether
-a particular fungicide spray program will be able to effectively and
-economically control the disease \citep{ChuangJeger1987} or not.
-
-There are efforts to apply machine learning methods to support
-decision-making in agriculture, including the control of crop
-diseases. For example, \cite{Camargo2012} present an intelligent
-system for the assessment of crop disorders, \cite{Huang2010}
-introduce a plant virus identification method based on neural networks
-with an evolutionary preprocessing stage, \cite{Kim2014} summarize in
-their survey crop pests prediction methods using regression and
-machine learning approaches, while \cite{Zhao2013} present an
-intelligent agricultural forecasting system based on wireless sensor
-networks.
-
-In this work, we compare five machine learning techniques to predict
-the development rate of the black Sigatoka disease: support vector
-regression (SVR), echo state networks (ESN), ridge regression,
-elastic-net regression and ordinary least squares linear regression.
-%
-The main contribution of this work is the selection of the best
-machine learning technique to forecast the black Sigatoka development
-rate according to the metric indicated below.
-
-\TODO{Definitivamente ese no es el aporte principal del artículo.
-  Cuando esté listo el resto hay que volver aquí.  Creo que va a ser
-  algo como mostrar la relevancia de la etapa de preprocesamiento, o
-  mostrar que a pesar de ser sistemas complejos, con variables
-  caóticas, lo mejor en este caso son predictores lineales, o algo por
-  el estilo.}
-
-The outline of the paper is as follows: Section~\ref{sec:techs}
-summarizes the machine learning techniques used in this research and
-presents the main related works. In Section~\ref{sec:data} we present
-the methodology used in this study and describe data used for its
-verification.  The results and their discussion are presented in
-section~\ref{sec:results}.  The Section~\ref{sec:concl} concludes this
-article and presents lines for future works.
-
 \section{Machine learning techniques and Related works}
@@
 where $w$ and $b$ are respectively the weight vector and the intercept
-of the model, and they are selected to find an optimal fit to the data
-available in $D$ in the sense of ...?
-\TODO{optimal respecto a qué?.  Hay muchísimas posibilidades de optimalidad!}
+of the model, and they are selected to find an \todofn{optimal}%
+{optimal respecto a qué?.  Hay muchísimas posibilidades de optimalidad!}
+%
+fit to the data available in $D$ \review{in the sense of ...?}
 
@@
 The weights between the reservoir and the output layer $(W_out)$ are subject to changes of a supervised learning algorithm to correct the degree of error generated by the entire system \citep{Lukose2012}.
-
-\subsection{Related works}
-Now, we present the main and recently works related with our study.
-
-Huang et al.
-\citep{Huang2010}  surveyed the development of soft computing techniques in agricultural and biological engineering, including fuzzy logic, artificial neural networks, genetic algorithms, bayesian inference and decision trees.
-
-A related work, proposed by Romero \citep{Romero1995} relies on regression models using a stepwise procedure to predict incubation and latency times of black Sigatoka.
-The author performed experiments on two farms located in Costa Rica (La Rita and Waldeck, the same as those used in this study but with different names).
-The study used data from: December 1993 to August 1995.
-Romero concluded that the model to predict the incubation period accounted a $R^2$ of 69\% in his observed data but it was not a good predictor when it was validated against an independent dataset (cross validation).
-For latency, he developed two models that accounted a $R^2$ of 78\% in the observed data, however, when validated against an independent dataset (cross validation), for La Rita obtained an adjusted $R^2$ of 82\%, and for Weldeck, none of the models satisfactorily predicted the latent period; and then those predictions were not shown \citep{Romero1995}.
-
-Glezakos et al.
-\citep{Glezakos2010} proposed to use Genetic Algorithms (GA) and Neural Networks (NN) to identify plant virus (Tobacco Rattle Virus (TRV) and the Cucumber Green Mottle Mosaic Virus (CGMMV)).
-This is achieved by the development of ana- lytical tools of evolutionary adaptive width, propelled by Genetic Algorithms (GAs) and Neural Networks (NNs).
-The method was tested against some of the most commonly used classifiers in machine learning (Bayes, Trees and k-NN) via cross-validation and proved its potential towards the identification.
-
-In the agricultural context, Alves et al.
-\citep{Alves2011} used  geoinformation techniques to develop predictive models to study the areas of risk to soybean rust in soybean, coffee leaf rust in coffee, and black Sigatoka in banana, considering Brazil’s climatic characterization and the distribution of soybean, coffee and banana crops.
-Temperature and rainfall data were obtained for the period from 1950 to 2000, and of simulations for 2020, 2050 and 2080 using the SRES A2 climate change scenarios.
-Using principal components analysis, a single variable was generated based on 57 variables, in order to determine an index explaining 87\%, 88\% and 90\% of the variability of soybean, coffee and banana crops, respectively, in municipal districts across Brazil.
-The climatic model was used to generate the zoning of the three plant diseases, using temperature and leaf wetness as input.
-Areas of favorability for the diseases were plotted against the main coffee, soybean and banana growing areas in Brazil.
-This methodology enabled the visualization of the changes in areas favorable for epidemics under possible future scenarios of climate change.
-
-Other applications of machine learning methods in precision agriculture include the use of support vector regression to predict carcass weight in beef cattle in advance to the slaughter  \citep{Alonso2013}, machine learning assessments of soil drying for agricultural planning \citep{Coopersmith2014}, and early detection and classification of plant diseases with support vector machines based on hyperspectral reflectance \citep{Rumpf2010}.
-
-Furthermore, there have been attempts to generate software tools.
-Camargo et al.
-\citep{Camargo2012} presented an information system for the assessment of plant disorders (Isacrodi).
-They proposed that experts will attain a much better accuracy than the Isacrodi classifier, particularly when provided with samples from the affected crop.
-However, those cases where such expertise is not available, they suggest that Isacrodi can provide valuable support to farmers.
-Isacordi includes 15 crop disorders, but the black Sigatoka no is one of them.
-The prediction process is based on multi-class support vector machines.
-
-Regarding the prediction of the develpment of the black Sigatoka with machine learning methods, Bendini et al.
-\citep{Bendini2013}  presented a study about the risk analysis of black Sigatoka occurrence based on polynomial models.
-A case study was developed in a commercial banana plantation located in Jacupiranga, Brazil.
-It was monitored weekly during the period from February to December 2005.
-Data included the weekly monitoring of the disease’s evolution stage, time series of meteorological data and remote sensing data.
-They obtained a model to estimate the evolution of the disease from satellite imagery.
-This model relates gray levels (NC) of the band 2 images of the Landsat-5 satellite, with the progress status or disease severity (EE).
-The authors claim to reach an $R^2$ of 90\%.
-
-Also there are works related to banana fruit.
-Soares et al.
-\citep{Soares2014} apply two techniques: artificial neural networks (ANNs) and multiple linear regression (MLR) in banana plant to predict the yield, their results show that the neural network proved to be more accurate in forecasting the weight of the bunch in comparison to the multiple linear regressions in terms of the mean prediction-error $(MPE = 1.40)$, mean square deviation $(MSD = 2.29)$ and coefficient of determination $(R^2 = 91\%)$.
-
-In general, the machine learning methods applied to predict the evolution of plant diseases, can be classified in two main approaches: 1) Those whose main inputs are images, and 2) Those whose main inputs are environmental and biological variables.
-Our study focuses in the second case.
-
-\section{Specification of data and methodology}
-\label{sec:data}
-
-In this section, we present the data and how we worked to obtain our results.
-
-\subsection{Data}
-We use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (previously called Waldeck and located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
-The banana type is Musa AAA, subgroup Cavendish, cv.
-Grande Naine.
-The \tablename $.$\ref{tabla1} shows the variables available.
-
-\begin{table}[h] 
-\caption{Variables used in the study} 
-\label{tabla1} 
-\centering
-\begin{tabular}{c|c} 
-\hline
-\bfseries Variable & \bfseries Meaning0'0' \\ 
-\hline\hline 
-$T_{a_{max}}$ &	Max air temperature \\
-$T_{a_{min}}$ &	Min air temperature \\
-$\overline{T}_{a}$	 & Mean air temperature \\
-$\overline{H}$	&  Mean Humidity \\
-$H_{min}$ &	Min humidity \\
-$H_{max}$	& Max humidity \\
-$\overline{R}$	& Mean Solar radiation \\
-$P$	& Sum precipitation  \\
-$W_{max}$	& Max speed wind \\
-$\overline{W}$	& Mean speed wind  \\
-$L_2$	& Biological warning system – Leaf 2 \\
-$L_3$	& Biological warning system – Leaf 3 \\
-$L_4$	& Biological warning system – Leaf 4 \\
-$E_s$	& Biological warning system – Evolution Stage \\
-\hline
-\end{tabular} 
-\end{table}
-
-The value to be predicted in all cases was $E_s$, that is the total measure of the biological warning system.
-
-The data on the biological warning system are collected once a week.
-Although Corbana has meteorological stations that take data every five minutes, for these experiments, weekly averages generated by nearby stations to each of the farms were used.
-
-The time intervals used for this study were: La Rita, week 48 of 2002 to week 17 of the 2015 (647 weeks) and for 28 Miles, week 37 of 2003 to week 18 of 2015 (605 weeks).
-
-\subsection{Data preprocessing}
-Usually, data need preprocessing to prepare for use. In this section we explain that process.
-
-In 28 Miles farm, 1\% of the data were missing, while in La Rita was 2.25\%. To fill-in the missing values we use spline interpolation. The data collected did not exhibit outliers.
-
-Due the fact that the variables measure meteorological or biological process, they are discretized in order to reflect trends in the data, i.e.
-the continuous values are not directly used.
-The coefficient of variation $C_v(x)$ of each variable x was used to determine the number $n$ of discretization levels.
-$$n= \lfloor 100 \ C_v(x) \rfloor$$
-where $\lfloor \ \rfloor$  is the round operator.
-
-Each discretization range was uniformly partitioned.
-Besides enabling the capture of tendencies, the discretization removes the effect of small variations in the data collection, either by inaccuracies of the instruments (meteorological variables) or by subjective bias introduced by the human who collects the data (biological warning system).
-
-Each feature was scaled to fit in a range between 0 and 1. The variable to be predicted was not scaled.
-
-\subsection{Evaluation criteria}
-
-Although there are many types of indicators to assess the quality of the prediction, we selected the determination coefficient $(R^2)$ and the Root Mean Square Error $(RMSE)$.
-
-Given $n$ records, let be $y$ the actual value of the series, $\hat{y}$ the predicted value and $\acute{y}$ the mean of the observed data.
-
-$$ \bar{y} = \frac{1}{n} \sum_{i=1}^{n} y_i $$
-
-$$ S_e^2 = \frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}$$
-
-$$ S_R^2 = \frac{\sum_{i=1}^{n} {(\hat{y}_i-\bar{y}_i)}^2 }{n}$$
-
-$$ S_y^2 = S_R^2 + S_e^2$$
-
-$$ R^2 = \frac{S_R^2}{S_y^2}$$
-
-$$ RMSE = \sqrt{\frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}}$$
-
-This decision is supported by the widespread use in machine learning and agriculture areas \citep{Soares2014}, \citep{Soares2013}, \citep{Ibrahim2014} and \citep{Demir2014}.
-
-\subsection{Programming environment}
-
-We use the python programming language with the Integrated Development Environment (IDE) Spyder \citep{Continuum2015}, particularly with the libraries pandas \citep{mckinneypandas2010} and numpy \citep{vanderWalt2011}.
-For SVR, ridge and ordinary least squares regressions, we used sklearn \citep{scikitlearn2011} and for ESN the python-based code of Dr.
-$Luko\breve{s} evi \breve{c} ius$ \citep{Lukose2012} on which the necessary were done for adjustments for the experiments of this work.
-The computer used a processor Intel(R) Core i7-4800MQ CPU 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
-
-\subsection{Methodology}
-The selection of the technique and its parametrisation was performed in two stages that they are presented below.
-
-{\bf Phase one }
-
-In the phase one, we did ten-fold-cross-validation on a set of  machine learning methods and different configurations:
-
-\begin{itemize}
-\item 	Patterns: n by m, where n from 1 to 8 and m from 1 to 2.
-
-\item Methods: support vector regression with the kernels functions: linear, gaussian and sigmoid; echo state networks; ordinary least squares linear regression, ridge regression and elastic-net regression.
-
-\item Variables included in the model:
-\begin{itemize}
-\item All variables.
-\item From the set $\{ \overline{T}_{a} , \overline{H}, P , \overline{W}  \}$ use the subsets with one, two or four elements.
-These variables are according to experts the ones having most impact on the disease development \citep{MarinVargas1995}.
-\end{itemize}
-
-\end{itemize}
-
-{\bf Phase two }
-
-In the second phase, the best configurations obtained in phase one are used to validate with the last 50 and 100 weeks.
-
-\section{Results and discussion}
-\label{sec:results}
-
-In this section we present the main results for phase and their discussion.
-
-{\bf Phase one }
-
-\figref{figura4} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Though La Rita obtains different results in magnitude than 28 Millas, the trend is similar.
-In both farms, the best results are for linear models, second position is occupied for Echo State Networks and  SVR with gausian and sigmoid kernels are the worst results.
-In linear models, to predict one week ahead is better than two weeks ahead, and this is better than three weeks ahead.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_algortihm}
- \caption{Phase one - Best $R^2$ for each algorithm} 
- \label{figura4} 
-\end{figure}
-
-\figref{figura5} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-In general, to predict one week ahead is better than two weeks ahead and so on.
-The number of weeks consider in the observation range in the pattern is not the main discriminant factor, but it is clear that we get better $R^2$ for one week ahead than two weeks ahead and so on.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase one - Best $R^2$ for each observation range} 
- \label{figura5} 
-\end{figure}
-
-\figref{figura6} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-The better results are obtained with $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms of similarly.
-You can note that the use of all variables in the model or the inclusion of the four variables suggest for expert criteria do not improve significantly the results, then the use of more sensors do not assure a better result.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
- \caption{Phase one - Best $R^2$ for each variable combination} 
- \label{figura6} 
-\end{figure}
-
-\figref{figura7} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
-The Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better $RMSE$ than La Rita.
-This situation arise because $RMSE$ considers errors only with respect the prediction and in 28 Millas the average of Stage of Evolution is $4316.16$, unlike, in La Rita the average is $5507.30$.
-So, in La Rita we obtains higher errors in absolute values.
-$R2$ is a relative metric between 0 thru 1 and it is less sensitive to absolute values.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.8]{Phase_one_R2_RMSE}
- \caption{Phase one - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura7} 
-\end{figure}
-
-The Pareto frontier for the La Rita farm is composed by 96 elements. The \tablename $.$\ref{tabla2} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase one} 
-\label{tabla2} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $\overline{W}$ &	1 to 1  & 36 & 64.25\% & 714.51 \\
- &	2 to 1  & 6 & 62.97\% & 695.10 \\
-\hline 
-All  & 1 to 1  & 18 & 62.98\% & 701.95 \\
-   & 2 to 1  & 12 & 61.76\% & 679.92 \\
-    & 3 to 1  & 6 & 60.60\% & 676.42 \\
-    & 5 to 1  &  2 & 60.37\% & 672.39 \\
-\hline    
-$\overline{T}_{a}$ & 1 to 1  & 12  & 63.60\% & 708.77 \\
-       &	2 to 1  & 4 & 62.23\% & 689.55 \\
-\hline
-\end{tabular} 
-\end{table}
-
-Similarly, the Pareto frontier for the 28 Millas farm is composed by 75 elements. The \tablename $.$\ref{tabla3} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - 28 Millas - Phase one} 
-\label{tabla3} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $\overline{W}$ & 1 to 1 & 8 & 57.80\% & 438.09 \\
-\hline 
-All   &	9 to 1 & 2 & 50.93\% & 397.93 \\
-  & 10 to 1	 & 2 & 50.97\% & 398.81 \\
-  &	8 to 1 & 6 & 51.62\% & 398.93 \\
-  &	7 to 1 & 2 & 52.25\% & 400.28 \\
-  &	6 to 1 & 2 & 53.16\% & 404.14 \\
-  &	4 to 1 & 2 & 54.32\% & 407.54 \\
-\hline    
-$\overline{T}_{a}$ & 1 to 1  & 8  & 59.09\% & 439.44 \\
-\hline
-Pair $\overline{T}_{a}$ $\overline{H}$ & 1 to 1	 & 8 & 57.51\% & 428.61 \\
- &	2 to 1 & 20 & 56.91\% & 414.37 \\
- &	3 to 1 & 3 & 54.41\% & 411.55 \\
- &	4 to 1 & 3 & 53.34\% & 406.65 \\
-\hline
-Pair $\overline{T}_{a}$ $P$ & 3 to 1 & 9 & 56.23\% & 422.76 \\
-\hline
-\end{tabular} 
-\end{table}
-
-We can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week.
-
-{\bf Phase two }
-
-In the second phase, the best configurations obtained in phase one were used to validate with the last 50 and 100 weeks.
-
-\figref{figura8} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Even if linear models continue with good $R^2$, Echo State Networks improve their scores because in $50$ and $100$ last weeks validation, we are in presence of climate change, then the behaviour is less lineal.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_algortihm}
- \caption{Phase two - Best $R^2$ for each algorithm} 
- \label{figura8} 
-\end{figure}
-
-\figref{figura9} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-The results confirms that, in general, to predict one week ahead is better than two weeks ahead, two than three and so on, this for both farms.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase two - Best $R^2$ for each observation range} 
- \label{figura9} 
-\end{figure}
-
-\figref{figura10} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-This results confirm that $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms are the best variables combinations.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
- \caption{Phase two - Best $R^2$ for each variable combination} 
- \label{figura10} 
-\end{figure}
-
-\figref{figura11} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.  You can note that the behaviour of $R^2$ and $RMSE$ is similar to the phase one.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.8]{Phase_two_R2_RMSE}
- \caption{Phase two - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura11} 
-\end{figure}
-
-In phase two, the Pareto frontier for the La Rita farm is composed by 2 elements. The \tablename $.$\ref{tabla4} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase two} 
-\label{tabla4} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-All & 1 to 1  & 2 & 69.33\% & 353.33 \\
-\hline
-\end{tabular} 
-\end{table}
-
-In 28 Millas Farm, the Pareto frontier is composed by 1 element. The \tablename $.$\ref{tabla5} details the result.
-
-\begin{table}[h] 
-\caption{Pareto frontier - 28 Millas - Phase two} 
-\label{tabla5} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $P$ & 1 to 1 & 1 & 61.36\% & 339.89 \\
-\hline
-\end{tabular} 
-\end{table}
-
-Again, similar to phase one, we can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week, one week to predict one week ahead with combinations of variables listed above.
-
-\section{Conclusions}
-\label{sec:concl}
-
-\subsection{Future work}
-
-\section{Acknowledgements}
-
-The authors would like to thank Corbana S.A. for providing the data for this research.
-
-\section{References}
-
-\bibliography{references}
-
-\end{document}

### 2016-06-11 intro.tex
Cambios de redacción y reagrupamiento de ideas que estaban aisladas.  Quedan algunas cosas por completar para cerrar la idea

@@
 This disease attacks the plant leaves producing a rapid deterioration
-of the leaf area, affects the growth and productivity of the plants
-due to the impairment of the photosynthetic process, causes a
-reduction in the quality of the fruit, and promotes premature
-ripening of bunches, which is
-the major cause of product losses associated with the black
-Sigatoka. \figref{fig:diseasestages} shows three stages of
-this disease.
+of the leaf area. It affects the growth and productivity of the plants
+due to the impairment of the photosynthetic process.  Furthermore, it
+causes a reduction in the quality of the fruit, and promotes premature
+ripening of bunches, which is the major cause of product losses
+associated with the black Sigatoka.
+
+For these reasons, warning systems have been developed to detect the
+disease and monitor its progress.  For instance, the early warning
+system developed by \citet{ganry1983} and modified by
+\citet{ganry1972} for the control of the yellow Sigatoka in Cameroon,
+was later adapted by \citet{Ternesien1985} and \citet{foure1988} for
+the black Sigatoka.
+
+This biological warning system is based on weekly observations of the
+disease progression on young leaves of the plant.
+%
+\figref{fig:diseasestages} shows an example of three progressive
+stages of the black Sigatoka.
 
@@
 
-Phytopathological studies point out that precipitation, temperature,
-relative humidity and wind are the main climatic variables that affect
-the development of this disease \citep{MarinVargas1995}.
-
-The climate has a major effect on the development of the black
-Sigatoka.  Precipitation, temperature, relative humidity and wind are
-the main climatic variables affecting the development of this disease
-\citep{MarinVargas1995}.
-
-The early warning system developed by Ganry and Meyer
-\citep{ganry1972} and modified by Ganry and Laville \citep{ganry1983}
-to control yellow Sigatoka in Cameroon, was later adapted by
-Ternesien \citep{Ternesien1985} and Fouré \citep{foure1988} for the
-black Sigatoka.
-
-This bilogical warning system is based on weekly quantified
-observations of the disease progression on young leaves of the plant,
-according to Fouré's scale of the symptom stages.
-%
-Numeric coefficients are used to describe the degree of incidence and
-the severity of the disease development.  These coefficients are used
-to calculate two variables: gross sum and state of evolution.
+The disease progression is then quantified according to Fouré's scale
+of the symptom stages \citep{foure1988} by means of numeric
+coefficients that describe the degree of incidence and the severity of
+the disease development.  These coefficients are then used to
+calculate two variables: gross sum and state of evolution.
 
@@
 %
-Although \todofn{threshold levels}% 
-{No queda claro threshold levels sobre qué variables!?  Tampoco
-  queda claro que quiere decir 'fluctuation' como medida de decisión}
-%
-were initially suggested as a guide to spray
-schedules, the fluctuation of these two variables
-was found to better define appropiate times to spray
+Decades ago threshold levels on these variables were used as a guide
+to plan the spray schedules.  Nowadays the fluctuation of these two
+variables seems to better suggest appropriate times to spray
 \citep{Marinetal2003}.
@@
 %
-This represents a cost per hectare per year in the range between US\$1600 and
-US\$2000; about 0.64--0.80 cents of the production costs for a
-18.14\,kg box.
+This represents a cost per hectare per year in the range from US\$1600
+to US\$2000; about 0.64--0.80 cents of the production costs for a
+18.14\,kg box, which overall corresponds to 10\%--12\% of the total
+production costs.
+
+The past and present rates of disease development can in principle be
+used to predict its future behavior and to determine if a particular
+fungicide spray program will be able to effectively treat the disease
+in an economically affordable way \citep{ChuangJeger1987}.
 %
-Overall, this represents 10\%--12\% of the total production costs.
+Phytopathological studies point out that climate has a major effect on
+the development of the black Sigatoka, where the main variables
+affecting it are precipitation, temperature, relative humidity and
+wind \citep{MarinVargas1995}.  Hence, it can be expected that patterns
+in these variables correlate with the disease development.
 
-The past and present rates of disease development can in principle be used
-to predict its future behavior and to determine whether
-a particular fungicide spray program will be able to effectively and
-economically control the disease \citep{ChuangJeger1987} or not.
-
-There are efforts to apply machine learning methods to support
-decision-making in agriculture, including the control of crop
-diseases. For example, \cite{Camargo2012} present an intelligent
+There exist already several efforts to apply machine learning methods
+to support decision-making in agriculture, including the control of
+crop diseases. For example, \cite{Camargo2012} present an intelligent
 system for the assessment of crop disorders, \cite{Huang2010}
@@
 
+\TODO{Importantísimo es explicar aquí brevemente por qué esos otros
+  trabajos no se aplican al caso de la Sigatoka, o qué es lo que han
+  hecho mal, que este trabajo sí hace bien.}
+
 In this work, we compare five machine learning techniques to predict
@@
 regression (SVR), echo state networks (ESN), ridge regression,
-elastic-net regression and ordinary least squares linear regression.
-%
+elastic-net regression and ordinary least squares linear regression,
+using input variables such as \review{completar las variables de
+  entrada usada} to predict \review{completar las variables de salida}.
+
 \todofn{The main contribution of this work}%
@@
 
-The outline of the paper is as follows: Section~\ref{sec:techs}
-summarizes the machine learning techniques used in this research and
-presents the main related works. In Section~\ref{sec:data} we present
-the methodology used in this study and describe data used for its
-verification.  The results and their discussion are presented in
-section~\ref{sec:results}.  The Section~\ref{sec:concl} concludes this
-article and presents lines for future works.
+The outline of the paper is as follows: Section~\ref{sec:related}
+presents related works and Section~\ref{sec:techs} summarizes the
+machine learning techniques selected for the analysis. In
+Section~\ref{sec:data} we present the methodology used in this study
+and describe data used for its verification.  The results and their
+discussion are presented in section~\ref{sec:results}.  The
+Section~\ref{sec:concl} concludes this article and presents lines for
+future works.

### 2016-06-12 intro.tex
Revision de los trabajos relacionados, y reagrupamiento con ideas de la introduccion.  Redundancias eliminadas y alguna informacion de mas eliminada

@@
 affecting it are precipitation, temperature, relative humidity and
-wind \citep{MarinVargas1995}.  Hence, it can be expected that patterns
-in these variables correlate with the disease development.
-
-There exist already several efforts to apply machine learning methods
-to support decision-making in agriculture, including the control of
-crop diseases. For example, \cite{Camargo2012} present an intelligent
-system for the assessment of crop disorders, \cite{Huang2010}
-introduce a plant virus identification method based on neural networks
-with an evolutionary preprocessing stage, \cite{Kim2014} summarize in
-their survey crop pests prediction methods using regression and
-machine learning approaches, while \cite{Zhao2013} present an
-intelligent agricultural forecasting system based on wireless sensor
-networks.
-
-\TODO{Importantísimo es explicar aquí brevemente por qué esos otros
-  trabajos no se aplican al caso de la Sigatoka, o qué es lo que han
-  hecho mal, que este trabajo sí hace bien.}
+wind \citep{MarinVargas1995}.  It can be expected that patterns in
+these variables correlate with the disease development and hence its
+automated discovery can support decision-making in the control of crop
+diseases.
 

### 2016-06-12 methodology.tex
Revision de los trabajos relacionados, y reagrupamiento con ideas de la introduccion.  Redundancias eliminadas y alguna informacion de mas eliminada

@@
 \subsection{Data}
-We use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (previously called Waldeck and located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
-The banana type is Musa AAA, subgroup Cavendish, cv.
-Grande Naine.
-The \tablename $.$\ref{tabla1} shows the variables available.
+We use data acquired in two research farms of Corbana in Costa Rica:
+\emph{28 Millas} located in Matina and previously called
+\emph{Waldeck}, and \emph{La Rita} located in Pococí, both in the
+province of Limón, Costa Rica. The banana type is Musa AAA, subgroup
+Cavendish, cv. Grande Naine. The \reftab{tabla1} shows the
+variables available.
 

### 2016-06-12 related.tex
Cambios menores
Revision de los trabajos relacionados, y reagrupamiento con ideas de la introduccion.  Redundancias eliminadas y alguna informacion de mas eliminada

@@
 
-Huang et al.
-\citep{Huang2010}  surveyed the development of soft computing techniques in agricultural and biological engineering, including fuzzy logic, artificial neural networks, genetic algorithms, bayesian inference and decision trees.
+Several efforts have been made to apply machine learning techniques in
+the automated discovery of relationships between environmental
+variables and quantified descriptors for variables of agricultural 
+interest such as the progress of diseases.
+% 
+\citet{Huang2010} summarize in their survey the
+development of soft computing techniques in agricultural and
+biological engineering, including fuzzy logic, artificial neural
+networks, genetic algorithms, bayesian inference and decision trees.
+%
+Similarly, \citet{Kim2014} survey more recent prediction
+methods for crop pests using regression and machine learning approaches.
 
-A related work, proposed by Romero \citep{Romero1995} relies on regression models using a stepwise procedure to predict incubation and latency times of black Sigatoka.
-The author performed experiments on two farms located in Costa Rica (La Rita and Waldeck, the same as those used in this study but with different names).
-The study used data from: December 1993 to August 1995.
-Romero concluded that the model to predict the incubation period accounted a $R^2$ of 69\% in his observed data but it was not a good predictor when it was validated against an independent dataset (cross validation).
-For latency, he developed two models that accounted a $R^2$ of 78\% in the observed data, however, when validated against an independent dataset (cross validation), for La Rita obtained an adjusted $R^2$ of 82\%, and for Weldeck, none of the models satisfactorily predicted the latent period; and then those predictions were not shown \citep{Romero1995}.
+\TODO{Falta indicar qué concluyen los dos surveys anteriores}
 
-Glezakos et al.
-\citep{Glezakos2010} proposed to use Genetic Algorithms (GA) and Neural Networks (NN) to identify plant virus (Tobacco Rattle Virus (TRV) and the Cucumber Green Mottle Mosaic Virus (CGMMV)).
-This is achieved by the development of ana- lytical tools of evolutionary adaptive width, propelled by Genetic Algorithms (GAs) and Neural Networks (NNs).
-The method was tested against some of the most commonly used classifiers in machine learning (Bayes, Trees and k-NN) via cross-validation and proved its potential towards the identification.
+In general, the machine learning methods applied to predict the
+evolution of plant diseases, can be classified in two main approaches:
+1)~those whose main inputs are images, and 2)~Those whose main inputs
+are environmental and biological variables. Our study focuses in the
+second case.
 
-In the agricultural context, Alves et al.
-\citep{Alves2011} used  geoinformation techniques to develop predictive models to study the areas of risk to soybean rust in soybean, coffee leaf rust in coffee, and black Sigatoka in banana, considering Brazil’s climatic characterization and the distribution of soybean, coffee and banana crops.
-Temperature and rainfall data were obtained for the period from 1950 to 2000, and of simulations for 2020, 2050 and 2080 using the SRES A2 climate change scenarios.
-Using principal components analysis, a single variable was generated based on 57 variables, in order to determine an index explaining 87\%, 88\% and 90\% of the variability of soybean, coffee and banana crops, respectively, in municipal districts across Brazil.
-The climatic model was used to generate the zoning of the three plant diseases, using temperature and leaf wetness as input.
-Areas of favorability for the diseases were plotted against the main coffee, soybean and banana growing areas in Brazil.
-This methodology enabled the visualization of the changes in areas favorable for epidemics under possible future scenarios of climate change.
+% %
+% For example, 
+% 
+% Me parece que aquí hay un error: Huang2010 es el survey, y sería mucha
+% coincidencia que fuera lo mismo que el \citet{Glezakos2010} ¿o no?
+%
+% \cite{Huang2010} introduce a plant virus
+% identification method based on neural networks with an evolutionary
+% preprocessing stage, 
+%
 
-Other applications of machine learning methods in precision agriculture include the use of support vector regression to predict carcass weight in beef cattle in advance to the slaughter  \citep{Alonso2013}, machine learning assessments of soil drying for agricultural planning \citep{Coopersmith2014}, and early detection and classification of plant diseases with support vector machines based on hyperspectral reflectance \citep{Rumpf2010}.
+\citet{Romero1995} relied on regression models using a stepwise
+procedure to predict incubation and disease latency periods for the
+black Sigatoka.
+% 
+He collected environmental data from two different farms in Costa Rica
+between December 1993 and August 1995.
+%
+The prediction models reached coefficients of determination $R^2$ of
+69\% or 78\% on the observed data for the incubation and disease
+latency periods, respectively; however, the cross validation on
+independent data sets failed.
 
-Furthermore, there have been attempts to generate software tools.
-Camargo et al.
-\citep{Camargo2012} presented an information system for the assessment of plant disorders (Isacrodi).
-They proposed that experts will attain a much better accuracy than the Isacrodi classifier, particularly when provided with samples from the affected crop.
-However, those cases where such expertise is not available, they suggest that Isacrodi can provide valuable support to farmers.
-Isacordi includes 15 crop disorders, but the black Sigatoka no is one of them.
-The prediction process is based on multi-class support vector machines.
+More recently, \citet{Glezakos2010} used genetic algorithms (GA) and
+neural networks (NN) to identify the Tobacco Rattle Virus (TRV) and
+the Cucumber Green Mottle Mosaic Virus (CGMMV).
+%
+The method was tested against some of the most commonly used
+classifiers in machine learning (Bayes classifiers, decision trees and
+$k$-nearest neighbors) via cross-validation and proved their
+applicability in these kind of problems.
 
-Regarding the prediction of the develpment of the black Sigatoka with machine learning methods, Bendini et al.
-\citep{Bendini2013}  presented a study about the risk analysis of black Sigatoka occurrence based on polynomial models.
-A case study was developed in a commercial banana plantation located in Jacupiranga, Brazil.
-It was monitored weekly during the period from February to December 2005.
-Data included the weekly monitoring of the disease’s evolution stage, time series of meteorological data and remote sensing data.
-They obtained a model to estimate the evolution of the disease from satellite imagery.
-This model relates gray levels (NC) of the band 2 images of the Landsat-5 satellite, with the progress status or disease severity (EE).
-The authors claim to reach an $R^2$ of 90\%.
+\citet{Alves2011} used geoinformation techniques to
+develop predictive models in the study of risk areas to soybean rust,
+coffee leaf rust, and banana black Sigatoka, under consideration of
+Brazil’s climatic characteristics and the distribution of soybean,
+coffee and banana crops.
+%
+Temperature and rainfall data were acquired for the period from 1950
+to 2000, and simulated data were generated for 2020, 2050 and 2080
+using the SRES~A2 climate change scenarios.
+%
+Using principal components analysis, a single variable was generated
+as a linear combination of 57 input variables, in order to determine
+an index explaining 87\%, 88\% and 90\% of the
+%
+\todofn{variability}{variabilidad de qué?}  
+%
+of soybean, coffee and
+banana crops, respectively, in municipal districts across Brazil.
+%
+The climatic model was used to generate the zoning of the three plant
+diseases, using temperature and leaf wetness as input.
+%
+%Redunda con lo que sigue:
+%Areas favorable for the diseases were plotted against the main coffee,
+%soybean and banana growing areas in Brazil.
+%
+This methodology enabled the visualization of the changes in areas
+favorable for epidemics under possible future scenarios of climate
+change.
 
-Also there are works related to banana fruit.
-Soares et al.
-\citep{Soares2014} apply two techniques: artificial neural networks (ANNs) and multiple linear regression (MLR) in banana plant to predict the yield, their results show that the neural network proved to be more accurate in forecasting the weight of the bunch in comparison to the multiple linear regressions in terms of the mean prediction-error $(MPE = 1.40)$, mean square deviation $(MSD = 2.29)$ and coefficient of determination $(R^2 = 91\%)$.
+Other applications of machine learning methods in precision
+agriculture include the use of support vector regression to predict
+carcass weight in beef cattle in advance to the slaughter
+\citep{Alonso2013}, machine learning assessments of soil drying for
+agricultural planning \citep{Coopersmith2014}, and early detection and
+classification of plant diseases with support vector machines based on
+hyperspectral reflectance \citep{Rumpf2010}.
 
-In general, the machine learning methods applied to predict the evolution of plant diseases, can be classified in two main approaches: 1) Those whose main inputs are images, and 2) Those whose main inputs are environmental and biological variables.
-Our study focuses in the second case.
+Furthermore, there have been attempts to generate software
+tools. \citet{Camargo2012} presented an information
+system for the assessment of plant disorders (Isacrodi).
+%
+They showed that human experts will attain a much accurate assessment
+than the Isacrodi classifier, particularly when provided with samples
+from the affected crop. However, in those cases where such expertise
+is not available, the authors suggest that Isacrodi can still provide
+valuable support to farmers.
+%
+Isacordi includes 15 crop disorders, but the black Sigatoka is none
+of them. The prediction process is based on multi-class support vector
+machines.
+
+Regarding the prediction of the black Sigatoka disease development
+with machine learning methods, \citet{Bendini2013} presented a study
+on the risk analysis of its occurrence based on polynomial models.
+%
+A case study was developed in a commercial banana plantation located
+in Jacupiranga, Brazil. It was monitored weekly from February to
+December 2005.
+%
+The data included the weekly monitoring of the disease’s evolution
+stage, time series of meteorological data and remote sensing
+data.
+%
+They obtained a model to estimate the evolution of the disease from
+satellite imagery. This model relates gray levels (NC) of the band 2
+images of the Landsat-5 satellite, with the progress status or disease
+severity (EE). The authors claim to reach an $R^2$ of 90\%.
+
+There are also works related to the banana fruit. \citet{Soares2014}
+apply two techniques: artificial neural networks (ANNs) and multiple
+linear regression (MLR) in banana plant to predict the yield.
+%
+Their results show that the neural network is more accurate in
+forecasting the weight of the bunch in comparison to the multiple
+linear regressors in terms of the mean prediction-error $(MPE =
+1.40)$, mean square deviation $(MSD = 2.29)$ and coefficient of
+determination $(R^2 = 91\%)$.
+
+% Este no calza, porque no se ve qué es lo de machine learning que usa...
+%
+%\citet{Zhao2013} present an intelligent agricultural
+%forecasting system based on wireless sensor networks.
+
+\TODO{Importantísimo es explicar aquí brevemente por qué esos otros
+  trabajos no se aplican al caso de la Sigatoka, o qué es lo que han
+  hecho mal, que este trabajo sí hace bien.}

### 2016-06-18 techniques.tex
Editando las primeras 3 técnicas.  Unifiqué la notación pero hay que ver si se logra esa unificación en todo el paper

@@
-\section{Machine learning techniques and Related works}
+\section{Compared regression techniques}
 \label{sec:techs}
 
-In this section we summarize the machine learning techniques used and we present the main related works.
+In the prediction of the development rate of the black Sigatoka, we
+compare techniques such as least squares or ridge regression, commonly
+encountered in the agricultural literature with machine learning
+methods such as support vector regression, elastic regression and echo
+state networks, where the parameter space of each technique is also
+taken into account.
+
+\subsection{Ordinary least squares regression}
+
+Given a data set 
+\begin{equation}
+  D=\left\{(\vct{x}_i,y_i) \mid i=1\ldots n\right\}
+\end{equation}
+composed of the $d$-dimensional%
+\footnote{Without loss of generality assume that the first component
+  of every vector $\vct{x}_i$ is always $1$.}
+%
+feature vectors $\vct{x}_i\in\setR^d$ and the corresponding responses
+$y_i$.
+%
+The ordinary least squares regression (OLSR) fits a linear model
+$\tilde{y}_i = f(\vct{x}_i) = \vct{w}^T\vct{x}_i$ such that the sum of
+squares of the residuals $(\tilde{y}_i-y_i)$ is minimized.
+%
+Let $\mat{X}$ be the $n\times{}d$ feature matrix containing the data
+samples in its rows and $\vct{y}$ contain all the responses $y_i$
+corresponding to each row, then the least squares regression finds
+\begin{equation*}
+  \hat{\vct{w}} =
+  \arg\min_{\vct{w}}\left\| \mat{X}\vct{w} - \vct{y}\right\|_2^2
+\end{equation*}
+which is solved by means of the pseudoinverse
+($\hat{\vct{w}}=\mat{X}^\dagger\vct{y}$) or equivalently by the singular
+value decomposition of $\mat{X}$ \citep{Press2007}.
+
+\TODO{Alex, no sé hasta donde hiciste en las pruebas ese ``arreglo''
+  de que los datos tengan todos una entrada igual a 1.  De no ser así,
+  el método no puede encontrar el offset, y sería como forzar que el
+  modelo tenga que pasar por cero...}
+
+\subsection{Ridge regression}
+
+In contrast to the OLSR, the ridge regression (RR) adds a term to
+penalize large weights \citep{scikitlearn2011}:
+\begin{equation*}
+\hat{\vct{w}} = 
+\arg\min_{\vct{w}} \left\| \mat{X}\vct{w} - \vct{y} \right\|_2^2 + 
+                 \alpha \left\|\vct{w}\right\|_2^2 
+\end{equation*}
+where the parameter $\alpha>0$ controls how strong is the shrinking of
+the estimates towards zero.  This shrinkage introduces some bias but
+helps to reduce the variance of the estimate.
+%
+The solution is given by $\hat{\vct{w}}=(\mat{X}^T\mat{X} +
+\lambda\mat{I})^{-1}\mat{X}^T \vct{y}$.
 
@@
 From the perspective of Support Vector Regression (SVR) the regression
-function $y = f(s)$ for a given dataset $D=\{(s_i,y_i)\}_{i=1}^n$ is
-represented as a linear function of the form \citep{Wei2013}:
-\begin{equation*}
-  f(s)=w^Ts+b
-\end{equation*}
-where $w$ and $b$ are respectively the weight vector and the intercept
-of the model, and they are selected to find an \todofn{optimal}%
-{optimal respecto a qué?.  Hay muchísimas posibilidades de optimalidad!}
+function is usually stated as 
+\begin{equation}
+\label{eq:svrfunc}
+  \tilde{y} = f(\vct{x}) = \vct{w}^T\vct{x} + b
+\end{equation}
+The weights are selected in a convex optimization problem
+\citep{Smola2004}:
+\begin{align*}
+  \text{minimize}   & \frac{1}{2}\|\vct{w}\|^2 + C\sum_{i=1}^n(\xi_i+\xi_i^\ast \\
+  \text{subject to} & 
+  \begin{cases}
+    y_i - \vct{w}^T\vct{x}_i - b &\leq \epsilon + \xi_i    \\
+    \vct{w}^T\vct{x}_i + b -y_i  &\leq \epsilon + \xi_i^\ast\\
+    \xi_i,\xi_i^\ast               &\geq 0
+  \end{cases}
+\end{align*}
+where $\epsilon$ is the maximal allowed deviation of the targets
+$\tilde{y}_i$ from $y_i$, the slack variables $\xi_i$ and $\xi_i^\ast$
+allow to cope with otherwise unfeasible constraints for the
+optimization problem, and the constant $C>0$ controls the trade-off
+between the flatness of $f$ and the tolerance to deviations larger
+than $\epsilon$.
+
+Note that since OLSR and RR use a squared error function, data
+outliers will have a strong influence on the resulting weights
+$\vct{w}$.  On the SVR formulation, however, the usage of the $L_1$
+norm and the slack variables considerably restrict or completely block
+the influence of those outliers.
+
+The SVR problem is reformulated by means of the dual optimization
+problem into
+\begin{equation}
+  \label{eq:svrw}
+  \vct{w}=\sum_{i=1}^n(\alpha_i-\alpha_i^\ast)\vct{x}_i
+\end{equation}
+where $\alpha_i,\alpha_i^\ast\in[0,C]$ are Lagrange multipliers
+subject to $\sum_{i=1}^n(\alpha_i-\alpha_i^\ast)=0$.
 %
-fit to the data available in $D$ \review{in the sense of ...?}
+In this so-called \emph{Support Vector expansion} the weights are
+expressed as a linear combination of the data set patterns
+$\vct{x}_i$.
+%
+Inserting (\ref{eq:svrw}) in (\ref{eq:svrfunc}) leads to
+\begin{equation}
+  \label{eq:svrf}
+ f(\vct{x}) = \sum_{i=1}^n(\alpha_i-\alpha_i^\ast)\vct{x}_i^T\vct{x} + b
+\end{equation}
+The Lagrange multipliers $\alpha_i,\alpha_i^\ast$ are both non-zero
+only for those data points where $|f(\vct{x}_i) - y_i|\geq\epsilon$.
+Hence, the expansion of $\vct{w}$ in terms of $\vct{x}_i$ is sparse.
+Those data points with non-vanishing coefficients are called
+\emph{Support Vectors} \citep{Wei2013}.
 
-For nonlinear cases, the $p$-dimensional input vectors are mapped by a
-nonlinear function $\phi : R^p\rightarrow F$ onto a feature space
-$F$.  After the nonlinear mapping, the regression function evolves to a
-pervasive form:
-\begin{equation*}
-f(s)=w^T \phi (s)+b
-\end{equation*}
+Additionally, in (\ref{eq:svrf}) it is possible to employ the
+\emph{kernel trick} and replace the terms $\vct{x}_i^T\vct{x}$ with
+the evaluation of any Mercer kernel
+$k(\vct{x}_i,\vct{x})=\phi(\vct{x}_i)^T\phi(\vct{x})$, where
+$\phi(\vct{x})$ is a non-linear mapping of the input space onto a
+higher (even infinite) dimensional feature space.
+%
+The kernel evaluation draws unnecessary the explicit evaluation of the
+non-linear mapping.
+%
+This allows to solve non-linear regressions in the input space by
+implicitly mapping through the kernel the samples into the higher
+dimensional space, where the linear regression occurs \citep{Alonso2013}.
 
-In SVR the generalization ability is measured by the Euclidean norm of
-$w$, and the empirical error is measured by the $\epsilon$-insensitive
-loss function, given by \citep{Cristiani2005}
-
-\TODO{si $l$ es una función, es función de qué, de $s$? $l(s)$? o de
-  $y$? o de todos $l(y,s;\epsilon)$?}
-
-\begin{equation*}
-l= {\left| y - f(s) \right| }_{\epsilon } = 
-\begin{cases}
-   0 & {\left| y - f(s) \right| } \leq \epsilon \\
-   \left| y - f(s) \right| - \epsilon & \text{otherwise}
-\end{cases}
-\end{equation*}
-which ignores the error if the difference between the prediction value
-and the actual value is smaller than $\epsilon$.
-%
-The $\epsilon$-insensitive loss function allows to find the
-coefficients $w$ and $b$ by solving a convex optimization problem,
-which balances the empirical error and the generalization ability.
-%
-Then, the optimization problem to identify the regression model can be
-formulated as \citep{Wei2013}:
-%
-\begin{equation} 
-\begin{aligned}
-& \underset{}{\text{minimize}}
-& & J(w,\xi_i , \xi_i^* ) = \frac{1}{2}   \Bigr| \Bigr| w \Bigr| \Bigr|^2 + C \sum_{i=1}^{n} (\xi_i , \xi_i^* )    \\
-& \text{subject to}
-& & \begin{array}{lcl} 
-y_i - w^T \phi (s) - b  \leq \epsilon + \xi_i   \\
-w^T \phi (s) + b - y_i \leq \epsilon + \xi_i^* & i= 1,2,...,n \\
-\xi_i , \xi_i^* \geq 0  \\
-\end{array}
-\end{aligned}
-\end{equation}  
-where $C$ denotes the penalty parameter between empirical and
-generalization errors, and $\xi_i , \xi_i^*$ are slack
-variables. \figref{figura2} shows this situation.
-
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{SVR}
- \caption{$\epsilon$-insensitive loss function \citep{Wei2013}.} 
- \label{figura2} 
-\end{figure}
-
-The solution of this optimization problem by the Lagrange method is given by:
-\begin{equation*}
-f(s) = w^T \phi (s) + b = \sum_{i=1}^{n} (\alpha_i - \alpha_i^*) K (s,s_i) + b
-\end{equation*}
-where $\alpha_i - \alpha_i^*$ are the Lagrange multipliers of the
-optimization problem’s dual form and $K(s_i,s_j )$ is the kernel
-function satisfying the Mercer condition, and holds:
-\begin{equation*}
-K(s_i,s_j ) = \big \langle  \phi(s_i) , \phi(s_j)  \big \rangle
-\end{equation*}
-Operations in the kernel function $K(s,s_i )$ are performed in the
-input space rather than in the potentially high dimensional feature
-space of $\phi$ \citep{Alonso2013}.
-
-\subsection{Ordinary least squares regression}
-
-This method fits a linear model with coefficients $w = (w_1,..,w_p)$
-to minimize the residual sum of squares between the observed responses
-in the dataset, and the responses predicted by the linear
-approximation. Mathematically it solves a problem of the form
-\citep{scikitlearn2011}:
-$$\min_{w} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  $$
-where $X$ denotes the features matrix.
-
-According Pedregosa et al.
-\citep{scikitlearn2011}  the coefficient estimates for Ordinary Least Squares rely on the independence of the model terms.
-When terms are correlated and the columns of the design matrix $X$ have an approximate linear dependence, the design matrix becomes close to singular and as a result, the least-squares estimate becomes highly sensitive to random errors in the observed response, producing a large variance.
-This situation of multicollinearity can arise, for example, when data are collected without an experimental design.
-
-\subsection{Ridge regression}
-The ridge regression addresses some of the problems of ordinary least squares regression by imposing a penalty on the size of the coefficients.
-The ridge coefficients minimize a penalized residual sum of squares  \citep{scikitlearn2011}:
-
-$$\min_{w} { \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-Here, $\alpha \ \textgreater \  0$ is a complexity parameter that controls the amount of shrinkage: the larger the value of $\alpha$, the greater the amount of shrinkage and thus the coefficients become more robust to collinearity.
+\TODO{Habría que insertar aquí los kernels utilizados en los experimentos}
 
 \subsection{Elastic-Net regression}
+
 Elastic-Net is a linear regression model trained with $L1$ and $L2$ prior as regularizer.

### 2016-06-19 techniques.tex
Tecnicas reacomodadas y referencias originales citadas.  El orden cambió para que quede más claro como las técnicas se van armando agregando cosas

@@
 The ordinary least squares regression (OLSR) fits a linear model
-$\tilde{y}_i = f(\vct{x}_i) = \vct{w}^T\vct{x}_i$ such that the sum of
+$\tilde{y}_i = f(\vct{x}_i) = \dotp{\vct{w},\vct{x}_i}$ such that the sum of
 squares of the residuals $(\tilde{y}_i-y_i)$ is minimized.
 %
-Let $\mat{X}$ be the $n\times{}d$ feature matrix containing the data
-samples in its rows and $\vct{y}$ contain all the responses $y_i$
-corresponding to each row, then the least squares regression finds
+Let $\mat{X}$ be the $n\times{}d$ feature matrix containing the $i$-th
+data sample $\vct{x}_i^T$ in its $i$-th row and $\vct{y}$ contain all
+the responses $y_i$ corresponding to each row, then the least squares
+regression finds
+\begin{equation}
+\label{eq:problem}
+  \hat{\vct{w}} =
+  \arg\min_{\vct{w}} E(\vct{w})
+\end{equation}
+with the error function
 \begin{equation*}
-  \hat{\vct{w}} =
-  \arg\min_{\vct{w}}\left\| \mat{X}\vct{w} - \vct{y}\right\|_2^2
+  E(\vct{w}) =  \left\| \mat{X}\vct{w} - \vct{y}\right\|_2^2
 \end{equation*}
-which is solved by means of the pseudoinverse
-($\hat{\vct{w}}=\mat{X}^\dagger\vct{y}$) or equivalently by the singular
-value decomposition of $\mat{X}$ \citep{Press2007}.
+The solution is found by means of the pseudoinverse
+$\hat{\vct{w}}=\mat{X}^+\vct{y}=(\mat{X}^T\mat{X})^{-1}\mat{X}^T\vct{y}$
+or equivalently by the singular value decomposition of $\mat{X}$
+\citep{Press2007}.
 
@@
   el método no puede encontrar el offset, y sería como forzar que el
-  modelo tenga que pasar por cero...}
+  modelo tenga que pasar por cero...  a menos por supuesto que la
+  constante se haya sacado de la ecuación al optimizar...}
 
@@
 
-In contrast to the OLSR, the ridge regression (RR) adds a term to
-penalize large weights \citep{scikitlearn2011}:
+In contrast to the OLSR, for the ridge regression (RR)
+\citet{Hoerl1988} proposed to add a term to penalize large weights
+into the error function
 \begin{equation*}
-\hat{\vct{w}} = 
-\arg\min_{\vct{w}} \left\| \mat{X}\vct{w} - \vct{y} \right\|_2^2 + 
+  E(\vct{w}) = 
+  \left\| \mat{X}\vct{w} - \vct{y} \right\|_2^2 + 
   \alpha \left\|\vct{w}\right\|_2^2 
@@
 %
-The solution is given by $\hat{\vct{w}}=(\mat{X}^T\mat{X} +
+The solution of the optimization problem (\ref{eq:problem}) in this
+case is given by $\hat{\vct{w}}=(\mat{X}^T\mat{X} +
 \lambda\mat{I})^{-1}\mat{X}^T \vct{y}$.
+
+\subsection{Elastic net regression}
+
+Instead of $L_2$ regularization prior
+($\alpha\left\|\vct{w}\right\|_2^2$) included in the ridge regression,
+\citet{Tibshirani1996} used an $L_1$ term ($\lambda
+\left\|\vct{w}\right\|_1$) for his lasso estimator, which permits to
+select a subset of the available features by zeroing the weights of
+the deselected features.
+%
+If the dimension $d$ of the data is larger than the number $n$ of data
+samples, lasso will select a maximum of $d$ variables.
+
+The elastic net regression (ENR) of \citet{Zou2005} combines both
+$L_1$ and $L_2$ priors of the ridge and lasso estimators such that the
+error function is now
+\begin{equation*}
+  E(\vct{w}) = 
+  \left\| \mat{X}\vct{w} - \vct{y} \right\|_2^2 + 
+  \alpha \left\|\vct{w}\right\|_2^2 +
+  \lambda \left\|\vct{w}\right\|_1 
+\end{equation*}
+%
+This combination of priors still allows to learn a sparse model with only a 
+few weights being non-zero like in the case of lasso, but still
+maintaining the regularization properties of the ridge regression
+\citep{scikitlearn2011}.
+
+The elastic net is useful when multiple features are correlated: lasso
+will likely pick one of these at random, while the elastic net will
+still likely pick both.
+
+\TODO{Algo que no entiendo es que las elastic net son en realidad una
+  generalización del ridge, Lasso y OLSR.  Así que en los experimentos
+  con las selección adecuada de parámetros este método debería
+  comportarse al menos igual o mejor que esos otros regresores!  No sé
+  hasta donde sea justificable usar este Y los otros, porque este ES
+  los otros...}
+
+\TODO{Alex, veo que usas mucho \citep{scikitlearn2011} para los
+  métodos, pero ese es solo el paper de un tool y no los proponentes
+  originales de los métodos.  Usualmente uno hace referencia a algún
+  artículo, libro o tutorial donde ojalá los que propusieron el método
+  son los que lo explican.  Ahí metí entonces en las referencias otros
+  artículos por ese motivo.}
 
@@
 \TODO{Estoy seguro que los lectores van a solicitar reducir esto y
-  dejar solo la referencia.  No hay nada nuevo en el resto del paper
-  que requiera toda esta información, así que hay que resumirlo a nada
-  más lo esencial. }
+  dejar solo la referencia.  Dejé solo lo que considero relevante,
+  pero podés reducirlo más, si querés.}
 
 From the perspective of Support Vector Regression (SVR) the regression
-function is usually stated as 
+function is usually formulated as 
 \begin{equation}
 \label{eq:svrfunc}
-  \tilde{y} = f(\vct{x}) = \vct{w}^T\vct{x} + b
+  \tilde{y} = f(\vct{x}) = \dotp{\vct{w},\vct{x}} + b
 \end{equation}
@@
   \begin{cases}
-    y_i - \vct{w}^T\vct{x}_i - b &\leq \epsilon + \xi_i    \\
-    \vct{w}^T\vct{x}_i + b -y_i  &\leq \epsilon + \xi_i^\ast\\
+    y_i - \dotp{\vct{w},\vct{x}_i} - b &\leq \epsilon + \xi_i    \\
+    \dotp{\vct{w},\vct{x}_i} + b -y_i  &\leq \epsilon + \xi_i^\ast\\
     \xi_i,\xi_i^\ast               &\geq 0
@@
 where $\epsilon$ is the maximal allowed deviation of the targets
-$\tilde{y}_i$ from $y_i$, the slack variables $\xi_i$ and $\xi_i^\ast$
-allow to cope with otherwise unfeasible constraints for the
-optimization problem, and the constant $C>0$ controls the trade-off
-between the flatness of $f$ and the tolerance to deviations larger
-than $\epsilon$.
+$\tilde{y}_i$ from the responses $y_i$, the slack variables $\xi_i$
+and $\xi_i^\ast$ allow to cope with otherwise unfeasible constraints
+for the optimization problem, and the constant $C>0$ controls the
+trade-off between the flatness of $f$ and the tolerance to deviations
+larger than $\epsilon$.
 
-Note that since OLSR and RR use a squared error function, data
+Note that since OLSR, RR and ENR use a squared error function, data
 outliers will have a strong influence on the resulting weights
@@
 The SVR problem is reformulated by means of the dual optimization
-problem into
+problem into \citep{Smola2004}
 \begin{equation}
@@
   \label{eq:svrf}
- f(\vct{x}) = \sum_{i=1}^n(\alpha_i-\alpha_i^\ast)\vct{x}_i^T\vct{x} + b
+ f(\vct{x}) = \sum_{i=1}^n(\alpha_i-\alpha_i^\ast)\dotp{\vct{x}_i,\vct{x}} + b
 \end{equation}
@@
 Additionally, in (\ref{eq:svrf}) it is possible to employ the
-\emph{kernel trick} and replace the terms $\vct{x}_i^T\vct{x}$ with
+\emph{kernel trick} and replace the terms $\dotp{\vct{x}_i,\vct{x}}$ with
 the evaluation of any Mercer kernel
-$k(\vct{x}_i,\vct{x})=\phi(\vct{x}_i)^T\phi(\vct{x})$, where
+$k(\vct{x}_i,\vct{x})=\dotp{\phi(\vct{x}_i),\phi(\vct{x})}$, where
 $\phi(\vct{x})$ is a non-linear mapping of the input space onto a
@@
 The kernel evaluation draws unnecessary the explicit evaluation of the
-non-linear mapping.
-%
-This allows to solve non-linear regressions in the input space by
-implicitly mapping through the kernel the samples into the higher
-dimensional space, where the linear regression occurs \citep{Alonso2013}.
+non-linear mapping, and it allows to solve non-linear regressions in
+the input space by implicitly mapping the samples through the kernel
+into the higher dimensional space, where the linear regression occurs
+\citep{Alonso2013}.
 
@@
 
-\subsection{Elastic-Net regression}
+\subsection{Echo State Networks (ESN)}
 
-Elastic-Net is a linear regression model trained with $L1$ and $L2$ prior as regularizer.
-This combination allows for learning a sparse model where few of the weights are non-zero like Lasso, while still maintaining the regularization properties of Ridge  \citep{scikitlearn2011}.
-The convex combination of $L1$ and $L2$ is controled by using the $l1_{ratio}$ parameter.
+Recurrent neural networks (RNN) are capable of learning temporal
+patterns by feeding neuron outputs back into lower layers. Their
+training (usually by means of error backpropagation) is in general
+slow.
+%
+Echo state networks (ESN) are a particular type of recurrent neural
+network with a sparsely connected random hidden layer where only the
+weights of the output neurons are changed at training.  The randomly
+selected weights at the input and middle layers (called \emph{reservoir})
+reproduce temporal patterns (\emph{echoes}) that the output layer
+learns to select during the training \citep{Lukose2009}.
 
-Elastic-Net is useful when there are multiple features which are correlated with one another. Lasso is likely to pick one of these at random, while elastic-net is likely to pick both.
-A practical advantage of trading-off between Lasso and Ridge is it allows Elastic-Net to inherit some of Ridge’s stability under rotation.
-The objective function to minimize is \citep{scikitlearn2011}:
+For a given training input signal $u(n) \in \setR^{N_u}$ a desired target
+output signal $y^{target}(n) \in \setR^{N_y}$ is known.
+%
+Here $n = 1,\ldots,T$ is the discrete time and \todofn{$T$ is the number of
+data points in the training dataset.}{No puede ser!}
 
-$$\min_{w} { \frac{1}{2n_{samples}} \Bigr| \Bigr| Xw - y \Bigr| \Bigr|_2^2  + \alpha \rho \Bigr| \Bigr| w \Bigr| \Bigr|_1 + \frac{\alpha (1- \rho)}{2} \Bigr| \Bigr| w \Bigr| \Bigr|_2^2 } $$
-
-\subsection{Echo State Networks (ESN)}
-Recurrent Neural Networks (RNN) are useful for temporal patterns, but when they are trained with backpropagation methods, they are very slow.
-Echo State Network (ESN) is an alternative training method to solve that problem.
-ESN is based on the observation that if a random RNN possesses certain algebraic properties, training only a linear readout from it is often sufficient to achieve excellent performance in practical applications \citep{Lukose2009}.
-For a given training input signal $u(n)  \in R^{N_u}$ a desired target output signal $y^{target}(n) \in R^{N_y}$
-is known.
-Here $n = 1, .
-.
-.
-,T$ is the discrete time and $T$ is the number of data points in the training dataset.
-The task is to learn a model with output $y(n) \in R^{N_y}$, where $y(n)$ matches $y^target(n)$ as well as possible, minimizing an error measure $E(y,y^target)$, and, more importantly, generalizes well to unseen data.
-The untrained RNN part of an ESN is called a dynamical reservoir, and the resulting states x(n) are termed echoes of its input history \citep{Lukose2012}.
-Finally, these signals are sent to an output layer as shown in the \figref{figura3}.
-\begin{figure}[h] 
- \centering
- \includegraphics[scale=.9]{Reservorio}
- \caption{An echo state network \citep{Lukose2012}.} 
- \label{figura3} 
-\end{figure}
-
-The connections between the different elements of an Echo State Network have weights randomly generated.
-The weights of the internal connections of the reservoir $(W)$ as well as the weights of the input layer $(W_in)$, after being generated are set statically during all stages of implementation of the algorithm.
-The weights between the reservoir and the output layer $(W_out)$ are subject to changes of a supervised learning algorithm to correct the degree of error generated by the entire system \citep{Lukose2012}.
+The training seeks to learn a model with output $y(n) \in
+\setR^{N_y}$, where $y(n)$ matches $y^{target}(n)$ as close as
+possible, by means of the minimization of an error measure
+$E(y,y^{target})$ such that it alse generalizes well to unseen data
+\citep{Lukose2012}.

### 2016-07-02 methodology.tex
Metodología revisada.  Hay bastantes cambios que hacer para que quede formalmente consistente y completa

--- 2016-06-11 methodology.tex
@@
 
-In this section, we present the data and how we worked to obtain our results.
+Since the suitability of a machine learning technique to a particular
+problem is entirely depend on the nature of the data, we describe in
+this section, first, the data set employed in the study, followed by
+the methodology to compare the chosen techniques under consideration
+of their parameter space.
 
 \subsection{Data}
-We use data acquired in two research farms of Corbana in Costa Rica: 1) 28 Millas (previously called Waldeck and located at Matina) and La Rita (located at Pococí), both in the province of Limón, Costa Rica.
-The banana type is Musa AAA, subgroup Cavendish, cv.
-Grande Naine.
-The \tablename $.$\ref{tabla1} shows the variables available.
 
+The data used for the current study was acquired in two research farms
+of Corbana in Costa Rica%
+%
+\footnote{Both farms were also use in the study of \citet{Romero1995}.  Back
+  then, \emph{La Rita} was referred to as \emph{Waldeck}.}
+%
+: \emph{28 Millas} located in the region of Matina, and \emph{La Rita}
+located in Pococí, both in the province of Limón, Costa Rica.
+%
+Both farms produce banana fruit \emph{Musa} sp.\ AAA group 'Grande
+Naine' (Cavendish subgroup).
+
+The available input and output variables are summarized in
+\tabref{tab:variables}.
+%
 \begin{table}[h] 
-\caption{Variables used in the study} 
-\label{tabla1} 
 \centering
-\begin{tabular}{c|c} 
+\begin{tabular}{l|l|c} 
 \hline
-\bfseries Variable & \bfseries Meaning0'0' \\ 
+\textbf{Symbol}  & \textbf{Description} & \textbf{Units} \\ 
 \hline\hline 
-$T_{a_{max}}$ &	Max air temperature \\
-$T_{a_{min}}$ &	Min air temperature \\
-$\overline{T}_{a}$	 & Mean air temperature \\
-$\overline{H}$	&  Mean Humidity \\
-$H_{min}$ &	Min humidity \\
-$H_{max}$	& Max humidity \\
-$\overline{R}$	& Mean Solar radiation \\
-$P$	& Sum precipitation  \\
-$W_{max}$	& Max speed wind \\
-$\overline{W}$	& Mean speed wind  \\
-$L_2$	& Biological warning system – Leaf 2 \\
-$L_3$	& Biological warning system – Leaf 3 \\
-$L_4$	& Biological warning system – Leaf 4 \\
-$E_s$	& Biological warning system – Evolution Stage \\
+$T_{a_{max}}$       & Maximal air temperature & $[^\circ$C$]$ \\
+$T_{a_{min}}$       & Minimal air temperature & $[^\circ$C$]$ \\
+$\overline{T}_{a}$ & Mean air temperature    & $[^\circ$C$]$ \\
+$\overline{H}$    & Mean humidity           & ???  \\
+$H_{min}$          & Minimal humidity        & ???  \\
+$H_{max}$          & Maximal humidity        & ???  \\
+$\overline{R}$    & Mean solar radiation    & $[$W/m$^2$$]$ \\
+$P$               & Sum precipitation       & ??? \\
+$W_{max}$          & Maximal wind speed      & $[$km/h$]$ \\
+$\overline{W}$    & Mean speed wind         & $[$km/h$]$ \\
+$L_2$             & Biological warning system – Leaf 2 & ???\\
+$L_3$             & Biological warning system – Leaf 3 & ???\\
+$L_4$             & Biological warning system – Leaf 4 & ???\\
+\hline
+$E_s$             & Biological warning system – Evolution Stage  & ???\\
 \hline
 \end{tabular} 
+\caption{Variables available for the learning algorithms} 
+\label{tab:variables} 
 \end{table}
 
-The value to be predicted in all cases was $E_s$, that is the total measure of the biological warning system.
+The data was captured for La Rita between the 48\,th week of
+2002 to the 17\,th week of 2015 (647 weeks); for 28 Miles the
+data was captured between the 37\,th week of 2003 and the
+18\,th week of 2015 (605 weeks).
+%
+The data on the biological warning system were collected once a
+week.
 
-The data on the biological warning system are collected once a week.
-Although Corbana has meteorological stations that take data every five minutes, for these experiments, weekly averages generated by nearby stations to each of the farms were used.
+The meteorological stations of Corbana acquire data every five
+minutes.
+%
+\todofn{However, for the current study weekly averages were used,}
+{Hay que explicar por qué? $\rightarrow$ reducción de ruido, etc.}
+%
+computed on the data collected by nearby stations in each farm.
 
-The time intervals used for this study were: La Rita, week 48 of 2002 to week 17 of the 2015 (647 weeks) and for 28 Miles, week 37 of 2003 to week 18 of 2015 (605 weeks).
+\TODO{Alex, Falta explicar cada variable y sus
+  unidades y rangos usuales, lo que es relevante luego cuando se hable
+  de la normalización.
+
+  Humidity ¿cuál se usa? ¿la relativa? ¿la absoluta?, ¿la
+  específica?  Las otras unidades las supuse, así que hay que revisar
+  que estén bien!
+
+  No tengo idea qué es el ``Sum precipitation'' y en qué unidades
+  estaría.  Parece no ser algo estándar o medido directamente, ¿o sí?
+  así que amerita una explicación aquí.
+
+  Dijiste que las variables se promedian para una semana.  ¿Qué
+  significan entonces las variables que dicen ``mean''?
+}
+
+\TODO{The mean variables ... }
+
+The value to be predicted in all cases is the evolution stage $E_s$,
+which corresponds to the total measure of the biological warning
+system.
 
 \subsection{Data preprocessing}
-Usually, data need preprocessing to prepare for use. In this section we explain that process.
 
-In 28 Miles farm, 1\% of the data were missing, while in La Rita was 2.25\%. To fill-in the missing values we use spline interpolation. The data collected did not exhibit outliers.
+Data taken on real farms during more than a decade is expected to
+contain outliers, noise and missing samples.  These problems are
+caused by human errors or by technical defects on the instruments
+used.  
+%
+In the preprocessing step described in this section these problems
+need to be detected and fixed before moving them to the next
+processing stages.
 
-Due the fact that the variables measure meteorological or biological process, they are discretized in order to reflect trends in the data, i.e.
-the continuous values are not directly used.
-The coefficient of variation $C_v(x)$ of each variable x was used to determine the number $n$ of discretization levels.
-$$n= \lfloor 100 \ C_v(x) \rfloor$$
-where $\lfloor \ \rfloor$  is the round operator.
+In the farm 28 Miles 1\% and in La Rita 2.25\% of the data were missing.
+%
+To fill-in the missing values \todofn{spline interpolation}{falta una
+  referencia} was used.
+%
+The data collected did not exhibit outliers.
 
-Each discretization range was uniformly partitioned.
-Besides enabling the capture of tendencies, the discretization removes the effect of small variations in the data collection, either by inaccuracies of the instruments (meteorological variables) or by subjective bias introduced by the human who collects the data (biological warning system).
+Due the fact that the variables measure meteorological or biological
+processes, they are discretized \todofn{in order to reflect data trends}%
+%
+{Formalmente la discretización NO captura tendencias, sino la derivada
+  de la función en el tiempo.  Para que no se capture ruido uno
+  simplemente usa un filtro pasabajas primero (los estadísticos le
+  dicen diferente, pero no sé cómo).  En otras palabras, si querés la
+  tendencia, con usar la derivada de un filtro gaussiano sobre los
+  datos ya se ve la tendencia muy bien con el signo de esa derivada.
+  Con la varianza del filtro uno ajusta qué nivel de detalle le
+  interesa, o lo qué es lo mismo, cuanto tiempo quiere usar para el
+  pronóstico de tendencias (variable crece, está igual o decrece).  El
+  valor de salida de la derivada filtrada también se puede discretizar
+  si solo interesa esa tendencia de forma difusa, pero quizá para la
+  regresión es mejor mantenerla continua (si está filtrada, con un
+  filtro o mejor aún con un banco de filtros, al estilo wavelets).
 
-Each feature was scaled to fit in a range between 0 and 1. The variable to be predicted was not scaled.
+  Eso es TOTALMENTE independiente de la cuantificación.  Ahora, un
+  cuantificador tiene en principio un efecto de filtro pasabajas, pero
+  eso reduce el ruido y en realidad no dice nada de tendencias.  En
+  resumen, formalmente la cuantificación en sí no refleja nada de un
+  comportamiento temporal, y ``tendencia'' es en principio un
+  comportamiento temporal!}.
+%
+The variation range of each variable is uniformly discretized.
+%
+This discretization removes the effect of small variations in the data,
+either by inaccuracies of the instruments (meteorological variables)
+or by subjective bias introduced by the human collecting the data
+(biological warning system).
+
+The coefficient of variation
+$C_v(x)={\sqrt{E((x-E(x))^2)}}/{E(x)}$ of each variable $x$ is
+used to determine the number $n$ of discretization levels as
+\todofn{$n=\lfloor 100 \ C_v(x) \rfloor$,}%
+%
+{Hay algo que no entiendo.  ¿De dónde sale el 100?  ¿Por qué 100?
+  Así como está esto, el mismo rango de variación de temperatura
+  produce diferentes $n$, con solo un cambio en el promedio!  Eso es
+  muy extraño.  Por otro lado, los rangos de las variables son tan
+  diferentes, que el número $n$ va a terminar siendo casi cualquier
+  cosa.  Yo hubiese esperado que la desviación estándar se use para
+  discretizar las variables.  Otro problema formal más serio, es que
+  el coeficiente de variación solo se puede usar/tiene sentido si la
+  escala de medición es de razón, y la temperatura, la precipitación y
+  la humedad son usualmente mediciones de intervalo! Los $L_i$ ni tan
+  siquiera sé qué son porque creo que son algo subjetivos a la persona
+  que los toma ¿no?, aunque creo que iban con el porcentaje de área
+  afectado, lo que también sería medición de intervalo!}
+%
+where $\lfloor \cdot \rfloor$ is the round operator.
+
+Each feature \todofn{was scaled}{¿escaladas cómo? ¿$sx$? ¿$sx+b$?
+  ¿$sf(x)$?} to fit in the range 0 and 1. The variable $E_s$ to be
+predicted was not scaled.
+
+\TODO{Aquí faltan la fórmula general de normalización usada.  Lo que
+  sigue lo estoy asumiendo, pero puede ser que no sea así}
+
+Each variable $x\in[x_{\min},x_{\max}]$ was normalized into the
+interval $[0,1]$ with the linear map $x_n = mx+b$ with
+$m=1/(x_{\max}-x_{\min})$ and $b=-mx_{\min}$.
+
+\TODO{Explicar brevemente la razón de la normalización, ojalá con una
+  referencia bibliográfica}
 
@@
 
-Although there are many types of indicators to assess the quality of the prediction, we selected the determination coefficient $(R^2)$ and the Root Mean Square Error $(RMSE)$.
+Although there are many types of indicators to assess the quality of
+the prediction, here the coefficient of determination $(R^2)$ and the
+Root Mean Square Error $(RMSE)$.
+%
+This decision is supported by the widespread use of the former
+indicator in the agriculture and the latter in machine learning
+\citep{Soares2013,Soares2014,Ibrahim2014,Demir2014}.
 
-Given $n$ records, let be $y$ the actual value of the series, $\hat{y}$ the predicted value and $\acute{y}$ the mean of the observed data.
+Given $n$ records $y_i$, $i=1\ldots{}n$ of the actual outcome of a
+process. The mean $\bar{y}$ of the observed data is given by
+\begin{equation*}
+  \bar{y} = \frac{1}{n} \sum_{i=1}^{n} y_i
+\end{equation*}
+Let $\hat{y}_i$ be the predicted value for $y_i$. Then, the mean
+square error (MSE) $S_e^2$ and the \review{¿cómo se llama esto?
+  ¿unexplained variance?} $S_R^2$ are estimated as
+\begin{align*}
+S_e^2 &= \frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n} &
+S_R^2 &= \frac{\sum_{i=1}^{n} {(\hat{y}_i-\bar{y})}^2 }{n}
+\end{align*}
+The root mean square error is defined as $RMSE = \sqrt{S_e^2}$ and the
+coefficient of determination is
+\begin{equation*}
+R^2 = \frac{S_R^2}{S_R^2 + S_e^2}
+\end{equation*}
 
-$$ \bar{y} = \frac{1}{n} \sum_{i=1}^{n} y_i $$
-
-$$ S_e^2 = \frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}$$
-
-$$ S_R^2 = \frac{\sum_{i=1}^{n} {(\hat{y}_i-\bar{y}_i)}^2 }{n}$$
-
-$$ S_y^2 = S_R^2 + S_e^2$$
-
-$$ R^2 = \frac{S_R^2}{S_y^2}$$
-
-$$ RMSE = \sqrt{\frac{\sum_{i=1}^{n} {(y_i-\hat{y}_i)}^2 }{n}}$$
-
-This decision is supported by the widespread use in machine learning and agriculture areas \citep{Soares2014}, \citep{Soares2013}, \citep{Ibrahim2014} and \citep{Demir2014}.
+\TODO{Favor revisar.  Vi una representación en términos de $R^2=1 -
+  MSE/\sigma^2$ (con $\sigma^2$ la varianza), que no sé si está bien o
+  no, pero lo mejor será poner la fuente de donde se tomó la anterior
+  definición.  Había en el $S_R^2$ un $\bar{y}_i$ pero el promedio no
+  depende de $i$ por lo que no tenía sentido.  Supuse que era el
+  promedio, pero no sé.  El cuadrado en las $S^2$ ¿está bien?, o
+  indirectamente estás diciendo que $RMSE=S_e$, que sería una mejor
+  notación matemática (dejando RMSE solo como abreviatura textual)!}
 
@@
 
-We use the python programming language with the Integrated Development Environment (IDE) Spyder \citep{Continuum2015}, particularly with the libraries pandas \citep{mckinneypandas2010} and numpy \citep{vanderWalt2011}.
-For SVR, ridge and ordinary least squares regressions, we used sklearn \citep{scikitlearn2011} and for ESN the python-based code of Dr.
-$Luko\breve{s} evi \breve{c} ius$ \citep{Lukose2012} on which the necessary were done for adjustments for the experiments of this work.
-The computer used a processor Intel(R) Core i7-4800MQ CPU 2.70GHz, 16.0 GB RAM, running Windows 8 Pro.
+We use the Python programming language \todofn{with the Integrated
+  Development Environment (IDE)}%
+{El IDE es para la reproducción de resultados irrelevante, y por lo
+  tanto información circunstancial que es mejor evitar.  Lo relevante
+  es lo que tenga efectos directos en los resultados.  Si yo prefiero
+  usar un editor de texto corriente para programar, el restulado final
+  es el mismo, si uso el mismo compilador/intérprete.  Eso es la
+  información relevante: ¿qué versión de Python concretamente? ¿Se
+  compiló el Python o se usó interpretado?  ¿Cuáles versiones de
+  Pandas y Numpy en concreto se usaron?} %
+Spyder \citep{Continuum2015}, particularly with the libraries Pandas
+\citep{mckinneypandas2010} and Numpy \citep{vanderWalt2011}.
+
+The implementation for SVR, ridge and ordinary least squares
+regressions in scikit-learn \citep{scikitlearn2011} were used.
+%
+Adjustments to the ESN implementation code of \citet{Lukose2012} were
+necessary to allow its integration into our experimental
+framework.
+
+All experiments were performed on a PC computer with an
+Intel\textsuperscript{\textregistered} Core i7-4800MQ processor,
+2.70\,GHz, 16.0\,GB RAM, under the operating system MS Windows 8 Pro.
 
 \subsection{Methodology}
-The selection of the technique and its parametrisation was performed in two stages that they are presented below.
 
-{\bf Phase one }
+The evaluation of the techniques under consideration of their
+parameter space, was performed in two stages, described below.
 
-In the phase one, we did ten-fold-cross-validation on a set of  machine learning methods and different configurations:
+\subsubsection{Phase one}
+
+In the phase one, ten-fold-cross-validation on the total set of
+machine learning methods under a subset of configurations was evaluated:
 
 \begin{itemize}
-\item 	Patterns: n by m, where n from 1 to 8 and m from 1 to 2.
-
-\item Methods: support vector regression with the kernels functions: linear, gaussian and sigmoid; echo state networks; ordinary least squares linear regression, ridge regression and elastic-net regression.
-
+\item Patterns: $n\times{}m$, with $n=1\ldots{}8$ and $m=1\ldots{}2$.
+\TODO{No se entiende que quiere decir eso!!}
+\item Methods: support vector regression with the kernels linear,
+  gaussian and sigmoid; echo state networks; ordinary least squares
+  linear regression, ridge regression and elastic-net regression.
+  \TODO{Cada uno de esos métodos tiene a su vez un espacio
+    paramétrico.  Eso tiene que quedar claro aquí: cómo se barrió el
+    espacio paramétrico, para saber el nivel de detalle de la prueba!
+    Eso es importantísimo, porque si no no se puede decir nada
+    respecto a los métodos, sino solo respecto la configuración
+    concreta probada!}
 \item Variables included in the model:
@@
 \item All variables.
-\item From the set $\{ \overline{T}_{a} , \overline{H}, P , \overline{W}  \}$ use the subsets with one, two or four elements.
-These variables are according to experts the ones having most impact on the disease development \citep{MarinVargas1995}.
+\item From the set $\{ \overline{T}_{a} , \overline{H}, P ,
+  \overline{W} \}$ use the subsets with one, two or four
+  elements. These variables have the largest impact on the disease
+  development \citep{MarinVargas1995}.
 \end{itemize}
@@
 
-{\bf Phase two }
+\subsubsection{Phase two}
 
-In the second phase, the best configurations obtained in phase one are used to validate with the last 50 and 100 weeks.
+In the second phase, the best configurations obtained in phase one are
+used to validate with the last 50 and 100 weeks.
+
+\TODO{Hmm, ¿entonces cuántas/cuáles semanas se usaron para la fase uno?}

### 2016-07-02 techniques.tex
Metodología revisada.  Hay bastantes cambios que hacer para que quede formalmente consistente y completa

@@
 possible, by means of the minimization of an error measure
-$E(y,y^{target})$ such that it alse generalizes well to unseen data
+$E(y,y^{target})$ such that it also generalizes well to unseen data
 \citep{Lukose2012}.

### 2016-07-03 intro.tex
Modificada Introduccion

@@
 elastic-net regression and ordinary least squares linear regression,
-using input variables such as \review{completar las variables de
-  entrada usada} to predict \review{completar las variables de salida}.
+using input variables such as: air temperature, relative humidity, solar radiation, precipitation and wind speed \review{completar las variables de
+  entrada usada - Modificado} to predict the evolution stage in the biological warning system\review{completar las variables de salida - Modificado}.
 
@@
   caóticas, lo mejor en este caso son predictores lineales, o algo por
-  el estilo.} %
+  el estilo - Pendiente} %
 % 

### 2016-07-03 related.tex
Modificacion 1 parte related works

@@
 development of soft computing techniques in agricultural and
-biological engineering, including fuzzy logic, artificial neural
-networks, genetic algorithms, bayesian inference and decision trees.
+biological engineering, especially in the soil and water context for crop management and decision support in precision agriculture, including fuzzy logic, artificial neural
+networks, genetic algorithms, Bayesian inference and decision trees. They do not present numeric results of each paper, only mention the main idea.
 %
 Similarly, \citet{Kim2014} survey more recent prediction
-methods for crop pests using regression and machine learning approaches.
+methods for crop pests using regression and machine learning approaches. Nor do they provide numerical results of each paper.
 
-\TODO{Falta indicar qué concluyen los dos surveys anteriores}
+\TODO{Falta indicar qué concluyen los dos surveys anteriores - Modificado: como indico, en realidad ellos mencionan qué métodos han usado en los paper, pero no dan R2 o rmse, ni detallan resultados en general}
 

### 2016-07-04 related.tex
Modificacion 2 related works

@@
 independent data sets failed.
+%
+In contrast, our proposal presents a model that can be generalizable to other farms that have data.
 
@@
 applicability in these kind of problems.
+%
+These authors do not prove their methods in Sigatoka disease and they do classification. Instead we do regression.
 
@@
 as a linear combination of 57 input variables, in order to determine
-an index explaining 87\%, 88\% and 90\% of the
+an index explaining 87\%, 88\% and 90\% of the data
 %
-\todofn{variability}{variabilidad de qué?}  
+\todofn{variability}{variabilidad de qué? - Comentario: de los datos de cada uno }  
 %
@@
 change.
+%
+How intermediate result, they characterized the monociclic process of the black Sigatoka using  nonlinear regression.
+Although they do not present the detailed results, it no seem that they wanted to predict the progression of the black Sigatoka one, two or more periods ahead.
 

### 2016-07-05 related.tex
Modificado 3 de related works - Ready

@@
 %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Nota. la cita para lo de plant virus identificacion es  \citet{Glezakos2010}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 How intermediate result, they characterized the monociclic process of the black Sigatoka using  nonlinear regression.
-Although they do not present the detailed results, it no seem that they wanted to predict the progression of the black Sigatoka one, two or more periods ahead.
+Although they do not present the detailed results, it no seem that they wanted to predict the progression of the black Sigatoka in one, two or more periods ahead, how we do.
 
@@
 
+Although these studies have their contribution, none proposed the kind of preprocessing that we present, nor pose how to predict more than one period ahead without trying to predict climate.
+
 % Este no calza, porque no se ve qué es lo de machine learning que usa...
@@
   trabajos no se aplican al caso de la Sigatoka, o qué es lo que han
-  hecho mal, que este trabajo sí hace bien.}
+  hecho mal, que este trabajo sí hace bien. - Modificado}

### 2016-07-05 techniques.tex
Modificado 1 en techniques

@@
   modelo tenga que pasar por cero...  a menos por supuesto que la
-  constante se haya sacado de la ecuación al optimizar...}
+  constante se haya sacado de la ecuación al optimizar...
+
+  Comentario:  Pero si por feature scaling los datos están entre -1 y 1, con 0 en medio?}
 
@@
   hasta donde sea justificable usar este Y los otros, porque este ES
-  los otros...}
+  los otros...
+
+  Comentario: En realidad la idea era mostrar los diversos resultados, pero me suena bien reducirlo a ElasticNet (optimizado los parámetros) lo cual reduce resultados y gráficos en dos (quitando ridge y OLSR).
+Cuando hablamos confirmamos y lo quito.
+}
 
@@
   son los que lo explican.  Ahí metí entonces en las referencias otros
-  artículos por ese motivo.}
+  artículos por ese motivo.
+
+  Comentario: muchas gracias..}
 
@@
 
+Kernels used in this work were:
+
+Linear kernel: 
+\begin{equation*}
+K(\vct{x}_i,\vct{x}_j)={\vct{x}_i}^T \vct{x}_j
+\end{equation*}
+
+RBF: 
+\begin{equation*}
+K(\vct{x}_i,\vct{x}_j)= exp{\vct{x}_i}^T \vct{x}_j
+\end{equation*}
+
+%K(x_i,x_j )=  exp((||x_i- x_j | |^2)/(2σ^2 ))
+%donde σ es el parámetro del modelo Gausiano.
+
+Sigmoid:
+%K(x_i,x_j )=  tanh[-c+ (x_i x_j)/σ^2 ]
+%Con c ≥0  y σ^2 es el vector de escalamiento
+
 \TODO{Habría que insertar aquí los kernels utilizados en los experimentos}

### 2016-07-11 intro.tex
Modificaciones en intro related y techniques

@@
 elastic-net regression and ordinary least squares linear regression,
-using input variables such as: air temperature, relative humidity, solar radiation, precipitation and wind speed \review{completar las variables de
-  entrada usada - Modificado} to predict the evolution stage in the biological warning system\review{completar las variables de salida - Modificado}.
+using input variables such as: air temperature, relative humidity, solar radiation, precipitation and wind speed \review{LISTO - completar las variables de
+  entrada usada} to predict the evolution stage in the biological warning system\review{LISTO - completar las variables de salida - Modificado}.
 

### 2016-07-11 techniques.tex
Typos fixed
Modificaciones en intro related y techniques

@@
 
-  Comentario:  Pero si por feature scaling los datos están entre -1 y 1, con 0 en medio?}
+  COMENTARIO:  Modificado según lo conversado}
 
@@
 
-  Comentario: En realidad la idea era mostrar los diversos resultados, pero me suena bien reducirlo a ElasticNet (optimizado los parámetros) lo cual reduce resultados y gráficos en dos (quitando ridge y OLSR).
-Cuando hablamos confirmamos y lo quito.
-}
+  Comentario: Revisando la documentación de la bibliteca (ver abajo), parece que por implementación no debería excluirse OLSR por separado.
+  Estos son dos parámetros de las ElasticNet
+alpha : float.
+Constant that multiplies the penalty terms.
+Defaults to 1.0 See the notes for the exact mathematical meaning of this parameter.
+alpha = 0 is equivalent to an ordinary least square, solved by the LinearRegression object.
+For numerical reasons, using alpha = 0 with the Lasso object is not advised and you should prefer the LinearRegression object.
+l1ratio    The ElasticNet mixing parameter, with 0 < = l1ratio < = 1.
+For l1ratio = 0 the penalty is an L2 penalty.
+For l1ratio = 1 it is an L1 penalty.
+For 0 < l1ratio < 1, the penalty is a combination of L1 and L2 }
 
@@
   dejar solo la referencia.  Dejé solo lo que considero relevante,
-  pero podés reducirlo más, si querés.}
+  pero podés reducirlo más, si querés.
+
+  Comentario: Ok, gracias }
 
@@
 \begin{align*}
-  \text{minimize}   & \frac{1}{2}\|\vct{w}\|^2 + C\sum_{i=1}^n(\xi_i+\xi_i^\ast \\
-  \text{subject to} & 
+  \text{minimize} \quad  & \frac{1}{2}\|\vct{w}\|^2 + C\sum_{i=1}^n(\xi_i+\xi_i^\ast) \\
+  \text{subject to} \quad & 
   \begin{cases}
@@
 
-RBF: 
+Radial basis function (RBF): 
 \begin{equation*}
-K(\vct{x}_i,\vct{x}_j)= exp{\vct{x}_i}^T \vct{x}_j
+K(\vct{x}_i,\vct{x}_j) = exp\left( \frac{ \| \vct{x}_i - \vct{x}_j \|^2}{2 \sigma^2} \right)
 \end{equation*}
-
-%K(x_i,x_j )=  exp((||x_i- x_j | |^2)/(2σ^2 ))
-%donde σ es el parámetro del modelo Gausiano.
+where $\sigma$ is the parameter of gausian model.
 
 Sigmoid:
-%K(x_i,x_j )=  tanh[-c+ (x_i x_j)/σ^2 ]
-%Con c ≥0  y σ^2 es el vector de escalamiento
+\begin{equation*}
+K(\vct{x}_i,\vct{x}_j) = tanh \left[ -c + \frac{ \vct{x}_i  \vct{x}_j }{ \sigma^2} \right]
+\end{equation*}
+with $c \geq 0$  and $\sigma^2$ is the scaling vector.
 
-\TODO{Habría que insertar aquí los kernels utilizados en los experimentos}
+\TODO{LISTO - Habría que insertar aquí los kernels utilizados en los experimentos}
 
@@
 Here $n = 1,\ldots,T$ is the discrete time and \todofn{$T$ is the number of
-data points in the training dataset.}{No puede ser!}
+data points in the training dataset.}{No puede ser! -- Así está en el paper original, lo puse en el repositorio}
 

### 2016-07-12 methodology.tex
Typos fixed
Methodology 12-julio

@@
 %
-Both farms produce banana fruit \emph{Musa} sp.\ AAA group 'Grande
+Both farms produce banana fruit \emph{Musa} sp.\ AAA group `Grande
 Naine' (Cavendish subgroup).
@@
 $\overline{T}_{a}$ & Mean air temperature    & $[^\circ$C$]$ \\
-$\overline{H}$    & Mean humidity           & ???  \\
-$H_{min}$          & Minimal humidity        & ???  \\
-$H_{max}$          & Maximal humidity        & ???  \\
-$\overline{R}$    & Mean solar radiation    & $[$W/m$^2$$]$ \\
-$P$               & Sum precipitation       & ??? \\
-$W_{max}$          & Maximal wind speed      & $[$km/h$]$ \\
-$\overline{W}$    & Mean speed wind         & $[$km/h$]$ \\
-$L_2$             & Biological warning system – Leaf 2 & ???\\
-$L_3$             & Biological warning system – Leaf 3 & ???\\
-$L_4$             & Biological warning system – Leaf 4 & ???\\
+$\overline{H}$    & Mean relative humidity           & $[0 - 100]$   \\
+$H_{min}$          & Minimal relative humidity        & $[0 - 100]$  \\
+$H_{max}$          & Maximal relative humidity        & $[0 - 100]$  \\
+$\overline{R}$    & Mean solar radiation    & $[W/m^2]$ \\
+$P$               & Precipitation       & $[mm]$ \\
+$W_{max}$          & Maximal wind speed      & $[m/s]$ \\
+$\overline{W}$    & Mean speed wind         & $[m/s]$ \\
 \hline
-$E_s$             & Biological warning system – Evolution Stage  & ???\\
+$E_s$             & Biological warning system – Evolution Stage  & $>0$\\
 \hline
@@
 The value to be predicted in all cases is the evolution stage $E_s$,
-which corresponds to the total measure of the biological warning
-system.
+which is a measure of the level of disease progression.
 

### 2016-09-25 methodology.tex
2016-08-24
todo

@@
 %
-\todofn{However, for the current study weekly averages were used,}
+\todofn{However, for the current study weekly data were used.}
 {Hay que explicar por qué? $\rightarrow$ reducción de ruido, etc.}
@@
 computed on the data collected by nearby stations in each farm.
+Experiments were carry out with daily periodicity in meteorological variables and the results proved do not improve the prediction.
+Besides, weekly data pretend to diminish noise due sensor accuracy, missing values and outliers no detected.
 
@@
   significan entonces las variables que dicen ``mean''?
+
+  COMENTARIO: Modifiqué estas explicaciones para que quedara más claro el detalle de las variables.  Me dirás qué te parece, si las dejamos así o amplío.
 }

### 2016-12-10 Ver 1_PaperSigatoka2015/related.tex
2016-12-09-1

--- 2016-07-05 related.tex
@@
 
-\TODO{Falta indicar qué concluyen los dos surveys anteriores - Modificado: como indico, en realidad ellos mencionan qué métodos han usado en los paper, pero no dan R2 o rmse, ni detallan resultados en general}
+\TODO{MODIFICADO - Falta indicar qué concluyen los dos surveys anteriores ------ como indico, en realidad ellos mencionan qué métodos han usado en los paper, pero no dan R2 o rmse, ni detallan resultados en general}
 
@@
 %
-\todofn{variability}{variabilidad de qué? - Comentario: de los datos de cada uno }  
+\todofn{variability}{variabilidad de qué? - MODIFICADO: de los datos de cada uno }  
 %
@@
 
-\TODO{Importantísimo es explicar aquí brevemente por qué esos otros
+\TODO{MODIFICADO - Importantísimo es explicar aquí brevemente por qué esos otros
   trabajos no se aplican al caso de la Sigatoka, o qué es lo que han
-  hecho mal, que este trabajo sí hace bien. - Modificado}
+  hecho mal, que este trabajo sí hace bien}

### 2016-12-10 methodology.tex
2016-12-09-1

@@
 Besides, weekly data pretend to diminish noise due sensor accuracy, missing values and outliers no detected.
-
-\TODO{Alex, Falta explicar cada variable y sus
-  unidades y rangos usuales, lo que es relevante luego cuando se hable
-  de la normalización.
-
-  Humidity ¿cuál se usa? ¿la relativa? ¿la absoluta?, ¿la
-  específica?  Las otras unidades las supuse, así que hay que revisar
-  que estén bien!
-
-  No tengo idea qué es el ``Sum precipitation'' y en qué unidades
-  estaría.  Parece no ser algo estándar o medido directamente, ¿o sí?
-  así que amerita una explicación aquí.
-
-  Dijiste que las variables se promedian para una semana.  ¿Qué
-  significan entonces las variables que dicen ``mean''?
-
-  COMENTARIO: Modifiqué estas explicaciones para que quedara más claro el detalle de las variables.  Me dirás qué te parece, si las dejamos así o amplío.
-}
-
-\TODO{The mean variables ... }
 

### 2016-12-10 Ver 1_PaperSigatoka2015/fig/2016-02-15-Resume archivos de validación.py

@@ -1,0 +1,106 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Mon Nov 23 10:42:31 2015
+
+@author: luis
+"""
+
+import pandas as pd
+import numpy as np
+import os
+import math
+
+
+tTamTotal1 = 55
+tTamUsar1 = 50
+tTamTotal2 = 105
+tTamUsar2 = 100
+tDirectorio = "D:\Cambios\Doctorado\CORRIDAS DEL PROGRAMA\Corbana\Sigatoka - Semanal - EstadoEvolucion\\2016-02-17"
+tArchivos = [   ["2016-02-16-Sigatoka - La Rita=01= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - La Rita=02= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - La Rita=03= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - La Rita=04= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=05= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - 28 Millas=06= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - 28 Millas=07= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=08= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=09= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - 28 Millas=10= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=11= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - 28 Millas=12= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=13= - 55",tTamTotal1, tTamUsar1],
+                ["2016-02-16-Sigatoka - 28 Millas=14= - 105",tTamTotal2, tTamUsar2],
+                ["2016-02-16-Sigatoka - 28 Millas=15= - 55",tTamTotal1, tTamUsar1],
[…]

### 2016-12-10 Ver 1_PaperSigatoka2015/fig/2016-02-19-Graficar Frente de Pareto.py

@@ -1,0 +1,175 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Feb 04 18:38:48 2016
+
+@author: lcalvo
+"""
+
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as pl
+import operator
+import os
+
+
+#########################################################################    
+def guarde_Frente_Pareto(pDatos, pTipos, maxX, maxY, pCategX, pCategY, pCol):
+# maxX : Si True es porque se quiere maximizar, si False se desea minimizar
+# maxY : Si True es porque se quiere maximizar, si False se desea minimizar    
+    global gNombreGlobal  
+    for tL in pTipos:
+        tParte = pDatos[(pDatos[pCol] == tL )]
+        if len(tParte) > 0:    
+            myListX =  tParte.sort([pCategX], ascending=(not maxX)) 
+            myListY =  tParte.sort([pCategY], ascending=(not maxY)) 
+            p_front= pd.DataFrame({})
+            if maxX:
+                p_front = myListY.head(1)
+                orden = myListY
+            else:
+                p_front = myListX.head(1)
+                orden = myListX                
+            tLargo = len(orden)
+            for pair in range(1,tLargo):
+                if maxY:
+                    if maxX:
+                        if (orden.iloc[pair][pCategX] >= p_front.tail(1).iloc[-1][pCategX]) : 
+                            p_front = p_front.append(orden.iloc[pair]) 
+                    else: # if maxX
+                        if (orden.iloc[pair][pCategY] >= p_front.tail(1).iloc[-1][pCategY]):                             
+                            p_front = p_front.append(orden.iloc[pair])                 
+                else: # if maxY
+                    if maxX:
+                        if (orden.iloc[pair][pCategY] <= p_front.tail(1).iloc[-1][pCategY]):
+                            p_front = p_front.append(orden.iloc[pair])                     
+                    else: # if maxX
+                        if (orden.iloc[pair][pCategX] <= p_front.tail(1).iloc[-1][pCategX]): 
+                            p_front = p_front.append(orden.iloc[pair]) 
+        else:
[…]

### 2016-12-10 related.tex
2016-12-09-02

@@
 methods for crop pests using regression and machine learning approaches. Nor do they provide numerical results of each paper.
-
-\TODO{Falta indicar qué concluyen los dos surveys anteriores - Modificado: como indico, en realidad ellos mencionan qué métodos han usado en los paper, pero no dan R2 o rmse, ni detallan resultados en general}
-
+%
 In general, the machine learning methods applied to predict the

### 2017-04-29 intro.tex
28 abril

--- 2016-06-12 intro.tex
@@
 stages of the black Sigatoka.
-
+%	 
 \begin{figure}[ht] 
@@
 
-In this work, we compare five machine learning techniques to predict
+In this work, we compare four machine learning techniques to predict
 the development rate of the black Sigatoka disease: support vector
-regression (SVR), echo state networks (ESN), ridge regression,
-elastic-net regression and ordinary least squares linear regression,
-using input variables such as \review{completar las variables de
-  entrada usada} to predict \review{completar las variables de salida}.
+regression (SVR), echo state networks (ESN), elastic net regression 
+and ordinary least squares linear regression,
+using input variables: maximal air temperature, minimal air temperature,
+ mean air temperature, mean relative humidity, minimal relative humidity, 
+ maximal relative humidity, mean solar radiation, sum of precipitation, 
+ maximal wind speed and mean speed wind, to predict the evolution stage 
+ in the biological warning system.
 
-\todofn{The main contribution of this work}%
-{Definitivamente ese no es el aporte principal del artículo.  Cuando
-  esté listo el resto hay que volver aquí.  Creo que va a ser algo
-  como mostrar la relevancia de la etapa de preprocesamiento, o
-  mostrar que a pesar de ser sistemas complejos, con variables
-  caóticas, lo mejor en este caso son predictores lineales, o algo por
-  el estilo.} %
-% 
- is the selection of the best machine learning technique
-to forecast the black Sigatoka development rate according to the
-metric indicated below.
+ The main findings were: 1) The highest R2 was 60\% 2) The highest R2 
+ were reached with linear models like support vector regression with 
+ linear kernel, 3) As little as three meteorological variables can 
+ be used because of the correlations detected among variables.
 
@@
 future works.
+%

### 2017-04-29 methodology.tex
2016-12-09-01
28 abril

@@
 %
-\todofn{However, for the current study weekly data were used.}
-{Hay que explicar por qué? $\rightarrow$ reducción de ruido, etc.}
-%
 computed on the data collected by nearby stations in each farm.
@@
 %
-To fill-in the missing values \todofn{spline interpolation}{falta una
-  referencia} was used.
+To fill-in the missing values spline interpolation was used. \TODO{Falta poner una cita de spline interpolation}
 %
 The data collected did not exhibit outliers.
-
-Due the fact that the variables measure meteorological or biological
-processes, they are discretized \todofn{in order to reflect data trends}%
-%
-{Formalmente la discretización NO captura tendencias, sino la derivada
-  de la función en el tiempo.  Para que no se capture ruido uno
-  simplemente usa un filtro pasabajas primero (los estadísticos le
-  dicen diferente, pero no sé cómo).  En otras palabras, si querés la
-  tendencia, con usar la derivada de un filtro gaussiano sobre los
-  datos ya se ve la tendencia muy bien con el signo de esa derivada.
-  Con la varianza del filtro uno ajusta qué nivel de detalle le
-  interesa, o lo qué es lo mismo, cuanto tiempo quiere usar para el
-  pronóstico de tendencias (variable crece, está igual o decrece).  El
-  valor de salida de la derivada filtrada también se puede discretizar
-  si solo interesa esa tendencia de forma difusa, pero quizá para la
-  regresión es mejor mantenerla continua (si está filtrada, con un
-  filtro o mejor aún con un banco de filtros, al estilo wavelets).
-
-  Eso es TOTALMENTE independiente de la cuantificación.  Ahora, un
-  cuantificador tiene en principio un efecto de filtro pasabajas, pero
-  eso reduce el ruido y en realidad no dice nada de tendencias.  En
-  resumen, formalmente la cuantificación en sí no refleja nada de un
-  comportamiento temporal, y ``tendencia'' es en principio un
-  comportamiento temporal!}.
-%
-The variation range of each variable is uniformly discretized.
-%
-This discretization removes the effect of small variations in the data,
-either by inaccuracies of the instruments (meteorological variables)
-or by subjective bias introduced by the human collecting the data
-(biological warning system).
-
-The coefficient of variation
-$C_v(x)={\sqrt{E((x-E(x))^2)}}/{E(x)}$ of each variable $x$ is
-used to determine the number $n$ of discretization levels as
-\todofn{$n=\lfloor 100 \ C_v(x) \rfloor$,}%
-%
-{Hay algo que no entiendo.  ¿De dónde sale el 100?  ¿Por qué 100?
-  Así como está esto, el mismo rango de variación de temperatura
-  produce diferentes $n$, con solo un cambio en el promedio!  Eso es
-  muy extraño.  Por otro lado, los rangos de las variables son tan
-  diferentes, que el número $n$ va a terminar siendo casi cualquier
-  cosa.  Yo hubiese esperado que la desviación estándar se use para
-  discretizar las variables.  Otro problema formal más serio, es que
-  el coeficiente de variación solo se puede usar/tiene sentido si la
-  escala de medición es de razón, y la temperatura, la precipitación y
-  la humedad son usualmente mediciones de intervalo! Los $L_i$ ni tan
-  siquiera sé qué son porque creo que son algo subjetivos a la persona
-  que los toma ¿no?, aunque creo que iban con el porcentaje de área
-  afectado, lo que también sería medición de intervalo!}
-%
-where $\lfloor \cdot \rfloor$ is the round operator.
-
-Each feature \todofn{was scaled}{¿escaladas cómo? ¿$sx$? ¿$sx+b$?
-  ¿$sf(x)$?} to fit in the range 0 and 1. The variable $E_s$ to be
-predicted was not scaled.
-
-\TODO{Aquí faltan la fórmula general de normalización usada.  Lo que
-  sigue lo estoy asumiendo, pero puede ser que no sea así}
 
@@
 
-\TODO{Explicar brevemente la razón de la normalización, ojalá con una
-  referencia bibliográfica}
+The variable $E_s$ to be predicted was not normalized.
+This normalization step was made because learning schemes, like regression methods, 
+deals only with ratio scales because they calculate the distance between two instances based on the 
+values of their attributes \citet{Witten2011}.
 
@@
 Let $\hat{y}_i$ be the predicted value for $y_i$. Then, the mean
-square error (MSE) $S_e^2$ and the \review{¿cómo se llama esto?
-  ¿unexplained variance?} $S_R^2$ are estimated as
+square error (MSE) $S_e^2$ and the unexplained variance $S_R^2$ 
+are estimated as \citet{Pennstate2017}
 \begin{align*}
@@
 
-\TODO{Favor revisar.  Vi una representación en términos de $R^2=1 -
-  MSE/\sigma^2$ (con $\sigma^2$ la varianza), que no sé si está bien o
-  no, pero lo mejor será poner la fuente de donde se tomó la anterior
-  definición.  Había en el $S_R^2$ un $\bar{y}_i$ pero el promedio no
-  depende de $i$ por lo que no tenía sentido.  Supuse que era el
-  promedio, pero no sé.  El cuadrado en las $S^2$ ¿está bien?, o
-  indirectamente estás diciendo que $RMSE=S_e$, que sería una mejor
-  notación matemática (dejando RMSE solo como abreviatura textual)!}
-
 \subsection{Programming environment}
 
-We use the Python programming language \todofn{with the Integrated
-  Development Environment (IDE)}%
-{El IDE es para la reproducción de resultados irrelevante, y por lo
-  tanto información circunstancial que es mejor evitar.  Lo relevante
-  es lo que tenga efectos directos en los resultados.  Si yo prefiero
-  usar un editor de texto corriente para programar, el restulado final
-  es el mismo, si uso el mismo compilador/intérprete.  Eso es la
-  información relevante: ¿qué versión de Python concretamente? ¿Se
-  compiló el Python o se usó interpretado?  ¿Cuáles versiones de
-  Pandas y Numpy en concreto se usaron?} %
-Spyder \citep{Continuum2015}, particularly with the libraries Pandas
-\citep{mckinneypandas2010} and Numpy \citep{vanderWalt2011}.
+We use Python programming language in its interpreted 3.5.2 version,
+ particularly with the libraries; Pandas (0.18.1 version)
+\citep{mckinneypandas2010} and Numpy (1.11.1 version) \citep{vanderWalt2011}.
 
-The implementation for SVR, ridge and ordinary least squares
+The implementation for SVR, elastic net and ordinary least squares
 regressions in scikit-learn \citep{scikitlearn2011} were used.
@@
 framework.
-
-All experiments were performed on a PC computer with an
-Intel\textsuperscript{\textregistered} Core i7-4800MQ processor,
-2.70\,GHz, 16.0\,GB RAM, under the operating system MS Windows 8 Pro.
 

### 2017-04-29 related.tex
2017-01-13-prueba
28 abril

@@
 second case.
-
-% %
-% For example, 
 %
-% Me parece que aquí hay un error: Huang2010 es el survey, y sería mucha
-% coincidencia que fuera lo mismo que el \citet{Glezakos2010} ¿o no?
-%
-% \cite{Huang2010} introduce a plant virus
-% identification method based on neural networks with an evolutionary
-% preprocessing stage, 
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Nota. la cita para lo de plant virus identificacion es  \citet{Glezakos2010}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \citet{Romero1995} relied on regression models using a stepwise
@@
 as a linear combination of 57 input variables, in order to determine
-an index explaining 87\%, 88\% and 90\% of the data
-%
-\todofn{variability}{variabilidad de qué? - MODIFICADO: de los datos de cada uno }  
-%
+an index explaining 87\%, 88\% and 90\% of the data variability 
 of soybean, coffee and 
@@
 diseases, using temperature and leaf wetness as input.
-%
-%Redunda con lo que sigue:
-%Areas favorable for the diseases were plotted against the main coffee,
-%soybean and banana growing areas in Brazil.
 %
@@
 Although these studies have their contribution, none proposed the kind of preprocessing that we present, nor pose how to predict more than one period ahead without trying to predict climate.
-
-% Este no calza, porque no se ve qué es lo de machine learning que usa...
 %
-%\citet{Zhao2013} present an intelligent agricultural
-%forecasting system based on wireless sensor networks.
-
-\TODO{MODIFICADO - Importantísimo es explicar aquí brevemente por qué esos otros
-  trabajos no se aplican al caso de la Sigatoka, o qué es lo que han
-  hecho mal, que este trabajo sí hace bien}

### 2017-04-29 techniques.tex
26-08-2016
28 abril

--- 2016-07-02 techniques.tex
@@
 In the prediction of the development rate of the black Sigatoka, we
-compare techniques such as least squares or ridge regression, commonly
+compare techniques such as least squares regression and elastic-net regression, commonly
 encountered in the agricultural literature with machine learning
-methods such as support vector regression, elastic regression and echo
+methods such as support vector regression and echo
 state networks, where the parameter space of each technique is also
@@
 
-\TODO{Alex, no sé hasta donde hiciste en las pruebas ese ``arreglo''
-  de que los datos tengan todos una entrada igual a 1.  De no ser así,
-  el método no puede encontrar el offset, y sería como forzar que el
-  modelo tenga que pasar por cero...  a menos por supuesto que la
-  constante se haya sacado de la ecuación al optimizar...}
-
-\subsection{Ridge regression}
-
-In contrast to the OLSR, for the ridge regression (RR)
-\citet{Hoerl1988} proposed to add a term to penalize large weights
-into the error function
-\begin{equation*}
-  E(\vct{w}) = 
-  \left\| \mat{X}\vct{w} - \vct{y} \right\|_2^2 + 
-  \alpha \left\|\vct{w}\right\|_2^2 
-\end{equation*}
-where the parameter $\alpha>0$ controls how strong is the shrinking of
-the estimates towards zero.  This shrinkage introduces some bias but
-helps to reduce the variance of the estimate.
-%
-The solution of the optimization problem (\ref{eq:problem}) in this
-case is given by $\hat{\vct{w}}=(\mat{X}^T\mat{X} +
-\lambda\mat{I})^{-1}\mat{X}^T \vct{y}$.
-
 \subsection{Elastic net regression}
@@
 Instead of $L_2$ regularization prior
-($\alpha\left\|\vct{w}\right\|_2^2$) included in the ridge regression,
+($\alpha\left\|\vct{w}\right\|_2^2$) included in the ridge regression (RR),
 \citet{Tibshirani1996} used an $L_1$ term ($\lambda
@@
 
-\TODO{Algo que no entiendo es que las elastic net son en realidad una
-  generalización del ridge, Lasso y OLSR.  Así que en los experimentos
-  con las selección adecuada de parámetros este método debería
-  comportarse al menos igual o mejor que esos otros regresores!  No sé
-  hasta donde sea justificable usar este Y los otros, porque este ES
-  los otros...}
-
-\TODO{Alex, veo que usas mucho \citep{scikitlearn2011} para los
-  métodos, pero ese es solo el paper de un tool y no los proponentes
-  originales de los métodos.  Usualmente uno hace referencia a algún
-  artículo, libro o tutorial donde ojalá los que propusieron el método
-  son los que lo explican.  Ahí metí entonces en las referencias otros
-  artículos por ese motivo.}
-
 \subsection{Support Vector Regression (SVR)}
-
-\TODO{Estoy seguro que los lectores van a solicitar reducir esto y
-  dejar solo la referencia.  Dejé solo lo que considero relevante,
-  pero podés reducirlo más, si querés.}
 
@@
 
-\TODO{Habría que insertar aquí los kernels utilizados en los experimentos}
+Kernels used in this work were:
+
+Linear kernel: 
+\begin{equation*}
+K(\vct{x}_i,\vct{x}_j)={\vct{x}_i}^T \vct{x}_j
+\end{equation*}
+
+Radial basis function (RBF): 
+\begin{equation*}
+K(\vct{x}_i,\vct{x}_j) = exp\left( \frac{ \| \vct{x}_i - \vct{x}_j \|^2}{2 \sigma^2} \right)
+\end{equation*}
+where $\sigma$ is the parameter of gausian model.
+
+Sigmoid:
+\begin{equation*}
+K(\vct{x}_i,\vct{x}_j) = tanh \left[ -c + \frac{ \vct{x}_i  \vct{x}_j }{ \sigma^2} \right]
+\end{equation*}
+with $c \geq 0$  and $\sigma^2$ is the scaling vector.
 
@@
 %
-Here $n = 1,\ldots,T$ is the discrete time and \todofn{$T$ is the number of
-data points in the training dataset.}{No puede ser!}
+Here $n = 1,\ldots,T$ is the discrete time and $T$ is the number of
+data points in the training dataset.
 
@@
 \citep{Lukose2012}.
+%

### 2017-04-29 methodology.tex
28 abril 2

@@
 %
-To fill-in the missing values spline interpolation was used. \TODO{Falta poner una cita de spline interpolation}
+To fill-in the missing values spline interpolation was used \citet{Alglib2017}.
 %
@@
 
-The evaluation of the techniques under consideration of their
-parameter space, was performed in two stages, described below.
-
-\subsubsection{Phase one}
-
-In the phase one, ten-fold-cross-validation on the total set of
+Ten-fold-cross-validation on the total set of
 machine learning methods under a subset of configurations was evaluated:
@@
 \end{itemize}
-
-\subsubsection{Phase two}
-
-In the second phase, the best configurations obtained in phase one are
-used to validate with the last 50 and 100 weeks.
-
-\TODO{Hmm, ¿entonces cuántas/cuáles semanas se usaron para la fase uno?}

### 2017-04-30 methodology.tex
29-4 9-46

@@
 \subsection{Methodology}
-
-Ten-fold-cross-validation on the total set of
-machine learning methods under a subset of configurations was evaluated:
+When the amount of data for training and testing is limited, is recommended
+to use cross-validation \citep{Witten2011}. We used ten-fold-cross-validation 
+on the total set. El diseño experimental combinó los siguientes factores:
 
 \begin{itemize}
-\item Patterns: $n\times{}m$, with $n=1\ldots{}8$ and $m=1\ldots{}2$.
-\TODO{No se entiende que quiere decir eso!!}
-\item Methods: support vector regression with the kernels linear,
+\item Patrones de semanas: Se formaron patrones desde tres semanas continuas
+de observación para predecir una semana adelante, hasta doce semanas de
+observación para predecir tres semanas adelante. Por tanto, $n\times{}m$, 
+with $n=1\ldots{}12$ and $m=1\ldots{}3$.
+
+\item techniques: support vector regression with the kernels linear,
   gaussian and sigmoid; echo state networks; ordinary least squares
-  linear regression, ridge regression and elastic-net regression.
-  \TODO{Cada uno de esos métodos tiene a su vez un espacio
-    paramétrico.  Eso tiene que quedar claro aquí: cómo se barrió el
-    espacio paramétrico, para saber el nivel de detalle de la prueba!
-    Eso es importantísimo, porque si no no se puede decir nada
-    respecto a los métodos, sino solo respecto la configuración
-    concreta probada!}
+  linear regression and elastic-net regression.
+
 \item Variables included in the model:
@@
 \item From the set $\{ \overline{T}_{a} , \overline{H}, P ,
-  \overline{W} \}$ use the subsets with one, two or four
+  \overline{W} \}$ use the subsets with one, two, three or four
   elements. These variables have the largest impact on the disease

### 2017-04-30 methodology.tex
29-4 10-47

@@
 de observación para predecir una semana adelante, hasta doce semanas de
-observación para predecir tres semanas adelante. Por tanto, $n\times{}m$, 
+observación para predecir tres semanas adelante. Por tanto, $n\times{}m$ combinations, 
 with $n=1\ldots{}12$ and $m=1\ldots{}3$.
 
-\item techniques: support vector regression with the kernels linear,
-  gaussian and sigmoid; echo state networks; ordinary least squares
-  linear regression and elastic-net regression.
+\item Techniques: Se utilizaron las siguientes técnicas, revisando el espacio
+paramétrico especificado a continuación:
+\begin{itemize}
+\item SVR with linear kernel: C [0.001, 1, 10, 100, 1000], epsilon [0.0, 0.4, 0.9].
+\item SVR with gaussian kernel: C [0.001, 1, 10, 100, 1000], epsilon [0.0, 0.4, 0.9],
+gamma [0.0, 0.4, 0.9].
+\item SVR with sigmoid kernel: C [0.001, 1, 10, 100, 1000], epsilon [0.0, 0.4, 0.9],
+gamma [0.0, 0.4, 0.9], coef0 [0.0, 0.5, 5, 10].
+\item Echo state networks: LeakingRate [0.02 .. 0.9], neurons [1\% .. 90\%] de la 
+cardinalidad del training set, InitLen [ 0.1 .. 0.8 ].
+\item Ordinary least squares linear regression: No parameters.
+\item Elastic-net regression: alpha [0 .. 0.9], l1_ratio [0 .. 1.0].
+\end{itemize}
 
@@
 \end{itemize}
-
 \end{itemize}

### 2017-04-30 methodology.tex
30-4 00-52

--- 2017-04-30 methodology.tex
@@
 de observación para predecir una semana adelante, hasta doce semanas de
-observación para predecir tres semanas adelante. Por tanto, $n\times{}m$, 
+observación para predecir tres semanas adelante. Por tanto, $n\times{}m$ combinations, 
 with $n=1\ldots{}12$ and $m=1\ldots{}3$.
 
-\item techniques: support vector regression with the kernels linear,
-  gaussian and sigmoid; echo state networks; ordinary least squares
-  linear regression and elastic-net regression.
+\item Techniques: Se utilizaron las siguientes técnicas, revisando el espacio
+paramétrico especificado a continuación:
+\begin{itemize}
+\item SVR with linear kernel: $C$ $[0.001, 1, 10, 100, 1000]$, $epsilon$ $[0.0, 0.4, 0.9]$.
+\item SVR with gaussian kernel: $C$ $[0.001, 1, 10, 100, 1000]$, $epsilon$ $[0.0, 0.4, 0.9]$,
+$gamma$ $[0.0, 0.4, 0.9]$.
+\item SVR with sigmoid kernel: $C$ $[0.001, 1, 10, 100, 1000]$, $epsilon$ $[0.0, 0.4, 0.9]$,
+$gamma$ $[0.0, 0.4, 0.9]$, $coef0$ $[0.0, 0.5, 5, 10]$.
+\item Echo state networks: $LeakingRate$ $[0.02 .. 0.9]$, $neurons$ $[1\% .. 90\%]$ de la 
+cardinalidad del training set, $InitLen$ $[ 0.1 .. 0.8 ]$.
+\item Ordinary least squares linear regression: No parameters.
+\item Elastic-net regression: $alpha$ $[0 .. 0.9]$, $l1\_ratio$ $[0 .. 1.0]$.
+\end{itemize}
 

### 2017-04-30 results.tex
30-4 00-52

@@
 
-In this section we present the main results for phase and their discussion.
-
-{\bf Phase one }
+In this section the main results are presented and discussed.
 
@@
  \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_algortihm}
+ \includegraphics[scale=.5]{2017-01-15-Algorithms-R2}
  \caption{Phase one - Best $R^2$ for each algorithm} 
@@
  \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_each_number_of_observation_range_in_the_patterm}
+ \includegraphics[scale=.5]{2017-01-15-Periods-R2}
  \caption{Phase one - Best $R^2$ for each observation range} 
@@
  \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
+ \includegraphics[scale=.5]{2017-01-15-Variables-R2}
  \caption{Phase one - Best $R^2$ for each variable combination} 
@@
 We can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week.
-
-{\bf Phase two }
-
-In the second phase, the best configurations obtained in phase one were used to validate with the last 50 and 100 weeks.
-
-\figref{figura8} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Even if linear models continue with good $R^2$, Echo State Networks improve their scores because in $50$ and $100$ last weeks validation, we are in presence of climate change, then the behaviour is less lineal.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_algortihm}
- \caption{Phase two - Best $R^2$ for each algorithm} 
- \label{figura8} 
-\end{figure}
-
-\figref{figura9} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-The results confirms that, in general, to predict one week ahead is better than two weeks ahead, two than three and so on, this for both farms.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_two_Best_R2_for_each_number_of_observation_range_in_the_patterm}
- \caption{Phase two - Best $R^2$ for each observation range} 
- \label{figura9} 
-\end{figure}
-
-\figref{figura10} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-This results confirm that $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms are the best variables combinations.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Phase_one_Best_R2_for_variables}
- \caption{Phase two - Best $R^2$ for each variable combination} 
- \label{figura10} 
-\end{figure}
-
-\figref{figura11} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.  You can note that the behaviour of $R^2$ and $RMSE$ is similar to the phase one.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.8]{Phase_two_R2_RMSE}
- \caption{Phase two - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura11} 
-\end{figure}
-
-In phase two, the Pareto frontier for the La Rita farm is composed by 2 elements. The \tablename $.$\ref{tabla4} shows the composition about variables and observation ranges.
-
-\begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase two} 
-\label{tabla4} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-All & 1 to 1  & 2 & 69.33\% & 353.33 \\
-\hline
-\end{tabular} 
-\end{table}
-
-In 28 Millas Farm, the Pareto frontier is composed by 1 element. The \tablename $.$\ref{tabla5} details the result.
-
-\begin{table}[h] 
-\caption{Pareto frontier - 28 Millas - Phase two} 
-\label{tabla5} 
-\centering
-\begin{tabular}{c|c|c|c|c} 
-\hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
-\hline\hline 
-Pair $\overline{T}_{a}$ $P$ & 1 to 1 & 1 & 61.36\% & 339.89 \\
-\hline
-\end{tabular} 
-\end{table}
-
-Again, similar to phase one, we can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week, one week to predict one week ahead with combinations of variables listed above.

### 2017-05-01 results.tex
30-4 2

@@
 
-In this section the main results are presented and discussed.
+Realizados los experimentos, la \tablename $.$\ref{tabla2} muestra los RMSE obtenidos, en la cual se puede observar que las ESN obtienen un resultado muy diferente a las demás técnicas.
+El RMSE obtenido por las ESN es cercano a 20 veces el de las otras técnicas, es por ello que de este punto en adelante es descartado de la comparación pues definitivamente no es opción con el conjunto de datos disponible.
+Este resultado de las ESN se explica en que como estas redes parten de estados aleatorios para irse ajustando en el proceso de entrenamiento, la cantidad de datos disponibles no logra que la red neuronal se entrene adecuadamente.
+%
+\begin{table}[h] 
+\caption{Promedio y desviación estándar de los RMSE obtenidos por Finca} 
+\label{tabla2} 
+\centering
+\begin{tabular}{c|c|c|c} 
+\hline
+\bfseries Farm & \bfseries Technique & \bfseries $RMSE_mean$ & \bfseries $RMSE_std$\\ 
+\hline\hline 
+28 Millas  	& Elastic net  & 463.56	& 	82.39  \\
+			& SVR with linear kernel  & 465.92  & 85.70  \\
+			& SVR with Gaussian kernel  & 466.63 & 89.53  \\						
+			& Linear regression  & 468.25 & 83.57  \\									
+			& SVR with sigmoid kernel & 552.81 & 123.82  \\									
+			& ESN  & 11085.47 &  7965.66 \\												
+\hline 
+La Rita  	& SVR with linear kernel  & 816.29	& 	216.43  \\
+			& Elastic net   &  817.98 &  211.14 \\
+			& SVR with Gaussian kernel  & 820.92 & 231.21  \\						
+			& Linear regression  & 823.55 &  213.24 \\									
+			& SVR with sigmoid kernel & 1070.58 &  331.89 \\									
+			& ESN  & 8329.72 & 5435.54  \\												
+\hline    
+\end{tabular} 
+\end{table}
+%
+La \figref{figura4} muestra los box plots con respecto al RMSE para el resto de técnicas (excluída las ESN).
+Se aprecia que a excepción de las SVR with sigmoid kernel, las otras técnicas poseen compartamientos muy similares.
+Aunque el promedio del RMSE es diferente entre la Rica (cercano a 450)y 28 Millas (cercano a ñps820), la tendencia a preferir las mismas técnicas se mantiene en ambas fincas.
 
-\figref{figura4} shows the best $R^2$ for each algorithm in the experiment.
+\begin{figure}[H] 
+ \centering
+ \includegraphics[scale=.8]{Usado_2017-04-30_Sigatoka_RMSE_Boxplot}
+ \caption{Box plots de los RMSE para cada una de las Fincas} 
+ \label{figura4} 
+\end{figure}
+
+The \figref{figura5} shows the best $R^2$ for each algorithm in the experiment.
 Results are group by farm.
@@
  \caption{Phase one - Best $R^2$ for each algorithm} 
- \label{figura4} 
+ \label{figura5} 
 \end{figure}
 
-\figref{figura5} presents, for one, two and three weeks ahead, the best $R^2$.
+\figref{figura6} presents, for one, two and three weeks ahead, the best $R^2$.
 Results are group by farm.
@@
  \caption{Phase one - Best $R^2$ for each observation range} 
- \label{figura5} 
+ \label{figura6} 
 \end{figure}
 
-\figref{figura6} shows the best $R^2$ for each variables combination.
+\figref{figura7} shows the best $R^2$ for each variables combination.
 Results are group by farm.
@@
  \caption{Phase one - Best $R^2$ for each variable combination} 
- \label{figura6} 
+ \label{figura7} 
 \end{figure}
 
-\figref{figura7} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
+\figref{figura8} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
 The Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better $RMSE$ than La Rita.
@@
  \centering
- \includegraphics[scale=.8]{Phase_one_R2_RMSE}
- \caption{Phase one - Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura7} 
+ \includegraphics[scale=.8]{Usado_2017-04-30_Sigatoka_R2_RMSE}
+ \caption{Pareto frontier for $R^2$ and $RMSE$} 
+ \label{figura8} 
 \end{figure}
 
-The Pareto frontier for the La Rita farm is composed by 96 elements. The \tablename $.$\ref{tabla2} shows the composition about variables and observation ranges.
+The Pareto frontier for the La Rita farm is composed by 96 elements. The \tablename $.$\ref{tabla3} shows the composition about variables and observation ranges.
 
@@
 \caption{Composition of the Pareto frontier - La Rita - Phase one} 
-\label{tabla2} 
+\label{tabla3} 
 \centering
@@
 
-Similarly, the Pareto frontier for the 28 Millas farm is composed by 75 elements. The \tablename $.$\ref{tabla3} shows the composition about variables and observation ranges.
+Similarly, the Pareto frontier for the 28 Millas farm is composed by 75 elements. The \tablename $.$\ref{tabla4} shows the composition about variables and observation ranges.
 
@@
 \caption{Composition of the Pareto frontier - 28 Millas - Phase one} 
-\label{tabla3} 
+\label{tabla4} 
 \centering

### 2017-05-01 results.tex
1 de mayo

@@
 Se aprecia que a excepción de las SVR with sigmoid kernel, las otras técnicas poseen compartamientos muy similares.
-Aunque el promedio del RMSE es diferente entre la Rica (cercano a 450)y 28 Millas (cercano a ñps820), la tendencia a preferir las mismas técnicas se mantiene en ambas fincas.
-
+Aunque el promedio del RMSE es diferente entre la Rica (cercano a 450)y 28 Millas (cercano a 820), la tendencia a preferir las mismas técnicas se mantiene en ambas fincas.
+%
 \begin{figure}[H] 
@@
 \end{figure}
-
-The \figref{figura5} shows the best $R^2$ for each algorithm in the experiment.
-Results are group by farm.
-Though La Rita obtains different results in magnitude than 28 Millas, the trend is similar.
-In both farms, the best results are for linear models, second position is occupied for Echo State Networks and  SVR with gausian and sigmoid kernels are the worst results.
-In linear models, to predict one week ahead is better than two weeks ahead, and this is better than three weeks ahead.
-
+%
+Now, the \figref{figura5} compares the mean of the $RMSE$ for each algorithm in the experiment, pero para la predicción de 1, 2 ó 3 semanas adelante con respecto a la información de las variables climatológocas y del preaviso biológico.
+Además, to predict one week ahead obtiene un RMSE más bajo que  two weeks ahead, and this is better than three weeks ahead.
+%
 \begin{figure}[H] 
  \centering
- \includegraphics[scale=.5]{2017-01-15-Algorithms-R2}
+ \includegraphics[scale=.5]{Usado_Algorithms-RMSE}
  \caption{Phase one - Best $R^2$ for each algorithm} 

### 2017-05-01 results.tex
1-5 1.00

@@
 
-Realizados los experimentos, la \tablename $.$\ref{tabla2} muestra los RMSE obtenidos, en la cual se puede observar que las ESN obtienen un resultado muy diferente a las demás técnicas.
-El RMSE obtenido por las ESN es cercano a 20 veces el de las otras técnicas, es por ello que de este punto en adelante es descartado de la comparación pues definitivamente no es opción con el conjunto de datos disponible.
-Este resultado de las ESN se explica en que como estas redes parten de estados aleatorios para irse ajustando en el proceso de entrenamiento, la cantidad de datos disponibles no logra que la red neuronal se entrene adecuadamente.
+Realizados los experimentos, la \tablename $.$\ref{tabla2} muestra los $RMSE$ obtenidos, en la cual se puede observar que las ESN obtienen un resultado muy diferente a las demás técnicas.
+El $RMSE$ promdio obtenido por las ESN $11085.47$ es cercano a 20 veces el de las otras técnicas.
+Este resultado de las ESN se explica debido a que estas redes neuronales recurrentes parten de estados aleatorios para irse ajustando en el proceso de entrenamiento, y por tanto, los valores obtenidos son reflejo de que la cantidad de datos disponibles no logra que la red neuronal se ajuste.
 %
@@
 %
-La \figref{figura4} muestra los box plots con respecto al RMSE para el resto de técnicas (excluída las ESN).
-Se aprecia que a excepción de las SVR with sigmoid kernel, las otras técnicas poseen compartamientos muy similares.
-Aunque el promedio del RMSE es diferente entre la Rica (cercano a 450)y 28 Millas (cercano a 820), la tendencia a preferir las mismas técnicas se mantiene en ambas fincas.
+Adicionalmente, la \figref{figura4} muestra los box plots con respecto a los $RMSE$ obtenidos.
+Se grafica ESN por separado debido a la diferencia de escala.
+Con respecto a las otras técnicas, la regresión lineal, Elastic net, SVR with gaussian kernel y SVR with linear kernel presentan compartamientos muy similares y aunque el promedio del $RMSE$ es diferente entre las fincas, La Rica cercano a 450 y 28 Millas cercano a 820, el comportaniento relativo de las técnicas es el mismo.
+Por su parte, SVR with sigmoid kernel presenta un promedio de $RMSE$ muy inferior a ESN, pero superior a las otras cuatro técnicas.
 %
@@
  \centering
- \includegraphics[scale=.8]{Usado_2017-04-30_Sigatoka_RMSE_Boxplot}
+ \includegraphics[scale=.8]{Usado_2017-04-30_Sigatoka_RMSE_Boxplot_4}
  \caption{Box plots de los RMSE para cada una de las Fincas} 
@@
 %
-Now, the \figref{figura5} compares the mean of the $RMSE$ for each algorithm in the experiment, pero para la predicción de 1, 2 ó 3 semanas adelante con respecto a la información de las variables climatológocas y del preaviso biológico.
-Además, to predict one week ahead obtiene un RMSE más bajo que  two weeks ahead, and this is better than three weeks ahead.
-%
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{Usado_Algorithms-RMSE}
- \caption{Phase one - Best $R^2$ for each algorithm} 
- \label{figura5} 
-\end{figure}
-
-\figref{figura6} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-In general, to predict one week ahead is better than two weeks ahead and so on.
-The number of weeks consider in the observation range in the pattern is not the main discriminant factor, but it is clear that we get better $R^2$ for one week ahead than two weeks ahead and so on.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{2017-01-15-Periods-R2}
- \caption{Phase one - Best $R^2$ for each observation range} 
- \label{figura6} 
-\end{figure}
-
-\figref{figura7} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-The better results are obtained with $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms of similarly.
-You can note that the use of all variables in the model or the inclusion of the four variables suggest for expert criteria do not improve significantly the results, then the use of more sensors do not assure a better result.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{2017-01-15-Variables-R2}
- \caption{Phase one - Best $R^2$ for each variable combination} 
- \label{figura7} 
-\end{figure}
-
-\figref{figura8} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
-The Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better $RMSE$ than La Rita.
+\figref{figura5} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
+La Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better $RMSE$ than La Rita.
 This situation arise because $RMSE$ considers errors only with respect the prediction and in 28 Millas the average of Stage of Evolution is $4316.16$, unlike, in La Rita the average is $5507.30$.
 So, in La Rita we obtains higher errors in absolute values.
-$R2$ is a relative metric between 0 thru 1 and it is less sensitive to absolute values.
-
+$R^2$  is a relative metric between 0 thru 1 and it is less sensitive to absolute values.
+%
 \begin{figure}[H] 
@@
  \caption{Pareto frontier for $R^2$ and $RMSE$} 
- \label{figura8} 
+ \label{figura5} 
 \end{figure}
 
-The Pareto frontier for the La Rita farm is composed by 96 elements. The \tablename $.$\ref{tabla3} shows the composition about variables and observation ranges.
+The Pareto frontier for the La Rita farm is composed by 5 elements. The \tablename $.$\ref{tabla3} shows the composition about variables, observation ranges, techniques and weeks ahead.
 
 \begin{table}[h] 
-\caption{Composition of the Pareto frontier - La Rita - Phase one} 
+\caption{Composition of the Pareto frontier - La Rita} 
 \label{tabla3} 
 \centering
-\begin{tabular}{c|c|c|c|c} 
+\begin{tabular}{c|c|c|c|c|c} 
 \hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
+\bfseries Variable & \bfseries Observation range & \bfseries Weeks ahead & \bfseries Technique &\bfseries $R^2$ & \bfseries $RMSE$\\ 
 \hline\hline 
+  &   &  &  &   &  \\
+
 Pair $\overline{T}_{a}$ $\overline{W}$ &	1 to 1  & 36 & 64.25\% & 714.51 \\
@@
 
+Now, the \figref{figura5} compares the mean of the $RMSE$ for each algorithm in the experiment, pero para la predicción de 1, 2 ó 3 semanas adelante con respecto a la información de las variables climatológocas y del preaviso biológico.
+Además, to predict one week ahead obtiene un RMSE más bajo que  two weeks ahead, and this is better than three weeks ahead.
+%
+\begin{figure}[H] 
+ \centering
+ \includegraphics[scale=.5]{Usado_Algorithms-RMSE}
+ \caption{$RMSE$ for each algorithm} 
+ \label{figura5} 
+\end{figure}
+
+\figref{figura6} presents, for one, two and three weeks ahead, the best $R^2$.
+Results are group by farm.
+In general, to predict one week ahead is better than two weeks ahead and so on.
+The number of weeks consider in the observation range in the pattern is not the main discriminant factor, but it is clear that we get better $R^2$ for one week ahead than two weeks ahead and so on.
+
+\begin{figure}[H] 
+ \centering
+ \includegraphics[scale=.5]{2017-01-15-Periods-R2}
+ \caption{Phase one - Best $R^2$ for each observation range} 
+ \label{figura6} 
+\end{figure}
+
+\figref{figura7} shows the best $R^2$ for each variables combination.
+Results are group by farm.
+The better results are obtained with $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms of similarly.
+You can note that the use of all variables in the model or the inclusion of the four variables suggest for expert criteria do not improve significantly the results, then the use of more sensors do not assure a better result.
+
+\begin{figure}[H] 
+ \centering
+ \includegraphics[scale=.5]{2017-01-15-Variables-R2}
+ \caption{Phase one - Best $R^2$ for each variable combination} 
+ \label{figura7} 
+\end{figure}
+
 We can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week.

### 2017-05-01 conclusions.tex
2017-01-13-prueba
1-5 4.52

@@
 
+The main findings were: 1) The highest R2 was 60\% 2) The highest R2 were reached with linear models like support vector regression with linear kernel, 3) As little as three meteorological variables can be used because of the correlations detected among variables.
+
 \subsection{Future work}
+Currently this research is focused on the extension of these findings to other farms and the extension of the model to be applied in an early warning system, which does not require the regression task.

### 2017-05-01 results.tex
1-5 4.52

@@
 \hline\hline 
-  &   &  &  &   &  \\
-
-Pair $\overline{T}_{a}$ $\overline{W}$ &	1 to 1  & 36 & 64.25\% & 714.51 \\
- &	2 to 1  & 6 & 62.97\% & 695.10 \\
+$\overline{T}_{a}$ $\overline{W}$ &	1  & 1 & SVR with linear kernel & 693.39 & 59.4\% \\
+										&	2  & 1 & SVR with linear kernel & 683.22 & 59.4\% \\
 \hline 
-All  & 1 to 1  & 18 & 62.98\% & 701.95 \\
-   & 2 to 1  & 12 & 61.76\% & 679.92 \\
-    & 3 to 1  & 6 & 60.60\% & 676.42 \\
-    & 5 to 1  &  2 & 60.37\% & 672.39 \\
+$\overline{T}_{a}$  &  5 & 1 &  SVR with linear kernel & 679.25 & 59.2\% \\
 \hline 
-$\overline{T}_{a}$ & 1 to 1  & 12  & 63.60\% & 708.77 \\
-       &	2 to 1  & 4 & 62.23\% & 689.55 \\
+$\overline{T}_{a}$ $\overline{H}$ &	5  & 1 & Elastic net & 677.49 & 57.8\% \\
+										&	5  & 1 & SVR with linear kernel & 677.37 & 59.1\% \\
 \hline										
@@
 \end{table}
-
-Similarly, the Pareto frontier for the 28 Millas farm is composed by 75 elements. The \tablename $.$\ref{tabla4} shows the composition about variables and observation ranges.
+%
+Similarly, the Pareto frontier for the 28 Millas farm is composed by 15 elements. The \tablename $.$\ref{tabla4} shows the composition about variables and observation ranges.
 
 \begin{table}[h] 
-\caption{Composition of the Pareto frontier - 28 Millas - Phase one} 
+\caption{Composition of the Pareto frontier - 28 Millas} 
 \label{tabla4} 
 \centering
-\begin{tabular}{c|c|c|c|c} 
+\begin{tabular}{c|c|c|c|c|c} 
 \hline
-\bfseries Variable & \bfseries Observation range & \bfseries Quantity & \bfseries Max $R^2$ & \bfseries Min $RMSE$\\ 
+\bfseries Variable & \bfseries Observation range & \bfseries Weeks ahead & \bfseries Technique &\bfseries $R^2$ & \bfseries $RMSE$\\ 
 \hline\hline 
-Pair $\overline{T}_{a}$ $\overline{W}$ & 1 to 1 & 8 & 57.80\% & 438.09 \\
+$\overline{T}_{a}$ $\overline{H}$ $\overline{W}$ $P$  &	9  & 1  & Elastic net & 398.65 & 48.5\% \\
+													&	8  & 1 & Elastic net & 399.57 & 48.8\% \\
+													&	7  & 1 & Elastic net & 399.65 & 49.1\% \\													
 \hline 													
-All   &	9 to 1 & 2 & 50.93\% & 397.93 \\
-  & 10 to 1	 & 2 & 50.97\% & 398.81 \\
-  &	8 to 1 & 6 & 51.62\% & 398.93 \\
-  &	7 to 1 & 2 & 52.25\% & 400.28 \\
-  &	6 to 1 & 2 & 53.16\% & 404.14 \\
-  &	4 to 1 & 2 & 54.32\% & 407.54 \\
+$\overline{T}_{a}$ $\overline{H}$ $\overline{W}$  &	9  & 1  & SVR with linear kernel & 399.86 & 49.3\% \\
+												&	7  & 1  & SVR with linear kernel & 400.87 & 50.3\% \\
+												&	6  & 1  & SVR with linear kernel & 403.40 & 50.7\% \\
+												&	3  & 1  & SVR with gaussian kernel & 407.25 & 90.7\% \\													
 \hline 													
-$\overline{T}_{a}$ & 1 to 1  & 8  & 59.09\% & 439.44 \\
+$\overline{T}_{a}$ $\overline{H}$  &	4  & 1  & Elastic net & 400.65 & 49.6\% \\
+									&	1  & 1  & SVR with linear kernel & 412.26 & 51.8\% \\
+									&	2  & 1  & SVR with linear kernel & 409.18 & 51.2\% \\			
 \hline 																												
-Pair $\overline{T}_{a}$ $\overline{H}$ & 1 to 1	 & 8 & 57.51\% & 428.61 \\
- &	2 to 1 & 20 & 56.91\% & 414.37 \\
- &	3 to 1 & 3 & 54.41\% & 411.55 \\
- &	4 to 1 & 3 & 53.34\% & 406.65 \\
+$\overline{T}_{a}$ $\overline{H}$ $P$  & 2  & 1  & SVR with linear kernel & 410.71 & 51.2\% \\
 \hline 																												
-Pair $\overline{T}_{a}$ $P$ & 3 to 1 & 9 & 56.23\% & 422.76 \\
+All  & 7  & 1  & SVR with linear kernel & 426.19 & 51.9\% \\
+     & 5  & 1  & SVR with linear kernel & 428.66 & 52.3\% \\
+     & 5 & 1  & Elastic net & 433.03 & 52.4\% \\
+     &  5 & 1  & Linear regression & 450.48 & 53.0\% \\          
 \hline	
@@
 \end{table}
-
-Now, the \figref{figura5} compares the mean of the $RMSE$ for each algorithm in the experiment, pero para la predicción de 1, 2 ó 3 semanas adelante con respecto a la información de las variables climatológocas y del preaviso biológico.
-Además, to predict one week ahead obtiene un RMSE más bajo que  two weeks ahead, and this is better than three weeks ahead.
+%
+Now, the \figref{figura6} compares the mean of the $RMSE$ for each technique in the experiment.
+Se muestra el $RMSE$ para la predicción de 1, 2 ó 3 semanas adelante con respecto a la información de las variables climatológicas y del preaviso biológico observadas.
+Se aprecia que to predict one week ahead obtiene un $RMSE$ más bajo que two weeks ahead, and this is better than three weeks ahead, lo cual es lo esperado pues entre más semanas adelante se desee predecir, hay más incertidumbre.
 %
@@
  \caption{$RMSE$ for each algorithm} 
- \label{figura5} 
+ \label{figura6} 
 \end{figure}
-
-\figref{figura6} presents, for one, two and three weeks ahead, the best $R^2$.
-Results are group by farm.
-In general, to predict one week ahead is better than two weeks ahead and so on.
-The number of weeks consider in the observation range in the pattern is not the main discriminant factor, but it is clear that we get better $R^2$ for one week ahead than two weeks ahead and so on.
-
+%
+\figref{figura7} shows the the mean $RMSE$ for each variables combination.
+The better results are obtained with $\overline{T}_{a}$, $\overline{W}$ and $P$.
+Incluir más variables en el modelo no asegura mejorar la métrica, lo cual es importante pues este resultado sugiere que the use of more sensors do not assure a better result of prediction.
+%
 \begin{figure}[H] 
  \centering
- \includegraphics[scale=.5]{2017-01-15-Periods-R2}
- \caption{Phase one - Best $R^2$ for each observation range} 
- \label{figura6} 
-\end{figure}
-
-\figref{figura7} shows the best $R^2$ for each variables combination.
-Results are group by farm.
-The better results are obtained with $\overline{T}_{a}$ and the combination of $\overline{T}_{a}$ with $\overline{W}$, in both farms of similarly.
-You can note that the use of all variables in the model or the inclusion of the four variables suggest for expert criteria do not improve significantly the results, then the use of more sensors do not assure a better result.
-
-\begin{figure}[H] 
- \centering
- \includegraphics[scale=.5]{2017-01-15-Variables-R2}
- \caption{Phase one - Best $R^2$ for each variable combination} 
+ \includegraphics[scale=.5]{Usado_Variables-RMSE}
+ \caption{$RMSE$ for each variable combination} 
  \label{figura7} 
 \end{figure}
-
-We can conclude that the best configuration in both farms is to consider the climate and the evolution stage of the current week to predict the evolution stage of the next week.
+%

### 2017-05-01 2017-04-30-Generador de graficos.py

@@ -1,0 +1,61 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Sun Apr 30 15:36:42 2017
+
+@author: lcalvo
+"""
+
+import pandas as pd
+import os
+import matplotlib.pyplot as pl
+
+pl.style.use('ggplot')
+
+os.chdir("D:/Usuarios/Lcalvo/Documents/Cambios/Doctorado/PAPERS EN PROCESO/PaperSigatoka2015")
+tArchivoEntrada = "Usado - 2017-04-30-Datos totales.xlsx"
+tArchivoSalida = "Usado_2017_"
+
+tDatos = pd.read_excel(tArchivoEntrada, header=0 )
+tParte = tDatos[['RMSE', 'Techniques', 'Farm']] 
+
+tBoxPlot = tDatos[['RMSE', 'Techniques', 'Farm']] 
+tBoxPlot = tBoxPlot[tBoxPlot.Techniques == "ESN"]
+
+fig, axes = pl.subplots(nrows=1, ncols=2, sharey=True)
+'''
+tParte1 = tParte[tParte.Farm == "28 Millas"]
+tParte2 = tParte[tParte.Farm == "La Rita"]
+
+tParte1 = pd.DataFrame(tParte1.groupby(['Techniques'])['RMSE'].mean())
+
+tParte2 = tParte2[tParte2.Techniques != "ESN"]
+tParte2 = pd.DataFrame(tParte2.groupby(['Techniques'])['RMSE'].mean())
+
+ax1 = tParte1.plot(ax=axes[0], sort_columns=True,legend=True, kind="bar")
+ax2 = tParte2.plot(ax=axes[1], sort_columns=True, kind="bar")
+
+ax1.set(xlabel='28 Millas', ylabel='RMSE')
+ax2.set(xlabel='La Rita', ylabel='RMSE')
+pl.plot()
+pl.tight_layout()
+pl.savefig( tArchivoSalida+ ".pdf")
+'''
+
+#######################################################
+# BoxPlot
+#######################################################
+tBox1 = tBoxPlot[tBoxPlot.Farm == "28 Millas"]
+tBox2 = tBoxPlot[tBoxPlot.Farm == "La Rita"]
+
+bx1 = tBox1.boxplot(ax=axes[0],column=['RMSE'], by='Techniques', fontsize =6)
[…]

### 2017-05-01 2017-05-01-Generador de graficos.py

@@ -1,0 +1,75 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Sun Apr 30 15:36:42 2017
+
+@author: lcalvo
+"""
+
+import pandas as pd
+import os
+import matplotlib.pyplot as pl
+
+pl.style.use('ggplot')
+
+os.chdir("D:/Usuarios/Lcalvo/Documents/Cambios/Doctorado/PAPERS EN PROCESO/PaperSigatoka2015")
+tArchivoEntrada = "Usado - 2017-04-30-Datos totales.xlsx"
+tArchivoSalida = "Usado_2017_"
+
+tDatos = pd.read_excel(tArchivoEntrada, header=0 )
+tParte = tDatos[['RMSE', 'Techniques', 'Farm']] 
+
+tBoxPlot = tDatos[['RMSE', 'Techniques', 'Farm']] 
+tBoxPlotE = tDatos[['RMSE', 'Techniques', 'Farm']] 
+tBoxPlot = tBoxPlot[tBoxPlot.Techniques != "ESN"]
+tBoxPlotE = tBoxPlotE[tBoxPlotE.Techniques == "ESN"]
+
+fig, axes = pl.subplots(nrows=2, ncols=2 ) #  sharey=True
+'''
+tParte1 = tParte[tParte.Farm == "28 Millas"]
+tParte2 = tParte[tParte.Farm == "La Rita"]
+
+tParte1 = pd.DataFrame(tParte1.groupby(['Techniques'])['RMSE'].mean())
+
+tParte2 = tParte2[tParte2.Techniques != "ESN"]
+tParte2 = pd.DataFrame(tParte2.groupby(['Techniques'])['RMSE'].mean())
+
+ax1 = tParte1.plot(ax=axes[0], sort_columns=True,legend=True, kind="bar")
+ax2 = tParte2.plot(ax=axes[1], sort_columns=True, kind="bar")
+
+ax1.set(xlabel='28 Millas', ylabel='RMSE')
+ax2.set(xlabel='La Rita', ylabel='RMSE')
+pl.plot()
+pl.tight_layout()
+pl.savefig( tArchivoSalida+ ".pdf")
+'''
+
+#######################################################
+# BoxPlot
+#######################################################
+tBox1 = tBoxPlot[tBoxPlot.Farm == "28 Millas"]
[…]

### 2017-05-02 conclusions.tex
1-5 6.59

--- 2016-06-11 conclusions.tex
@@
 \label{sec:concl}
+Este estudio presentó la comparación de varias técnicas de machine learning en la predicción de the la tasa de desarrollo de la Sigatoka negra. Al respecto se presentan las conclusionez al respecto:
+\begin{itemize}
+\item 
+\item 
+\item 
+\item
+
+\end{itemize}
+
+A comparison of machine learning techniques
+The main findings were: 1) The highest R2 was 60\% 2) The highest R2 were reached with linear models like support vector regression with linear kernel, 3) As little as three meteorological variables can be used because of the correlations detected among variables.
 
 \subsection{Future work}
+Currently this research is focused on the extension of these findings to other farms and the extension of the model to be applied in an early warning system, which does not require the regression task.

### 2017-05-02 conclusions.tex
1-5 10.45pm

--- 2016-06-11 conclusions.tex
@@
-\section{Conclusions}
-\label{sec:concl}
-
-\subsection{Future work}
+\section{Conclusions and Future work}
+Este estudio presentó la comparación de varias técnicas de machine learning en la predicción de la tasa de desarrollo de la enfermedad llamada Sigatoka negra. The following conclusions can be drawn:
+\begin{enumerate}
+\item Si bien los valores absolutos del $RMSE$ y el $R^2$ difieren entre las dos fincas en estudio, se mostró que al seleccopmar la técnica, la cantidad de semanas de observación, la cantidad de semanas hacia adelante a predecir y las variables a incluir en el modelo, coinciden los resultados.
+Lo cual es muestra de que estas características estan asociadas al fenómeno más que a una finca en particular.
+\item Los menores $RMWE$ were reached with linear models.
+\item  As little as three meteorological variables can be used because of the correlations detected among variables. 
+\item Las ESN obtuvieron $RMWE$ muy altos, lo cual es producto de la cardinalidad del conjunto de datos, que no es suficiente para que la red neuronal se ajuste durante el entrenamiento.
+\item Valores de $R^2$ cercanos al 60\% se alcanzaron en las pruebas.
+\end{enumerate}
+%
+Future work of this research is focused on:
+\begin{enumerate}
+\item The extension of these findings to other farms.
+\item The extension of the model to be applied in an early warning system, which does not require the regression task.
+\item The use of image to increase the feature sets.
+\end{enumerate}

### 2017-05-02 methodology.tex
1-5 10.45pm

@@
 \begin{itemize}
-\item Patrones de semanas: Se formaron patrones desde tres semanas continuas
+\item Patrones de semanas: Se formaron patrones desde una semana
 de observación para predecir una semana adelante, hasta doce semanas de
 observación para predecir tres semanas adelante. Por tanto, $n\times{}m$ combinations, 
-with $n=1\ldots{}12$ and $m=1\ldots{}3$.
+with $n=1\ldots{}12$ and $m=1\ldots{}3$. \figref{figura8} shows the concept.
+%
+\begin{figure}[H] 
+ \centering
+ \includegraphics[scale=.7]{Usado_2017-04-30_Weeks_Patrones}
+ \caption{Week patterns explanation process} 
+ \label{figura8} 
+\end{figure}
+%
 

### 2017-05-02 results.tex
1-5 10.45pm

@@
 Realizados los experimentos, la \tablename $.$\ref{tabla2} muestra los $RMSE$ obtenidos, en la cual se puede observar que las ESN obtienen un resultado muy diferente a las demás técnicas.
-El $RMSE$ promdio obtenido por las ESN $11085.47$ es cercano a 20 veces el de las otras técnicas.
+El $RMSE$ promedio obtenido por las ESN $11085.47$ es cercano a 20 veces el de las otras técnicas.
 Este resultado de las ESN se explica debido a que estas redes neuronales recurrentes parten de estados aleatorios para irse ajustando en el proceso de entrenamiento, y por tanto, los valores obtenidos son reflejo de que la cantidad de datos disponibles no logra que la red neuronal se ajuste.
@@
 Se grafica ESN por separado debido a la diferencia de escala.
-Con respecto a las otras técnicas, la regresión lineal, Elastic net, SVR with gaussian kernel y SVR with linear kernel presentan compartamientos muy similares y aunque el promedio del $RMSE$ es diferente entre las fincas, La Rica cercano a 450 y 28 Millas cercano a 820, el comportaniento relativo de las técnicas es el mismo.
+Con respecto a las otras técnicas, la regresión lineal, Elastic net, SVR with gaussian kernel y SVR with linear kernel presentan compartamientos muy similares y aunque el promedio del $RMSE$ es diferente entre las fincas, La Rita cercano a 450 y 28 Millas cercano a 820, el comportaniento relativo de las técnicas es el mismo.
 Por su parte, SVR with sigmoid kernel presenta un promedio de $RMSE$ muy inferior a ESN, pero superior a las otras cuatro técnicas.
@@
 \figref{figura5} shows the Pareto frontier for each farm with respect to $R^2$ and $RMSE$.
+The red dots correspond to the observations belonging to Pareto frontier.
 La Rita obtains upper $R^2$ with respect to 28 Millas, but 28 Millas obtains better $RMSE$ than La Rita.
@@
 \hline
-\bfseries Variable & \bfseries Observation range & \bfseries Weeks ahead & \bfseries Technique &\bfseries $R^2$ & \bfseries $RMSE$\\ 
+\bfseries Variable & \bfseries Observation  & \bfseries Weeks & \bfseries Technique &\bfseries $RMSE$ & \bfseries $R^2$\\ 
+                   & \bfseries range - weeks & \bfseries ahead   & & & \\
 \hline\hline 
@@
 Se muestra el $RMSE$ para la predicción de 1, 2 ó 3 semanas adelante con respecto a la información de las variables climatológicas y del preaviso biológico observadas.
-Se aprecia que to predict one week ahead obtiene un $RMSE$ más bajo que two weeks ahead, and this is better than three weeks ahead, lo cual es lo esperado pues entre más semanas adelante se desee predecir, hay más incertidumbre.
+Se aprecia que to predict one week ahead obtiene un $RMSE$ más bajo que two weeks ahead, and this is lower than three weeks ahead, lo cual es lo esperado pues entre más semanas adelante se desee predecir (en el futuro), hay más incertidumbre.
 %

### 2017-05-10 conclusions.tex
09-5-2017 cambios en tamaño de gráficos y conclusiones en prosa

--- 2017-05-01 conclusions.tex
@@
-\section{Conclusions}
-\label{sec:concl}
-
-The main findings were: 1) The highest R2 was 60\% 2) The highest R2 were reached with linear models like support vector regression with linear kernel, 3) As little as three meteorological variables can be used because of the correlations detected among variables.
-
-\subsection{Future work}
-Currently this research is focused on the extension of these findings to other farms and the extension of the model to be applied in an early warning system, which does not require the regression task.
+\section{Conclusions and Future work}
+Este estudio presentó la comparación de varias técnicas de machine learning en la predicción de la tasa de desarrollo de la enfermedad llamada Sigatoka negra.
+Al respecto, una primera conclusión es que si bien los valores absolutos del $RMSE$ y el $R^2$ difieren entre las dos fincas en estudio, se mostró que al seleccopmar la técnica, la cantidad de semanas de observación, la cantidad de semanas hacia adelante a predecir y las variables a incluir en el modelo, coinciden los resultados.
+Lo cual es muestra de que estas características estan asociadas al fenómeno más que a una finca en particular.
+%
+Además, se pudo mostrar que los menores $RMWE$ were reached with linear models.  
+%
+Con respecto a cuántas y cuáles variables incluir en el modelo con el fin de obtener un $RMSE$ menor y/o un $R^2$ mayor, se concluye que as little as three meteorological variables can be used because of the correlations detected among variables.
+%
+Por otra parte, se concluyó que las ESN obtuvieron $RMWE$ muy altos, lo cual es producto de la cardinalidad del conjunto de datos, que no es suficiente para que la red neuronal se ajuste durante el entrenamiento.
+%
+Finalmente se destaca que los valores de $R^2$ cercanos al 60\% se alcanzaron en las pruebas.
+%
+Como trabajo futuro de esta investigación se tiene the extension of these findings to other farms, the extension of the model to be applied in an early warning system, which does not require the regression task, and the use of image to increase the feature sets.

### 2017-05-10 2017-05-01-Generador de graficos.py

@@ -23,8 +23,8 @@ tBoxPlotE = tDatos[['RMSE', 'Techniques', 'Farm']]
 tBoxPlot = tBoxPlot[tBoxPlot.Techniques != "ESN"]
 tBoxPlotE = tBoxPlotE[tBoxPlotE.Techniques == "ESN"]
 
-fig, axes = pl.subplots(nrows=2, ncols=2 ) #  sharey=True
 '''
+fig, axes = pl.subplots(nrows=2, ncols=2 ) #  sharey=True
 tParte1 = tParte[tParte.Farm == "28 Millas"]
 tParte2 = tParte[tParte.Farm == "La Rita"]
 
@@ -46,6 +46,7 @@ pl.savefig( tArchivoSalida+ ".pdf")
 #######################################################
 # BoxPlot
 #######################################################
+'''
 tBox1 = tBoxPlot[tBoxPlot.Farm == "28 Millas"]
 tBox2 = tBoxPlot[tBoxPlot.Farm == "La Rita"]
 tBox3 = tBoxPlotE[tBoxPlotE.Farm == "28 Millas"]
@@ -66,6 +67,20 @@ bx1.set_xticklabels(tBox1.Techniques.unique(), rotation=70)
 bx2.set_xticklabels(tBox2.Techniques.unique(), rotation=70)
 bx3.set_xticklabels(tBox3.Techniques.unique(), rotation=0)
 bx4.set_xticklabels(tBox4.Techniques.unique(), rotation=0)
+'''
+
+fig, axes = pl.subplots(nrows=1, ncols=2 ) #  sharey=True
+tBox1 = tBoxPlot[tBoxPlot.Farm == "28 Millas"]
+tBox2 = tBoxPlot[tBoxPlot.Farm == "La Rita"]
+
+bx1 = tBox1.boxplot(ax=axes[0],column=['RMSE'], by='Techniques', fontsize =6)
+bx2 = tBox2.boxplot(ax=axes[1],column=['RMSE'], by='Techniques', fontsize =6)
+
+bx1.set(xlabel='28 Millas', ylabel='')
+bx2.set(xlabel='La Rita', ylabel='')
+
+bx1.set_xticklabels(tBox1.Techniques.unique(), rotation=70)
+bx2.set_xticklabels(tBox2.Techniques.unique(), rotation=70)
 
 pl.plot(title ="")
 fig.suptitle('')
